### 2021-06-09 extras/outline/outline.rmd
initialize

---
title: "Paper outline"
output:
    bookdown::pdf_document2: default
geometry: margin=1.0in
toc: false
font-size: 12pt
header-includes:
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)

d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      format = "f",
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```

# Abstract (200 words)

- Objective
- Study Design and Setting
- Results
- Conclusion

# Introduction

  - Generalizing overall treatment effects estimated in RCTs to individuals
    often is problematic (Rothwell, 1995; Kravitz, 2004; Rothwell, 2005;)
  - Subgroup analyses (SGAs) play central role for the assessment of HTE,
    despite well-known issues (references).
  - Several guidelines have been proposed to outline the appropriate settings of
    SGAs (references).
  - Earlier work has provided framework separating predictive HTE approaches
    from descriptive ones in an attempt to complement previous categorizations
    of exploratory-confirmatory SGAs (Varadhan, 2013)
  - More recent work (PATH, EnE) focuses on predictive HTE based on risk
    prediction modeling and differentiates it from causal inference approaches
      - Risk is a mathematical determinant of risk (explain that!!), so
        focusing on risk-based approaches provides a viable option.
      - There are earlier implementations, mainly assuming constant relative
        treatment effect (Califf, 1997; Dorresteijn, 2011)
  - In parallel, methods for the evaluation of predictive HTE have been
    transported from the prediction modeling field (c for benefit, else?)
  - Even though stratifying on risk is an improvement, as it is less prone to
    overfitting compared to effect modeling approaches (van Klaveren, 2019), it
    still may not be appropriate for individualizing benefit predictions. Jumps
    at cut-offs are not realistic.
  - This results in need for smoother types of analyses for HTE...

# Methods

  - Describe different approaches to individualize
  - Describe base-case scenario
  - Describe deviations from base-case scenario
  - Evaluation methods
      - Evaluation is performed on validation dataset
      - Metrics description

# Results

## Simulation results

## GUSTO trial results

# Discussion

# Figures and Tables

- 6 in total (both figures **and** tables)

### 2021-06-09 extras/protocol/protocol.rmd
initialize

---
title: "Protocol"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2: default
geometry: margin=1.0in
toc: false
font-size: 12pt
header-includes:
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jamia.csl
---

```{r, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)

d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      format = "f",
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```

# Introduction
Risk-based HTE analyses typically result in risk stratum specific estimates of treatment effects. Different approaches to individualize treatment effects have been proposed, but it is unclear which approach performs best under different circumstances. We aim  to compare different approaches to  estimating individualized treatment effects.

For all patients we observe covariates $x_1,\dots,x_8$, of which $4$ are continuous and $4$ are binary. More specifically,

\begin{equation*}
x_1,\dots,x_4 \sim N(0, 1)
\end{equation*}
\begin{equation*}
x_5,\dots,x_8 \sim B(1, 0.2)
\end{equation*}

We first, generate the binary outcomes $y$ for the untreated patients ($t_x=0$), based on

\begin{equation} 
P(y\given \bm{x}, t_x=0) = g(\beta_0 + \beta_1x_1+\dots+\beta_8x_8) = g(lp_0),
(\#eq:p0)
\end{equation}

where $$g(x) = \frac{e^x}{1+e^x}$$

For treated patients, outcomes are generated from:

\begin{equation}
P(y\given\bm{x}, t_x=1) = g(lp_1)
(\#eq:p1)
\end{equation}

where $$lp_1 = \gamma_2(lp_0-c)^2+\gamma_1(lp_0-c)+\gamma_0$$

# Settings
## Base-case scenario
The base-case scenario assumes a constant odds ratio of $0.8$ in favor of treatment. The simulated datasets are of size $n=4250$, where treatment is allocated at random using a 50/50 split (80% power for the detection of an unadjusted OR of 0.8, assuming an event rate of 20% in the untreated arm). Outcome incidence in the untreated population is set at $20\%$. For the development of the prediction model we use the model defined in \@ref(eq:p0) including a constant treatment effect. When doing predictions, $t_x$ is set to $0$. The value of the true $\bm{\beta}$ is such that the above prediction model has an AUC of $0.75$.

The previously defined targets are achieved when $\bm{\beta}=(-2.08, 0.49,\dots,0.49)^t$. For the derivations in the treatment arm we use $\bm{\gamma}=(\log(0.8), 1, 0)^t$.

## Deviations from base-case
We deviate from the base-case scenario in two ways. First, we alter the overall target settings of sample size, overall treatment effect and prediction model AUC. In a second stage, we consider settings that violate the assumption of a constant relative treatment effect, using a model-based approach.

For the first part, we consider:

* Sample size:
  + $n=1064$
  + $n=17000$
* Overall treatment effect:
  + $OR=0.5$
  + $OR=1$
* Prediction performance:
  + $AUC=0.65$
  + $AUC=0.85$

We set the true risk model coefficients to be $\bm{\beta} = \big(-1.63, 0.26,\dots,0.26\big)^t$ for $AUC=0.65$ and $\bm{\beta} = \big(-2.7, 0.82,\dots,0.82\big)^t$ for $AUC=0.85$. In both cases, $\beta_0$ is selected so that an event rate of $20\%$ is maintained in the control arm.

For the second part linear and quadratic deviations from the assumption of
constant relative effect are considered. We also consider different intensity
levels of these deviations. The settings for these deviations are defined in
supplementary Table \@ref(tab:tab1) and result in the effects of Figure
\@ref(fig:fig1).

```{r fig1, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '50%'}
knitr::include_graphics(here::here("figures/deviate_linear_08.png"))
knitr::include_graphics(here::here("figures/deviate_quadratic_08.png"))
```

In Figure \@ref(fig:fig2) the absolute benefits observed based on different settings are presented. The base-case scenario is also presented as a reference.

```{r fig2, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '50%'}
knitr::include_graphics(here::here("figures/deviate_linear_absolute_08.png"))
knitr::include_graphics(here::here("figures/deviate_quadratic_absolute_08.png"))
```

Finally, we consider 3 additional scenarios of interaction of individual covariates with treatment. These scenarios include a 4 weak interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2 weak and 2 strong interactions (Table \@ref(tab:tab2)).

# Risk modeling
Merging treatment arms, we develop prediction models including a constant relative treatment effect:

\begin{equation}
E\{y\given x,t_x\} = P(y\given x, t_x) = g(\beta_0+\beta_1x_1+\dots+\beta_8x_8+\gamma t_x)
\end{equation}
(\#eq:risk)
Individualized predictions are derived setting $t_x=0$.

# Approaches to individualize benefit predictions

## Risk stratification
Derive a prediction model using the same approach as above and divide the population in equally sized risk-based subgroups. Estimate subgroup-specific absolute benefit from the observed absolute differences. Subject-specific benefit predictions are made by attributing to individuals their corresponding subgroup-specific estimate.

## Constant treatment effect
Assuming a constant relative treatment effect, fit the adjusted model in \@ref(eq:risk). Then, an estimate of absolute benefit can be derived from $$\hat{f}_{\text{benefit}}(lp\given \bm{x}, \hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$

## Linear interaction
The assumption of constant relative treatment effect is relaxed modeling a linear interaction of treatment with the risk linear predictor: 
$$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} = g\big(lp+(\delta_0+\delta_1lp)t_x\big)$$
We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) = g(lp) - g\big(\delta_0+(1+\delta_1)lp\big)$$

## Non-linear interaction
Finally, we drop the linearity assumption and predict absolute benefit by taking the difference between smooth fits, separately derived in each treatment arm:
$$f_{\text{benefit}}(lp\given\bm{x},\hat{\bm{\beta}}) = \hat{f}_{\text{smooth}}(lp\given \bm{x}, \hat{\bm{\beta}}, t_x=0) - \hat{f}_{\text{smooth}}(lp\given \bm{x}, \hat{\bm{\beta}}, t_x=1)$$

We consider three different approaches to smoothing:

- Loess
- Restricted cubic splines
- Local likelihood

# Evaluation
Assuming that $\tau(\bm{x})=E\{y\given\bm{x}, t_x=0\} - E\{y\given\bm{x},t_x=1\}$ is the true benefit for each patient and $\hat{\tau}(x)$ is the estimated benefit from a method under study, the ideal loss function to use for the considered methods would be the unobservable root mean squared error $E\big\{(\hat{\tau} - \tau)^2\given \bm{x}\big\}$. However, in simulations the true patient-level benefit is available. Therefore, we will estimate the RMSE from
$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$
We also compare the predictive performance of the methods under study. We assess
discrimination using the c-for-benefit statistic described in
[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the larger
treatment arm without a pair. We define *observed* pair-specific treatment
benefit as the difference of observed outcomes between the untreated and the
treated patient of each pair. Pair-specific *predicted* benefit is defined as
the average of predicted benefits within each pair. Then, c-for-benefit is
defined as the probability that from two randomly chosen matched patient pairs
with unequal observed benefit, the pair with greater observed benefit also has a
higher predicted benefit.

We evaluate calibration in a similar manner, using the integrated calibration
index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
benefit, observed benefits are regressed on the predicted benefits using a
locally weighted scatterplot smoother (loess). The ICI is the area between the
loess fit and the diagonal, with values closer to $0$ indicating better
calibration.

In all cases, the evaluation metrics are calculated in a large simulated
super-population of 500,000 patients.

# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

\newpage
# Supplement
```{r paok, echo=FALSE, eval=TRUE, warning=FALSE}
table <- read_csv(
  here("data/processed/analysisIds.csv"),
  col_types = "iffiiddddddddddddd"
) %>%
  select(-"type")

kableExtra::kable(
  table,
  format = "latex",
  escape = FALSE,
  longtable   = TRUE,
  align       = "r",
  booktabs    = TRUE,
  col.names   = c(
    "Scenario",
    "Effect",
    "N",
    "AUC",
    paste0("b", 0:8),
    paste0("g", 0:2),
    "c"
  ),
  digits      = 2,
  format.args = list(
    big.mar      = ",",
    decimal.mark = "."
  )
) %>%
  kableExtra::group_rows(
    group_label = "Constant treatment effect",
    start_row   = 10,
    italic      = TRUE,
    end_row     = 27
  ) %>%
  kableExtra::group_rows(
    group_label = "Linear deviation",
    start_row   = 28,
    italic      = TRUE,
    end_row     = 45
  ) %>%
  kableExtra::group_rows(
    group_label = "Quadratic deviation",
    start_row   = 46,
    italic      = TRUE,
    end_row     = 63
  ) %>%
  kableExtra::landscape() %>%
  kableExtra::add_header_above(
    c(
      "Analysis ID"           = 4,
      "Baseline risk"         = 9,
      "True treatment effect" = 4
      )
  )
```

\newpage
```{r tab2, echo=FALSE, eval=TRUE, warning=FALSE}
asd <- read_csv(
  here("data/processed/analysisIdsInteractions.csv"),
  col_types = "iffiidddddddddddddd"
) %>%
  select(-c("type", "g0"))

  knitr::kable(
    asd,
    format      = "latex",
    longtable   = TRUE,
    align       = "r",
    booktabs    = TRUE,
    col.names   = c(
      "Scenario",
      "Effect",
      "N",
      "AUC",
      paste0("b", 0:8),
      "g1", "g2", "g5", "g6"
    ),
    digits      = 2,
    format.args = list(
      big.mar      = ",",
      decimal.mark = "."
    ),
    caption    = ""
  ) %>%
  kableExtra::landscape() %>%
  kableExtra::add_header_above(
    c(
      "Analysis ID"           = 4,
      "Baseline risk"         = 9,
      "Coefficient in\ntreatment arm" = 4
      )
  )
```

### 2021-06-09 submission/manuscript.rmd
initialize

---
title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}

knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\thispagestyle{empty}
\vspace{35mm}

Alexandros Rekkas${^1}$, Peter R. Rijnbeek${^1}$, ..., Ewout W. Steyerberg${^2}$, 
David van Klaveren${^3}$

\vspace{40mm}

$^1$ Department of Medical Informatics, Erasmus University Medical Center,
Rotterdam, Netherlands

$^2$ Department of Biomedical Data Sciences, Leiden University Medical Center,
Leiden, Netherlands

$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam,
Netherlands

\newpage
\newpage
\thispagestyle{empty}
# Abstract {-}
\singlespacing 
**Objective:** Simulation study to compare different risk-based approaches to
estimating individualized treatment effects within the RCT setting. **Study
Design and Setting:** Starting from a base case scenario that assumes a true
constant treatment effect, we considered a total of 66 scenarios for introducing
non-constant effects and evaluating methods under different sample sizes and
baseline risk prediction performance. We compared 7 methods for predicting
absolute benefit: A constant treatment effect model, a risk stratified approach,
a model including a linear interaction of the baseline risk linear predictor
with treatment, 3 restricted cubic spline smoothing models of increasing
flexibility (3, 4 and 5 knots) and an adaptive model selection method based on
Akaike's Information Criterion. We evaluated performance using root mean squared
error, discrimination for benefit and calibration for benefit (i.e., observed
vs. predicted risk difference in treated vs. untreated). **Results:** The model
including a linear interaction of the risk linear predictor with treatment had
adequate performance that was robust under the majority of the simulation
scenarios. Methods using restricted cubic spline smoothing required larger
sample sizes and higher prediction AUC to achieve adequate performance. The
adaptive approach's performance was comparable to the performance of the best
model in each scenario. **Conclusion:** In most cases using a model just
including a linear interaction of the risk linear predictor with treatment
adequately predicts absolute benefit.

\newpage 
\doublespacing

\linenumbers

# Introduction

Within the setting of patient-centered outcomes research, predictive approaches
for assessing heterogeneity of treatment effects (HTE) aim at the development of
models predicting either individualized effects or which of two (or more)
treatments is better for an individual [@Varadhan2013]. In prior work, we
divided such methods in three broader categories based on the reference class
used for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. Risk-modeling approaches use prediction of
baseline risk as the reference; treatment effect modeling approaches also model
treatment-covariate interactions, in addition to risk factors; optimal treatment
regime approaches focus on developing treatment assignment rules and therefore
rely heavily on modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers in the form of
treatment-covariate interactions often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, "blinded" to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. However, even though estimates at the risk subgroup level are
accurate, this does not apply on the individual level, especially for patients
with predicted risk at the boundaries of the risk intervals. Therefore, the
risk-stratified approach should be used for exploring and presenting an overview
of HTE, while inferences on the individual level should be made with caution.

We aimed to provide an overview of methods that can be used to move from a
risk-stratified approach to a continuous one using common smoothing techniques.
These methods extend the risk-based framework of predictive HTE analyses to
allow predictions on the individual level, within the RCT setting. We carried
out a simulation study to compare the performance of these methods under
different settings of increasing non-linearity of treatment effects. Finally, we
carried out an application on real data as a demonstration of the considered
techniques.

# Methods

## Simulation scenarios

In the simulated datasets of the base-case scenario treatment was allocated at
random using a 50/50 split. For each patient we simulated $8$ baseline
covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$.
Outcomes for patients in the control arm were generated from a logistic
regression model including all baseline covariates. Coefficient values were
such, so that the prediction model had an AUC of $0.75$ and an event rate of
$20\%$ in the control arm was achieved. Outcomes in the treatment arm were
created using the same logistic regression model, including a constant treatment
effect odds ratio (OR) of $0.8$. The generated samples of the base-case scenario were
of size $n=4,250$ ($80\%$ power for the detection of an unadjusted OR of
$0.8$).

We evaluated the effect of sample size considering additional scenarios with
sample sizes of $1,064$ and $17,000$. We also evaluated the effect of prediction
performance, adjusting the baseline covariate coefficients, so that AUC values
of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
$500,000$ patients.

A true logistic regression model with a constant treatment effect (constant OR)
implies that outcome risk in the treatment arm is a straight line parallel to
the first diagonal on the *log-odds* scale, with distance equal to $\log(\text{OR})$. 
We assessed the effect of stronger and absent relative treatment effects
($\text{OR}=0.5$ or $\text{OR}=1$). We also relaxed the assumption that the
line should be parallel to the diagonal, considering moderate and stronger
linear deviations. Finally, we dropped the assumption of linearity allowing for
quadratic deviations (Figure \@ref(fig:fig1)).

```{r fig1, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width='50%'}
knitr::include_graphics(here::here("figures/deviationsManuscript.png"))
knitr::include_graphics(here::here("figures/deviationsAbsoluteManuscript.png"))
```

We also considered scenarios with treatment-covariate interactions. These
scenarios include 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of $66$ scenarios. The exact settings for each
scenario are available in the supplementary material.

## Individualized risk-based benefit predictions

All methods assume that a risk prediction model is available and can be used to
assign individualized predictions. For the simulations we developed the
prediction models internally and blinded to treatment using logistic regression
including main effects for all baseline covariates and treatment. Predictions on
individuals were made setting treatment to $0$.

The **stratified HTE method** was suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute effects are estimated using the
differences in event rates between treatments within risk quarters. We
considered this approach as a reference, expecting it to perform worse than the
other candidates, as its objective is not individual benefit prediction.

We also considered a set of **linear methods**. We fit separate models within
treatment arms using only the treatment indicator and the linear predictor of
the internal risk prediction model. In the simpler case, we assume a constant
relative treatment effect (OR). Absolute benefit is then estimated from
$\text{expit}(lp +\log(\text{OR}))$, 
where $\text{expit}(x)=\frac{e^x}{1+e^x}$
and *lp* is the linear predictor of the prediction model. A different approach
fits a logistic regression using treatment, risk linear predictor and their
interaction within each treatment arm. In this case, absolute benefit is
estimated from
$\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)$.
We will refer to this method as the linear interaction approach.

Finally, we used restricted cubic splines (RCS) to relax the linearity assumption on
the effect of the linear predictor [@Harrell1988]. We compared the results
for 3, 4 and 5 knots when fitting the splines to introduce increasing
flexibility to the methods considered.

## Evaluation metrics

# Results

## Simulations

```{r, echo=FALSE, warning=FALSE, message=FALSE}
rmse_scenarios_10_12_16 <- readr::read_csv(
  file = here::here(file.path("data/processed", "rmse.csv"))
) %>%
  dplyr::filter(scenarioId %in% c(10, 12, 16))
```
Under the base case of constant relative treatment effect, the model assuming a
constant treatment effect had the lowest median RMSE, regardless of true
prediction AUC and sample size (Figure \@ref(fig:rmseconstant)). Linear
interaction models demonstrated comparable performance. Among the RCS smoothing
methods, the one fitted with 3 knots always performed best, while the increased
flexibility achieved when increasing the knots resulted in overfitting and worse
performance. The adaptive approach under all scenarios performed similar to the
model with smaller RMSE in all scenarios.

```{r rmseconstant, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_constant.png"))
```

When we introduced deviations from the base case of constant relative treatment
effects, while keeping fixed both the sample size (N = 4250) and the true
prediction AUC (0.75) the linear interaction model had the lowest RMSE (Figure
\@ref(fig:rmsebase); Panels A, D, and G). When these deviations were moderate
(Figure \@ref(fig:rmsebase); Panel A) the constant treatment effect model had
comparable performance to the linear interaction model. This can be attributed
to the fact that such deviations are quite mild and absolute benefits maintain
similar patterns across baseline risk. On the contrary, when strong quadratic
deviations were considered the constant effect model's RMSE sharply increased,
while the more flexible method of RCS smoothing (3 knots) preformed very well
(Figure \@ref(fig:rmsebase); Panel G). Again, increasing the number of knots
increased RMSE, indicating overfitting.

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE when strong quadratic deviations from the base
case of constant relative treatment effects were assumed (Figure
\@ref(fig:rmsebase); Panel H). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \@ref(fig:rmsebase); Panels B and E). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations. We observed similar
results when we increased the sample size to 17000, while keeping the true
prediction AUC constant at 0.75 (Figure \@ref(fig:rmsebase); Panels C, F, and I).

```{r rmsebase, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_base.png"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
rmse_scenarios_10_64_66_65 <- readr::read_csv(
  file = here::here(file.path("data/processed", "rmse.csv"))
) %>%
  dplyr::filter(scenarioId %in% c(10, 64, 66, 65))
```
When focusing on the different scenarios where true treatment-covariate
interactions were considered all methods had similar RMSE performance, (Figure
\@ref(fig:rmseInteractions); Panels B, C and D). In case of strong
treatment-covariate interactions the constant effect model had slightly
increased RMSE
(`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`])
compared to the other methods. The linear interaction model with the risk linear
predictor had the lowest RMSE
(`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`]).

```{r rmseInteractions, echo=FALSE, fig.cap="Root mean squared error of the different methods when focusing on the effect treatment-covariate interactions. In all scenarios sample size is fixed to 4250 and the true prediction AUC is 0.75. \\textit{Panel A} presents the results for the base case scenario, where no interactions are present. \\textit{Panel B} presents the results for the scenario with 4 weak interactions. \\textit{Panel C} presents the results for the scenario with mixed interactions (2 weak and 2 strong). \\textit{Panel D} presents the results for the scenario with 4 strong interactions.", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/rmse_interactions.png"))
```

All candidate methods demonstrated comparable discrimination for benefit in all
scenarios where linear and quadratic deviations from the base case of constant
treatment effect were considered (Figure \@ref(fig:discrimination)). However,
models including a linear interaction with the risk linear predictor tended to
present much lower variability compared to all other model-based and smoothing
approaches. We also observed an increasing trend of discrimination for benefit
variability with increasing number of restricted cubic spline knots in all
scenarios. This is evidence that the increased flexibility of these methods
often led to overfitting.

```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the different methods. \\textit{Panel A} presents the results when moderate linear deviations from the base case scenario of constant treatment effects are considered, sample size is 4250 and true prediction AUC is 0.75. \\textit{Panel B} presents the results when moderate linear deviations are considered and true prediction AUC is 0.85. Sample size is 4250. \\textit{Panel C} presents the results when moderate linear deviations are considered and the sample size is 0.75. \\textit{Panel D} presents the results when strong linear deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85. \\textit{Panel E} presents results for true prediction AUC of 0.85 and \\textit{Panel F} for sample size of 17000 (assuming moderate linear deviations from base case). \\textit{Panel G} presents the results when strong quadratic deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85. \\textit{Panel H} presents the results for true prediction AUC of 0.85 and \\textit{Panel I} for sample size of 17000 (assuming strong quadratic deviations from base case)", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

When focusing on calibration for benefit, the linear interaction model had the
lowest median ICI for benefit in the majority of the scenarios except for the
scenarios where moderate linear deviations from the base case were considered.
In that case constant treatment effect models demonstrated the best performance,
very comparable to the linear interaction model's performance, nonetheless
(Figure \@ref(fig:calibration); Panels A, B, and C).

```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the different methods. \\textit{Panel A} presents the results when moderate linear deviations from the base case scenario of constant treatment effects are considered, sample size is 4250 and true prediction AUC is 0.75. \\textit{Panel B} presents the results when moderate linear deviations are considered and true prediction AUC is 0.85. Sample size is 4250. \\textit{Panel C} presents the results when moderate linear deviations are considered and the sample size is 0.75. \\textit{Panel D} presents the results when strong linear deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85. \\textit{Panel E} presents results for true prediction AUC of 0.85 and \\textit{Panel F} for sample size of 17000 (assuming moderate linear deviations from base case). \\textit{Panel G} presents the results when strong quadratic deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85. \\textit{Panel H} presents the results for true prediction AUC of 0.85 and \\textit{Panel I} for sample size of 17000 (assuming strong quadratic deviations from base case)", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/calibration_base.png"))
```

## Real data

\newpage
# References
\nolinenumbers
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

### 2021-06-09 code/CalibrationBase.R

@@ -1,0 +1,113 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(
+  28, 30, 34,
+  37, 39, 43,
+  55, 57, 61
+)
+
+metric <- "calibration"
+titles <- c(
+  "**A.** Moderate linear deviation",
+  "**B.** AUC: 0.85; N: 4250",
+  "**C.** AUC: 0.75; N: 17000",
+  "**D.** Strong linear deviation",
+  "**E.** AUC: 0.85; N: 4250",
+  "**F.** AUC: 0.75; N: 17000",
+  "**G.** Strong quadratic deviation",
+  "**H.** AUC: 0.85; N: 4250",
+  "**I.** AUC: 0.75; N: 17000"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, 6),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9.5),
+      plot.title      = ggtext::element_markdown(size = 14)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  plotList[[5]],
+  plotList[[6]],
+  plotList[[7]],
[…]

### 2021-06-09 code/DiscriminationBase.R

@@ -1,0 +1,90 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(
+  28, 30, 34,
+  37, 39, 43,
+  55, 57, 61
+)
+
+metric <- "discrimination"
+titles <- c(
+  "**A.** Moderate linear deviation",
+  "**B.** AUC: 0.85; N: 4250",
+  "**C.** AUC: 0.75; N: 17000",
+  "**D.** Strong linear deviation",
+  "**E.** AUC: 0.85; N: 4250",
+  "**F.** AUC: 0.75; N: 17000",
+  "**G.** Strong quadratic deviation",
+  "**H.** AUC: 0.85; N: 4250",
+  "**I.** AUC: 0.75; N: 17000"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(.52, .62),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9),
+      plot.title      = ggtext::element_markdown(size = 12)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  plotList[[5]],
+  plotList[[6]],
+  plotList[[7]],
+  plotList[[8]],
+  plotList[[9]],
+  nrow = 3,
+  ncol = 3,
+  left = "C-statistic for benefit"
+)
+  ggplot2::ggsave(
+    file.path("figures", "discrimination_base.tiff"), 
+    plot = pp,
+    dpi = 1200,
+    width = 11, 
[…]

### 2021-06-09 code/MergeResults.R

@@ -1,0 +1,16 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+source("code/helpers/TmpFiles.R")
+
+scenarios <- length(list.files("data/raw", pattern = "scenario"))
+if (!dir.exists(".scratch/tmp")) {
+  dir.create(".scratch/tmp", recursive = TRUE)
+}
+lapply(1:scenarios, createTmpFiles, scenarioDir = "data/raw/", tmpDir = ".scratch/tmp/")
+lapply(
+  c("rmse", "discrimination", "calibration", "concordance"), 
+  mergeTmpFiles, 
+  tmpDir = ".scratch/tmp",
+  saveDir = "data/processed"
+  )

### 2021-06-09 code/PlotDeviations.R

@@ -1,0 +1,152 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Plots the theoretical deviations from the base case scenario
+# Output:
+#   figures/deviate_linear_08.png
+#   figures/deviate_quadratic_08.png
+#   figures/deviate_linear_absolute_08.png
+#   figures/deviate_quadratic_absolute_08.png
+# Depends:
+#   code/PlotGammas.R
+
+
+
+source("code/helpers/PlotGammas.R")
+
+library(tidyverse)
+library(SmoothHte)
+library(SimulateHte)
+library(SimulationEvaluationHte)
+library(dplyr)
+
+analysisIds <- readr::read_csv(
+  "data/processed/analysisIds.csv",
+  col_types = "iffiiddddddddddddd"
+)
+
+analysisIdsInteractions <- readr::read_csv(
+  "data/processed/analysisIdsInteractions.csv",
+  col_types = "icfiidddddddddddddd"
+)
+
+generatePlot <- function(scenarios) {
+  
+  gammas <- list()
+  for (i in seq_along(scenarios)) {
+    idSettings <- analysisIds %>%
+      filter(scenario == scenarios[i])
+    gammas[[i]] <- list(
+      g0 = idSettings$g0,
+      g1 = idSettings$g1,
+      g2 = idSettings$g2,
+      c  = idSettings$c
+    )
+  }
+  
+  names(gammas) <- c(
+    "gammas_moderate",
+    "gammas_high"
+  )
+  
+  p <- constantEffectPlot(or = .8, label = "base-case") +
+    plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed") +
+    plotSquare(
+      g0    = gammas$gammas_moderate$g0,
+      g1    = gammas$gammas_moderate$g1,
+      g2    = gammas$gammas_moderate$g2,
+      l     = gammas$gammas_moderate$c,
+      label = "moderate"
+    ) +
+    plotSquare(
+      g0    = gammas$gammas_high$g0,
+      g1    = gammas$gammas_high$g1,
+      g2    = gammas$gammas_high$g2,
+      l     = gammas$gammas_high$c,
+      label = "high"
+    ) +
+    xlab("Linear predictor in control arm") +
+    ylab("Linear predictor in treatment arm") +
+    xlim(-5, 2) +
+    scale_color_manual(
+      values = c("black", "#66c2a5", "#fc8d62", "#8da0cb"),
[…]

### 2021-06-09 code/PlotDeviationsManuscript.R

@@ -1,0 +1,190 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+source("code/helpers/PlotGammas.R")
+
+library(tidyverse)
+library(SmoothHte)
+library(SimulateHte)
+library(SimulationEvaluationHte)
+library(dplyr)
+
+analysisIds <- readr::read_csv(
+  "data/processed/analysisIds.csv",
+  col_types = "iffiiddddddddddddd"
+)
+
+analysisIdsInteractions <- readr::read_csv(
+  "data/processed/analysisIdsInteractions.csv",
+  col_types = "icfiidddddddddddddd"
+)
+
+generatePlot <- function(scenarios) {
+  
+  gammas <- list()
+  for (i in seq_along(scenarios)) {
+    idSettings <- analysisIds %>%
+      filter(scenario == scenarios[i])
+    gammas[[i]] <- list(
+      g0 = idSettings$g0,
+      g1 = idSettings$g1,
+      g2 = idSettings$g2,
+      c  = idSettings$c
+    )
+  }
+  
+  names(gammas) <- c(
+    "gammas_base_case",
+    "gammas_linear_moderate",
+    "gammas_linear_high",
+    "gammas_quadratic_high"
+  )
+  
+  p <- constantEffectPlot(or = .8, label = "base-case") +
+    plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed") +
+    plotSquare(
+      g0    = gammas$gammas_linear_moderate$g0,
+      g1    = gammas$gammas_linear_moderate$g1,
+      g2    = gammas$gammas_linear_moderate$g2,
+      l     = gammas$gammas_linear_moderate$c,
+      label = "linear-moderate"
+    ) +
+    plotSquare(
+      g0    = gammas$gammas_linear_high$g0,
+      g1    = gammas$gammas_linear_high$g1,
+      g2    = gammas$gammas_linear_high$g2,
+      l     = gammas$gammas_linear_high$c,
+      label = "linear-strong"
+    ) +
+    plotSquare(
+      g0    = gammas$gammas_quadratic_high$g0,
+      g1    = gammas$gammas_quadratic_high$g1,
+      g2    = gammas$gammas_quadratic_high$g2,
+      l     = gammas$gammas_quadratic_high$c,
+      label = "quadratic-strong"
+    ) +
+    xlab("Linear predictor in control arm") +
+    ylab("Linear predictor in treatment arm") +
[…]

### 2021-06-09 code/RmseAuc.R

@@ -1,0 +1,76 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(12, 30, 39, 57)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
+  "**C.** Strong linear deviation",
+  "**D.** Strong quadratic deviation"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .11),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9),
+      plot.title      = ggtext::element_markdown(size = 12)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = "Root mean squared error"
+)
+  ggplot2::ggsave(
+    file.path("figures", "rmse_auc.tiff"), 
+    plot = pp,
+    width = 8.5, 
+    height = 7,
+    dpi = 1200,
+    compression = "lzw"
+  )
+  
+  ggplot2::ggsave(
+    file.path("figures", "rmse_auc.png"), 
+    plot = pp,
+    width = 8.5, 
+    height = 7
+  )
+  

### 2021-06-09 code/RmseBase.R

@@ -1,0 +1,113 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(
+  28, 30, 34,
+  37, 39, 43,
+  55, 57, 61
+)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** Moderate linear deviation",
+  "**B.** AUC: 0.85; N: 4250",
+  "**C.** AUC: 0.75; N: 17000",
+  "**D.** Strong linear deviation",
+  "**E.** AUC: 0.85; N: 4250",
+  "**F.** AUC: 0.75; N: 17000",
+  "**G.** Strong quadratic deviation",
+  "**H.** AUC: 0.85; N: 4250",
+  "**I.** AUC: 0.75; N: 17000"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, 10),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9.5),
+      plot.title      = ggtext::element_markdown(size = 14)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  plotList[[5]],
+  plotList[[6]],
+  plotList[[7]],
[…]

### 2021-06-09 code/RmseSampleSize.R

@@ -1,0 +1,76 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(16, 34, 43, 61)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
+  "**C.** Strong linear deviation",
+  "**D.** Strong quadratic deviation"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .07),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9),
+      plot.title      = ggtext::element_markdown(size = 12)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = "Root mean squared error"
+)
+  ggplot2::ggsave(
+    file.path("figures", "rmse_sample_size.tiff"), 
+    plot = pp,
+    dpi = 1200,
+    width = 8.5, 
+    height = 7,
+    compression = "lzw"
+  )
+  
+  ggplot2::ggsave(
+    file.path("figures", "rmse_sample_size.png"), 
+    plot = pp,
+    width = 8.5, 
+    height = 7
+  )
+  

### 2021-06-09 code/SimulationScript.R

@@ -1,0 +1,209 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Script to be run for each simulation scenario
+# Depends:
+#   data/processed/analysisIds.rds
+# Output:
+#   data/processed/scenario_#/
+# Note:
+#   Needs a data/raw/scenario_# as input
+
+args <- commandArgs(trailingOnly = TRUE)
+if (length(args) != 1) stop("Requires scenario id")
+
+library(tidyverse)
+library(SmoothHte)
+library(SimulateHte)
+library(SimulationEvaluationHte)
+library(dplyr)
+
+analysisIds <- readr::read_csv(
+  "data/processed/analysisIds.csv",
+  col_types = "iffiiddddddddddddd"
+)
+
+analysisIdsInteractions <- readr::read_csv(
+  "data/processed/analysisIdsInteractions.csv",
+  col_types = "icfiidddddddddddddd"
+)
+
+selectedScenario <- as.numeric(
+  stringr::str_extract(args[1], "[0-9]+")
+)
+
+if (selectedScenario <= 63) {
+  idSettings <- analysisIds %>%
+    filter(scenario == selectedScenario)
+  createF1 <- function(c) function(x) x - c
+  createF2 <- function(c) function(x) (x - c)^2
+  treatmentEffectSettings <- SimulateHte::createTreatmentEffectSettings(
+    type = "lp",
+    modelSettings = SimulateHte::createModelSettings(
+      constant = idSettings$g0,
+      modelMatrix = matrix(c(1, 1)),
+      transformationSettings = list(
+        createF1(idSettings$c),
+        createF2(idSettings$c)
+      ),
+      coefficients = c(
+        idSettings$g1,
+        idSettings$g2
+      )
+    )
+  )
+} else {
+  idSettings <- analysisIdsInteractions %>%
+    filter(scenario == selectedScenario)
+  treatmentEffectSettings <- SimulateHte::createTreatmentEffectSettings(
+    type = "covariates",
+    modelSettings = SimulateHte::createModelSettings(
+      constant = idSettings$g0,
+      modelMatrix = rbind(
+        c(1, rep(0, 7)),
+        c(0, 1, rep(0, 6)),
+        c(rep(0, 4), 1, 0, 0, 0),
+        c(rep(0, 5), 1, 0, 0)
+      ),
+      transformationSettings = list(
+        identity,
+        identity,
+        identity,
+        identity
+      ),
[…]

### 2021-06-09 code/WriteAnalysisIds.R

@@ -1,0 +1,95 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(readr)
+
+effect       <- c("moderate", "high")
+typeOfEffect <- c("constant", "linear", "quadratic")
+n            <- c("4250", "1064", "17000")
+auc          <- c("75", "65", "85")
+
+grid1 <- expand.grid(
+  type       = typeOfEffect,
+  effectSize = effect,
+  sampleSize = n,
+  auc        = auc
+) %>%
+  arrange(type, effectSize, sampleSize, auc)
+
+effect       <- "absent"
+typeOfEffect <- "constant"
+
+grid2 <- expand.grid(
+  type       = typeOfEffect,
+  effectSize = effect,
+  sampleSize = n,
+  auc        = auc
+) %>%
+  arrange(type, effectSize, sampleSize, auc)
+
+table <- bind_rows(grid2, grid1) %>%
+  mutate(
+    b0 = case_when(
+      auc == "75" ~ -2.08,
+      auc == "65" ~ -1.63,
+      auc == "85" ~ -2.7
+    ),
+    b1 = case_when(
+      auc == "75" ~ .49,
+      auc == "65" ~ .26,
+      auc == "85" ~ .82
+    ),
+    b2 = b1,
+    b3 = b1,
+    b4 = b1,
+    b5 = b1,
+    b6 = b1,
+    b7 = b1,
+    b8 = b1,
+    g0 = case_when(
+      (type == "constant"   &  effectSize == "absent"                )   ~ 0,
+      (type == "constant"   &  effectSize == "moderate"              )   ~ log(.8),
+      (type == "constant"   &  effectSize == "high"                  )   ~ log(.5),
+      (type == "linear"     &  effectSize == "moderate" & auc == "75")   ~ -.29,
+      (type == "linear"     &  effectSize == "moderate" & auc == "65")   ~ -.35,
+      (type == "linear"     &  effectSize == "moderate" & auc == "85")   ~ -.37,
+      (type == "linear"     &  effectSize == "high"     & auc == "75")   ~ -.44,
+      (type == "linear"     &  effectSize == "high"     & auc == "65")   ~ -.54,
+      (type == "linear"     &  effectSize == "high"     & auc == "85")   ~ -.50,
+      (type == "quadratic"  &  effectSize == "moderate" & auc == "75")   ~ -5.02,
+      (type == "quadratic"  &  effectSize == "moderate" & auc == "65")   ~ -5.02,
[…]

### 2021-06-09 code/WriteAnalysisIdsInteractions.R

@@ -1,0 +1,50 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(readr)
+
+effect <- c("weak", "strong", "mixed")
+typeOfEffect <- "interaction"
+n <- 4250
+auc <- 75
+
+table <- tibble(
+  type       = typeOfEffect,
+  effectSize = effect,
+  sampleSize = n,
+  auc        = auc
+) %>%
+  mutate(
+    b0 = -2.08,
+    b1 = .49,
+    b2 = .49,
+    b3 = .49,
+    b4 = .49,
+    b5 = .49,
+    b6 = .49,
+    b7 = .49,
+    b8 = .49,
+    g0 = log(.8),
+    g1 = case_when(
+      (effect == "weak" | effect == "mixed") ~ -.19,
+      (effect == "strong"                  ) ~ -.49
+    ),
+    g2 = case_when(
+      (effect == "weak"                      ) ~ -.19,
+      (effect == "mixed" | effect == "strong") ~ -.49
+    ),
+    g5 = case_when(
+      (effect == "weak" | effect == "mixed") ~ -.19,
+      (effect == "strong"                  ) ~ -.49
+    ),
+    g6 = case_when(
+      (effect == "weak"                      ) ~ -.19,
+      (effect == "mixed" | effect == "strong") ~ -.49
+    ),
+    scenario = 63 + 1:n()
+  ) %>%
+  select(scenario, everything())
+
+# saveRDS(table, "data/processed/analysisIds.rds")
+readr::write_csv(table, "data/processed/analysisIdsInteractions.csv")
+

### 2021-06-09 code/helpers/CreateManuscriptPlots.R

@@ -1,0 +1,80 @@
+createPlot <- function(
+  data,
+  metric,
+  title,
+  limits,
+  pointSize = 1.5
+) {
+  
+  yAxis <- dplyr::case_when(
+    metric == "rmse"           ~ "RMSE",
+    metric == "discrimination" ~ "Discrimination",
+    metric == "calibration"    ~ "Calibration"
+  )
+  
+  tmp <- data %>%
+    dplyr::select(
+      c(
+        "constant_treatment_effect", 
+        "stratified", 
+        "linear_predictor", 
+        "rcs_3_knots",
+        "rcs_4_knots",
+        "rcs_5_knots", 
+        "adaptive"
+      )
+    ) %>% 
+    reshape2::melt()
+  levels(tmp$variable) <- case_when(
+    levels(tmp$variable) == "linear_predictor"          ~ "Linear predictor",
+    levels(tmp$variable) == "constant_treatment_effect" ~ "Constant\ntreatment effect",
+    levels(tmp$variable) == "stratified"                ~ "Stratified",
+    levels(tmp$variable) == "rcs_3_knots"               ~ "RCS\n(3 knots)",
+    levels(tmp$variable) == "rcs_4_knots"               ~ "RCS\n(4 knots)",
+    levels(tmp$variable) == "rcs_5_knots"               ~ "RCS\n(5 knots)",
+    levels(tmp$variable) == "adaptive"                  ~ "Adaptive",
+    TRUE                                                ~ levels(tmp$variable)
+  )
+  
+  plot <- ggplot2::ggplot(
+    data = tmp,
+    ggplot2::aes(
+      x    = variable, 
+      y    = value, 
+      fill = variable
+    )
+  ) +
+    ggplot2::geom_boxplot(
+      outlier.size = pointSize
+    ) +
+    ggplot2::scale_fill_manual(
+      values = c(
+        "#66c2a5",
+        "#fc8d62", 
+        "#8da0cb", 
+        "#e78ac3", 
+        "#a6d854", 
+        "#ffd92f", 
+        "#e5c494"
+      ),
+      breaks = c(
+        "Constant\ntreatment effect", 
+        "Stratified", 
+        "Linear predictor", 
+        "RCS\n(3 knots)",
+        "RCS\n(4 knots)",
+        "RCS\n(5 knots)", 
+        "Adaptive"
+        )
+    ) +
+    ggplot2::ylab(yAxis) +
[…]

### 2021-06-09 code/helpers/TmpFiles.R

@@ -1,0 +1,121 @@
+library(tidyverse)
+
+createTmpFiles <- function(scenarioId, scenarioDir, tmpDir) {
+  
+  evaluation <- readRDS(
+    file.path(
+      scenarioDir,
+      paste(
+        "scenario",
+        scenarioId,
+        sep = "_"
+      ),
+      "evaluation.rds"
+    )
+  )
+  
+  evaluation <- evaluation %>% map(~mutate(., scenarioId = scenarioId))
+  
+  readr::write_csv(
+    evaluation$rmse,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "rmse",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$discrimination,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "discrimination",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$calibration,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "calibration",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$calibration,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "concordance",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  return("done")
+  
+}
+
+
+mergeTmpFiles <- function(metric, tmpDir, saveDir) {
+  files <- list.files(
+    path=tmpDir,
+    full.names = TRUE, 
+    pattern = paste(
+      "tmp",
+      metric,
+      sep = "_"
+    )
+  ) 
+  
+  tmp <- files %>%
+    lapply(readr::read_csv, col_types = cols(.default = "d")) %>% 
+    bind_rows %>%
+    readr::write_csv(
+      file.path(
+        saveDir,
+        paste(metric, "csv", sep = ".")
+      )
+    )
+  
+  file.remove(files)
+  
+  message(paste("Merged results for:", metric))
+  
+}

### 2021-06-09 extras/shiny/global.R

@@ -1,0 +1,69 @@
+library(tidyverse)
+rmse <- readr::read_csv(
+  "data/rmse.csv",
+  col_types = readr::cols(.default = "d")
+)
+discrimination <- readr::read_csv(
+  "data/discrimination.csv",
+  col_types = readr::cols(.default = "d")
+)
+calibration <- readr::read_csv(
+  "data/calibration.csv",
+  col_types = readr::cols(.default = "d")
+)
+
+analysisIds1 <- readr::read_csv(
+  "data/analysisIds.csv",
+  col_types = "icciiddddddddddddd"
+) %>%
+  select(scenario, type, effectSize, sampleSize, auc)
+
+analysisIds2 <- readr::read_csv(
+  "data/analysisIdsInteractions.csv",
+  col_types = "icciidddddddddddddd"
+) %>%
+  select(scenario, type, effectSize, sampleSize, auc)
+
+analysisIds <- bind_rows(analysisIds1, analysisIds2)
+
+
+
+
+createPlot2 <- function(data) {
+  tmp <- data %>%
+    reshape2::melt()
+  levels(tmp$variable) <- case_when(
+    levels(tmp$variable) == "linear_predictor"          ~ "linear predictor",
+    levels(tmp$variable) == "constant_treatment_effect" ~ "constant treatment\neffect",
+    TRUE                                                ~ levels(tmp$variable)
+  )
+  plotly::plot_ly(
+    x = ~ variable,
+    y = ~ value,
+    color = ~ variable,
+    colors = c(
+      "#1b9e77",
+      "#d95f02",
+      "#7570b3",
+      "#e7298a",
+      "#66a61e",
+      "#e6ab02",
+      "#a6761d"
+    ),
+    data = tmp
+  ) %>%
+    plotly::add_trace(
+      type = "box",
+      hoverinfo = "y+q"
+    ) %>%
+    plotly::layout(
+      yaxis = list(
+        hoverformat = ".3f",
+        title = ""
+      ),
+      xaxis = list(
+        title = ""
+      )
+    )
+}
+

### 2021-06-09 extras/shiny/server.R

@@ -1,0 +1,173 @@
+shiny::shinyServer(
+  function(
+    input,
+    output,
+    session
+  ) {
+    
+    selectedVals                       <- shiny::reactiveValues()
+    selectedVals$effect                <- "moderate"
+    selectedVals$nPatients             <- 4250
+    selectedVals$predictionPerformance <- 75
+    
+    observeEvent(
+      input$type,
+      {
+        req(
+          input$type, 
+          input$effect, 
+          input$nPatients, 
+          input$predictionPerformance
+        )
+        
+        selectedVals$effect                <- input$effect
+        selectedVals$nPatients             <- input$nPatients
+        selectedVals$predictionPerformance <- input$predictionPerformance
+      }
+    )
+    
+    output$effectInput <- shiny::renderUI(
+      {
+        if (input$type != "interaction") {
+          shiny::selectInput(
+            inputId  = "effect",
+            label    = "Effect",
+            choices  = c(
+              "absent",
+              "moderate",
+              "high"
+            ),
+            selected = selectedVals$effect
+          )
+        } else {
+          shiny::selectInput(
+            inputId  = "effect",
+            label    = "Effect",
+            choices  = c(
+              "weak",
+              "strong",
+              "mixed"
+            ),
+            selected = "weak"
+          )
+        }
+      }
+    )
+    
+    output$toggleNPatients <- shiny::renderUI(
+      {
+        if (input$type != "interaction") {
+          shiny::selectInput(
+            inputId  = "nPatients",
+            label    = "Number of patients",
+            choices  = c(
+              1064,
+              4250,
+              17000
+            ),
+            selected = selectedVals$nPatients
+          )
+        }
+      }
+    )
+    
+    output$togglePredictionPerformance <- shiny::renderUI(
+      {
+        if (input$type != "interaction") {
+          shiny::selectInput(
+            inputId  = "predictionPerformance",
+            label    = "Prediction model AUC",
+            choices  = c(
+              65,
+              75,
+              85
+            ),
+            selected = selectedVals$predictionPerformance
+          )
+        }
+      }
+    )
+    
+    currentScenario <- reactive(
+      {
+        if (input$type != "interaction") {
+          analysisIds %>%
+            filter(
+              type       == input$type,
+              effectSize == input$effect,
+              sampleSize == input$nPatients,
[…]

### 2021-06-09 code/GetGustoData.sh
retrieve gusto data

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+wget --output-document data/raw/gusto.rda -nc http://hbiostat.org/data/gusto.rda

### 2021-06-09 code/GustoPlot.R
add gusto plot

@@ -1,0 +1,207 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(rms)
+library(SmoothHte)
+
+load("data/raw/gusto.rda")
+maxRisk <- .4      # the maximum risk to be plotted
+
+gusto <- gusto %>%
+  tibble() %>%
+  filter(tx != "SK+tPA") %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+prediction <- lrm(
+  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+  data = gusto,
+  maxit = 99
+)
+
+riskLinearPredictor <- predict(
+  prediction,
+  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
+)
+
+gusto <- gusto %>%
+  mutate(
+    riskLinearPredictor = riskLinearPredictor
+  )
+
+constantModel <- fitModelBasedHte(
+  data     = gusto, 
+  settings = createModelBasedSettings(type = "treatment")
+)
+
+stratified <- fitStratifiedHte(
+  data     = gusto,
+  settings = createStratifiedSettings()
+)
+
+linearModel <- fitModelBasedHte(
+  data     = gusto,
+  settings = createModelBasedSettings()
+)
+
+rcs3Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings()
+)
+
+rcs4Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings(nKnots = 4)
+)
+
+rcs5Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings(nKnots = 5)
+)
+
+
+
+plot <- ggplot() +
+  stat_function(
+    data = data.frame(
+      x     = c(0, maxRisk), 
+      label = "constant"
+    ),
+    aes(
+      x     = x, 
+      color = label
+    ),
+    fun  = predictBenefitModelBasedHte,
+    args = list(
+      modelBasedFit = constantModel
+    )
+  ) +
+  geom_pointrange(
+    data = stratified$data,
+    aes(
+      x     = meanRisk, 
+      y     = estimate, 
+      ymin  = lower, 
+      ymax  = upper,
+      color = "stratified"
+    ),
+    key_glyph = "rect"
+  ) +
+  stat_function(
+    data = data.frame(
+      x = c(0, maxRisk), 
[…]

### 2021-06-09 submission/manuscript.rmd
add gusto plot
remove deviations plor and rerun manuscript creation

@@
 linear deviations. Finally, we dropped the assumption of linearity allowing for
-quadratic deviations (Figure \@ref(fig:fig1)).
-
-```{r fig1, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width='50%'}
-knitr::include_graphics(here::here("figures/deviationsManuscript.png"))
-knitr::include_graphics(here::here("figures/deviationsAbsoluteManuscript.png"))
-```
+quadratic deviations.
 
@@
 ## Real data
+```{r gusto, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/gusto.png"))
+```
 

### 2021-06-09 code/GustoPlot.R

@@ -67,7 +67,7 @@ plot <- ggplot() +
   stat_function(
     data = data.frame(
       x     = c(0, maxRisk), 
-      label = "constant"
+      label = "Constant treatment effect"
     ),
     aes(
       x     = x, 
@@ -85,14 +85,14 @@ plot <- ggplot() +
       y     = estimate, 
       ymin  = lower, 
       ymax  = upper,
-      color = "stratified"
+      color = "Stratified"
     ),
     key_glyph = "rect"
   ) +
   stat_function(
     data = data.frame(
       x = c(0, maxRisk), 
-      label = "linear predictor"
+      label = "Linear interaction"
     ),
     aes(
       x     = x,
@@ -169,9 +169,9 @@ plot <- ggplot() +
       "#ffd92f"
     ),
     breaks = c(
-      "constant",
-      "stratified",
-      "linear predictor",
+      "Constant treatment effect",
+      "Stratified",
+      "Linear interaction",
       "RCS 3 knots",
       "RCS 4 knots",
       "RCS 5 knots"
@@ -185,8 +185,8 @@ plot <- ggplot() +
     axis.text.x = ggplot2::element_text(size = 12),
     axis.text.y = ggplot2::element_text(size = 12),
     axis.title = ggplot2::element_text(size = 14),
-    legend.text = element_text(size = 12),
-    legend.position = c(.9, .2)
+    legend.text = element_text(size = 10),
+    legend.position = c(.85, .2)
   )
   
  ggplot2::ggsave(

### 2021-06-09 submission/manuscript.rmd
new set of figures; text not updated

@@
 created using the same logistic regression model, including a constant treatment
-effect odds ratio (OR) of $0.8$. The generated samples of the base-case scenario were
-of size $n=4,250$ ($80\%$ power for the detection of an unadjusted OR of
+effect odds ratio (OR) of $0.8$. The generated samples of the base-case scenario
+were of size $n=4,250$ ($80\%$ power for the detection of an unadjusted OR of
 $0.8$).
@@
 constant treatment effect had the lowest median RMSE, regardless of true
-prediction AUC and sample size (Figure \@ref(fig:rmseconstant)). Linear
-interaction models demonstrated comparable performance. Among the RCS smoothing
-methods, the one fitted with 3 knots always performed best, while the increased
-flexibility achieved when increasing the knots resulted in overfitting and worse
-performance. The adaptive approach under all scenarios performed similar to the
-model with smaller RMSE in all scenarios.
+prediction AUC and sample size. Linear interaction models demonstrated
+comparable performance. Among the RCS smoothing methods, the one fitted with 3
+knots always performed best, while the increased flexibility achieved when
+increasing the knots resulted in overfitting and worse performance. The adaptive
+approach under all scenarios performed similar to the model with smaller RMSE in
+all scenarios.
 
-```{r rmseconstant, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_constant.png"))
+```{r rmsebase, echo=FALSE, fig.cap="RMSE base case", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/rmse_base.png"))
+```
+
+```{r rmseauc, echo=FALSE, fig.cap="AUC 0.85", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/rmse_auc.png"))
+```
+
+```{r rmsesamplesize, echo=FALSE, fig.cap="Sample size 17000", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
 ```
@@
 
-```{r rmsebase, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_base.png"))
-```
-
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
@@
 ```
+
 When focusing on the different scenarios where true treatment-covariate
-interactions were considered all methods had similar RMSE performance, (Figure
-\@ref(fig:rmseInteractions); Panels B, C and D). In case of strong
-treatment-covariate interactions the constant effect model had slightly
-increased RMSE
+interactions were considered all methods had similar RMSE performance. In case
+of strong treatment-covariate interactions the constant effect model had
+slightly increased RMSE
 (`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`])
@@
 (`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`]).
-
-```{r rmseInteractions, echo=FALSE, fig.cap="Root mean squared error of the different methods when focusing on the effect treatment-covariate interactions.
-In all scenarios sample size is fixed to 4250 and the true prediction AUC is 0.75.
-\\textit{Panel A} presents the results for the base case scenario, where no interactions are present.
-\\textit{Panel B} presents the results for the scenario with 4 weak interactions.
-\\textit{Panel C} presents the results for the scenario with mixed interactions (2 weak and 2 strong).
-\\textit{Panel D} presents the results for the scenario with 4 strong interactions.", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_interactions.png"))
-```
 
@@
 
-```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the different methods.
-\\textit{Panel A} presents the results when moderate linear deviations from the base case scenario of constant treatment effects are considered, sample size is 4250 and true prediction AUC is 0.75.
-\\textit{Panel B} presents the results when moderate linear deviations are considered and true prediction AUC is 0.85.
-Sample size is 4250.
-\\textit{Panel C} presents the results when moderate linear deviations are considered and the sample size is 0.75.
-\\textit{Panel D} presents the results when strong linear deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85.
-\\textit{Panel E} presents results for true prediction AUC of 0.85 and \\textit{Panel F} for sample size of 17000 (assuming moderate linear deviations from base case).
-\\textit{Panel G} presents the results when strong quadratic deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85.
-\\textit{Panel H} presents the results for true prediction AUC of 0.85 and \\textit{Panel I} for sample size of 17000 (assuming strong quadratic deviations from base case)", fig.show="hold", out.width = '100%'}
+```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit base case.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/discrimination_base.png"))
@@
 
-```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the different methods.
-\\textit{Panel A} presents the results when moderate linear deviations from the base case scenario of constant treatment effects are considered, sample size is 4250 and true prediction AUC is 0.75.
-\\textit{Panel B} presents the results when moderate linear deviations are considered and true prediction AUC is 0.85.
-Sample size is 4250.
-\\textit{Panel C} presents the results when moderate linear deviations are considered and the sample size is 0.75.
-\\textit{Panel D} presents the results when strong linear deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85.
-\\textit{Panel E} presents results for true prediction AUC of 0.85 and \\textit{Panel F} for sample size of 17000 (assuming moderate linear deviations from base case).
-\\textit{Panel G} presents the results when strong quadratic deviations from the base case are considered, sample size is 4250 and true prediction AUC is 0.85.
-\\textit{Panel H} presents the results for true prediction AUC of 0.85 and \\textit{Panel I} for sample size of 17000 (assuming strong quadratic deviations from base case)", fig.show="hold", out.width = '100%'}
+```{r calibration, echo=FALSE, fig.cap="Calibration for benefit base case.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/calibration_base.png"))

### 2021-06-09 code/CalibrationBase.R

@@ -8,23 +8,14 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(
-  28, 30, 34,
-  37, 39, 43,
-  55, 57, 61
-)
+scenarios <- c(10, 28, 37, 55)
 
 metric <- "calibration"
 titles <- c(
-  "**A.** Moderate linear deviation",
-  "**B.** AUC: 0.85; N: 4250",
-  "**C.** AUC: 0.75; N: 17000",
-  "**D.** Strong linear deviation",
-  "**E.** AUC: 0.85; N: 4250",
-  "**F.** AUC: 0.75; N: 17000",
-  "**G.** Strong quadratic deviation",
-  "**H.** AUC: 0.85; N: 4250",
-  "**I.** AUC: 0.75; N: 17000"
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
+  "**C.** Strong linear deviation",
+  "**D.** Strong quadratic deviation"
 )
 
 metricFile <- paste(metric, "csv", sep = ".")
@@ -77,13 +68,8 @@ pp <- gridExtra::grid.arrange(
   plotList[[2]],
   plotList[[3]],
   plotList[[4]],
-  plotList[[5]],
-  plotList[[6]],
-  plotList[[7]],
-  plotList[[8]],
-  plotList[[9]],
-  nrow = 3,
-  ncol = 3,
+  nrow = 2,
+  ncol = 2,
   left = grid::textGrob(
     expression(
       paste(

### 2021-06-09 code/DiscriminationBase.R

@@ -8,23 +8,14 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(
-  28, 30, 34,
-  37, 39, 43,
-  55, 57, 61
-)
+scenarios <- c(10, 28, 37, 55)
 
 metric <- "discrimination"
 titles <- c(
-  "**A.** Moderate linear deviation",
-  "**B.** AUC: 0.85; N: 4250",
-  "**C.** AUC: 0.75; N: 17000",
-  "**D.** Strong linear deviation",
-  "**E.** AUC: 0.85; N: 4250",
-  "**F.** AUC: 0.75; N: 17000",
-  "**G.** Strong quadratic deviation",
-  "**H.** AUC: 0.85; N: 4250",
-  "**I.** AUC: 0.75; N: 17000"
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
+  "**C.** Strong linear deviation",
+  "**D.** Strong quadratic deviation"
 )
 
 metricFile <- paste(metric, "csv", sep = ".")
@@ -43,7 +34,7 @@ for (i in seq_along(scenarios)) {
     data = tmp,
     metric = metric,
     title = titles[i],
-    limits = c(.52, .62),
+    limits = c(.5, .57),
     pointSize = .5
   ) +
     ggplot2::theme(
@@ -63,13 +54,8 @@ pp <- gridExtra::grid.arrange(
   plotList[[2]],
   plotList[[3]],
   plotList[[4]],
-  plotList[[5]],
-  plotList[[6]],
-  plotList[[7]],
-  plotList[[8]],
-  plotList[[9]],
-  nrow = 3,
-  ncol = 3,
+  nrow = 2,
+  ncol = 2,
   left = "C-statistic for benefit"
 )
   ggplot2::ggsave(

### 2021-06-09 code/GustoPlot.R

@@ -5,7 +5,7 @@ library(rms)
 library(SmoothHte)
 
 load("data/raw/gusto.rda")
-maxRisk <- .4      # the maximum risk to be plotted
+maxRisk <- .3      # the maximum risk to be plotted
 
 gusto <- gusto %>%
   tibble() %>%
@@ -185,8 +185,8 @@ plot <- ggplot() +
     axis.text.x = ggplot2::element_text(size = 12),
     axis.text.y = ggplot2::element_text(size = 12),
     axis.title = ggplot2::element_text(size = 14),
-    legend.text = element_text(size = 10),
-    legend.position = c(.85, .2)
+    legend.text = element_text(size = 9),
+    legend.position = c(.2, .8)
   )
   
  ggplot2::ggsave(

### 2021-06-09 code/RmseAuc.R

@@ -20,10 +20,23 @@ titles <- c(
 
 metricFile <- paste(metric, "csv", sep = ".")
 
+f <- function(x) x * 100
+
 processed <- readr::read_csv(
   file = file.path("data/processed", metricFile)
-) 
-
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
 plotList <- list()
 
 for (i in seq_along(scenarios)) {
@@ -34,7 +47,7 @@ for (i in seq_along(scenarios)) {
     data = tmp,
     metric = metric,
     title = titles[i],
-    limits = c(0, .11),
+    limits = c(0, 11),
     pointSize = .5
   ) +
     ggplot2::theme(
@@ -56,7 +69,16 @@ pp <- gridExtra::grid.arrange(
   plotList[[4]],
   nrow = 2,
   ncol = 2,
-  left = "Root mean squared error"
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
 )
   ggplot2::ggsave(
     file.path("figures", "rmse_auc.tiff"), 

### 2021-06-09 code/RmseBase.R

@@ -8,23 +8,14 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(
-  28, 30, 34,
-  37, 39, 43,
-  55, 57, 61
-)
+scenarios <- c(10, 28, 37, 55)
 
 metric <- "rmse"
 titles <- c(
-  "**A.** Moderate linear deviation",
-  "**B.** AUC: 0.85; N: 4250",
-  "**C.** AUC: 0.75; N: 17000",
-  "**D.** Strong linear deviation",
-  "**E.** AUC: 0.85; N: 4250",
-  "**F.** AUC: 0.75; N: 17000",
-  "**G.** Strong quadratic deviation",
-  "**H.** AUC: 0.85; N: 4250",
-  "**I.** AUC: 0.75; N: 17000"
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
+  "**C.** Strong linear deviation",
+  "**D.** Strong quadratic deviation"
 )
 
 metricFile <- paste(metric, "csv", sep = ".")
@@ -77,17 +68,12 @@ pp <- gridExtra::grid.arrange(
   plotList[[2]],
   plotList[[3]],
   plotList[[4]],
-  plotList[[5]],
-  plotList[[6]],
-  plotList[[7]],
-  plotList[[8]],
-  plotList[[9]],
-  nrow = 3,
-  ncol = 3,
+  nrow = 2,
+  ncol = 2,
   left = grid::textGrob(
     expression(
       paste(
-        "Root mean squared error (x ", 
+        "Root mean squared error (x", 
         10^-2, 
         ")"
       )
@@ -99,15 +85,15 @@ pp <- gridExtra::grid.arrange(
     file.path("figures", "rmse_base.tiff"), 
     plot = pp,
     dpi = 1200,
-    width = 11, 
-    height = 11,
+    width = 8.5, 
+    height = 7,
     compression = "lzw"
   )
   
   ggplot2::ggsave(
     file.path("figures", "rmse_base.png"), 
     plot = pp,
-    width = 11, 
-    height = 11
+    width = 8.5, 
+    height = 7
   )
   

### 2021-06-09 code/RmseSampleSize.R

@@ -20,9 +20,23 @@ titles <- c(
 
 metricFile <- paste(metric, "csv", sep = ".")
 
+f <- function(x) x * 100
+
 processed <- readr::read_csv(
   file = file.path("data/processed", metricFile)
-) 
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
 
 plotList <- list()
 
@@ -34,7 +48,7 @@ for (i in seq_along(scenarios)) {
     data = tmp,
     metric = metric,
     title = titles[i],
-    limits = c(0, .07),
+    limits = c(0, 7),
     pointSize = .5
   ) +
     ggplot2::theme(
@@ -56,7 +70,16 @@ pp <- gridExtra::grid.arrange(
   plotList[[4]],
   nrow = 2,
   ncol = 2,
-  left = "Root mean squared error"
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
 )
   ggplot2::ggsave(
     file.path("figures", "rmse_sample_size.tiff"), 

### 2021-06-10 submission/manuscript.rmd
first draft of GUSTO results

@@
 library(kableExtra)
+library(SmoothHte)
+library(rms)
 
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-rmse_scenarios_10_12_16 <- readr::read_csv(
-  file = here::here(file.path("data/processed", "rmse.csv"))
-) %>%
-  dplyr::filter(scenarioId %in% c(10, 12, 16))
-```
-Under the base case of constant relative treatment effect, the model assuming a
-constant treatment effect had the lowest median RMSE, regardless of true
-prediction AUC and sample size. Linear interaction models demonstrated
-comparable performance. Among the RCS smoothing methods, the one fitted with 3
-knots always performed best, while the increased flexibility achieved when
-increasing the knots resulted in overfitting and worse performance. The adaptive
-approach under all scenarios performed similar to the model with smaller RMSE in
-all scenarios.
+The model including a constant relative treatment effect had the lowest median
+RMSE in scenarios where the base case of true constant relative treatment effect
+(OR = 0.8, N = 4250 and AUC = 0.75) or moderate relative deviations were
+considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
+strong linear and quadratic deviations from the base case the linear interaction
+model performed best (Figure \@ref(fig:rmsebase); Panels B and C). Only in the
+case of strong quadratic deviations models including RCS smoothing (3 knots)
+performed equally well to the linear interaction method. Increasing the number
+of knots in RCS smoothing resulted in higher error rates across all scenarios.
+The performance of the adaptive approach was quite comparable with the best
+performing model in all scenarios.
 
@@
 
-```{r rmseauc, echo=FALSE, fig.cap="AUC 0.85", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_auc.png"))
-```
+When we increased the sample size (N = 17000), in all scenarios the model more
+in agreement with the underlying settings had the lowest error rates. Under the
+base case of constant relative treatment effects (OR = 0.8), the model including
+a constant relative treatment effect had the lowest RMSE (Figure
+\@ref(fig:rmsesamplesize); Panel A). When introducing moderate and strong linear
+deviations the linear interaction model performed best (Figure
+\@ref(fig:rmsesamplesize); Panels B and C). In the case of true strong quadratic
+deviations from the base case of constant relative effects, the more flexible
+RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
+flexibility of RCS smoothing with higher number of knots resulted in overfitting
+and worse performance (Figure \@ref(fig:rmsesamplesize); Panel D).
 
@@
 
-When we introduced deviations from the base case of constant relative treatment
-effects, while keeping fixed both the sample size (N = 4250) and the true
-prediction AUC (0.75) the linear interaction model had the lowest RMSE (Figure
-\@ref(fig:rmsebase); Panels A, D, and G). When these deviations were moderate
-(Figure \@ref(fig:rmsebase); Panel A) the constant treatment effect model had
-comparable performance to the linear interaction model. This can be attributed
-to the fact that such deviations are quite mild and absolute benefits maintain
-similar patterns across baseline risk. On the contrary, when strong quadratic
-deviations were considered the constant effect model's RMSE sharply increased,
-while the more flexible method of RCS smoothing (3 knots) preformed very well
-(Figure \@ref(fig:rmsebase); Panel G). Again, increasing the number of knots
-increased RMSE, indicating overfitting.
-
 When we increased the true prediction AUC to 0.85, models including RCS
 smoothing had the lowest RMSE when strong quadratic deviations from the base
-case of constant relative treatment effects were assumed (Figure
-\@ref(fig:rmsebase); Panel H). However, with milder deviations, the linear
+case of true constant relative treatment effects were assumed (Figure
+\@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
 interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
-being a close second (Figure \@ref(fig:rmsebase); Panels B and E). Increasing the
+being a close second (Figure \@ref(fig:rmseauc); Panels B and C). Increasing the
 number of knots of RCS smoothing resulted in increased RMSE, which was less
-pronounced in the case of strong quadratic deviations. We observed similar
-results when we increased the sample size to 17000, while keeping the true
-prediction AUC constant at 0.75 (Figure \@ref(fig:rmsebase); Panels C, F, and I).
+pronounced in the case of strong quadratic deviations.
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-rmse_scenarios_10_64_66_65 <- readr::read_csv(
-  file = here::here(file.path("data/processed", "rmse.csv"))
-) %>%
-  dplyr::filter(scenarioId %in% c(10, 64, 66, 65))
+```{r rmseauc, echo=FALSE, fig.cap="AUC 0.85", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/rmse_auc.png"))
 ```
-
-When focusing on the different scenarios where true treatment-covariate
-interactions were considered all methods had similar RMSE performance. In case
-of strong treatment-covariate interactions the constant effect model had
-slightly increased RMSE
-(`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`])
-compared to the other methods. The linear interaction model with the risk linear
-predictor had the lowest RMSE
-(`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`]).
 
@@
 ## Real data
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+load(here::here("data/raw/gusto.rda"))
+gusto <- gusto %>%
+  tibble() %>%
+  filter(!is.na(tpa))
+
+treatmentArms <- gusto %>%
+  group_by(tpa) %>%
+  summarise(n = n())
+```
+
+We demonstrate the different methods for individualizing treatment benefits
+using data from `r nrow(gusto)` patients with an acute myocardial infarction
+(MI) included in the GUSTO-I trial.
+`r treatmentArms %>% filter(tpa == 1) %>% select(n)` patients were randomized to
+tissue plasminogen activator (tPA) treatment and
+`r treatmentArms %>% filter(tpa == 0) %>% select(n)` were randomized to
+streptokinase. The outcome of interest was 30-day mortality, recorded for all
+patients.
+
+In line with (Califf 1997; Steyerberg 2000), we fitted a logistic regression
+model with 6 baseline covariates, i.e. age, Killip class, systolic blood
+pressure, heart rate, an indicator of previous MI, and the location of MI, to
+predict 30-day mortality risk. A constant effect of treatment was included in
+the model. When deriving risk predictions for individuals we set the treatment
+indicator to 0. More information on model development can be found in the
+supplement.
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+gusto <- gusto %>%
+  tibble() %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+prediction <- lrm(
+  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+  data = gusto,
+  maxit = 99
+)
+
+riskLinearPredictor <- predict(
+  prediction,
+  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
+)
+
+gusto <- gusto %>%
+  mutate(
+    riskLinearPredictor = riskLinearPredictor
+  )
+
+constantModel <- fitModelBasedHte(
+  data     = gusto, 
+  settings = createModelBasedSettings(type = "treatment")
+)
+
+tmp <- gusto %>%
+  mutate(
+    predictedBenefit = predictBenefitModelBasedHte(
+      p             = plogis(.$riskLinearPredictor), 
+      modelBasedFit = constantModel
+    )
+  )
+
+constantC   <- calculateCForBenefit(tmp)
+constantIci <- calculateCalibrationForBenefit(tmp)
+
+stratified <- fitStratifiedHte(
+  data     = gusto,
+  settings = createStratifiedSettings()
+)
+
+linearModel <- fitModelBasedHte(
+  data     = gusto,
+  settings = createModelBasedSettings()
+)
+
+tmp <- gusto %>%
+  mutate(
+    predictedBenefit = predictBenefitModelBasedHte(
+      p             = plogis(.$riskLinearPredictor), 
+      modelBasedFit = linearModel
+    )
+  )
+
+linearC   <- calculateCForBenefit(tmp)
+linearIci <- calculateCalibrationForBenefit(tmp)
+
+rcs3Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings()
+)
+
+tmp <- gusto %>%
+  mutate(
+    predictedBenefit = predictSmoothBenefit(
+      p          = plogis(riskLinearPredictor), 
+      smoothFit  = rcs3Model
+    )
+  )
+
+rcs3C   <- calculateCForBenefit(tmp)
+rcs3Ici <- calculateCalibrationForBenefit(tmp)
+
+rcs4Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings(nKnots = 4)
+)
+
+tmp <- gusto %>%
+  mutate(
+    predictedBenefit = predictSmoothBenefit(
+      p          = plogis(riskLinearPredictor), 
+      smoothFit  = rcs4Model
+    )
+  )
+
+rcs4C   <- calculateCForBenefit(tmp)
+rcs4Ici <- calculateCalibrationForBenefit(tmp)
+
+rcs5Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings(nKnots = 5)
+)
+
+tmp <- gusto %>%
+  mutate(
+    predictedBenefit = predictSmoothBenefit(
+      p          = plogis(.$riskLinearPredictor), 
+      smoothFit  = rcs5Model
+    )
+  )
+
+rcs5C   <- calculateCForBenefit(tmp)
+rcs5Ici <- calculateCalibrationForBenefit(tmp)
+
+gustoPerformance <- tibble(
+  discrimination = c(
+    constantC, linearC, rcs3C,
+    rcs4C, rcs5C
+  ),
+  calibration = c(
+    constantIci$ici, linearIci$ici, rcs3Ici$ici,
+    rcs4Ici$ici, rcs5Ici$ici
+  ),
+  method = c(
+    "constant treatment effect",
+    "linear interaction",
+    "RCS smoothing with 3 knots",
+    "RCS smoothing with 4 knots",
+    "RCS smoothing with 5 knots"
+  )
+)
+
+cBenefit <- gustoPerformance %>% arrange(discrimination)
+
+```
+
+We used the risk linear predictor to fit the the proposed methods under study
+for individualizing absolute benefit predictions. All methods had quite
+comparable results, in the sense that we predicted increasing benefits for
+patients with higher baseline risk predictions. In terms of c-for benefit,
+validated internally, all models had quite comparable performance ranging from
+0.519 (RCS smoothing with 3 knots) to 0.542 (RCS smoothing with 4 knots).
+Similar conclusions could be drawn in terms of ICI-for-benefit which ranged
+from 0.0039 (linear interaction approach) to 0.0053 (RCS smoothing with 4
+knots).
+
+The adaptive approach picked the
+model with RCS smoothing with 4 knots, which had quite comparable performance to
+the smooth fit with 5 knots. However, for lower baseline risk these 2 models
+predicted implausible benefits and maybe should be avoided when applying such
+models in practice. The linear interaction model and the model with RCS
+smoothing (3 knots) made very similar predictions, while also followed quite
+closely the evolution of the stratified estimates. In this case, we observed
+that even using a simple model with a constant relative treatment effect can
+result in very similar benefit predictions as other more complex approaches.
+
 ```{r gusto, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}

### 2021-06-10 code/GustoPlot.R

@@ -186,7 +186,7 @@ plot <- ggplot() +
     axis.text.y = ggplot2::element_text(size = 12),
     axis.title = ggplot2::element_text(size = 14),
     legend.text = element_text(size = 9),
-    legend.position = c(.2, .8)
+    legend.position = c(.17, .85)
   )
   
  ggplot2::ggsave(

### 2021-06-15 submission/manuscript.rmd
fixing typos

@@
   - \usepackage[left]{lineno}
+  - \usepackage{amssymb}
+  - \usepackage{bm}
+  - \usepackage{booktabs}
+  - \newcommand\given[1][]{\:#1\vert\:}
 editor_options: 
@@
 
-In the simulated datasets of the base-case scenario treatment was allocated at
-random using a 50/50 split. For each patient we simulated $8$ baseline
-covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$.
-Outcomes for patients in the control arm were generated from a logistic
-regression model including all baseline covariates. Coefficient values were
-such, so that the prediction model had an AUC of $0.75$ and an event rate of
-$20\%$ in the control arm was achieved. Outcomes in the treatment arm were
-created using the same logistic regression model, including a constant treatment
-effect odds ratio (OR) of $0.8$. The generated samples of the base-case scenario
-were of size $n=4,250$ ($80\%$ power for the detection of an unadjusted OR of
-$0.8$).
+In the simulated datasets based on the base-case scenario treatment was
+allocated at random using a 50/50 split. For each patient we simulated $8$
+baseline covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and
+$x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes for patients in the control arm were
+generated from a logistic regression model including all baseline covariates.
+Coefficient values were such, so that the prediction model had an AUC of $0.75$
+and an event rate of $20%$ in the control arm was achieved. Outcomes in the
+treatment arm were created using the same logistic regression model, including a
+constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
+base-case scenario were of size $n=4,250$ ($80%$ power for the detection of an
+unadjusted OR of $0.8$).
 
@@
 the first diagonal on the *log-odds* scale, with distance equal to $\log(\text{OR})$. 
-We assessed the effect of stronger and absent relative treatment effects
-($\text{OR}=0.5$ or $\text{OR}=1$). We also relaxed the assumption that the
-line should be parallel to the diagonal, considering moderate and stronger
-linear deviations. Finally, we dropped the assumption of linearity allowing for
-quadratic deviations.
+(See Figure XX of the supplementary material). We assessed the effect of
+stronger and absent relative treatment effects ($\text{OR}=0.5$ or
+$\text{OR}=1$). We also relaxed the assumption that the line should be parallel
+to the diagonal, considering moderate and stronger linear deviations. Finally,
+we dropped the assumption of linearity allowing for quadratic deviations.
 
@@
 considered this approach as a reference, expecting it to perform worse than the
-other candidates, as its objective is not individual benefit prediction.
+other candidates, as its objective is not individualized benefit prediction.
 
@@
 fits a logistic regression using treatment, risk linear predictor and their
-interaction within each treatment arm. In this case, absolute benefit is
-estimated from
+interaction. In this case, absolute benefit is estimated from
 $\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)$.
@@
 
+Assuming that $\tau(\bm{x})=E\{y\given\bm{x}, t_x=0\} - E\{y\given\bm{x},t_x=1\}$ is the true benefit for each patient and $\hat{\tau}(x)$ is the estimated benefit from a method under study, the ideal loss function to use for the considered methods would be the unobservable root mean squared error $E\big\{(\hat{\tau} - \tau)^2\given \bm{x}\big\}$.
+However, in simulations the true patient-level benefit is available.
+Therefore, we estimated RMSE from
+$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$
+
+We also compared the predictive performance of the methods under study. We
+assessed discrimination using the c-for-benefit statistic described in
+[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
+predicted benefit and then are matched 1:1, dropping patients in the larger
+treatment arm without a pair. We define *observed* pair-specific treatment
+benefit as the difference of observed outcomes between the untreated and the
+treated patient of each pair. Pair-specific *predicted* benefit is defined as
+the average of predicted benefits within each pair. Then, c-for-benefit is
+defined as the probability that from two randomly chosen matched patient pairs
+with unequal observed benefit, the pair with greater observed benefit also has a
+higher predicted benefit.
+
+We evaluated calibration in a similar manner, using the integrated calibration
+index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
+benefit, observed benefits are regressed on the predicted benefits using a
+locally weighted scatterplot smoother (loess). The ICI is the area between the
+loess fit and the diagonal, with values closer to $0$ indicating better
+calibration.
+
 # Results
@@
 RMSE in scenarios where the base case of true constant relative treatment effect
-(OR = 0.8, N = 4250 and AUC = 0.75) or moderate relative deviations were
+(OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
 considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
@@
 The performance of the adaptive approach was quite comparable with the best
-performing model in all scenarios.
+performing model in each scenario.
 
@@
 
-When we increased the sample size (N = 17000), in all scenarios the model more
+When we increased the sample size (N = `r 17000`), in all scenarios the model more
 in agreement with the underlying settings had the lowest error rates. Under the
@@
 When we increased the true prediction AUC to 0.85, models including RCS
-smoothing had the lowest RMSE when strong quadratic deviations from the base
-case of true constant relative treatment effects were assumed (Figure
+smoothing had the lowest RMSE in the presence of strong quadratic deviations
+from the base case of true constant relative treatment effects (Figure
 \@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
@@
 scenarios where linear and quadratic deviations from the base case of constant
-treatment effect were considered (Figure \@ref(fig:discrimination)). However,
-models including a linear interaction with the risk linear predictor tended to
-present much lower variability compared to all other model-based and smoothing
-approaches. We also observed an increasing trend of discrimination for benefit
-variability with increasing number of restricted cubic spline knots in all
-scenarios. This is evidence that the increased flexibility of these methods
-often led to overfitting.
+treatment effect were considered (N=`r 4250`; AUC = `r 0.75`). However, models
+including a linear interaction with the risk linear predictor tended to present
+much lower variability compared to all other model-based and smoothing
+approaches (Figure \@ref(fig:discrimination)). We also observed an increasing
+trend of discrimination for benefit variability with increasing number of
+restricted cubic spline knots in all scenarios.
 
@@
 
-In line with (Califf 1997; Steyerberg 2000), we fitted a logistic regression
+In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a logistic regression
 model with 6 baseline covariates, i.e. age, Killip class, systolic blood

### 2021-06-15 code/GetGustoData.sh

@@ -1,3 +1,3 @@
 #!/bin/bash
 
-wget --output-document data/raw/gusto.rda -nc http://hbiostat.org/data/gusto.rda
+[[ ! -f "data/raw/gusto.rda" ]] && wget --output-document data/raw/gusto.rda -nc http://hbiostat.org/data/gusto.rda || echo -e "Nothing to be done"

### 2021-06-15 submission/manuscript.rmd
proof reading sections

@@
 comparable results, in the sense that we predicted increasing benefits for
-patients with higher baseline risk predictions. In terms of c-for benefit,
-validated internally, all models had quite comparable performance ranging from
+patients with higher baseline risk predictions. In terms of c-for benefit 
+(validated internally) all models had quite comparable performance ranging from
 0.519 (RCS smoothing with 3 knots) to 0.542 (RCS smoothing with 4 knots).

### 2021-06-15 submission/manuscript.rmd
add proper figure captions

@@
 
-```{r rmsebase, echo=FALSE, fig.cap="RMSE base case", fig.show="hold", out.width = '100%'}
+```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_base.png"))
@@
 
-```{r rmsesamplesize, echo=FALSE, fig.cap="Sample size 17000", fig.show="hold", out.width = '100%'}
+```{r rmsesamplesize, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
@@
 
-```{r rmseauc, echo=FALSE, fig.cap="AUC 0.85", fig.show="hold", out.width = '100%'}
+```{r rmseauc, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_auc.png"))
@@
 
-```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit base case.", fig.show="hold", out.width = '100%'}
+```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/discrimination_base.png"))
@@
 
-```{r calibration, echo=FALSE, fig.cap="Calibration for benefit base case.", fig.show="hold", out.width = '100%'}
+```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/calibration_base.png"))
@@
 
-```{r gusto, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
+```{r gusto, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/gusto.png"))
 ```
+
+# Discussion
 

### 2021-06-16 submission/manuscript.rmd
first complete version of manuscript

@@
 
-In the simulated datasets based on the base-case scenario treatment was
+In the simulated datasets based on the base case scenario treatment was
 allocated at random using a 50/50 split. For each patient we simulated $8$
@@
 Coefficient values were such, so that the prediction model had an AUC of $0.75$
-and an event rate of $20%$ in the control arm was achieved. Outcomes in the
+and an event rate of $20\%$ in the control arm was achieved. Outcomes in the
 treatment arm were created using the same logistic regression model, including a
 constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
-base-case scenario were of size $n=4,250$ ($80%$ power for the detection of an
+base case scenario were of size $n=4,250$ ($80\%$ power for the detection of an
 unadjusted OR of $0.8$).
@@
 We evaluated the effect of sample size considering additional scenarios with
-sample sizes of $1,064$ and $17,000$. We also evaluated the effect of prediction
+sample sizes of `r 1064` and `r 17000`. We also evaluated the effect of prediction
 performance, adjusting the baseline covariate coefficients, so that AUC values
@@
 
-When focusing on calibration for benefit, the linear interaction model had the
-lowest median ICI for benefit in the majority of the scenarios except for the
-scenarios where moderate linear deviations from the base case were considered.
-In that case constant treatment effect models demonstrated the best performance,
-very comparable to the linear interaction model's performance, nonetheless
-(Figure \@ref(fig:calibration); Panels A, B, and C).
+When focusing on calibration, the linear interaction model had the lowest median
+ICI for benefit in the majority of scenarios except for the scenarios where
+moderate linear deviations from the base case were considered. In that case
+constant treatment effect models demonstrated the best performance, very
+comparable to the linear interaction model's performance, nonetheless (Figure
+\@ref(fig:calibration); Panels A, B, and C).
 
@@
 comparable results, in the sense that we predicted increasing benefits for
-patients with higher baseline risk predictions. In terms of c-for benefit 
+patients with higher baseline risk predictions. In terms of c-for-benefit 
 (validated internally) all models had quite comparable performance ranging from
@@
 models in practice. The linear interaction model and the model with RCS
-smoothing (3 knots) made very similar predictions, while also followed quite
+smoothing (3 knots) made very similar predictions and also followed quite
 closely the evolution of the stratified estimates. In this case, we observed
@@
 # Discussion
+We have carried out a simulation study comparing different approaches for
+individualized benefit prediction using baseline risk as the reference. These
+methods included a constant treatment effect model, a model including a simple
+linear interaction of the risk linear predictor with treatment, and models using
+smoothing with restricted cubic splines with 3, 4, and 5 knots. Performance was
+assessed evaluating error rates using RMSE and benefit prediction performance
+using some novel methods for evaluating discrimination and calibration for
+benefit.
+
+The linear interaction model proved to be flexible enough to adequately predict
+benefit under a wide range of scenarios. It had lower error rates and usually
+the highest c-for-benefit with the lower variability across simulation
+replications. Its benefit predictions were also well-calibrated across the
+majority of the simulation settings. Models with RCS smoothing with 3 knots
+performed very similarly, only outperformig the linear interaction model in the
+presence of strong quadratic deviations from the base case scenario of constant
+relative treatment effects. These simpler methods proved to have quite robust
+performance across all scenarios, which was not the case for more flexible
+approaches.
+
+Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry and this
+often led to overfitting with smaller sample sizes. This resulted in higher
+RMSE, increased variability of discrimination for benefit and worse-calibrated
+benefit predictions across simulation replications. Even with larger sample
+sizes and strong quadratic deviations from the base case scenario of constant
+relative treatment effects, these more flexible RCS smoothing methods did not
+outperform simpler RCS smoothing with 3 knots. More flexible approaches usually
+do not come with any added value, unless we expect extreme heterogeneity in
+treatment effects.
+
+The constant treatment effect model, despite having adequate performance even in
+the presence of weak treatment effect heterogeneity on the relative scale, it
+quickly broke down with stronger deviations from the base case scenario of
+constant effects. In these cases, the stratified approach generally had lower
+error rates compared to the constant treatment effect model. In concept, the
+stratified approach lies between the constant effects model and smoother
+approaches, only assuming constant treatment effects within strata of predicted
+risk and, therefore, is more sensitive to treatment effect heterogeneity.
+However, these stepwise benefit estimates can only be used as a demonstration of
+treatment effect heterogeneity, identifying patient subgroups where treatment
+benefits are expected to be higher or lower, and should not be used to make
+individualized absolute benefit predictions.
+
+Increasing the true prediction AUC by increasing the predictor coefficients of
+the true risk model reduced RMSE for all methods, similar to increasing the
+sample size. Higher AUC translates in higher variability of predicted risks,
+which, in turn, allows the considered methods to better follow the evolution of
+treatment benefit. Higher prediction AUC also led to higher absolute values of
+c-for-benefit. Even though, model performance appears to be very important for
+the performance of these methods, prediction model development was outside the
+scope of this work and has already been studied quite extensively
+[@vanKlaveren2019; @Burke2014; @Abadie2018].
+
+Risk-based approaches to predictive HTE assume that benefit is a function of
+baseline risk. A limitation of our study is that this was assumed to be the case
+in the majority of the simulation scenarios. Even though our main conclusions
+did not change when simulating true treatment-covariate interactions (results in
+supplement), RMSE had higher values. The linear interaction model under these
+scenarios again had the best overall performance under different sizes of the
+interaction effects. This can be explained by the fact that linear interactions
+with treatment were simulated, thus favoring the linear interaction model. In
+studies evaluating prediction model development, the underlying outcome
+generation process sometimes proved to affect performance of certain methods
+[@Austin2021]. This might be the case here as well and should be explored in
+more extensive future simulation studies.
+
+Recent years have seen an increased interest in predictive HTE approaches
+focusing on individualized benefit predictions. In our simulations we only
+focused on risk-based methods, using baseline risk as a reference in a two-stage
+approach to individualizing benefit predictions. However, there is a plethora of
+different methods, ranging from treatment effect modeling to tree-based
+approaches available in more recent literature [@Athey2019; @Lu2018;
+@Wager2018]. Simulations are also needed to assess relative performance and
+define the settings where these break down or outperform each other.
+
+In conclusion, we carried out a simulation study comparing different risk-based
+approaches to individualizing benefit predictions. Simpler models including a
+linear interaction with the risk linear predictor or RCS smoothing with 3 knots
+had adequate performance under a wide range scenarios. An adaptive approach based
+on selecting among the considered methods using AIC had satisfactory
+performance.
 

### 2021-06-16 submission/supplement.rmd
create supplement

--- 2021-06-09 extras/protocol/protocol.rmd
@@
 ---
-title: "Protocol"
+title: "Supplementary material"
 output:
     bookdown::pdf_document2: default
-    bookdown::word_document2: default
 geometry: margin=1.0in
@@
 bibliography: references.bib
-csl: jamia.csl
+csl: jce.csl
 ---
@@
 
-# Introduction
-Risk-based HTE analyses typically result in risk stratum specific estimates of treatment effects.
-Different approaches to individualize treatment effects have been proposed, but it is unclear which approach performs best under different circumstances.
-We aim  to compare different approaches to  estimating individualized treatment effects.
-
-For all patients we observe covariates $x_1,\dots,x_8$, of which $4$ are continuous and $4$ are binary. More specifically,
+# Simulation settings
+For all patients we observe covariates $x_1,\dots,x_8$, of which $4$ are
+continuous and $4$ are binary. More specifically,
 
@@
 
-# Settings
 ## Base-case scenario
@@
 
-# Risk modeling
+## Risk modeling
 Merging treatment arms, we develop prediction models including a constant relative treatment effect:
@@
 
-# Approaches to individualize benefit predictions
+## Approaches to individualize benefit predictions
 
-## Risk stratification
+### Risk stratification
 Derive a prediction model using the same approach as above and divide the population in equally sized risk-based subgroups.
@@
 
-## Constant treatment effect
+### Constant treatment effect
 Assuming a constant relative treatment effect, fit the adjusted model in \@ref(eq:risk).
@@
 
-## Linear interaction
+### Linear interaction
 The assumption of constant relative treatment effect is relaxed modeling a linear interaction of treatment with the risk linear predictor: 
@@
 
-## Non-linear interaction
+### Non-linear interaction
 Finally, we drop the linearity assumption and predict absolute benefit by taking the difference between smooth fits, separately derived in each treatment arm:
@@
 
-# Evaluation
-Assuming that $\tau(\bm{x})=E\{y\given\bm{x}, t_x=0\} - E\{y\given\bm{x},t_x=1\}$ is the true benefit for each patient and $\hat{\tau}(x)$ is the estimated benefit from a method under study, the ideal loss function to use for the considered methods would be the unobservable root mean squared error $E\big\{(\hat{\tau} - \tau)^2\given \bm{x}\big\}$.
-However, in simulations the true patient-level benefit is available.
-Therefore, we will estimate the RMSE from
-$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$
-We also compare the predictive performance of the methods under study. We assess
-discrimination using the c-for-benefit statistic described in
-[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
-predicted benefit and then are matched 1:1, dropping patients in the larger
-treatment arm without a pair. We define *observed* pair-specific treatment
-benefit as the difference of observed outcomes between the untreated and the
-treated patient of each pair. Pair-specific *predicted* benefit is defined as
-the average of predicted benefits within each pair. Then, c-for-benefit is
-defined as the probability that from two randomly chosen matched patient pairs
-with unequal observed benefit, the pair with greater observed benefit also has a
-higher predicted benefit.
-
-We evaluate calibration in a similar manner, using the integrated calibration
-index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
-benefit, observed benefits are regressed on the predicted benefits using a
-locally weighted scatterplot smoother (loess). The ICI is the area between the
-loess fit and the diagonal, with values closer to $0$ indicating better
-calibration.
-
-In all cases, the evaluation metrics are calculated in a large simulated
-super-population of 500,000 patients.
-
-# References
-\setlength{\parindent}{-0.25in}
-\setlength{\leftskip}{0.25in}
-\noindent
-<div id="refs"></div>
-\setlength{\parindent}{0in}
-\setlength{\leftskip}{0in}
-\noindent
-
-\newpage
-# Supplement
 ```{r paok, echo=FALSE, eval=TRUE, warning=FALSE}

### 2021-06-17 submission/supplement.rmd
add interaction plots

@@
 levels of these deviations. The settings for these deviations are defined in
-supplementary Table \@ref(tab:tab1) and result in the effects of Figure
+Table \@ref(tab:tab1) and result in the effects of Figure
 \@ref(fig:fig1).
@@
 
-We consider three different approaches to smoothing:
-
-- Loess
-- Restricted cubic splines
-- Local likelihood
-
-```{r paok, echo=FALSE, eval=TRUE, warning=FALSE}
-table <- read_csv(
+```{r tab1, echo=FALSE, eval=TRUE, warning=FALSE}
+ asd <- read_csv(
   here("data/processed/analysisIds.csv"),
@@
 kableExtra::kable(
-  table,
+  asd,
   format = "latex",
@@
   booktabs    = TRUE,
+  caption     = "",
   col.names   = c(
@@
 ```
+
+# Results in scenarios with interactions
+
+```{r test, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/rmse_interactions.png"))
+```
+
+```{r test2, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/discrimination_interactions.png"))
+```
+
+```{r test4, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
+knitr::include_graphics(here::here("figures/calibration_interactions.png"))
+```

### 2021-06-21 submission/manuscript.rmd
arxiv style

@@
 title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
+abstract: |
+  **Objective:** Compare different risk-based approaches to estimating
+  individualized treatment effects with simulations of the RCT setting. **Study
+  Design and Setting:** We compared different methods for predicting absolute
+  benefit on the basis of an available prognostic index (PI) for risk: A model
+  with the PI and a constant relative treatment effect; models including an
+  interaction of treatment assignment with the PI; 4 quarters of the PI; nonlinear
+  transformations of the PI with increasing flexibility (restricted cubic splines
+  with 3, 4 and 5 knots, respectively); an adaptive method, selecting the optimal
+  model based on Akaike’s Information Criterion. Starting from a base case
+  scenario (sample size 4,250, average treatment effect odds ratio 0.8, AUC of the
+  PI 0.75), we considered diverse assumptions of the data generation process of
+  the “true” treatment effect, introducing linear and quadratic interactions of
+  the PI with treatment. We analyzed the effect of varying sample size and
+  discriminative ability of the PI. We evaluated performance using root mean
+  squared error, discrimination for benefit and calibration for benefit. **Results:**
+  The model including a linear interaction of the risk linear predictor with
+  treatment had adequate performance that was robust under the majority of the
+  simulation scenarios. Methods using restricted cubic splines required larger
+  sample sizes and higher AUC of the PI to achieve adequate performance. The
+  adaptive approach’s performance was comparable to the performance of the best
+  model in each scenario. **Conclusion:** In most cases using a model just
+  including a linear interaction of the PI with treatment
+  adequately predicts absolute benefit.
+authors:
+  - name: Alexandros Rekkas
+    department: Department of Medical Informatics
+    affiliation: Erasmus Medical Center
+    location: Rotterdam, The Netherlands
+  - name: Peter R. Rijnbeek
+    department: Department of Medical Informatics
+    affiliation: Erasmus Medical Center
+    location: Rotterdam, The Netherlands
+  - name: Ewout W. Steyerberg
+    department: Department of Biomedical Data Sciences
+    affiliation: Leiden University Medical Center
+    location: Leiden, The Netherlands
+  - name: David van Klaveren
+    department: Department of Public Health
+    affiliation: Erasmus Medical Center
+    location: Rotterdam, The Netherlands
+keywords:
+  - test
+  - test that
 output: 
-    bookdown::pdf_document2: default
-    bookdown::word_document2:
-        reference_docx: reference.docx
+  rticles::arxiv_article:
+    keep_tex: true
 geometry: margin=1.0in
+date: false
 toc: false
@@
   - \newcommand\given[1][]{\:#1\vert\:}
+  - \date{}
 editor_options: 
@@
 ```
-\thispagestyle{empty}
-\vspace{35mm}
-
-Alexandros Rekkas${^1}$, Peter R. Rijnbeek${^1}$, ..., Ewout W. Steyerberg${^2}$, 
-David van Klaveren${^3}$
-
-\vspace{40mm}
-
-$^1$ Department of Medical Informatics, Erasmus University Medical Center,
-Rotterdam, Netherlands
-
-$^2$ Department of Biomedical Data Sciences, Leiden University Medical Center,
-Leiden, Netherlands
-
-$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam,
-Netherlands
-
-\newpage
-\newpage
-\thispagestyle{empty}
-# Abstract {-}
-\singlespacing 
-**Objective:** Simulation study to compare different risk-based approaches to
-estimating individualized treatment effects within the RCT setting. **Study
-Design and Setting:** Starting from a base case scenario that assumes a true
-constant treatment effect, we considered a total of 66 scenarios for introducing
-non-constant effects and evaluating methods under different sample sizes and
-baseline risk prediction performance. We compared 7 methods for predicting
-absolute benefit: A constant treatment effect model, a risk stratified approach,
-a model including a linear interaction of the baseline risk linear predictor
-with treatment, 3 restricted cubic spline smoothing models of increasing
-flexibility (3, 4 and 5 knots) and an adaptive model selection method based on
-Akaike's Information Criterion. We evaluated performance using root mean squared
-error, discrimination for benefit and calibration for benefit (i.e., observed
-vs. predicted risk difference in treated vs. untreated). **Results:** The model
-including a linear interaction of the risk linear predictor with treatment had
-adequate performance that was robust under the majority of the simulation
-scenarios. Methods using restricted cubic spline smoothing required larger
-sample sizes and higher prediction AUC to achieve adequate performance. The
-adaptive approach's performance was comparable to the performance of the best
-model in each scenario. **Conclusion:** In most cases using a model just
-including a linear interaction of the risk linear predictor with treatment
-adequately predicts absolute benefit.
-
-\newpage 
-\doublespacing
 
 \linenumbers
-
 # Introduction
@@
 
-In the simulated datasets based on the base case scenario treatment was
-allocated at random using a 50/50 split. For each patient we simulated $8$
-baseline covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and
-$x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes for patients in the control arm were
-generated from a logistic regression model including all baseline covariates.
-Coefficient values were such, so that the prediction model had an AUC of $0.75$
-and an event rate of $20\%$ in the control arm was achieved. Outcomes in the
-treatment arm were created using the same logistic regression model, including a
-constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
-base case scenario were of size $n=4,250$ ($80\%$ power for the detection of an
-unadjusted OR of $0.8$).
+In the simulated datasets based treatment was allocated at random using a 50/50
+split. For each patient we simulated $8$ baseline covariates, where
+$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes
+for patients in the control arm were generated from a logistic regression model
+including all baseline covariates. Coefficient values were such, so that the
+prediction model had an AUC of $0.75$ and an event rate of $20%$ in the control
+arm was achieved. Under the base case scenario, outcomes in the treatment arm
+were created using the same logistic regression model, including a constant
+treatment odds ratio (OR) of $0.8$. The generated samples of the base case
+scenario were 4,250 ($80\%$ power for the detection of an unadjusted OR of
+$0.8$).
 
@@
 
-A true logistic regression model with a constant treatment effect (constant OR)
-implies that outcome risk in the treatment arm is a straight line parallel to
-the first diagonal on the *log-odds* scale, with distance equal to $\log(\text{OR})$. 
-(See Figure XX of the supplementary material). We assessed the effect of
-stronger and absent relative treatment effects ($\text{OR}=0.5$ or
-$\text{OR}=1$). We also relaxed the assumption that the line should be parallel
-to the diagonal, considering moderate and stronger linear deviations. Finally,
-we dropped the assumption of linearity allowing for quadratic deviations.
-
-We also considered scenarios with treatment-covariate interactions. These
-scenarios include 4 weak interactions
+We simulated binary outcomes in the control arm based on 
+$\text{expit}(PI)$, where 
+$\text{expit}(x)= \frac{e^x}{1 + e^x}$. 
+In the treatment arm outcomes where generated from $\text{expit}(lp_1)$ with
+$$ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,$$
+where the coefficients $\gamma_0, \gamma_1$ and $\gamma_2$ along with the
+centering constant $c$ were set for each simulation scenario. In this way we
+were able to assess a wide variety of scenarios, ranging from true constant
+relative treatment effects to moderate and strong linear and quadratic
+deviations. We also considered scenarios with treatment-covariate interactions.
+These scenarios include 4 weak interactions
 ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
@@
 benefit as the difference of observed outcomes between the untreated and the
-treated patient of each pair. Pair-specific *predicted* benefit is defined as
+treated patient of each pair. Pair-specifi *predicted* benefit is defined as
 the average of predicted benefits within each pair. Then, c-for-benefit is
@@
 (OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
-considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
+considered (Figure \ref{fig:rmsebase}; Panel A). However, when we considered
 strong linear and quadratic deviations from the base case the linear interaction
-model performed best (Figure \@ref(fig:rmsebase); Panels B and C). Only in the
+model performed best (Figure \ref{fig:rmsebase}; Panels B and C). Only in the
 case of strong quadratic deviations models including RCS smoothing (3 knots)
@@
 a constant relative treatment effect had the lowest RMSE (Figure
-\@ref(fig:rmsesamplesize); Panel A). When introducing moderate and strong linear
+\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
 deviations the linear interaction model performed best (Figure
-\@ref(fig:rmsesamplesize); Panels B and C). In the case of true strong quadratic
+\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong quadratic
 deviations from the base case of constant relative effects, the more flexible
@@
 flexibility of RCS smoothing with higher number of knots resulted in overfitting
-and worse performance (Figure \@ref(fig:rmsesamplesize); Panel D).
+and worse performance (Figure \ref{fig:rmsesamplesize}; Panel D).
 
@@
 from the base case of true constant relative treatment effects (Figure
-\@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
+\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
 interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
-being a close second (Figure \@ref(fig:rmseauc); Panels B and C). Increasing the
+being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
 number of knots of RCS smoothing resulted in increased RMSE, which was less
@@
 much lower variability compared to all other model-based and smoothing
-approaches (Figure \@ref(fig:discrimination)). We also observed an increasing
+approaches (Figure \ref{fig:discrimination}). We also observed an increasing
 trend of discrimination for benefit variability with increasing number of
@@
 comparable to the linear interaction model's performance, nonetheless (Figure
-\@ref(fig:calibration); Panels A, B, and C).
+\ref{fig:calibration}; Panels A, B, and C).
 

### 2021-06-21 submission/manuscript.tex
arxiv style

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{graphicx}

\title{Smooth risk-based predictive approaches to treatment effect
heterogeneity: A simulation study}

\author{
    Alexandros Rekkas
   \\
    Department of Medical Informatics \\
    Erasmus Medical Center \\
  Rotterdam, The Netherlands \\
  \texttt{} \\
   \And
    Peter R. Rijnbeek
   \\
    Department of Medical Informatics \\
    Erasmus Medical Center \\
  Rotterdam, The Netherlands \\
  \texttt{} \\
   \And
    Ewout W. Steyerberg
   \\
    Department of Biomedical Data Sciences \\
    Leiden University Medical Center \\
  Leiden, The Netherlands \\
  \texttt{} \\
   \And
    David van Klaveren
   \\
    Department of Public Health \\
    Erasmus Medical Center \\
  Rotterdam, The Netherlands \\
  \texttt{} \\
  }

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\usepackage{amsmath}
\doublespacing
\usepackage[left]{lineno}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\newcommand\given[1][]{\:#1\vert\:}
\date{}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\def\tightlist{}

\begin{abstract}
\textbf{Objective:} Compare different risk-based approaches to
estimating individualized treatment effects with simulations of the RCT
setting. \textbf{Study Design and Setting:} We compared different
methods for predicting absolute benefit on the basis of an available
prognostic index (PI) for risk: A model with the PI and a constant
relative treatment effect; models including an interaction of treatment
assignment with the PI; 4 quarters of the PI; nonlinear transformations
of the PI with increasing flexibility (restricted cubic splines with 3,
4 and 5 knots, respectively); an adaptive method, selecting the optimal
model based on Akaike's Information Criterion. Starting from a base case
scenario (sample size 4,250, average treatment effect odds ratio 0.8,
AUC of the PI 0.75), we considered diverse assumptions of the data
generation process of the ``true'' treatment effect, introducing linear
and quadratic interactions of the PI with treatment. We analyzed the
effect of varying sample size and discriminative ability of the PI. We
evaluated performance using root mean squared error, discrimination for
benefit and calibration for benefit. \textbf{Results:} The model
including a linear interaction of the risk linear predictor with
treatment had adequate performance that was robust under the majority of
the simulation scenarios. Methods using restricted cubic splines
required larger sample sizes and higher AUC of the PI to achieve
adequate performance. The adaptive approach's performance was comparable
to the performance of the best model in each scenario.
\textbf{Conclusion:} In most cases using a model just including a linear
interaction of the PI with treatment adequately predicts absolute
benefit.
\end{abstract}

\keywords{
    test
   \and
    test that
  }

\linenumbers

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Within the setting of patient-centered outcomes research, predictive
approaches for assessing heterogeneity of treatment effects (HTE) aim at
the development of models predicting either individualized effects or
which of two (or more) treatments is better for an individual {[}1{]}.
In prior work, we divided such methods in three broader categories based
on the reference class used for defining patient similarity when making
individualized predictions or recommendations {[}2{]}. Risk-modeling
approaches use prediction of baseline risk as the reference; treatment
effect modeling approaches also model treatment-covariate interactions,
in addition to risk factors; optimal treatment regime approaches focus
on developing treatment assignment rules and therefore rely heavily on
modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers in the form of
treatment-covariate interactions often led to miscalibrated predictions
of benefit, while risk-based methods proved quite robust {[}5{]}. Most
often, risk-modeling approaches are carried out in two steps: first a
risk prediction model is developed externally or internally on the
entire RCT population, ``blinded'' to treatment; then the RCT population
is stratified using this prediction model to evaluate risk-based
treatment effect variation {[}6{]}. However, even though estimates at
the risk subgroup level are accurate, this does not apply on the
individual level, especially for patients with predicted risk at the
boundaries of the risk intervals. Therefore, the risk-stratified
approach should be used for exploring and presenting an overview of HTE,
while inferences on the individual level should be made with caution.

We aimed to provide an overview of methods that can be used to move from
a risk-stratified approach to a continuous one using common smoothing
techniques. These methods extend the risk-based framework of predictive
HTE analyses to allow predictions on the individual level, within the
RCT setting. We carried out a simulation study to compare the
performance of these methods under different settings of increasing
non-linearity of treatment effects. Finally, we carried out an
application on real data as a demonstration of the considered
techniques.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{simulation-scenarios}{%
\subsection{Simulation scenarios}\label{simulation-scenarios}}

In the simulated datasets based treatment was allocated at random using
a 50/50 split. For each patient we simulated \(8\) baseline covariates,
where \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Outcomes for patients in the control arm were generated from a logistic
regression model including all baseline covariates. Coefficient values
were such, so that the prediction model had an AUC of \(0.75\) and an
event rate of \(20%
\) in the control arm was achieved. Under the base case scenario,
outcomes in the treatment arm were created using the same logistic
regression model, including a constant treatment odds ratio (OR) of
\(0.8\). The generated samples of the base case scenario were 4,250
(\(80\%\) power for the detection of an unadjusted OR of \(0.8\)).

We evaluated the effect of sample size considering additional scenarios
with sample sizes of 1,064 and 17,000. We also evaluated the effect of
prediction performance, adjusting the baseline covariate coefficients,
so that AUC values of \(0.65\) and \(0.80\) were achieved when
validating in a simulated dataset of \(500,000\) patients.

We simulated binary outcomes in the control arm based on
\(\text{expit}(PI)\), where \(\text{expit}(x)= \frac{e^x}{1 + e^x}\). In
the treatment arm outcomes where generated from \(\text{expit}(lp_1)\)
with \[ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,\] where the
coefficients \(\gamma_0, \gamma_1\) and \(\gamma_2\) along with the
centering constant \(c\) were set for each simulation scenario. In this
way we were able to assess a wide variety of scenarios, ranging from
true constant relative treatment effects to moderate and strong linear
and quadratic deviations. We also considered scenarios with
treatment-covariate interactions. These scenarios include 4 weak
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of \(66\) scenarios. The exact settings
for each scenario are available in the supplementary material.

\hypertarget{individualized-risk-based-benefit-predictions}{%
\subsection{Individualized risk-based benefit
predictions}\label{individualized-risk-based-benefit-predictions}}

All methods assume that a risk prediction model is available and can be
used to assign individualized predictions. For the simulations we
developed the prediction models internally and blinded to treatment
using logistic regression including main effects for all baseline
covariates and treatment. Predictions on individuals were made setting
treatment to \(0\).

The \textbf{stratified HTE method} was suggested as an alternative to
traditional subgroup analyses. Patients are stratified into
equally-sized risk strata---in this case based on risk quartiles.
Absolute effects are estimated using the differences in event rates
between treatments within risk quarters. We considered this approach as
a reference, expecting it to perform worse than the other candidates, as
its objective is not individualized benefit prediction.

We also considered a set of \textbf{linear methods}. We fit separate
models within treatment arms using only the treatment indicator and the
linear predictor of the internal risk prediction model. In the simpler
case, we assume a constant relative treatment effect (OR). Absolute
benefit is then estimated from \(\text{expit}(lp +\log(\text{OR}))\),
where \(\text{expit}(x)=\frac{e^x}{1+e^x}\) and \emph{lp} is the linear
predictor of the prediction model. A different approach fits a logistic
regression using treatment, risk linear predictor and their interaction.
In this case, absolute benefit is estimated from
\(\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)\).
We will refer to this method as the linear interaction approach.

Finally, we used restricted cubic splines (RCS) to relax the linearity
assumption on the effect of the linear predictor {[}7{]}. We compared
the results for 3, 4 and 5 knots when fitting the splines to introduce
increasing flexibility to the methods considered.

\hypertarget{evaluation-metrics}{%
\subsection{Evaluation metrics}\label{evaluation-metrics}}

Assuming that
\(\tau(\bm{x})=E\{y\:\vert\:\bm{x}, t_x=0\} - E\{y\:\vert\:\bm{x},t_x=1\}\)
is the true benefit for each patient and \(\hat{\tau}(x)\) is the
estimated benefit from a method under study, the ideal loss function to
use for the considered methods would be the unobservable root mean
squared error \(E\big\{(\hat{\tau} - \tau)^2\:\vert\:\bm{x}\big\}\).
However, in simulations the true patient-level benefit is available.
Therefore, we estimated RMSE from
\[\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2\]

We also compared the predictive performance of the methods under study.
We assessed discrimination using the c-for-benefit statistic described
in {[}8{]}. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the
larger treatment arm without a pair. We define \emph{observed}
pair-specific treatment benefit as the difference of observed outcomes
between the untreated and the treated patient of each pair. Pair-specifi
\emph{predicted} benefit is defined as the average of predicted benefits
within each pair. Then, c-for-benefit is defined as the probability that
from two randomly chosen matched patient pairs with unequal observed
benefit, the pair with greater observed benefit also has a higher
predicted benefit.

We evaluated calibration in a similar manner, using the integrated
calibration index (ICI) for benefit {[}9{]}. After creating pairs based
on predicted benefit, observed benefits are regressed on the predicted
benefits using a locally weighted scatterplot smoother (loess). The ICI
is the area between the loess fit and the diagonal, with values closer
to \(0\) indicating better calibration.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{simulations}{%
\subsection{Simulations}\label{simulations}}

The model including a constant relative treatment effect had the lowest
median RMSE in scenarios where the base case of true constant relative
treatment effect (OR = 0.8, N = 4,250 and AUC = 0.75) or moderate
relative deviations were considered (Figure \ref{fig:rmsebase}; Panel
A). However, when we considered strong linear and quadratic deviations
from the base case the linear interaction model performed best (Figure
\ref{fig:rmsebase}; Panels B and C). Only in the case of strong
quadratic deviations models including RCS smoothing (3 knots) performed
equally well to the linear interaction method. Increasing the number of
knots in RCS smoothing resulted in higher error rates across all
scenarios. The performance of the adaptive approach was quite comparable
with the best performing model in each scenario.

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:rmsebase}
\end{figure}

When we increased the sample size (N = 17,000), in all scenarios the
model more in agreement with the underlying settings had the lowest
error rates. Under the base case of constant relative treatment effects
(OR = 0.8), the model including a constant relative treatment effect had
the lowest RMSE (Figure \ref{fig:rmsesamplesize}; Panel A). When
introducing moderate and strong linear deviations the linear interaction
model performed best (Figure \ref{fig:rmsesamplesize}; Panels B and C).
In the case of true strong quadratic deviations from the base case of
constant relative effects, the more flexible RCS smoothing models (3
knots) had the lowest RMSE. Again, the increased flexibility of RCS
smoothing with higher number of knots resulted in overfitting and worse
performance (Figure \ref{fig:rmsesamplesize}; Panel D).

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_sample_size} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:rmsesamplesize}
\end{figure}

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE in the presence of strong quadratic
deviations from the base case of true constant relative treatment
effects (Figure \ref{fig:rmseauc}; Panel D). However, with milder
deviations, the linear interaction model had the lowest RMSE with the
RCS smoothing methods (3 knots) being a close second (Figure
\ref{fig:rmseauc}; Panels B and C). Increasing the number of knots of
RCS smoothing resulted in increased RMSE, which was less pronounced in
the case of strong quadratic deviations.

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_auc} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:rmseauc}
\end{figure}

All candidate methods demonstrated comparable discrimination for benefit
in all scenarios where linear and quadratic deviations from the base
case of constant treatment effect were considered (N=4,250; AUC = 0.75).
However, models including a linear interaction with the risk linear
predictor tended to present much lower variability compared to all other
model-based and smoothing approaches (Figure \ref{fig:discrimination}).
We also observed an increasing trend of discrimination for benefit
variability with increasing number of restricted cubic spline knots in
all scenarios.

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/discrimination_base} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:discrimination}
\end{figure}

When focusing on calibration, the linear interaction model had the
lowest median ICI for benefit in the majority of scenarios except for
the scenarios where moderate linear deviations from the base case were
considered. In that case constant treatment effect models demonstrated
the best performance, very comparable to the linear interaction model's
performance, nonetheless (Figure \ref{fig:calibration}; Panels A, B, and
C).

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/calibration_base} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. \textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:calibration}
\end{figure}

\hypertarget{real-data}{%
\subsection{Real data}\label{real-data}}

We demonstrate the different methods for individualizing treatment
benefits using data from 30,510 patients with an acute myocardial
infarction (MI) included in the GUSTO-I trial. 10,348 patients were
randomized to tissue plasminogen activator (tPA) treatment and 20,162
were randomized to streptokinase. The outcome of interest was 30-day
mortality, recorded for all patients.

In line with previous analyses {[}10,11{]}, we fitted a logistic
regression model with 6 baseline covariates, i.e.~age, Killip class,
systolic blood pressure, heart rate, an indicator of previous MI, and
the location of MI, to predict 30-day mortality risk. A constant effect
of treatment was included in the model. When deriving risk predictions
for individuals we set the treatment indicator to 0. More information on
model development can be found in the supplement.

We used the risk linear predictor to fit the the proposed methods under
study for individualizing absolute benefit predictions. All methods had
quite comparable results, in the sense that we predicted increasing
benefits for patients with higher baseline risk predictions. In terms of
c-for-benefit (validated internally) all models had quite comparable
performance ranging from 0.519 (RCS smoothing with 3 knots) to 0.542
(RCS smoothing with 4 knots). Similar conclusions could be drawn in
terms of ICI-for-benefit which ranged from 0.0039 (linear interaction
approach) to 0.0053 (RCS smoothing with 4 knots).

The adaptive approach picked the model with RCS smoothing with 4 knots,
which had quite comparable performance to the smooth fit with 5 knots.
However, for lower baseline risk these 2 models predicted implausible
benefits and maybe should be avoided when applying such models in
practice. The linear interaction model and the model with RCS smoothing
(3 knots) made very similar predictions and also followed quite closely
the evolution of the stratified estimates. In this case, we observed
that even using a simple model with a constant relative treatment effect
can result in very similar benefit predictions as other more complex
approaches.

\begin{figure}
\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/gusto} \caption{Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots. Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.}\label{fig:gusto}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We have carried out a simulation study comparing different approaches
for individualized benefit prediction using baseline risk as the
reference. These methods included a constant treatment effect model, a
model including a simple linear interaction of the risk linear predictor
with treatment, and models using smoothing with restricted cubic splines
with 3, 4, and 5 knots. Performance was assessed evaluating error rates
using RMSE and benefit prediction performance using some novel methods
for evaluating discrimination and calibration for benefit.

The linear interaction model proved to be flexible enough to adequately
predict benefit under a wide range of scenarios. It had lower error
rates and usually the highest c-for-benefit with the lower variability
across simulation replications. Its benefit predictions were also
well-calibrated across the majority of the simulation settings. Models
with RCS smoothing with 3 knots performed very similarly, only
outperformig the linear interaction model in the presence of strong
quadratic deviations from the base case scenario of constant relative
treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more
flexible approaches.

Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry
and this often led to overfitting with smaller sample sizes. This
resulted in higher RMSE, increased variability of discrimination for
benefit and worse-calibrated benefit predictions across simulation
replications. Even with larger sample sizes and strong quadratic
deviations from the base case scenario of constant relative treatment
effects, these more flexible RCS smoothing methods did not outperform
simpler RCS smoothing with 3 knots. More flexible approaches usually do
not come with any added value, unless we expect extreme heterogeneity in
treatment effects.

The constant treatment effect model, despite having adequate performance
even in the presence of weak treatment effect heterogeneity on the
relative scale, it quickly broke down with stronger deviations from the
base case scenario of constant effects. In these cases, the stratified
approach generally had lower error rates compared to the constant
treatment effect model. In concept, the stratified approach lies between
the constant effects model and smoother approaches, only assuming
constant treatment effects within strata of predicted risk and,
therefore, is more sensitive to treatment effect heterogeneity. However,
these stepwise benefit estimates can only be used as a demonstration of
treatment effect heterogeneity, identifying patient subgroups where
treatment benefits are expected to be higher or lower, and should not be
used to make individualized absolute benefit predictions.

Increasing the true prediction AUC by increasing the predictor
coefficients of the true risk model reduced RMSE for all methods,
similar to increasing the sample size. Higher AUC translates in higher
variability of predicted risks, which, in turn, allows the considered
methods to better follow the evolution of treatment benefit. Higher
prediction AUC also led to higher absolute values of c-for-benefit. Even
though, model performance appears to be very important for the
performance of these methods, prediction model development was outside
the scope of this work and has already been studied quite extensively
{[}5,12,13{]}.

Risk-based approaches to predictive HTE assume that benefit is a
function of baseline risk. A limitation of our study is that this was
assumed to be the case in the majority of the simulation scenarios. Even
though our main conclusions did not change when simulating true
treatment-covariate interactions (results in supplement), RMSE had
higher values. The linear interaction model under these scenarios again
had the best overall performance under different sizes of the
interaction effects. This can be explained by the fact that linear
interactions with treatment were simulated, thus favoring the linear
interaction model. In studies evaluating prediction model development,
the underlying outcome generation process sometimes proved to affect
performance of certain methods {[}14{]}. This might be the case here as
well and should be explored in more extensive future simulation studies.

Recent years have seen an increased interest in predictive HTE
approaches focusing on individualized benefit predictions. In our
simulations we only focused on risk-based methods, using baseline risk
as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}15--17{]}. Simulations are also needed to
assess relative performance and define the settings where these break
down or outperform each other.

In conclusion, we carried out a simulation study comparing different
risk-based approaches to individualizing benefit predictions. Simpler
models including a linear interaction with the risk linear predictor or
RCS smoothing with 3 knots had adequate performance under a wide range
scenarios. An adaptive approach based on selecting among the considered
methods using AIC had satisfactory performance.

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\nolinenumbers
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}

\noindent

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Varadhan2013}{}%
{[}1{]} Varadhan R, Segal JB, Boyd CM, Wu AW, Weiss CO. A framework for
the analysis of heterogeneity of treatment effect in~patient-centered
outcomes research. Journal of Clinical Epidemiology 2013;66:818--25.
\url{https://doi.org/10.1016/j.jclinepi.2013.02.009}.

\leavevmode\hypertarget{ref-Rekkas2020}{}%
{[}2{]} Rekkas A, Paulus JK, Raman G, Wong JB, Steyerberg EW, Rijnbeek
PR, et al. Predictive approaches to heterogeneous treatment effects: A
scoping review. BMC Medical Research Methodology 2020;20.
\url{https://doi.org/10.1186/s12874-020-01145-1}.

\leavevmode\hypertarget{ref-Kent2019}{}%
{[}3{]} Kent DM, Paulus JK, Klaveren D van, D'Agostino R, Goodman S,
Hayward R, et al. The predictive approaches to treatment effect
heterogeneity (PATH) statement. Annals of Internal Medicine 2019;172:35.
\url{https://doi.org/10.7326/m18-3667}.

\leavevmode\hypertarget{ref-PathEnE}{}%
{[}4{]} Kent DM, Klaveren D van, Paulus JK, D'Agostino R, Goodman S,
Hayward R, et al. The predictive approaches to treatment effect
heterogeneity (PATH) statement: Explanation and elaboration. Annals of
Internal Medicine 2019;172:W1. \url{https://doi.org/10.7326/m18-3668}.

\leavevmode\hypertarget{ref-vanKlaveren2019}{}%
{[}5{]} Klaveren D van, Balan TA, Steyerberg EW, Kent DM. Models with
interactions overestimated heterogeneity of treatment effects and were
prone to treatment mistargeting. Journal of Clinical Epidemiology
2019;114:72--83. \url{https://doi.org/10.1016/j.jclinepi.2019.05.029}.

\leavevmode\hypertarget{ref-Kent2010}{}%
{[}6{]} Kent DM, Rothwell PM, Ioannidis JP, Altman DG, Hayward RA.
Assessing and reporting heterogeneity in treatment effects in clinical
trials: A proposal. Trials 2010;11.
\url{https://doi.org/10.1186/1745-6215-11-85}.

\leavevmode\hypertarget{ref-Harrell1988}{}%
{[}7{]} Harrell FE, Lee KL, Pollock BG. Regression models in clinical
studies: Determining relationships between predictors and response. JNCI
Journal of the National Cancer Institute 1988;80:1198--202.
\url{https://doi.org/10.1093/jnci/80.15.1198}.

\leavevmode\hypertarget{ref-vanKlaveren2018}{}%
{[}8{]} Klaveren D van, Steyerberg EW, Serruys PW, Kent DM. The proposed
``concordance-statistic for benefit'' provided a useful metric when
modeling heterogeneous treatment effects. Journal of Clinical
Epidemiology 2018;94:59--68.
\url{https://doi.org/10.1016/j.jclinepi.2017.10.021}.

\leavevmode\hypertarget{ref-Austin2019}{}%
{[}9{]} Austin PC, Steyerberg EW. The integrated calibration index (ICI)
and related metrics for quantifying the calibration of logistic
regression models. Statistics in Medicine 2019;38:4051--65.
\url{https://doi.org/10.1002/sim.8281}.

\leavevmode\hypertarget{ref-Califf1997}{}%
{[}10{]} Califf RM, Woodlief LH, Harrell FE, Lee KL, White HD, Guerci A,
et al. Selection of thrombolytic therapy for individual patients:
Development of a clinical model. American Heart Journal 1997;133:630--9.
\url{https://doi.org/10.1016/s0002-8703(97)70164-9}.

\leavevmode\hypertarget{ref-Steyerberg2000}{}%
{[}11{]} Steyerberg EW, Bossuyt PMM, Lee KL. Clinical trials in acute
myocardial infarction: Should we adjust for baseline characteristics?
American Heart Journal 2000;139:745--51.
\url{https://doi.org/10.1016/s0002-8703(00)90001-2}.

\leavevmode\hypertarget{ref-Burke2014}{}%
{[}12{]} Burke JF, Hayward RA, Nelson JP, Kent DM. Using internally
developed risk models to assess heterogeneity in treatment effects in
clinical trials. Circulation: Cardiovascular Quality and Outcomes
2014;7:163--9. \url{https://doi.org/10.1161/circoutcomes.113.000497}.

\leavevmode\hypertarget{ref-Abadie2018}{}%
{[}13{]} Abadie A, Chingos MM, West MR. Endogenous stratification in
randomized experiments. The Review of Economics and Statistics
2018;100:567--80. \url{https://doi.org/10.1162/rest_a_00732}.

\leavevmode\hypertarget{ref-Austin2021}{}%
{[}14{]} Austin PC, Harrell FE, Steyerberg EW. Predictive performance of
machine and statistical learning methods: Impact of data-generating
processes on external validity in the ``large n, small p'' setting.
Statistical Methods in Medical Research 2021;30:1465--83.
\url{https://doi.org/10.1177/09622802211002867}.

\leavevmode\hypertarget{ref-Athey2019}{}%
{[}15{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.

\leavevmode\hypertarget{ref-Lu2018}{}%
{[}16{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
treatment effect in observational data using random forest methods.
Journal of Computational and Graphical Statistics 2018;27:209--19.
\url{https://doi.org/10.1080/10618600.2017.1356325}.

\leavevmode\hypertarget{ref-Wager2018}{}%
{[}17{]} Wager S, Athey S. Estimation and inference of heterogeneous
treatment effects using random forests. Journal of the American
Statistical Association 2018;113:1228--42.
\url{https://doi.org/10.1080/01621459.2017.1319839}.
\end{cslreferences}

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\noindent

\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}

### 2021-06-21 code/GetArxivStyle.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+[[ ! -f "submission/arxiv.sty" ]] && wget --output-document submission/arxiv.sty -nc https://raw.githubusercontent.com/kourgeorge/arxiv-style/master/arxiv.sty || echo -e "Nothing to be done"

### 2021-06-22 submission/manuscript.rmd
reduce abstract word count

@@
 ---
-title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
+title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
 abstract: |
-  **Objective:** Compare different risk-based approaches to estimating
-  individualized treatment effects with simulations of the RCT setting. **Study
-  Design and Setting:** We compared different methods for predicting absolute
-  benefit on the basis of an available prognostic index (PI) for risk: A model
-  with the PI and a constant relative treatment effect; models including an
-  interaction of treatment assignment with the PI; 4 quarters of the PI; nonlinear
-  transformations of the PI with increasing flexibility (restricted cubic splines
-  with 3, 4 and 5 knots, respectively); an adaptive method, selecting the optimal
-  model based on Akaike’s Information Criterion. Starting from a base case
-  scenario (sample size 4,250, average treatment effect odds ratio 0.8, AUC of the
-  PI 0.75), we considered diverse assumptions of the data generation process of
-  the “true” treatment effect, introducing linear and quadratic interactions of
-  the PI with treatment. We analyzed the effect of varying sample size and
-  discriminative ability of the PI. We evaluated performance using root mean
-  squared error, discrimination for benefit and calibration for benefit. **Results:**
-  The model including a linear interaction of the risk linear predictor with
-  treatment had adequate performance that was robust under the majority of the
-  simulation scenarios. Methods using restricted cubic splines required larger
-  sample sizes and higher AUC of the PI to achieve adequate performance. The
-  adaptive approach’s performance was comparable to the performance of the best
-  model in each scenario. **Conclusion:** In most cases using a model just
-  including a linear interaction of the PI with treatment
-  adequately predicts absolute benefit.
+  **Objective:** Compare risk-based methods for individualizing treatment 
+  effects with simulations in the RCT setting. **Study Design and Setting:** We
+  predicted absolute benefit based on an available prognostic index (PI) using:
+  a model with the PI and a constant relative treatment effect; a model 
+  including an interaction of treatment with the PI; 4 quarters of the PI; 
+  nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5
+  knots); an adaptive model selection method using Akaike’s Information 
+  Criterion. Starting from a base case scenario (sample size 4,250, constant 
+  odds ratio 0.8, AUC of the PI 0.75), we considered diverse assumptions by
+  introducing linear and quadratic interactions of the PI with treatment and 
+  varying sample size and discriminative ability of the PI. We evaluated
+  performance using root mean squared error, discrimination and calibration for 
+  benefit. **Results:** Models including a linear interaction of the PI with
+  treatment had adequate performance that was robust under most simulation 
+  scenarios. Restricted cubic splines required larger sample sizes and higher 
+  AUC of the PI to achieve adequate performance. The adaptive approach performed 
+  equivalently to the best-performing method in each scenario. **Conclusion:** 
+  Usually, models with just a linear interaction of the PI with treatment
+  adequately predict absolute benefit.
 authors:
@@
 keywords:
-  - test
-  - test that
+  - treatment effect heterogeneity
+  - absolute benefit
+  - prediction models
 output: 
@@
 
-In the simulated datasets based treatment was allocated at random using a 50/50
+In the simulated datasets based treatment was allocated at random using a 50:50
 split. For each patient we simulated $8$ baseline covariates, where
@@
 including all baseline covariates. Coefficient values were such, so that the
-prediction model had an AUC of $0.75$ and an event rate of $20%$ in the control
+prediction model had an AUC of $0.75$ and an event rate of $20\%$ in the control
 arm was achieved. Under the base case scenario, outcomes in the treatment arm
@@
 treatment odds ratio (OR) of $0.8$. The generated samples of the base case
-scenario were 4,250 ($80\%$ power for the detection of an unadjusted OR of
-$0.8$).
+scenario had sample size of 4,250 ($85%$ power for the detection of an
+unadjusted OR of $0.8$).
 
@@
 of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
-$500,000$ patients.
+500,000 patients.
 
-We simulated binary outcomes in the control arm based on 
-$\text{expit}(PI)$, where 
+We simulated binary outcomes in the control arm using true probabilities 
+$P(y=1|X) = \text{expit}(PI)$, where 
 $\text{expit}(x)= \frac{e^x}{1 + e^x}$. 
-In the treatment arm outcomes where generated from $\text{expit}(lp_1)$ with
+In the treatment arm outcomes where generated with probabilities
+$\text{expit}(lp_1)$ with
 $$ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,$$
@@
 All methods assume that a risk prediction model is available and can be used to
-assign individualized predictions. For the simulations we developed the
+assign individualized risk predictions. For the simulations we developed the
 prediction models internally and blinded to treatment using logistic regression
-including main effects for all baseline covariates and treatment. Predictions on
-individuals were made setting treatment to $0$.
+including main effects for all baseline covariates and treatment. Risk
+predictions on individuals were made setting treatment to $0$.
 
-The **stratified HTE method** was suggested as an alternative to traditional
-subgroup analyses. Patients are stratified into equally-sized risk strata---in
-this case based on risk quartiles. Absolute effects are estimated using the
-differences in event rates between treatments within risk quarters. We
+The *stratified HTE method* has been suggested as an alternative to
+traditional subgroup analyses. Patients are stratified into equally-sized risk
+strata---in this case based on risk quartiles. Absolute effects are estimated
+using the differences in event rates between treatments within risk quarters. We
 considered this approach as a reference, expecting it to perform worse than the
@@
 
-We also considered a set of **linear methods**. We fit separate models within
-treatment arms using only the treatment indicator and the linear predictor of
-the internal risk prediction model. In the simpler case, we assume a constant
-relative treatment effect (OR). Absolute benefit is then estimated from
-$\text{expit}(lp +\log(\text{OR}))$, 
+Another approach would be to assume *constant relative treatment effect* (OR) is
+constant. In that case, absolute benefit is estimated from
+$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$,
 where $\text{expit}(x)=\frac{e^x}{1+e^x}$
-and *lp* is the linear predictor of the prediction model. A different approach
-fits a logistic regression using treatment, risk linear predictor and their
-interaction. In this case, absolute benefit is estimated from
-$\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)$.
-We will refer to this method as the linear interaction approach.
+and *lp* is the linear predictor of the prediction model.
 
-Finally, we used restricted cubic splines (RCS) to relax the linearity assumption on
+A different approach fits a logistic regression using treatment, the prognostic
+index and their interaction. In this case absolute benefit is estimated from
+$\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
+We will refer to this method as the *linear interaction* approach.
+
+Finally, we used *restricted cubic splines* (RCS) to relax the linearity assumption on
 the effect of the linear predictor [@Harrell1988]. We compared the results
@@
 
-Assuming that $\tau(\bm{x})=E\{y\given\bm{x}, t_x=0\} - E\{y\given\bm{x},t_x=1\}$ is the true benefit for each patient and $\hat{\tau}(x)$ is the estimated benefit from a method under study, the ideal loss function to use for the considered methods would be the unobservable root mean squared error $E\big\{(\hat{\tau} - \tau)^2\given \bm{x}\big\}$.
-However, in simulations the true patient-level benefit is available.
-Therefore, we estimated RMSE from
+For evaluating the prediction error of the considered methods we used root mean
+squared error (RMSE), since both the true and the predicted benefits are known,
+given that this is a simulation study. More specifically, we calculated RMSE
+from
 $$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$
 
-We also compared the predictive performance of the methods under study. We
+We also compared the discriminative ability of the methods under study. We
 assessed discrimination using the c-for-benefit statistic described in
@@
 benefit as the difference of observed outcomes between the untreated and the
-treated patient of each pair. Pair-specifi *predicted* benefit is defined as
+treated patient of each pair. Pair-specific *predicted* benefit is defined as
 the average of predicted benefits within each pair. Then, c-for-benefit is
@@
 benefit, observed benefits are regressed on the predicted benefits using a
-locally weighted scatterplot smoother (loess). The ICI is the area between the
-loess fit and the diagonal, with values closer to $0$ indicating better
-calibration.
+locally weighted scatterplot smoother (loess). The ICI is the average absolute
+difference between predicted and smooth observed benefit. Values closer to $0$
+represent better calibration.
 
@@
 The model including a constant relative treatment effect had the lowest median
-RMSE in scenarios where the base case of true constant relative treatment effect
-(OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
-considered (Figure \ref{fig:rmsebase}; Panel A). However, when we considered
-strong linear and quadratic deviations from the base case the linear interaction
-model performed best (Figure \ref{fig:rmsebase}; Panels B and C). Only in the
-case of strong quadratic deviations models including RCS smoothing (3 knots)
+RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
+`r 4250` and AUC = 0.75) or moderate relative deviations were considered (Figure
+\ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
+quadratic deviations from a constant relative treatment effect the linear
+interaction model performed best (Figure \ref{fig:rmsebase}; Panels B and C).
+Only in the case of strong quadratic deviations models including RCS (3 knots)
 performed equally well to the linear interaction method. Increasing the number
-of knots in RCS smoothing resulted in higher error rates across all scenarios.
-The performance of the adaptive approach was quite comparable with the best
-performing model in each scenario.
+of knots in RCS resulted in higher error rates across all scenarios. The
+adaptive approach performed very similarly to the best performing model in each
+scenario.
 
@@
 
-When we increased the sample size (N = `r 17000`), in all scenarios the model more
-in agreement with the underlying settings had the lowest error rates. Under the
-base case of constant relative treatment effects (OR = 0.8), the model including
-a constant relative treatment effect had the lowest RMSE (Figure
+When we increased the sample size (N = `r 17000`), the model including a
+constant relative treatment effect had the lowest RMSE under the base case
+scenario of constant relative treatment effects (Figure
 \ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
@@
 flexibility of RCS smoothing with higher number of knots resulted in overfitting
-and worse performance (Figure \ref{fig:rmsesamplesize}; Panel D).
+and worse performance (Figure \\ref{fig:rmsesamplesize}; Panel D).
 

### 2021-06-22 submission/manuscript.tex
reduce abstract word count

@@
 
-\title{Smooth risk-based predictive approaches to treatment effect
+\title{Continuous risk-based predictive approaches to treatment effect
 heterogeneity: A simulation study}
@@
 \begin{abstract}
-\textbf{Objective:} Compare different risk-based approaches to
-estimating individualized treatment effects with simulations of the RCT
-setting. \textbf{Study Design and Setting:} We compared different
-methods for predicting absolute benefit on the basis of an available
-prognostic index (PI) for risk: A model with the PI and a constant
-relative treatment effect; models including an interaction of treatment
-assignment with the PI; 4 quarters of the PI; nonlinear transformations
-of the PI with increasing flexibility (restricted cubic splines with 3,
-4 and 5 knots, respectively); an adaptive method, selecting the optimal
-model based on Akaike's Information Criterion. Starting from a base case
-scenario (sample size 4,250, average treatment effect odds ratio 0.8,
-AUC of the PI 0.75), we considered diverse assumptions of the data
-generation process of the ``true'' treatment effect, introducing linear
-and quadratic interactions of the PI with treatment. We analyzed the
-effect of varying sample size and discriminative ability of the PI. We
-evaluated performance using root mean squared error, discrimination for
-benefit and calibration for benefit. \textbf{Results:} The model
-including a linear interaction of the risk linear predictor with
-treatment had adequate performance that was robust under the majority of
-the simulation scenarios. Methods using restricted cubic splines
-required larger sample sizes and higher AUC of the PI to achieve
-adequate performance. The adaptive approach's performance was comparable
-to the performance of the best model in each scenario.
-\textbf{Conclusion:} In most cases using a model just including a linear
-interaction of the PI with treatment adequately predicts absolute
+\textbf{Objective:} Compare risk-based methods for individualizing
+treatment effects with simulations in the RCT setting. \textbf{Study
+Design and Setting:} We predicted absolute benefit based on an available
+prognostic index (PI) using: a model with the PI and a constant relative
+treatment effect; a model including an interaction of treatment with the
+PI; 4 quarters of the PI; nonlinear transformations of the PI
+(restricted cubic splines with 3, 4 and 5 knots); an adaptive model
+selection method using Akaike's Information Criterion. Starting from a
+base case scenario (sample size 4,250, constant odds ratio 0.8, AUC of
+the PI 0.75), we considered diverse assumptions by introducing linear
+and quadratic interactions of the PI with treatment and varying sample
+size and discriminative ability of the PI. We evaluated performance
+using root mean squared error, discrimination and calibration for
+benefit. \textbf{Results:} Models including a linear interaction of the
+PI with treatment had adequate performance that was robust under most
+simulation scenarios. Restricted cubic splines required larger sample
+sizes and higher AUC of the PI to achieve adequate performance. The
+adaptive approach performed equivalently to the best-performing method
+in each scenario. \textbf{Conclusion:} Usually, models with just a
+linear interaction of the PI with treatment adequately predict absolute
 benefit.
@@
 \keywords{
-    test
+    treatment effect heterogeneity
    \and
-    test that
+    absolute benefit
+   \and
+    prediction models
   }
@@
 In the simulated datasets based treatment was allocated at random using
-a 50/50 split. For each patient we simulated \(8\) baseline covariates,
+a 50:50 split. For each patient we simulated \(8\) baseline covariates,
 where \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
@@
 were such, so that the prediction model had an AUC of \(0.75\) and an
-event rate of \(20%
-\) in the control arm was achieved. Under the base case scenario,
-outcomes in the treatment arm were created using the same logistic
-regression model, including a constant treatment odds ratio (OR) of
-\(0.8\). The generated samples of the base case scenario were 4,250
-(\(80\%\) power for the detection of an unadjusted OR of \(0.8\)).
+event rate of \(20\%\) in the control arm was achieved. Under the base
+case scenario, outcomes in the treatment arm were created using the same
+logistic regression model, including a constant treatment odds ratio
+(OR) of \(0.8\). The generated samples of the base case scenario had
+sample size of 4,250 (\(85%
+\) power for the detection of an unadjusted OR of \(0.8\)).
 
@@
 so that AUC values of \(0.65\) and \(0.80\) were achieved when
-validating in a simulated dataset of \(500,000\) patients.
+validating in a simulated dataset of 500,000 patients.
 
-We simulated binary outcomes in the control arm based on
-\(\text{expit}(PI)\), where \(\text{expit}(x)= \frac{e^x}{1 + e^x}\). In
-the treatment arm outcomes where generated from \(\text{expit}(lp_1)\)
-with \[ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,\] where the
+We simulated binary outcomes in the control arm using true probabilities
+\(P(y=1|X) = \text{expit}(PI)\), where
+\(\text{expit}(x)= \frac{e^x}{1 + e^x}\). In the treatment arm outcomes
+where generated with probabilities \(\text{expit}(lp_1)\) with
+\[ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,\] where the
 coefficients \(\gamma_0, \gamma_1\) and \(\gamma_2\) along with the
@@
 All methods assume that a risk prediction model is available and can be
-used to assign individualized predictions. For the simulations we
+used to assign individualized risk predictions. For the simulations we
 developed the prediction models internally and blinded to treatment
 using logistic regression including main effects for all baseline
-covariates and treatment. Predictions on individuals were made setting
-treatment to \(0\).
+covariates and treatment. Risk predictions on individuals were made
+setting treatment to \(0\).
 
-The \textbf{stratified HTE method} was suggested as an alternative to
+The \emph{stratified HTE method} has been suggested as an alternative to
 traditional subgroup analyses. Patients are stratified into
@@
 
-We also considered a set of \textbf{linear methods}. We fit separate
-models within treatment arms using only the treatment indicator and the
-linear predictor of the internal risk prediction model. In the simpler
-case, we assume a constant relative treatment effect (OR). Absolute
-benefit is then estimated from \(\text{expit}(lp +\log(\text{OR}))\),
-where \(\text{expit}(x)=\frac{e^x}{1+e^x}\) and \emph{lp} is the linear
-predictor of the prediction model. A different approach fits a logistic
-regression using treatment, risk linear predictor and their interaction.
-In this case, absolute benefit is estimated from
-\(\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)\).
-We will refer to this method as the linear interaction approach.
+Another approach would be to assume \emph{constant relative treatment
+effect} (OR) is constant. In that case, absolute benefit is estimated
+from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\), where
+\(\text{expit}(x)=\frac{e^x}{1+e^x}\) and \emph{lp} is the linear
+predictor of the prediction model.
 
-Finally, we used restricted cubic splines (RCS) to relax the linearity
-assumption on the effect of the linear predictor {[}7{]}. We compared
-the results for 3, 4 and 5 knots when fitting the splines to introduce
-increasing flexibility to the methods considered.
+A different approach fits a logistic regression using treatment, the
+prognostic index and their interaction. In this case absolute benefit is
+estimated from
+\(\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
+We will refer to this method as the \emph{linear interaction} approach.
+
+Finally, we used \emph{restricted cubic splines} (RCS) to relax the
+linearity assumption on the effect of the linear predictor {[}7{]}. We
+compared the results for 3, 4 and 5 knots when fitting the splines to
+introduce increasing flexibility to the methods considered.
 
@@
 
-Assuming that
-\(\tau(\bm{x})=E\{y\:\vert\:\bm{x}, t_x=0\} - E\{y\:\vert\:\bm{x},t_x=1\}\)
-is the true benefit for each patient and \(\hat{\tau}(x)\) is the
-estimated benefit from a method under study, the ideal loss function to
-use for the considered methods would be the unobservable root mean
-squared error \(E\big\{(\hat{\tau} - \tau)^2\:\vert\:\bm{x}\big\}\).
-However, in simulations the true patient-level benefit is available.
-Therefore, we estimated RMSE from
+For evaluating the prediction error of the considered methods we used
+root mean squared error (RMSE), since both the true and the predicted
+benefits are known, given that this is a simulation study. More
+specifically, we calculated RMSE from
 \[\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2\]
 
-We also compared the predictive performance of the methods under study.
+We also compared the discriminative ability of the methods under study.
 We assessed discrimination using the c-for-benefit statistic described
@@
 pair-specific treatment benefit as the difference of observed outcomes
-between the untreated and the treated patient of each pair. Pair-specifi
-\emph{predicted} benefit is defined as the average of predicted benefits
-within each pair. Then, c-for-benefit is defined as the probability that
-from two randomly chosen matched patient pairs with unequal observed
-benefit, the pair with greater observed benefit also has a higher
-predicted benefit.
+between the untreated and the treated patient of each pair.
+Pair-specific \emph{predicted} benefit is defined as the average of
+predicted benefits within each pair. Then, c-for-benefit is defined as
+the probability that from two randomly chosen matched patient pairs with
+unequal observed benefit, the pair with greater observed benefit also
+has a higher predicted benefit.
 
@@
 benefits using a locally weighted scatterplot smoother (loess). The ICI
-is the area between the loess fit and the diagonal, with values closer
-to \(0\) indicating better calibration.
+is the average absolute difference between predicted and smooth observed
+benefit. Values closer to \(0\) represent better calibration.
 
@@
 The model including a constant relative treatment effect had the lowest
-median RMSE in scenarios where the base case of true constant relative
-treatment effect (OR = 0.8, N = 4,250 and AUC = 0.75) or moderate
-relative deviations were considered (Figure \ref{fig:rmsebase}; Panel
-A). However, when we considered strong linear and quadratic deviations
-from the base case the linear interaction model performed best (Figure
-\ref{fig:rmsebase}; Panels B and C). Only in the case of strong
-quadratic deviations models including RCS smoothing (3 knots) performed
-equally well to the linear interaction method. Increasing the number of
-knots in RCS smoothing resulted in higher error rates across all
-scenarios. The performance of the adaptive approach was quite comparable
-with the best performing model in each scenario.
+median RMSE in scenarios with a true constant relative treatment effect
+(OR = 0.8, N = 4,250 and AUC = 0.75) or moderate relative deviations
+were considered (Figure \ref{fig:rmsebase}; Panel A). However, when we
+considered strong linear and quadratic deviations from a constant
+relative treatment effect the linear interaction model performed best
+(Figure \ref{fig:rmsebase}; Panels B and C). Only in the case of strong
+quadratic deviations models including RCS (3 knots) performed equally
+well to the linear interaction method. Increasing the number of knots in
+RCS resulted in higher error rates across all scenarios. The adaptive
+approach performed very similarly to the best performing model in each
+scenario.
 
@@
 
-When we increased the sample size (N = 17,000), in all scenarios the
-model more in agreement with the underlying settings had the lowest
-error rates. Under the base case of constant relative treatment effects
-(OR = 0.8), the model including a constant relative treatment effect had
-the lowest RMSE (Figure \ref{fig:rmsesamplesize}; Panel A). When
-introducing moderate and strong linear deviations the linear interaction
-model performed best (Figure \ref{fig:rmsesamplesize}; Panels B and C).
-In the case of true strong quadratic deviations from the base case of
-constant relative effects, the more flexible RCS smoothing models (3
-knots) had the lowest RMSE. Again, the increased flexibility of RCS
-smoothing with higher number of knots resulted in overfitting and worse
-performance (Figure \ref{fig:rmsesamplesize}; Panel D).
+When we increased the sample size (N = 17,000), the model including a
+constant relative treatment effect had the lowest RMSE under the base
+case scenario of constant relative treatment effects (Figure
+\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong
+linear deviations the linear interaction model performed best (Figure
+\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong
+quadratic deviations from the base case of constant relative effects,
+the more flexible RCS smoothing models (3 knots) had the lowest RMSE.
+Again, the increased flexibility of RCS smoothing with higher number of
+knots resulted in overfitting and worse performance (Figure
+\textbackslash ref\{fig:rmsesamplesize\}; Panel D).
 

### 2021-06-22 submission/manuscript.rmd
add corrections to results section

@@
 When we increased the sample size (N = `r 17000`), the model including a
-constant relative treatment effect had the lowest RMSE under the base case
-scenario of constant relative treatment effects (Figure
-\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
-deviations the linear interaction model performed best (Figure
-\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong quadratic
-deviations from the base case of constant relative effects, the more flexible
-RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
-flexibility of RCS smoothing with higher number of knots resulted in overfitting
-and worse performance (Figure \\ref{fig:rmsesamplesize}; Panel D).
+constant relative treatment effect had the lowest RMSE under the assumption of
+true constant relative treatment effects (Figure \ref{fig:rmsesamplesize};
+Panel A). Contrary to previous results with smaller sample sizes, when
+introducing moderate and strong linear deviations the linear interaction model
+performed best, outperforming the constant treatment effect model (Figure
+\ref{fig:rmsesamplesize}; Panels B and C). However, the linear interaction
+model was outperformed by the more flexible RCS models (3 knots) in the case of
+strong quadratic deviations. Again, the increased flexibility of RCS smoothing
+with higher number of knots resulted in overfitting and worse performance
+(Figure \ref{fig:rmsesamplesize}; Panel D).
 
@@
 
-All candidate methods demonstrated comparable discrimination for benefit in all
-scenarios where linear and quadratic deviations from the base case of constant
-treatment effect were considered (N=`r 4250`; AUC = `r 0.75`). However, models
-including a linear interaction with the risk linear predictor tended to present
-much lower variability compared to all other model-based and smoothing
-approaches (Figure \ref{fig:discrimination}). We also observed an increasing
-trend of discrimination for benefit variability with increasing number of
-restricted cubic spline knots in all scenarios.
+The constant effects model, the linear interaction model and models with RCS
+smoothing (3 knots) had the highest median c-for-benefit in the base case
+scenario and the scenarios where linear and quadratic deviations were
+considered. However, constant treatment effects and linear interaction models
+tended to present much lower variability compared to all other model-based and
+smoothing approaches (Figure \ref{fig:discrimination}). We also observed an
+increasing trend of discrimination for benefit variability and decreasing median
+values with increasing number of restricted cubic spline knots in all scenarios.
 
@@
 When focusing on calibration, the linear interaction model had the lowest median
-ICI for benefit in the majority of scenarios except for the scenarios where
-moderate linear deviations from the base case were considered. In that case
+ICI for benefit in the majority of scenarios except for the scenarios where no
+or moderate linear deviations from the base case were considered. In that case
 constant treatment effect models demonstrated the best performance, very
 comparable to the linear interaction model's performance, nonetheless (Figure
-\ref{fig:calibration}; Panels A, B, and C).
+\ref{fig:calibration}). However, under strong linear or
+quadratic deviations, the constant treatment effect model was very poorly
+calibrated (Figure \ref{fig:calibration}; Panels C and D).
 
@@
 
-## Real data
+## Case study
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
@@
 # Discussion
-We have carried out a simulation study comparing different approaches for
-individualized benefit prediction using baseline risk as the reference. These
-methods included a constant treatment effect model, a model including a simple
-linear interaction of the risk linear predictor with treatment, and models using
-smoothing with restricted cubic splines with 3, 4, and 5 knots. Performance was
-assessed evaluating error rates using RMSE and benefit prediction performance
-using some novel methods for evaluating discrimination and calibration for
-benefit.
 
@@
 majority of the simulation settings. Models with RCS smoothing with 3 knots
-performed very similarly, only outperformig the linear interaction model in the
+performed very similarly, only outperforming the linear interaction model in the
 presence of strong quadratic deviations from the base case scenario of constant
@@
 
-In conclusion, we carried out a simulation study comparing different risk-based
-approaches to individualizing benefit predictions. Simpler models including a
-linear interaction with the risk linear predictor or RCS smoothing with 3 knots
-had adequate performance under a wide range scenarios. An adaptive approach based
-on selecting among the considered methods using AIC had satisfactory
-performance.
+In conclusion, when comparing different risk-based approaches to individualizing
+benefit predictions simpler models including a linear interaction with the
+prognostic index or RCS smoothing with 3 knots had adequate performance under a
+wide range scenarios. An adaptive approach selecting among the considered
+methods with optimal AIC had satisfactory performance.
 

### 2021-06-22 submission/manuscript.tex
add corrections to results section

@@
 When we increased the sample size (N = 17,000), the model including a
-constant relative treatment effect had the lowest RMSE under the base
-case scenario of constant relative treatment effects (Figure
-\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong
-linear deviations the linear interaction model performed best (Figure
-\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong
-quadratic deviations from the base case of constant relative effects,
-the more flexible RCS smoothing models (3 knots) had the lowest RMSE.
-Again, the increased flexibility of RCS smoothing with higher number of
-knots resulted in overfitting and worse performance (Figure
-\textbackslash ref\{fig:rmsesamplesize\}; Panel D).
+constant relative treatment effect had the lowest RMSE under the
+assumption of true constant relative treatment effects (Figure
+\ref{fig:rmsesamplesize}; Panel A). Contrary to previous results with
+smaller sample sizes, when introducing moderate and strong linear
+deviations the linear interaction model performed best, outperforming
+the constant treatment effect model (Figure \ref{fig:rmsesamplesize};
+Panels B and C). However, the linear interaction model was outperformed
+by the more flexible RCS models (3 knots) in the case of strong
+quadratic deviations. Again, the increased flexibility of RCS smoothing
+with higher number of knots resulted in overfitting and worse
+performance (Figure \ref{fig:rmsesamplesize}; Panel D).
 
@@
 
-All candidate methods demonstrated comparable discrimination for benefit
-in all scenarios where linear and quadratic deviations from the base
-case of constant treatment effect were considered (N=4,250; AUC = 0.75).
-However, models including a linear interaction with the risk linear
-predictor tended to present much lower variability compared to all other
-model-based and smoothing approaches (Figure \ref{fig:discrimination}).
-We also observed an increasing trend of discrimination for benefit
-variability with increasing number of restricted cubic spline knots in
-all scenarios.
+The constant effects model, the linear interaction model and models with
+RCS smoothing (3 knots) had the highest median c-for-benefit in the base
+case scenario and the scenarios where linear and quadratic deviations
+were considered. However, constant treatment effects and linear
+interaction models tended to present much lower variability compared to
+all other model-based and smoothing approaches (Figure
+\ref{fig:discrimination}). We also observed an increasing trend of
+discrimination for benefit variability and decreasing median values with
+increasing number of restricted cubic spline knots in all scenarios.
 
@@
 lowest median ICI for benefit in the majority of scenarios except for
-the scenarios where moderate linear deviations from the base case were
-considered. In that case constant treatment effect models demonstrated
-the best performance, very comparable to the linear interaction model's
-performance, nonetheless (Figure \ref{fig:calibration}; Panels A, B, and
-C).
+the scenarios where no or moderate linear deviations from the base case
+were considered. In that case constant treatment effect models
+demonstrated the best performance, very comparable to the linear
+interaction model's performance, nonetheless (Figure
+\ref{fig:calibration}). However, under strong linear or quadratic
+deviations, the constant treatment effect model was very poorly
+calibrated (Figure \ref{fig:calibration}; Panels C and D).
 
@@
 
-\hypertarget{real-data}{%
-\subsection{Real data}\label{real-data}}
+\hypertarget{case-study}{%
+\subsection{Case study}\label{case-study}}
 
@@
 
-We have carried out a simulation study comparing different approaches
-for individualized benefit prediction using baseline risk as the
-reference. These methods included a constant treatment effect model, a
-model including a simple linear interaction of the risk linear predictor
-with treatment, and models using smoothing with restricted cubic splines
-with 3, 4, and 5 knots. Performance was assessed evaluating error rates
-using RMSE and benefit prediction performance using some novel methods
-for evaluating discrimination and calibration for benefit.
-
 The linear interaction model proved to be flexible enough to adequately
@@
 with RCS smoothing with 3 knots performed very similarly, only
-outperformig the linear interaction model in the presence of strong
+outperforming the linear interaction model in the presence of strong
 quadratic deviations from the base case scenario of constant relative
@@
 
-In conclusion, we carried out a simulation study comparing different
-risk-based approaches to individualizing benefit predictions. Simpler
-models including a linear interaction with the risk linear predictor or
-RCS smoothing with 3 knots had adequate performance under a wide range
-scenarios. An adaptive approach based on selecting among the considered
-methods using AIC had satisfactory performance.
+In conclusion, when comparing different risk-based approaches to
+individualizing benefit predictions simpler models including a linear
+interaction with the prognostic index or RCS smoothing with 3 knots had
+adequate performance under a wide range scenarios. An adaptive approach
+selecting among the considered methods with optimal AIC had satisfactory
+performance.
 

### 2021-06-22 code/helpers/CreateManuscriptPlots.R
switch to linear interaction in figures

@@ -26,7 +26,7 @@ createPlot <- function(
     ) %>% 
     reshape2::melt()
   levels(tmp$variable) <- case_when(
-    levels(tmp$variable) == "linear_predictor"          ~ "Linear predictor",
+    levels(tmp$variable) == "linear_predictor"          ~ "Linear interaction",
     levels(tmp$variable) == "constant_treatment_effect" ~ "Constant\ntreatment effect",
     levels(tmp$variable) == "stratified"                ~ "Stratified",
     levels(tmp$variable) == "rcs_3_knots"               ~ "RCS\n(3 knots)",
@@ -60,7 +60,7 @@ createPlot <- function(
       breaks = c(
         "Constant\ntreatment effect", 
         "Stratified", 
-        "Linear predictor", 
+        "Linear interaction", 
         "RCS\n(3 knots)",
         "RCS\n(4 knots)",
         "RCS\n(5 knots)", 

### 2021-06-23 submission/manuscript.rmd
update figure captions

@@
 
-```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
+\\textit{Panel A} presents the results of the scenario with true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_base.png"))
@@
 ```{r rmsesamplesize, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
@@
 ```{r rmseauc, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_auc.png"))
@@
 scenario and the scenarios where linear and quadratic deviations were
-considered. However, constant treatment effects and linear interaction models
-tended to present much lower variability compared to all other model-based and
-smoothing approaches (Figure \ref{fig:discrimination}). We also observed an
-increasing trend of discrimination for benefit variability and decreasing median
-values with increasing number of restricted cubic spline knots in all scenarios.
+considered. The constant treatment effect model and the linear interaction model
+tended to present much lower variability compared to all other approaches
+(Figure \ref{fig:discrimination}). We also observed an increasing trend of
+discrimination for benefit variability and decreasing median values with
+increasing number of restricted cubic spline knots in all scenarios.
 
 ```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/discrimination_base.png"))
@@
 ```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effectst; \\textit{Panel C} presents the results for strong linear deviations from of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/calibration_base.png"))

### 2021-06-23 submission/manuscript.tex
update figure captions

@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:rmsebase}
+\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
+\textit{Panel A} presents the results of the scenario with true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:rmsebase}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_sample_size} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:rmsesamplesize}
+\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations constant relative treatment effects.}\label{fig:rmsesamplesize}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_auc} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:rmseauc}
+\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:rmseauc}
 \end{figure}
@@
 case scenario and the scenarios where linear and quadratic deviations
-were considered. However, constant treatment effects and linear
-interaction models tended to present much lower variability compared to
-all other model-based and smoothing approaches (Figure
-\ref{fig:discrimination}). We also observed an increasing trend of
-discrimination for benefit variability and decreasing median values with
-increasing number of restricted cubic spline knots in all scenarios.
+were considered. The constant treatment effect model and the linear
+interaction model tended to present much lower variability compared to
+all other approaches (Figure \ref{fig:discrimination}). We also observed
+an increasing trend of discrimination for benefit variability and
+decreasing median values with increasing number of restricted cubic
+spline knots in all scenarios.
 
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/discrimination_base} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:discrimination}
+\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:discrimination}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/calibration_base} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from the base case of constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from the base case scenario of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from the base case scenario of constant relative treatment effects.}\label{fig:calibration}
+\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effectst; \textit{Panel C} presents the results for strong linear deviations from of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:calibration}
 \end{figure}

### 2021-06-23 submission/manuscript.rmd
produce word document

@@
     keep_tex: true
+  bookdown::word_document2: default
 geometry: margin=1.0in

### 2021-06-23 submission/manuscript.rmd
adjust manuscript to the reference docx file

--- 2021-06-23 submission/manuscript.rmd
@@
     keep_tex: true
+  bookdown::word_document2:
+    reference_docx: reference.docx
 geometry: margin=1.0in

### 2021-06-23 code/SimulationScript.R
adjust default settings

@@ -128,10 +128,10 @@ simulationSettings <- list(
 )
 
 analysisSettings <- SimulationEvaluationHte::createAnalysisSettings(
-  threads        = 4,
+  threads        = 2,
   seed           = 19910930,
-  replications   = 100,
-  validationSize = 1e5,
+  replications   = 500,
+  validationSize = 5e5,
   analysisId     = paste(
     "scenario",
     idSettings$scenario,

### 2021-06-23 submission/supplement.rmd
write interaction results

@@
 
-```{r test, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
+When we considered a set of 4 true linear treatment-covariate interactions the
+model containing a linear interaction with the prognostic index had the lowest
+median RMSE. We observed an increasing trend in prediction errors with
+increasing interaction intensity (Figure \@ref(fig:rmseInteractions)). The
+model with restricted cubic spline smoothing (3 knots) had very comparable
+performance to the linear interaction model. Increasing the flexibility of the
+smooth methods resulted in increasing median RMSE. These results may be
+explained by the fact that the interactions considered were linear, thus
+favoring the linear interaction model. More flexible approaches may be better
+suited for higher-order treatment-covariate interactions. Finally, the adaptive
+approach had adequate performance under all scenarios, resembling the
+performance of the best-performing approach every time.
+
+```{r rmseInteractions, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_interactions.png"))
@@
 
-```{r test2, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
+The constant treatment effects model, the linear interaction model and the model
+with RCS smoothing (3 knots) had the highest c-for-benefit across all scenarios
+(Figrue \@ref(fig:discriminationInteractions)). RCS smoothing with 4 or 5 knots
+did not improve performance. On the contrary, we observed an increasing trend in
+c-for-benefit variability, as was the case in the main text. The adaptive
+apprach again had statisfactory performance.
+
+```{r discriminationInteractions, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/discrimination_interactions.png"))
@@
 
-```{r test4, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
+Despite the very similar performance in terms of prediction errors, the linear
+interaction model resulted in better-calibrated benefit predictions compared to
+the rest of the methods (Figure \@ref(fig:calibrationInteractions)). The constant
+treatment effects model had the highest median ICI-for-benefit across all
+scenario settings, which became more pornounced with increasing
+treatment-covariate interaction intensity.
+
+```{r calibrationInteractions, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/calibration_interactions.png"))

### 2021-06-25 submission/supplement.rmd
add prediction model of GUSTO-I

@@
 ```
+
+# Case study
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+load(here::here("data/raw/gusto.rda"))
+gusto <- gusto %>%
+  tibble() %>%
+  filter(!is.na(tpa))
+
+gustoTable1 <- gusto %>%
+  mutate(
+    tpa = factor(
+      tpa,
+      levels = c(0, 1),
+      labels = c(
+        "Tissue plasminogen activator",
+        "Streptokinase"
+      )
+    ),
+    Age = age,
+    Outcome = factor(
+      day30,
+      levels = c(0, 1),
+      labels = c(
+        "Alive",
+        "Dead"
+      )
+    )
+  )
+
+table1::table1(
+  ~ sex + Age + Killip + sysbp + + miloc + pulse + pmi + Outcome | tpa,
+  data = gustoTable1
+)
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+gustoModelData <- gusto %>%
+  tibble() %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+prediction <- rms::lrm(
+  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + rms::lsp(pulse, 50) + pmi + miloc,
+  data = gustoModelData,
+  maxit = 99
+)
+
+predictionStats <- rms::val.prob(
+  y = gustoModelData$outcome, 
+  logit = prediction$linear.predictors,
+  logistic.cal = F,
+  statloc = FALSE
+)
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+table <- tibble(
+  Metric = c(
+    "AUC",
+    "Brier score",
+    "Emax",
+    "E90",
+    "Eavg"
+  ),
+  Value = c(
+    predictionStats[2],
+    predictionStats["Brier"],
+    predictionStats["Emax"],
+    predictionStats["E90"],
+    predictionStats["Eavg"]
+  )
+)
+
+ knitr::kable(
+    table,
+    format      = "latex",
+    align       = "r",
+    booktabs    = TRUE,
+    digits      = 4,
+    format.args = list(
+      big.mar      = ",",
+      decimal.mark = "."
+    ),
+    caption    = "Evaluation statistics of the prediction model for 30-day mortality in GUSTO-I."
+  ) 
+```

### 2021-06-28 submission/manuscript.tex
update figures and rebuild
changes in discussion after DvK comments

@@
 \begin{abstract}
-\textbf{Objective:} Compare risk-based methods for individualizing
-treatment effects with simulations in the RCT setting. \textbf{Study
-Design and Setting:} We predicted absolute benefit based on an available
-prognostic index (PI) using: a model with the PI and a constant relative
-treatment effect; a model including an interaction of treatment with the
-PI; 4 quarters of the PI; nonlinear transformations of the PI
-(restricted cubic splines with 3, 4 and 5 knots); an adaptive model
-selection method using Akaike's Information Criterion. Starting from a
-base case scenario (sample size 4,250, constant odds ratio 0.8, AUC of
-the PI 0.75), we considered diverse assumptions by introducing linear
-and quadratic interactions of the PI with treatment and varying sample
-size and discriminative ability of the PI. We evaluated performance
-using root mean squared error, discrimination and calibration for
-benefit. \textbf{Results:} Models including a linear interaction of the
-PI with treatment had adequate performance that was robust under most
-simulation scenarios. Restricted cubic splines required larger sample
-sizes and higher AUC of the PI to achieve adequate performance. The
-adaptive approach performed equivalently to the best-performing method
-in each scenario. \textbf{Conclusion:} Usually, models with just a
-linear interaction of the PI with treatment adequately predict absolute
-benefit.
+\textbf{Objective:} Baseline risk is an important determinant of the
+absolute effect of a treatment for an individual patient. We aimed to
+compare different risk-based methods for predicting individualized
+treatment effects with simulations of the RCT setting. \textbf{Study
+Design} and Setting: We sampled patients from a superpopulation with
+diverse assumptions for a baseline prognostic index of risk (PI), and
+for the shape of the interaction between the PI and treatment (no,
+linear or quadratic interaction). In each sample, we fitted different
+models for predicting absolute benefit: a model with the PI and a
+constant relative treatment effect and models including an interaction
+of treatment assignment with the PI, with 4 quarters of the PI, and with
+nonlinear transformations of the PI (restricted cubic splines with 3, 4
+and 5 knots). We also used an adaptive model selection approach based on
+Akaike's Information Criterion. We evaluated predictive performance in
+the superpopulation using root mean squared error and measures of
+discrimination and calibration for benefit. Starting from a base case
+scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI
+0.75), we varied the sample size, the strength of the treatment effect,
+and the discriminative ability of the PI. \textbf{Results:} Models
+including a linear interaction of the PI with treatment had (close to?)
+optimal performance under most simulation scenarios. More flexible
+models with restricted cubic splines required larger sample sizes and
+higher AUC of the PI to outperform the linear model. The adaptive
+approach performed similarly to the best-performing method in each
+scenario. \textbf{Conclusion:} Under most circumstances, a model with a
+linear interaction of the PI with treatment is better able to predict
+absolute treatment benefit.
 \end{abstract}
@@
 
-Within the setting of patient-centered outcomes research, predictive
-approaches for assessing heterogeneity of treatment effects (HTE) aim at
-the development of models predicting either individualized effects or
-which of two (or more) treatments is better for an individual {[}1{]}.
-In prior work, we divided such methods in three broader categories based
-on the reference class used for defining patient similarity when making
-individualized predictions or recommendations {[}2{]}. Risk-modeling
-approaches use prediction of baseline risk as the reference; treatment
-effect modeling approaches also model treatment-covariate interactions,
-in addition to risk factors; optimal treatment regime approaches focus
-on developing treatment assignment rules and therefore rely heavily on
-modeling treatment effect modifiers.
+Predictive approaches for assessing heterogeneity of treatment effects
+(HTE) aim at the development of models predicting either individualized
+effects or which of two (or more) treatments is better for an individual
+{[}1{]}. In prior work, we divided such methods in three broader
+categories based on the reference class used for defining patient
+similarity when making individualized predictions or recommendations
+{[}2{]}. Risk-modeling approaches use prediction of baseline risk as the
+reference; treatment effect modeling approaches also model
+treatment-covariate interactions, in addition to risk factors; optimal
+treatment regime approaches focus on developing treatment assignment
+rules and therefore rely heavily on modeling treatment effect modifiers.
 
@@
 option in the absence of well-established treatment effect modifiers
-{[}3,4{]}. In simulations, modeling of effect modifiers in the form of
-treatment-covariate interactions often led to miscalibrated predictions
+{[}3,4{]}. In simulations, modeling of effect modifiers, i.e.
+treatment-covariate interactions, often led to miscalibrated predictions
 of benefit, while risk-based methods proved quite robust {[}5{]}. Most
@@
 treatment effect variation {[}6{]}. However, even though estimates at
-the risk subgroup level are accurate, this does not apply on the
-individual level, especially for patients with predicted risk at the
-boundaries of the risk intervals. Therefore, the risk-stratified
-approach should be used for exploring and presenting an overview of HTE,
-while inferences on the individual level should be made with caution.
+the risk subgroup level may be accurate, these estimated do not apply to
+individual patients, especially for patients with predicted risk at the
+boundaries of the risk intervals. Hence, the risk-stratified approach is
+useful for exploring and presenting HTE, but is not useful for
+supporting treatment decisions for individual patients.
 
-We aimed to provide an overview of methods that can be used to move from
-a risk-stratified approach to a continuous one using common smoothing
-techniques. These methods extend the risk-based framework of predictive
-HTE analyses to allow predictions on the individual level, within the
-RCT setting. We carried out a simulation study to compare the
-performance of these methods under different settings of increasing
-non-linearity of treatment effects. Finally, we carried out an
-application on real data as a demonstration of the considered
-techniques.
+To individualize treatment effects, the recent PATH statement suggested
+various risk-based models including a prognostic index of baseline risk
+(PI) and treatment assignment {[}cite PATH !!{]}. We aimed to summarize
+and compare different risk-based models for predicting individualized
+treatment effects. We simulated RCT settings to compare the performance
+of these models under different assumptions of the relationship between
+baseline risk and treatment. We illustrated the different models by a
+case study of predicting individualized effects of tissue plasminogen
+activator (tPA) versus streptokinase treatment in patients with an acute
+myocardial infarction (MI).
 
@@
 
-The linear interaction model proved to be flexible enough to adequately
-predict benefit under a wide range of scenarios. It had lower error
-rates and usually the highest c-for-benefit with the lower variability
-across simulation replications. Its benefit predictions were also
-well-calibrated across the majority of the simulation settings. Models
-with RCS smoothing with 3 knots performed very similarly, only
-outperforming the linear interaction model in the presence of strong
-quadratic deviations from the base case scenario of constant relative
-treatment effects. These simpler methods proved to have quite robust
-performance across all scenarios, which was not the case for more
-flexible approaches.
+The linear interaction model proved to be superior to alternative
+approaches when predicting risk-based treatment benefit under a wide
+range of scenarios. It generally had lower mean squared error, lower
+e-for-benefit and higher c-for-benefit with lower variability across
+simulation replications. Models including restricted cubic splines with
+3 knots only outperformed the linear interaction model in the presence
+of strong quadratic deviations from a constant relative treatment
+effect.
 
-Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry
-and this often led to overfitting with smaller sample sizes. This
-resulted in higher RMSE, increased variability of discrimination for
-benefit and worse-calibrated benefit predictions across simulation
-replications. Even with larger sample sizes and strong quadratic
-deviations from the base case scenario of constant relative treatment
-effects, these more flexible RCS smoothing methods did not outperform
-simpler RCS smoothing with 3 knots. More flexible approaches usually do
-not come with any added value, unless we expect extreme heterogeneity in
-treatment effects.
+Including restricted cubic splines with 4 or 5 knots proved to be too
+flexible, as indicated by higher RMSE, increased variability of
+discrimination for benefit and worse calibration of benefit predictions.
+Even with larger sample sizes and strong quadratic deviations from the
+base case scenario of constant relative treatment effects, these more
+flexible restricted cubic splines did not outperform the simpler RCS
+with 3 knots. These more flexible approaches may only be helpful if we
+expect more extreme patterns of heterogeneous treatment effects.
 
 The constant treatment effect model, despite having adequate performance
-even in the presence of weak treatment effect heterogeneity on the
-relative scale, it quickly broke down with stronger deviations from the
-base case scenario of constant effects. In these cases, the stratified
-approach generally had lower error rates compared to the constant
-treatment effect model. In concept, the stratified approach lies between
-the constant effects model and smoother approaches, only assuming
-constant treatment effects within strata of predicted risk and,
-therefore, is more sensitive to treatment effect heterogeneity. However,
-these stepwise benefit estimates can only be used as a demonstration of
-treatment effect heterogeneity, identifying patient subgroups where
-treatment benefits are expected to be higher or lower, and should not be
-used to make individualized absolute benefit predictions.
+in the presence of weak treatment effect heterogeneity on the relative
+scale, quickly broke down with stronger deviations from constant
+relative treatment effects. In these cases, the stratified approach
+generally had lower error rates compared to the constant treatment
+effect model. In concept, the stratified approach lies between the
+constant effects model and smoother approaches, only assuming constant
+treatment effects within strata of predicted risk and, therefore, is
+more sensitive to treatment effect heterogeneity. Stepwise treatment
+benefit estimates are very useful for demonstrating treatment effect
+heterogeneity -- because estimating treatment effect requires groups of
+patients rather than individual patients --but are not helpful for
+making individualized absolute benefit predictions.
 
-Increasing the true prediction AUC by increasing the predictor
-coefficients of the true risk model reduced RMSE for all methods,
-similar to increasing the sample size. Higher AUC translates in higher
-variability of predicted risks, which, in turn, allows the considered
-methods to better follow the evolution of treatment benefit. Higher
-prediction AUC also led to higher absolute values of c-for-benefit. Even
-though, model performance appears to be very important for the
-performance of these methods, prediction model development was outside
-the scope of this work and has already been studied quite extensively
-{[}5,12,13{]}.
+Increasing the discriminative ability of the risk model -- by increasing
+the predictor coefficients of the true risk model -- reduced RMSE for
+all methods, similar to increasing the sample size. This increase in
+discriminative ability translates in higher variability of predicted
+risks, which, in turn, allows the considered methods to better follow
+the evolution of treatment benefit. As expected, the increase in
+discriminative ability of the risk model also led to higher values of
+c-for-benefit. Even though risk model performance is very important for
+the ability of risk-based methods to predict treatment benefit,
+prediction model development was outside the scope of this work and has
+already been studied extensively {[}5,12,13{]}.
 
-Risk-based approaches to predictive HTE assume that benefit is a
-function of baseline risk. A limitation of our study is that this was
-assumed to be the case in the majority of the simulation scenarios. Even
-though our main conclusions did not change when simulating true
-treatment-covariate interactions (results in supplement), RMSE had
-higher values. The linear interaction model under these scenarios again
-had the best overall performance under different sizes of the
-interaction effects. This can be explained by the fact that linear
-interactions with treatment were simulated, thus favoring the linear
-interaction model. In studies evaluating prediction model development,
-the underlying outcome generation process sometimes proved to affect
-performance of certain methods {[}14{]}. This might be the case here as
-well and should be explored in more extensive future simulation studies.
+Risk-based approaches to predictive HTE estimate treatment benefit as a
+function of baseline risk. A limitation of our study is that we assumed
+treatment benefit to be a function of baseline risk in the majority of
+the simulation scenarios. Nevertheless, our main conclusions did not
+change when we generated individual treatment-covariate interactions
+(supplementary table/figure x). Future simulation studies could explore
+the effect of more extensive deviations from risk-based treatment
+effects.
 
@@
 from treatment effect modeling to tree-based approaches available in
-more recent literature {[}15--17{]}. Simulations are also needed to
+more recent literature {[}14--16{]}. Simulations are also needed to
 assess relative performance and define the settings where these break
@@
 
-In conclusion, when comparing different risk-based approaches to
-individualizing benefit predictions simpler models including a linear
-interaction with the prognostic index or RCS smoothing with 3 knots had
-adequate performance under a wide range scenarios. An adaptive approach
-selecting among the considered methods with optimal AIC had satisfactory
-performance.
+In conclusion, when comparing different risk-based approaches to predict
+individualizing benefit a models including a linear treatment
+interaction with the prognostic performed best in a wide range of
+scenarios. More flexible models with restricted cubic splines required
+larger sample sizes and higher AUC of the PI to outperform the linear
+model. An adaptive approach, selecting the model with optimal AIC, had
+satisfactory performance.
 
@@
 
-\leavevmode\hypertarget{ref-Austin2021}{}%
-{[}14{]} Austin PC, Harrell FE, Steyerberg EW. Predictive performance of
-machine and statistical learning methods: Impact of data-generating
-processes on external validity in the ``large n, small p'' setting.
-Statistical Methods in Medical Research 2021;30:1465--83.
-\url{https://doi.org/10.1177/09622802211002867}.
-
 \leavevmode\hypertarget{ref-Athey2019}{}%
-{[}15{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
+{[}14{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
 Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.
@@
 \leavevmode\hypertarget{ref-Lu2018}{}%
-{[}16{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
+{[}15{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
 treatment effect in observational data using random forest methods.
@@
 \leavevmode\hypertarget{ref-Wager2018}{}%
-{[}17{]} Wager S, Athey S. Estimation and inference of heterogeneous
+{[}16{]} Wager S, Athey S. Estimation and inference of heterogeneous
 treatment effects using random forests. Journal of the American

### 2021-06-28 submission/manuscript.rmd

@@
 abstract: |
-  **Objective:** Compare risk-based methods for individualizing treatment 
-  effects with simulations in the RCT setting. **Study Design and Setting:** We
-  predicted absolute benefit based on an available prognostic index (PI) using:
-  a model with the PI and a constant relative treatment effect; a model 
-  including an interaction of treatment with the PI; 4 quarters of the PI; 
-  nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5
-  knots); an adaptive model selection method using Akaike’s Information 
-  Criterion. Starting from a base case scenario (sample size 4,250, constant 
-  odds ratio 0.8, AUC of the PI 0.75), we considered diverse assumptions by
-  introducing linear and quadratic interactions of the PI with treatment and 
-  varying sample size and discriminative ability of the PI. We evaluated
-  performance using root mean squared error, discrimination and calibration for 
-  benefit. **Results:** Models including a linear interaction of the PI with
-  treatment had adequate performance that was robust under most simulation 
-  scenarios. Restricted cubic splines required larger sample sizes and higher 
-  AUC of the PI to achieve adequate performance. The adaptive approach performed 
-  equivalently to the best-performing method in each scenario. **Conclusion:** 
-  Usually, models with just a linear interaction of the PI with treatment
-  adequately predict absolute benefit.
+ **Objective:** Baseline risk is an important determinant of the absolute effect of a treatment for an individual patient.
+We aimed to compare different risk-based methods for predicting individualized treatment effects with simulations of the RCT setting.
+**Study Design** and Setting: We sampled patients from a superpopulation with diverse assumptions for a baseline prognostic index of risk (PI), and for the shape of the interaction between the PI and treatment (no, linear or quadratic interaction).
+In each sample, we fitted different models for predicting absolute benefit: a model with the PI and a constant relative treatment effect and models including an interaction of treatment assignment with the PI, with 4 quarters of the PI, and with nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots).
+We also used an adaptive model selection approach based on Akaike’s Information Criterion.
+We evaluated predictive performance in the superpopulation using root mean squared error and measures of discrimination and calibration for benefit.
+Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size,  the strength of the treatment effect, and the discriminative ability of the PI.
+**Results:** Models including a linear interaction of the PI with treatment had (close to?) optimal performance under most simulation scenarios.
+More flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model.
+The adaptive approach performed similarly to the best-performing method in each scenario.
+**Conclusion:** Under most circumstances, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
 authors:
@@
 # Introduction
-
-Within the setting of patient-centered outcomes research, predictive approaches
-for assessing heterogeneity of treatment effects (HTE) aim at the development of
-models predicting either individualized effects or which of two (or more)
-treatments is better for an individual [@Varadhan2013]. In prior work, we
-divided such methods in three broader categories based on the reference class
-used for defining patient similarity when making individualized predictions or
-recommendations [@Rekkas2020]. Risk-modeling approaches use prediction of
-baseline risk as the reference; treatment effect modeling approaches also model
-treatment-covariate interactions, in addition to risk factors; optimal treatment
-regime approaches focus on developing treatment assignment rules and therefore
-rely heavily on modeling treatment effect modifiers.
+Predictive approaches for assessing heterogeneity of treatment effects (HTE) aim
+at the development of models predicting either individualized effects or which
+of two (or more) treatments is better for an individual [@Varadhan2013]. In
+prior work, we divided such methods in three broader categories based on the
+reference class used for defining patient similarity when making individualized
+predictions or recommendations [@Rekkas2020]. Risk-modeling approaches use
+prediction of baseline risk as the reference; treatment effect modeling
+approaches also model treatment-covariate interactions, in addition to risk
+factors; optimal treatment regime approaches focus on developing treatment
+assignment rules and therefore rely heavily on modeling treatment effect
+modifiers.
 
@@
 the absence of well-established treatment effect modifiers [@Kent2019;
-@PathEnE]. In simulations, modeling of effect modifiers in the form of
-treatment-covariate interactions often led to miscalibrated predictions of
+@PathEnE]. In simulations, modeling of effect modifiers, i.e.
+treatment-covariate interactions, often led to miscalibrated predictions of
 benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
@@
 prediction model is developed externally or internally on the entire RCT
-population, "blinded" to treatment; then the RCT population is stratified using
+population, “blinded” to treatment; then the RCT population is stratified using
 this prediction model to evaluate risk-based treatment effect variation
-[@Kent2010]. However, even though estimates at the risk subgroup level are
-accurate, this does not apply on the individual level, especially for patients
-with predicted risk at the boundaries of the risk intervals. Therefore, the
-risk-stratified approach should be used for exploring and presenting an overview
-of HTE, while inferences on the individual level should be made with caution.
+[@Kent2010]. However, even though estimates at the risk subgroup level may be
+accurate, these estimated do not apply to individual patients, especially for
+patients with predicted risk at the boundaries of the risk intervals. Hence, the
+risk-stratified approach is useful for exploring and presenting HTE, but is not
+useful for supporting treatment decisions for individual patients.
 
-We aimed to provide an overview of methods that can be used to move from a
-risk-stratified approach to a continuous one using common smoothing techniques.
-These methods extend the risk-based framework of predictive HTE analyses to
-allow predictions on the individual level, within the RCT setting. We carried
-out a simulation study to compare the performance of these methods under
-different settings of increasing non-linearity of treatment effects. Finally, we
-carried out an application on real data as a demonstration of the considered
-techniques.
+To individualize treatment effects, the recent PATH statement suggested various
+risk-based models including a prognostic index of baseline risk (PI) and
+treatment assignment [cite PATH !!]. We aimed to summarize and compare different
+risk-based models for predicting individualized treatment effects. We simulated
+RCT settings to compare the performance of these models under different
+assumptions of the relationship between baseline risk and treatment. We
+illustrated the different models by a case study of predicting individualized
+effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
+patients with an acute myocardial infarction (MI).
 
@@
 
-The linear interaction model proved to be flexible enough to adequately predict
-benefit under a wide range of scenarios. It had lower error rates and usually
-the highest c-for-benefit with the lower variability across simulation
-replications. Its benefit predictions were also well-calibrated across the
-majority of the simulation settings. Models with RCS smoothing with 3 knots
-performed very similarly, only outperforming the linear interaction model in the
-presence of strong quadratic deviations from the base case scenario of constant
-relative treatment effects. These simpler methods proved to have quite robust
-performance across all scenarios, which was not the case for more flexible
-approaches.
+The linear interaction model proved to be superior to alternative approaches
+when predicting risk-based treatment benefit under a wide range of scenarios. It
+generally had lower mean squared error, lower e-for-benefit and higher
+c-for-benefit with lower variability across simulation replications. Models
+including restricted cubic splines with 3 knots only outperformed the linear
+interaction model in the presence of strong quadratic deviations from a constant
+relative treatment effect.
 
-Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry and this
-often led to overfitting with smaller sample sizes. This resulted in higher
-RMSE, increased variability of discrimination for benefit and worse-calibrated
-benefit predictions across simulation replications. Even with larger sample
-sizes and strong quadratic deviations from the base case scenario of constant
-relative treatment effects, these more flexible RCS smoothing methods did not
-outperform simpler RCS smoothing with 3 knots. More flexible approaches usually
-do not come with any added value, unless we expect extreme heterogeneity in
-treatment effects.
+Including restricted cubic splines with 4 or 5 knots proved to be too flexible,
+as indicated by higher RMSE, increased variability of discrimination for benefit
+and worse calibration of benefit predictions. Even with larger sample sizes and
+strong quadratic deviations from the base case scenario of constant relative
+treatment effects, these more flexible restricted cubic splines did not
+outperform the simpler RCS with 3 knots. These more flexible approaches may only
+be helpful if we expect more extreme patterns of heterogeneous treatment
+effects.
 
-The constant treatment effect model, despite having adequate performance even in
-the presence of weak treatment effect heterogeneity on the relative scale, it
-quickly broke down with stronger deviations from the base case scenario of
-constant effects. In these cases, the stratified approach generally had lower
-error rates compared to the constant treatment effect model. In concept, the
-stratified approach lies between the constant effects model and smoother
-approaches, only assuming constant treatment effects within strata of predicted
-risk and, therefore, is more sensitive to treatment effect heterogeneity.
-However, these stepwise benefit estimates can only be used as a demonstration of
-treatment effect heterogeneity, identifying patient subgroups where treatment
-benefits are expected to be higher or lower, and should not be used to make
-individualized absolute benefit predictions.
+The constant treatment effect model, despite having adequate performance in the
+presence of weak treatment effect heterogeneity on the relative scale, quickly
+broke down with stronger deviations from constant relative treatment effects. In
+these cases, the stratified approach generally had lower error rates compared to
+the constant treatment effect model. In concept, the stratified approach lies
+between the constant effects model and smoother approaches, only assuming
+constant treatment effects within strata of predicted risk and, therefore, is
+more sensitive to treatment effect heterogeneity. Stepwise treatment benefit
+estimates are very useful for demonstrating treatment effect heterogeneity –
+because estimating treatment effect requires groups of patients rather than
+individual patients –but are not helpful for making individualized absolute
+benefit predictions.
 
-Increasing the true prediction AUC by increasing the predictor coefficients of
-the true risk model reduced RMSE for all methods, similar to increasing the
-sample size. Higher AUC translates in higher variability of predicted risks,
-which, in turn, allows the considered methods to better follow the evolution of
-treatment benefit. Higher prediction AUC also led to higher absolute values of
-c-for-benefit. Even though, model performance appears to be very important for
-the performance of these methods, prediction model development was outside the
-scope of this work and has already been studied quite extensively
-[@vanKlaveren2019; @Burke2014; @Abadie2018].
+Increasing the discriminative ability of the risk model – by increasing the
+predictor coefficients of the true risk model – reduced RMSE for all methods,
+similar to increasing the sample size. This increase in discriminative ability
+translates in higher variability of predicted risks, which, in turn, allows the
+considered methods to better follow the evolution of treatment benefit. As
+expected, the increase in discriminative ability of the risk model also led to
+higher values of c-for-benefit. Even though risk model performance is very
+important for the ability of risk-based methods to predict treatment benefit,
+prediction model development was outside the scope of this work and has already
+been studied extensively [@vanKlaveren2019; @Burke2014; @Abadie2018].
 
-Risk-based approaches to predictive HTE assume that benefit is a function of
-baseline risk. A limitation of our study is that this was assumed to be the case
-in the majority of the simulation scenarios. Even though our main conclusions
-did not change when simulating true treatment-covariate interactions (results in
-supplement), RMSE had higher values. The linear interaction model under these
-scenarios again had the best overall performance under different sizes of the
-interaction effects. This can be explained by the fact that linear interactions
-with treatment were simulated, thus favoring the linear interaction model. In
-studies evaluating prediction model development, the underlying outcome
-generation process sometimes proved to affect performance of certain methods
-[@Austin2021]. This might be the case here as well and should be explored in
-more extensive future simulation studies.
+Risk-based approaches to predictive HTE estimate treatment benefit as a function
+of baseline risk. A limitation of our study is that we assumed treatment benefit
+to be a function of baseline risk in the majority of the simulation scenarios.
+Nevertheless, our main conclusions did not change when we generated individual
+treatment-covariate interactions (supplementary table/figure x). Future
+simulation studies could explore the effect of more extensive deviations from
+risk-based treatment effects.
 
@@
 
-In conclusion, when comparing different risk-based approaches to individualizing
-benefit predictions simpler models including a linear interaction with the
-prognostic index or RCS smoothing with 3 knots had adequate performance under a
-wide range scenarios. An adaptive approach selecting among the considered
-methods with optimal AIC had satisfactory performance.
+In conclusion, when comparing different risk-based approaches to predict
+individualizing benefit a models including a linear treatment interaction with
+the prognostic performed best in a wide range of scenarios. More flexible models
+with restricted cubic splines required larger sample sizes and higher AUC of the
+PI to outperform the linear model. An adaptive approach, selecting the model
+with optimal AIC, had satisfactory performance.
 

### 2021-06-28 submission/supplement.rmd
work on gusto part

--- 2021-06-23 submission/supplement.rmd
@@
 ```
+
+# Case study
+
+We fitted a logistic regression model to predict 30-day mortality after a
+myocardial infarction using data from the GUSTO-I trial. The risk factors
+modeled were age, Killip class, systolic blood pressure heart rate, primary MI
+and MI location. We restricted systolic blood pressure values to 120 mmHg and
+modeled heart rate using linear spline transformation with a knot at 50 beats
+per minute. A brief overview of the distribution of baseline characteristics is
+presented in Table \@ref(tab:table1). The coefficients of the model can be seen
+in Table \@ref(tab:modelCoefficients). We used this model in the manuscript to
+derive baseline risk predictions to all patients of the study. For all patients
+we set the treatment indicator to 0, assuming a constant relative treatment
+effect. The model displayed very good internal validity (Table
+\@ref(tab:internalValidity)).
+
+```{r table1, echo=FALSE, warning=FALSE, message=FALSE}
+load(here::here("data/raw/gusto.rda"))
+gusto <- gusto %>%
+  tibble() %>%
+  filter(!is.na(tpa))
+
+gustoTable1 <- gusto %>%
+  mutate(
+    tpa = factor(
+      tpa,
+      levels = c(0, 1),
+      labels = c(
+        "Tissue plasminogen activator",
+        "Streptokinase"
+      )
+    ),
+    Age = age,
+    Outcome = factor(
+      day30,
+      levels = c(0, 1),
+      labels = c(
+        "Alive",
+        "Dead"
+      )
+    )
+  )
+
+res <- table1::table1(
+  ~ sex + Age + Killip + sysbp + + miloc + pulse + pmi + Outcome | tpa,
+  data = gustoTable1,
+  caption = "Baseline characteristics of the patients included in GUSTO-I trial."
+)
+
+table1::t1kable(res)
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Smooth calibration of the derived prediction model."}
+gustoModelData <- gusto %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+prediction <- glm(
+  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + rms::lsp(pulse, 50) + pmi + miloc,
+  data = gustoModelData,
+  maxit = 99,
+  family = "binomial"
+)
+
+predictionStats <- rms::val.prob(
+  y = gustoModelData$outcome, 
+  logit = prediction$linear.predictors,
+  logistic.cal = FALSE,
+  statloc = FALSE
+)
+```
+
+```{r modelCoefficients, echo=FALSE, warning=FALSE, message=FALSE}
+
+predictionSummary <- summary(prediction)
+table <- predictionSummary$coefficients
+
+rownames(table)[2] <- "Treatment"
+rownames(table)[3] <- "Age"
+rownames(table)[4] <- "KILLIP (II)"
+rownames(table)[5] <- "KILLIP (III)"
+rownames(table)[6] <- "KILLIP (IV)"
+rownames(table)[7] <- "Systolic blood pressure"
+rownames(table)[8] <- "Pulse rate"
+rownames(table)[9] <- "Pulse rate'"
+rownames(table)[10] <- "Prior MI (yes)"
+rownames(table)[11] <- "MI location (other)"
+rownames(table)[12] <- "MI location (anterior)"
+
+ knitr::kable(
+    table,
+    format      = "latex",
+    align       = "r",
+    booktabs    = TRUE,
+    digits      = 4,
+    format.args = list(
+      big.mar      = ",",
+      decimal.mark = "."
+    ),
+    caption    = "Coefficients of the prediction model for 30-day mortality after myocardial infarction (MI)."
+  )
+
+```
+
+```{r internalValidity, echo=FALSE, warning=FALSE, message=FALSE}
+table <- tibble(
+  Metric = c(
+    "AUC",
+    "Brier score",
+    "Emax",
+    "E90",
+    "Eavg"
+  ),
+  Value = c(
+    predictionStats[2],
+    predictionStats["Brier"],
+    predictionStats["Emax"],
+    predictionStats["E90"],
+    predictionStats["Eavg"]
+  )
+)
+
+ knitr::kable(
+    table,
+    format      = "latex",
+    align       = "r",
+    booktabs    = TRUE,
+    digits      = 4,
+    format.args = list(
+      big.mar      = ",",
+      decimal.mark = "."
+    ),
+    caption    = "Evaluation statistics of the prediction model for 30-day mortality in GUSTO-I."
+  ) 
+```

### 2021-06-29 submission/supplement.rmd
add performance metrics in gusto

@@
   - \newcommand\given[1][]{\:#1\vert\:}
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
+
 editor_options: 
@@
 ```
+
+\beginsupplement
 
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Smooth calibration of the derived prediction model."}
+```{r modelCoefficients, echo=FALSE, warning=FALSE, message=FALSE}
 gustoModelData <- gusto %>%
@@
 )
-
-predictionStats <- rms::val.prob(
-  y = gustoModelData$outcome, 
-  logit = prediction$linear.predictors,
-  logistic.cal = FALSE,
-  statloc = FALSE
-)
-```
-
-```{r modelCoefficients, echo=FALSE, warning=FALSE, message=FALSE}
 
@@
 
+```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Smooth calibration of the derived prediction model."}
+predictionStats <- rms::val.prob(
+  y = gustoModelData$outcome, 
+  logit = prediction$linear.predictors,
+  logistic.cal = FALSE,
+  statloc = FALSE
+)
+```
+
 ```{r internalValidity, echo=FALSE, warning=FALSE, message=FALSE}
@@
     format      = "latex",
-    align       = "r",
+    align       = "l",
     booktabs    = TRUE,
@@
 ```
+
+We used the derived prediction model to predict benefit of tPA to 30-day
+mortality using the data at hand. The models are presented in the main
+manuscript. All approaches displayed comaprable discrimination for benefit, with
+the constant treatment effect model, the linear interaction model and the RCS (5
+knots) model displaying the highest performance (Table
+\@ref(tab:gustoPerformanceMetrics)). However, the linear interaction model and
+the RCS (5 knots) model were worse calibrated. The constant treatment effect
+model also had the best performance among the considered methods in terms of
+calibration for benefit.
+
+```{r gustoPerformanceMetrics, echo=FALSE, warning=FALSE, message=FALSE}
+metrics <- readr::read_csv(
+  here::here("data/processed/gustoPerformanceMetrics.csv")
+)
+
+ knitr::kable(
+    metrics,
+    format      = "latex",
+    align       = "l",
+    booktabs    = TRUE,
+    digits      = 4,
+    format.args = list(
+      big.mar      = ",",
+      decimal.mark = "."
+    ),
+    col.names = c(
+      "Model",
+      "C",
+      "ICI"
+    ),
+    caption    = "Performance metrics of the different methods used to predict benefit of tPA on 30-day mortality using data from the GUSTO-I trial."
+  ) 
+```

### 2021-06-29 submission/supplement.rmd
update supplement

--- 2021-06-28 submission/supplement.rmd
@@
 
-In Figure \@ref(fig:fig2) the absolute benefits observed based on different settings are presented. The base-case scenario is also presented as a reference.
+In Figure \@ref(fig:fig2) the absolute benefits observed based on different
+settings are presented. The base-case scenario is also presented as a reference.
 
@@
 
-Finally, we consider 3 additional scenarios of interaction of individual covariates with treatment.
-These scenarios include a 4 weak interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2 weak and 2 strong interactions (Table \@ref(tab:tab2)).
+Finally, we consider 3 additional scenarios of interaction of individual
+covariates with treatment. These scenarios include a 4 weak interactions
+($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2 weak and 2 strong interactions (Table \@ref(tab:tab2)).
 
@@
 ### Risk stratification
-Derive a prediction model using the same approach as above and divide the population in equally sized risk-based subgroups.
-Estimate subgroup-specific absolute benefit from the observed absolute differences.
-Subject-specific benefit predictions are made by attributing to individuals their corresponding subgroup-specific estimate.
+Derive a prediction model using the same approach as above and divide the
+population in equally sized risk-based subgroups. Estimate subgroup-specific
+absolute benefit from the observed absolute differences. Subject-specific
+benefit predictions are made by attributing to individuals their corresponding
+subgroup-specific estimate.
 
 ### Constant treatment effect
-Assuming a constant relative treatment effect, fit the adjusted model in \@ref(eq:risk).
-Then, an estimate of absolute benefit can be derived from $$\hat{f}_{\text{benefit}}(lp\given \bm{x}, \hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
+Assuming a constant relative treatment effect, fit the adjusted model in
+\@ref(eq:risk). Then, an estimate of absolute benefit can be derived from$$\hat{f}_{\text{benefit}}(lp\given \bm{x}, \hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
 
 ### Linear interaction
-The assumption of constant relative treatment effect is relaxed modeling a linear interaction of treatment with the risk linear predictor: 
-$$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} = g\big(lp+(\delta_0+\delta_1lp)t_x\big)$$
-We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) = g(lp) - g\big(\delta_0+(1+\delta_1)lp\big)$$
+The assumption of constant relative treatment effect is relaxed modeling a
+linear interaction of treatment with the risk linear predictor:
+$$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} = g\big(lp+(\gamma_0+\gamma_1lp)t_x\big)$$
+We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) = g(lp) - g\big(\gamma_0+(1+\gamma_1)lp\big)$$
 
-### Non-linear interaction
-Finally, we drop the linearity assumption and predict absolute benefit by taking the difference between smooth fits, separately derived in each treatment arm:
-$$f_{\text{benefit}}(lp\given\bm{x},\hat{\bm{\beta}}) = \hat{f}_{\text{smooth}}(lp\given \bm{x}, \hat{\bm{\beta}}, t_x=0) - \hat{f}_{\text{smooth}}(lp\given \bm{x}, \hat{\bm{\beta}}, t_x=1)$$
+### Restricted cubic splines
+Finally, we drop the linearity assumption and predict absolute benefit using smoothing with restricted cubic splines with 3, 4, and 5 knots. More specifically, we fit the model:
+
+$$P(y=1\given lp, t_x) = g\big(\beta_0+\beta_{t_x}t_x+f_{RCS}(lp) + f_{RCS}(lp)\times t_x\big)$$
+where
+$$f_{RCS}(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$ with
+$$h_{j+1}(x)= (x-t_j)^3-(x-t_{k-1})_+\frac{t_k-t_j}{t_k-t_{k-1}}+(x-t_k)^3_+\frac{t_{k-1}-t_j}{t_k-t_{k-   1}}$$
+and $t_1,\dots,t_k$ are the selected knots. We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp \given \bm{x},\hat{\bm{\beta}})=P(y=1\given lp, t_x=0)-P(y=1\given lp,t_x=1)$$.
 

### 2021-06-30 submission/supplement.rmd
fix typo; closes #1
rebuild supplement

@@
   - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
-
 editor_options: 
@@
 ### Restricted cubic splines
-Finally, we drop the linearity assumption and predict absolute benefit using smoothing with restricted cubic splines with 3, 4, and 5 knots. More specifically, we fit the model:
+Finally, we drop the linearity assumption and predict absolute benefit using
+smoothing with restricted cubic splines with 3, 4, and 5 knots. More
+specifically, we fit the model:
 

### 2021-06-30 submission/manuscript.rmd

@@
 constant. In that case, absolute benefit is estimated from
-$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$,
-where $\text{expit}(x)=\frac{e^x}{1+e^x}$
-and *lp* is the linear predictor of the prediction model.
+$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$. This is the method
+usually considered when performing risk-based assessment of treatment effect
+heterogeneity in RCT data.
 
@@
 index and their interaction. In this case absolute benefit is estimated from
-$\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
+$\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
 We will refer to this method as the *linear interaction* approach.

### 2021-06-30 submission/manuscript.tex

@@
 % For Pandoc 2.11+
-\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
  {% don't indent paragraphs
@@
 \newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 effect} (OR) is constant. In that case, absolute benefit is estimated
-from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\), where
-\(\text{expit}(x)=\frac{e^x}{1+e^x}\) and \emph{lp} is the linear
-predictor of the prediction model.
+from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\). This is
+the method usually considered when performing risk-based assessment of
+treatment effect heterogeneity in RCT data.
 
@@
 estimated from
-\(\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
+\(\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
 We will refer to this method as the \emph{linear interaction} approach.

### 2021-07-05 submission/manuscript.rmd
include dvk's comments to discussion

@@
 ---
-title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
+title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit"
 abstract: |
- **Objective:** Baseline risk is an important determinant of the absolute effect of a treatment for an individual patient.
-We aimed to compare different risk-based methods for predicting individualized treatment effects with simulations of the RCT setting.
-**Study Design** and Setting: We sampled patients from a superpopulation with diverse assumptions for a baseline prognostic index of risk (PI), and for the shape of the interaction between the PI and treatment (no, linear or quadratic interaction).
-In each sample, we fitted different models for predicting absolute benefit: a model with the PI and a constant relative treatment effect and models including an interaction of treatment assignment with the PI, with 4 quarters of the PI, and with nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots).
-We also used an adaptive model selection approach based on Akaike’s Information Criterion.
-We evaluated predictive performance in the superpopulation using root mean squared error and measures of discrimination and calibration for benefit.
-Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size,  the strength of the treatment effect, and the discriminative ability of the PI.
-**Results:** Models including a linear interaction of the PI with treatment had (close to?) optimal performance under most simulation scenarios.
-More flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model.
-The adaptive approach performed similarly to the best-performing method in each scenario.
-**Conclusion:** Under most circumstances, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
+ **Objective:** To compare different risk-based methods for predicting individualized treatment effects using simulations of the RCT setting.
+**Study Design and Setting:** We sampled patients using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic).
+In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in 4 quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive model selection approach based on Akaike’s Information Criterion.
+We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit.
+Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample size, treatment effect strength, and the PI's discriminative ability.
+**Results:** Models including a PI by treatment interaction performed better under most simulation settings.
+Flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model.
+The adaptive approach performed similarly to the best-performing method in most scenarios.
+**Conclusion:** Usually, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
 authors:
@@
 these cases, the stratified approach generally had lower error rates compared to
-the constant treatment effect model. In concept, the stratified approach lies
-between the constant effects model and smoother approaches, only assuming
-constant treatment effects within strata of predicted risk and, therefore, is
-more sensitive to treatment effect heterogeneity. Stepwise treatment benefit
-estimates are very useful for demonstrating treatment effect heterogeneity –
-because estimating treatment effect requires groups of patients rather than
-individual patients –but are not helpful for making individualized absolute
-benefit predictions.
+the constant treatment effect model. Stepwise treatment benefit estimates are
+very useful for demonstrating treatment effect heterogeneity--because
+estimating treatment effect requires groups of patients rather than individual
+patients--but are not helpful for making individualized absolute benefit
+predictions.
 
 Increasing the discriminative ability of the risk model – by increasing the
-predictor coefficients of the true risk model – reduced RMSE for all methods,
-similar to increasing the sample size. This increase in discriminative ability
-translates in higher variability of predicted risks, which, in turn, allows the
-considered methods to better follow the evolution of treatment benefit. As
-expected, the increase in discriminative ability of the risk model also led to
-higher values of c-for-benefit. Even though risk model performance is very
-important for the ability of risk-based methods to predict treatment benefit,
-prediction model development was outside the scope of this work and has already
-been studied extensively [@vanKlaveren2019; @Burke2014; @Abadie2018].
+predictor coefficients of the true risk model – reduced RMSE for all methods.
+This increase in discriminative ability translates in higher variability of
+predicted risks, which, in turn, allows the considered methods to better capture
+absolute treatment benefits. As a consequence, the increase in discriminative
+ability of the risk model also led to higher values of c-for-benefit. Even
+though risk model performance is very important for the ability of risk-based
+methods to predict treatment benefit, prediction model development was outside
+the scope of this work and has already been studied extensively
+[@vanKlaveren2019; @Burke2014; @Abadie2018].
 
@@
 PI to outperform the linear model. An adaptive approach, selecting the model
-with optimal AIC, had satisfactory performance.
+with optimal AIC, had comparable performance to the best performing approach in
+most of the scenarios.
 

### 2021-07-05 submission/manuscript.tex
include dvk's comments to discussion

@@
 
-\title{Continuous risk-based predictive approaches to treatment effect
-heterogeneity: A simulation study}
+\title{Models including an interaction of treatment with baseline risk
+were better able to predict absolute benefit}
 
@@
 % For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
  {% don't indent paragraphs
@@
 \newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \begin{abstract}
-\textbf{Objective:} Baseline risk is an important determinant of the
-absolute effect of a treatment for an individual patient. We aimed to
-compare different risk-based methods for predicting individualized
-treatment effects with simulations of the RCT setting. \textbf{Study
-Design} and Setting: We sampled patients from a superpopulation with
-diverse assumptions for a baseline prognostic index of risk (PI), and
-for the shape of the interaction between the PI and treatment (no,
-linear or quadratic interaction). In each sample, we fitted different
-models for predicting absolute benefit: a model with the PI and a
-constant relative treatment effect and models including an interaction
-of treatment assignment with the PI, with 4 quarters of the PI, and with
-nonlinear transformations of the PI (restricted cubic splines with 3, 4
-and 5 knots). We also used an adaptive model selection approach based on
-Akaike's Information Criterion. We evaluated predictive performance in
-the superpopulation using root mean squared error and measures of
-discrimination and calibration for benefit. Starting from a base case
-scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI
-0.75), we varied the sample size, the strength of the treatment effect,
-and the discriminative ability of the PI. \textbf{Results:} Models
-including a linear interaction of the PI with treatment had (close to?)
-optimal performance under most simulation scenarios. More flexible
-models with restricted cubic splines required larger sample sizes and
-higher AUC of the PI to outperform the linear model. The adaptive
-approach performed similarly to the best-performing method in each
-scenario. \textbf{Conclusion:} Under most circumstances, a model with a
-linear interaction of the PI with treatment is better able to predict
-absolute treatment benefit.
+\textbf{Objective:} To compare different risk-based methods for
+predicting individualized treatment effects using simulations of the RCT
+setting. \textbf{Study Design and Setting:} We sampled patients using
+diverse assumptions for a baseline prognostic index of risk (PI) and the
+shape of its intereraction with treatment (none, linear or quadratic).
+In each sample we predicted absolute benefit using: models with the PI
+and a constant relative treatment effect; models including an
+interaction of treatment with the PI; stratification in 4 quarters of
+the PI; nonlinear transformations of the PI (restricted cubic splines
+with 3, 4 and 5 knots); an adaptive model selection approach based on
+Akaike's Information Criterion. We evaluated predictive performance
+using root mean squared error and measures of discrimination and
+calibration for benefit. Starting from a base case scenario (sample size
+4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample
+size, treatment effect strength, and the PI's discriminative ability.
+\textbf{Results:} Models including a PI by treatment interaction
+performed better under most simulation settings. Flexible models with
+restricted cubic splines required larger sample sizes and higher AUC of
+the PI to outperform the linear model. The adaptive approach performed
+similarly to the best-performing method in most scenarios.
+\textbf{Conclusion:} Usually, a model with a linear interaction of the
+PI with treatment is better able to predict absolute treatment benefit.
 \end{abstract}
@@
 generally had lower error rates compared to the constant treatment
-effect model. In concept, the stratified approach lies between the
-constant effects model and smoother approaches, only assuming constant
-treatment effects within strata of predicted risk and, therefore, is
-more sensitive to treatment effect heterogeneity. Stepwise treatment
-benefit estimates are very useful for demonstrating treatment effect
-heterogeneity -- because estimating treatment effect requires groups of
-patients rather than individual patients --but are not helpful for
-making individualized absolute benefit predictions.
+effect model. Stepwise treatment benefit estimates are very useful for
+demonstrating treatment effect heterogeneity--because estimating
+treatment effect requires groups of patients rather than individual
+patients--but are not helpful for making individualized absolute benefit
+predictions.
 
@@
 the predictor coefficients of the true risk model -- reduced RMSE for
-all methods, similar to increasing the sample size. This increase in
-discriminative ability translates in higher variability of predicted
-risks, which, in turn, allows the considered methods to better follow
-the evolution of treatment benefit. As expected, the increase in
-discriminative ability of the risk model also led to higher values of
-c-for-benefit. Even though risk model performance is very important for
-the ability of risk-based methods to predict treatment benefit,
-prediction model development was outside the scope of this work and has
-already been studied extensively {[}5,12,13{]}.
+all methods. This increase in discriminative ability translates in
+higher variability of predicted risks, which, in turn, allows the
+considered methods to better capture absolute treatment benefits. As a
+consequence, the increase in discriminative ability of the risk model
+also led to higher values of c-for-benefit. Even though risk model
+performance is very important for the ability of risk-based methods to
+predict treatment benefit, prediction model development was outside the
+scope of this work and has already been studied extensively
+{[}5,12,13{]}.
 
@@
 model. An adaptive approach, selecting the model with optimal AIC, had
-satisfactory performance.
+comparable performance to the best performing approach in most of the
+scenarios.
 

### 2021-07-05 submission/manuscript.rmd

@@
 ---
-title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit"
+title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit in simulations"
 abstract: |
- **Objective:** To compare different risk-based methods for predicting individualized treatment effects using simulations of the RCT setting.
-**Study Design and Setting:** We sampled patients using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic).
-In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in 4 quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive model selection approach based on Akaike’s Information Criterion.
+ **Objective:** To compare different risk-based methods predicting individualized treatment effects in RCT simulations.
+**Study Design and Setting:** We simulated data using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic).
+In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach using Akaike’s Information Criterion.
 We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit.
-Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample size, treatment effect strength, and the PI's discriminative ability.
+Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect strength, and the PI's discriminative ability.
 **Results:** Models including a PI by treatment interaction performed better under most simulation settings.
-Flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model.
+Flexible models required larger sample sizes and higher AUC of the PI to outperform the linear interaction model.
 The adaptive approach performed similarly to the best-performing method in most scenarios.

### 2021-07-05 submission/manuscript.tex

@@
 \title{Models including an interaction of treatment with baseline risk
-were better able to predict absolute benefit}
+were better able to predict absolute benefit in simulations}
 
@@
 \begin{abstract}
-\textbf{Objective:} To compare different risk-based methods for
-predicting individualized treatment effects using simulations of the RCT
-setting. \textbf{Study Design and Setting:} We sampled patients using
-diverse assumptions for a baseline prognostic index of risk (PI) and the
-shape of its intereraction with treatment (none, linear or quadratic).
-In each sample we predicted absolute benefit using: models with the PI
-and a constant relative treatment effect; models including an
-interaction of treatment with the PI; stratification in 4 quarters of
-the PI; nonlinear transformations of the PI (restricted cubic splines
-with 3, 4 and 5 knots); an adaptive model selection approach based on
-Akaike's Information Criterion. We evaluated predictive performance
-using root mean squared error and measures of discrimination and
-calibration for benefit. Starting from a base case scenario (sample size
-4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample
-size, treatment effect strength, and the PI's discriminative ability.
-\textbf{Results:} Models including a PI by treatment interaction
-performed better under most simulation settings. Flexible models with
-restricted cubic splines required larger sample sizes and higher AUC of
-the PI to outperform the linear model. The adaptive approach performed
-similarly to the best-performing method in most scenarios.
+\textbf{Objective:} To compare different risk-based methods predicting
+individualized treatment effects in RCT simulations. \textbf{Study
+Design and Setting:} We simulated data using diverse assumptions for a
+baseline prognostic index of risk (PI) and the shape of its
+intereraction with treatment (none, linear or quadratic). In each sample
+we predicted absolute benefit using: models with the PI and a constant
+relative treatment effect; models including an interaction of treatment
+with the PI; stratification in quarters of the PI; nonlinear
+transformations of the PI (restricted cubic splines with 3, 4 and 5
+knots); an adaptive approach using Akaike's Information Criterion. We
+evaluated predictive performance using root mean squared error and
+measures of discrimination and calibration for benefit. Starting from a
+base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
+the PI 0.75), we varied the sample size, the treatment effect strength,
+and the PI's discriminative ability. \textbf{Results:} Models including
+a PI by treatment interaction performed better under most simulation
+settings. Flexible models required larger sample sizes and higher AUC of
+the PI to outperform the linear interaction model. The adaptive approach
+performed similarly to the best-performing method in most scenarios.
 \textbf{Conclusion:} Usually, a model with a linear interaction of the

### 2021-07-05 code/CalibrationBase.R
fix figure text size; closes #6

@@ -85,15 +85,15 @@ pp <- gridExtra::grid.arrange(
     file.path("figures", "calibration_base.tiff"), 
     plot = pp,
     dpi = 1200,
-    width = 11, 
-    height = 10.5,
+    width = 8.5, 
+    height = 7,
     compression = "lzw"
   )
   
   ggplot2::ggsave(
     file.path("figures", "calibration_base.png"), 
     plot = pp,
-    width = 11, 
-    height = 10.5
+    width = 8.5, 
+    height = 7
   )
   

### 2021-07-05 code/RmseAuc.R

@@ -53,10 +53,10 @@ for (i in seq_along(scenarios)) {
     ggplot2::theme(
       legend.position = "none",
       axis.title.x    = ggplot2::element_blank(),
-      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
       axis.title.y    = ggplot2::element_blank(),
-      axis.text.y     = ggplot2::element_text(size = 9),
-      plot.title      = ggtext::element_markdown(size = 12)
+      axis.text.y     = ggplot2::element_text(size = 9.5),
+      plot.title      = ggtext::element_markdown(size = 14)
     )
   
   plotList[[i]] <- plot
@@ -83,9 +83,9 @@ pp <- gridExtra::grid.arrange(
   ggplot2::ggsave(
     file.path("figures", "rmse_auc.tiff"), 
     plot = pp,
+    dpi = 1200,
     width = 8.5, 
     height = 7,
-    dpi = 1200,
     compression = "lzw"
   )
   

### 2021-07-05 code/RmseSampleSize.R

@@ -54,11 +54,12 @@ for (i in seq_along(scenarios)) {
     ggplot2::theme(
       legend.position = "none",
       axis.title.x    = ggplot2::element_blank(),
-      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
       axis.title.y    = ggplot2::element_blank(),
-      axis.text.y     = ggplot2::element_text(size = 9),
-      plot.title      = ggtext::element_markdown(size = 12)
+      axis.text.y     = ggplot2::element_text(size = 9.5),
+      plot.title      = ggtext::element_markdown(size = 14)
     )
+  
   
   plotList[[i]] <- plot
 }

### 2021-07-12 submission/manuscript.rmd
incorporate DvK comments

@@
 
-In the simulated datasets based treatment was allocated at random using a 50:50
-split. For each patient we simulated $8$ baseline covariates, where
-$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes
-for patients in the control arm were generated from a logistic regression model
-including all baseline covariates. Coefficient values were such, so that the
-prediction model had an AUC of $0.75$ and an event rate of $20\%$ in the control
-arm was achieved. Under the base case scenario, outcomes in the treatment arm
-were created using the same logistic regression model, including a constant
-treatment odds ratio (OR) of $0.8$. The generated samples of the base case
-scenario had sample size of 4,250 ($85%$ power for the detection of an
-unadjusted OR of $0.8$).
+For each patient we generated 8 baseline covariates here
+$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
+was allocated using a 50:50 split. Outcomes for patients in the control arm were
+generated from a logistic regression model including all baseline covariates. In
+the base case scenario coefficient values were such, that the AUC of the
+logistic regression model was $0.75$ and the event rate in the control arm was
+$20\%$ (Supplement 1).
 
-We evaluated the effect of sample size considering additional scenarios with
-sample sizes of `r 1064` and `r 17000`. We also evaluated the effect of prediction
-performance, adjusting the baseline covariate coefficients, so that AUC values
-of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
-500,000 patients.
+While binary outcomes in the control arm were generated from Bernoulli variables
+with true probabilities $P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$, 
+the outcomes in the treatment arm based on true probabilities 
+$\text{expit}(lp_1)$, with $$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0$$
 
-We simulated binary outcomes in the control arm using true probabilities 
-$P(y=1|X) = \text{expit}(PI)$, where 
-$\text{expit}(x)= \frac{e^x}{1 + e^x}$. 
-In the treatment arm outcomes where generated with probabilities
-$\text{expit}(lp_1)$ with
-$$ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,$$
-where the coefficients $\gamma_0, \gamma_1$ and $\gamma_2$ along with the
-centering constant $c$ were set for each simulation scenario. In this way we
-were able to assess a wide variety of scenarios, ranging from true constant
-relative treatment effects to moderate and strong linear and quadratic
-deviations. We also considered scenarios with treatment-covariate interactions.
+The coefficients $\gamma_0,\gamma_1$ and $\gamma_2$ along with the
+centering constant $c$ were set for each simulation scenario to mimic a wide
+variety scenarios, ranging from true constant relative treatment effect
+($\gamma_1=\gamma_2=0$) to moderate and strong linear ($\gamma_2=0$) and
+quadratic deviations. We also considered scenarios with treatment-covariate
+interactions.
 These scenarios include 4 weak interactions
@@
 
+The sample size of the base case scenario was set to 4,250 ($80%$ power for the
+detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
+sample sizes of 1,064 and 17,000, respectively. We also evaluated the of worse
+or better discriminative ability for risk, adjusting the baseline covariate
+coefficients, such that the AUC of the regression model in the control arm
+was 0.65 and 0.85 respectively.
+
+Combining all these settings resulted in a simualation study of 66 scenarios
+(exact settings in the supplementary material).
+
 ## Individualized risk-based benefit predictions
 
-All methods assume that a risk prediction model is available and can be used to
-assign individualized risk predictions. For the simulations we developed the
-prediction models internally and blinded to treatment using logistic regression
-including main effects for all baseline covariates and treatment. Risk
-predictions on individuals were made setting treatment to $0$.
+All methods assume that a risk prediction model is available to assign risk
+predictions to individual patients. For the simulations we developed a
+prediction model internally, using logistic regression including main effects
+for all baseline covariates and treatment assignment. Risk predictions for
+individual patients were based on treatment assignment to the control arm, that
+is setting treatment assignment to 0.
 
-The *stratified HTE method* has been suggested as an alternative to
-traditional subgroup analyses. Patients are stratified into equally-sized risk
-strata---in this case based on risk quartiles. Absolute effects are estimated
-using the differences in event rates between treatments within risk quarters. We
-considered this approach as a reference, expecting it to perform worse than the
-other candidates, as its objective is not individualized benefit prediction.
+A *stratified HTE method* has been suggested as an alternative to traditional
+subgroup analyses. Patients are stratified into equally-sized risk strata---in
+this case based on risk quartiles. Absolute treatment effects within risk strata
+are estimated by the difference in event rate between patients in the control
+arm and patients in the treated arm. We considered this approach as a reference,
+expecting it to perform worse than the other candidates, as its objective is not
+to individualize benefit prediction.
 
-Another approach would be to assume *constant relative treatment effect* (OR) is
-constant. In that case, absolute benefit is estimated from
-$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$. This is the method
-usually considered when performing risk-based assessment of treatment effect
-heterogeneity in RCT data.
+Second, we considered a model which assumes *constant relative treatment effect* (constant OR)
+Hence, absolute benefit is predicted from
+$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.
 
-A different approach fits a logistic regression using treatment, the prognostic
-index and their interaction. In this case absolute benefit is estimated from
+Third, we considered a logistic regression model including treatment, the prognostic index, and their interaction. Absolute benefit is then estimated from 
 $\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
@@
 
-Finally, we used *restricted cubic splines* (RCS) to relax the linearity assumption on
-the effect of the linear predictor [@Harrell1988]. We compared the results
-for 3, 4 and 5 knots when fitting the splines to introduce increasing
-flexibility to the methods considered.
+Finally, we used *restricted cubic splines* (RCS) to relax the linearity
+assumption on the effect of the linear predictor [@Harrell1988]. We considered
+splines with 3, 4 and 5 knots to compare models with different levels of
+flexibility.
 
 ## Evaluation metrics
+We evaluated the predictive accuracy of the considered methods by the root mean
+squared error (RMSE):
 
-For evaluating the prediction error of the considered methods we used root mean
-squared error (RMSE), since both the true and the predicted benefits are known,
-given that this is a simulation study. More specifically, we calculated RMSE
-from
 $$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$
-
-We also compared the discriminative ability of the methods under study. We
-assessed discrimination using the c-for-benefit statistic described in
-[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
-predicted benefit and then are matched 1:1, dropping patients in the larger
-treatment arm without a pair. We define *observed* pair-specific treatment
-benefit as the difference of observed outcomes between the untreated and the
-treated patient of each pair. Pair-specific *predicted* benefit is defined as
-the average of predicted benefits within each pair. Then, c-for-benefit is
-defined as the probability that from two randomly chosen matched patient pairs
-with unequal observed benefit, the pair with greater observed benefit also has a
-higher predicted benefit.
+We compared the discriminative ability of the methods under study using
+c-for-benefit [@vanKlaveren2018]. The c-for-benefit represents the probability
+that from two randomly chosen matched patient pairs with unequal observed
+benefit, the pair with greater observed benefit also has a higher predicted
+benefit. To be able to calculate observed benefit, patients in each treatment
+arm are ranked based on their predicted benefit and then matched 1:1 across
+treatment arms. *Observed* treatment benefit is defined as the difference of
+observed outcomes between the untreated and the treated patient of each matched
+patient pair. *Predicted* benefit is defined as the average of predicted benefit
+within each matched patient pair.
 
 We evaluated calibration in a similar manner, using the integrated calibration
-index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
-benefit, observed benefits are regressed on the predicted benefits using a
-locally weighted scatterplot smoother (loess). The ICI is the average absolute
-difference between predicted and smooth observed benefit. Values closer to $0$
-represent better calibration.
+index (ICI) for benefit [@Austin2019]. The observed benefits are regressed on
+the predicted benefits using a locally weighted scatterplot smoother (loess).
+The ICI-for-benefit is the average absolute difference between predicted and
+smooth observed benefit. Values closer to $0$ represent better calibration.
 
@@
 RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
-`r 4250` and AUC = 0.75) or moderate relative deviations were considered (Figure
+`r 4250` and AUC = 0.75) or with moderate relative deviations (Figure
 \ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
@@
 performed equally well to the linear interaction method. Increasing the number
-of knots in RCS resulted in higher error rates across all scenarios. The
-adaptive approach performed very similarly to the best performing model in each
-scenario.
+of knots in RCS resulted in higher RMSE across all scenarios. The adaptive
+approach performed very similarly to the best performing model in each scenario.
 
@@
 true constant relative treatment effects (Figure \ref{fig:rmsesamplesize};
-Panel A). Contrary to previous results with smaller sample sizes, when
-introducing moderate and strong linear deviations the linear interaction model
-performed best, outperforming the constant treatment effect model (Figure
-\ref{fig:rmsesamplesize}; Panels B and C). However, the linear interaction
-model was outperformed by the more flexible RCS models (3 knots) in the case of
-strong quadratic deviations. Again, the increased flexibility of RCS smoothing
-with higher number of knots resulted in overfitting and worse performance
-(Figure \ref{fig:rmsesamplesize}; Panel D).
+Panel A). However, when introducing moderate and strong linear deviations the
+linear interaction model outperformed the constant treatment
+effect model (Figure \ref{fig:rmsesamplesize}; Panels B and C). Furthermore, the
+linear interaction model was outperformed by the more flexible RCS models (3
+knots) in the case of strong quadratic deviations. Again, the increased
+flexibility of RCS smoothing with higher number of knots resulted in overfitting
+and worse predictive accuracy (Figure \ref{fig:rmsesamplesize}; Panel D).
 
@@
 
-When we increased the true prediction AUC to 0.85, models including RCS
-smoothing had the lowest RMSE in the presence of strong quadratic deviations
+When we increased the AUC of the true prediction model to 0.85, models including
+RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
 from the base case of true constant relative treatment effects (Figure
@@
 tended to present much lower variability compared to all other approaches
-(Figure \ref{fig:discrimination}). We also observed an increasing trend of
-discrimination for benefit variability and decreasing median values with
-increasing number of restricted cubic spline knots in all scenarios.
+(Figure \ref{fig:discrimination}). With increasing number of RCS knots, we
+observed decreasing median values and increasing variability of the
+c-for-benefit in all scenarios.
 
@@
 
-When focusing on calibration, the linear interaction model had the lowest median
-ICI for benefit in the majority of scenarios except for the scenarios where no
+As for calibration, the linear interaction model had the lowest median
+ICI-for-benefit in the majority of scenarios except for the scenarios where no
 or moderate linear deviations from the base case were considered. In that case
-constant treatment effect models demonstrated the best performance, very
+constant treatment effect models demonstrated optimal calibration, very
 comparable to the linear interaction model's performance, nonetheless (Figure
-\ref{fig:calibration}). However, under strong linear or
-quadratic deviations, the constant treatment effect model was very poorly
-calibrated (Figure \ref{fig:calibration}; Panels C and D).
+\ref{fig:calibration}). With strong linear or quadratic deviations, the
+constant treatment effect model was poorly calibrated (Figure
+\ref{fig:calibration}; Panels C and D).
 
@@
 
-We used the risk linear predictor to fit the the proposed methods under study
-for individualizing absolute benefit predictions. All methods had quite
-comparable results, in the sense that we predicted increasing benefits for
-patients with higher baseline risk predictions. In terms of c-for-benefit 
-(validated internally) all models had quite comparable performance ranging from
-0.519 (RCS smoothing with 3 knots) to 0.542 (RCS smoothing with 4 knots).
-Similar conclusions could be drawn in terms of ICI-for-benefit which ranged
-from 0.0039 (linear interaction approach) to 0.0053 (RCS smoothing with 4
-knots).
-
-The adaptive approach picked the
-model with RCS smoothing with 4 knots, which had quite comparable performance to
-the smooth fit with 5 knots. However, for lower baseline risk these 2 models
-predicted implausible benefits and maybe should be avoided when applying such
-models in practice. The linear interaction model and the model with RCS
-smoothing (3 knots) made very similar predictions and also followed quite
-closely the evolution of the stratified estimates. In this case, we observed
-that even using a simple model with a constant relative treatment effect can
-result in very similar benefit predictions as other more complex approaches.
+We used the risk linear predictor to fit the proposed methods under study for
+individualizing absolute benefit predictions. All methods predicted increasing
+benefits for patients with higher baseline risk predictions, but the fitted
+patterns were clearly different. The adaptive approach selected the model with
+RCS smoothing with 4 knots. However, for very low baseline risk this model
+predicted decreasing benefit with increasing risk maybe somewhat too flexible.
+The more robust models, the linear interaction model or the model with RCS
+smoothing (3 knots), gave very similar benefit predictions, followed the
+evolution of the stratified estimates very closely and may therefore be
+preferable for use in clinical practice.
 

### 2021-07-12 submission/manuscript.tex
incorporate DvK comments

@@
 
-In the simulated datasets based treatment was allocated at random using
-a 50:50 split. For each patient we simulated \(8\) baseline covariates,
-where \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
-Outcomes for patients in the control arm were generated from a logistic
-regression model including all baseline covariates. Coefficient values
-were such, so that the prediction model had an AUC of \(0.75\) and an
-event rate of \(20\%\) in the control arm was achieved. Under the base
-case scenario, outcomes in the treatment arm were created using the same
-logistic regression model, including a constant treatment odds ratio
-(OR) of \(0.8\). The generated samples of the base case scenario had
-sample size of 4,250 (\(85%
-\) power for the detection of an unadjusted OR of \(0.8\)).
+For each patient we generated 8 baseline covariates here
+\(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
+Treatment was allocated using a 50:50 split. Outcomes for patients in
+the control arm were generated from a logistic regression model
+including all baseline covariates. In the base case scenario coefficient
+values were such, that the AUC of the logistic regression model was
+\(0.75\) and the event rate in the control arm was \(20\%\) (Supplement
+1).
 
-We evaluated the effect of sample size considering additional scenarios
-with sample sizes of 1,064 and 17,000. We also evaluated the effect of
-prediction performance, adjusting the baseline covariate coefficients,
-so that AUC values of \(0.65\) and \(0.80\) were achieved when
-validating in a simulated dataset of 500,000 patients.
+While binary outcomes in the control arm were generated from Bernoulli
+variables with true probabilities
+\(P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\), the outcomes in
+the treatment arm based on true probabilities \(\text{expit}(lp_1)\),
+with \[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]
 
-We simulated binary outcomes in the control arm using true probabilities
-\(P(y=1|X) = \text{expit}(PI)\), where
-\(\text{expit}(x)= \frac{e^x}{1 + e^x}\). In the treatment arm outcomes
-where generated with probabilities \(\text{expit}(lp_1)\) with
-\[ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,\] where the
-coefficients \(\gamma_0, \gamma_1\) and \(\gamma_2\) along with the
-centering constant \(c\) were set for each simulation scenario. In this
-way we were able to assess a wide variety of scenarios, ranging from
-true constant relative treatment effects to moderate and strong linear
-and quadratic deviations. We also considered scenarios with
-treatment-covariate interactions. These scenarios include 4 weak
+The coefficients \(\gamma_0,\gamma_1\) and \(\gamma_2\) along with the
+centering constant \(c\) were set for each simulation scenario to mimic
+a wide variety scenarios, ranging from true constant relative treatment
+effect (\(\gamma_1=\gamma_2=0\)) to moderate and strong linear
+(\(\gamma_2=0\)) and quadratic deviations. We also considered scenarios
+with treatment-covariate interactions. These scenarios include 4 weak
 interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
@@
 
+The sample size of the base case scenario was set to 4,250 (\(80%
+\) power for the detection of a marginal OR of 0.8). We evaluated the
+effect of smaller or larger sample sizes of 1,064 and 17,000,
+respectively. We also evaluated the of worse or better discriminative
+ability for risk, adjusting the baseline covariate coefficients, such
+that the AUC of the regression model in the control arm was 0.65 and
+0.85 respectively.
+
+Combining all these settings resulted in a simualation study of 66
+scenarios (exact settings in the supplementary material).
+
 \hypertarget{individualized-risk-based-benefit-predictions}{%
@@
 
-All methods assume that a risk prediction model is available and can be
-used to assign individualized risk predictions. For the simulations we
-developed the prediction models internally and blinded to treatment
-using logistic regression including main effects for all baseline
-covariates and treatment. Risk predictions on individuals were made
-setting treatment to \(0\).
+All methods assume that a risk prediction model is available to assign
+risk predictions to individual patients. For the simulations we
+developed a prediction model internally, using logistic regression
+including main effects for all baseline covariates and treatment
+assignment. Risk predictions for individual patients were based on
+treatment assignment to the control arm, that is setting treatment
+assignment to 0.
 
-The \emph{stratified HTE method} has been suggested as an alternative to
+A \emph{stratified HTE method} has been suggested as an alternative to
 traditional subgroup analyses. Patients are stratified into
 equally-sized risk strata---in this case based on risk quartiles.
-Absolute effects are estimated using the differences in event rates
-between treatments within risk quarters. We considered this approach as
-a reference, expecting it to perform worse than the other candidates, as
-its objective is not individualized benefit prediction.
+Absolute treatment effects within risk strata are estimated by the
+difference in event rate between patients in the control arm and
+patients in the treated arm. We considered this approach as a reference,
+expecting it to perform worse than the other candidates, as its
+objective is not to individualize benefit prediction.
 
-Another approach would be to assume \emph{constant relative treatment
-effect} (OR) is constant. In that case, absolute benefit is estimated
-from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\). This is
-the method usually considered when performing risk-based assessment of
-treatment effect heterogeneity in RCT data.
+Second, we considered a model which assumes \emph{constant relative
+treatment effect} (constant OR) Hence, absolute benefit is predicted
+from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).
 
-A different approach fits a logistic regression using treatment, the
-prognostic index and their interaction. In this case absolute benefit is
+Third, we considered a logistic regression model including treatment,
+the prognostic index, and their interaction. Absolute benefit is then
 estimated from
@@
 linearity assumption on the effect of the linear predictor {[}7{]}. We
-compared the results for 3, 4 and 5 knots when fitting the splines to
-introduce increasing flexibility to the methods considered.
+considered splines with 3, 4 and 5 knots to compare models with
+different levels of flexibility.
 
@@
 
-For evaluating the prediction error of the considered methods we used
-root mean squared error (RMSE), since both the true and the predicted
-benefits are known, given that this is a simulation study. More
-specifically, we calculated RMSE from
+We evaluated the predictive accuracy of the considered methods by the
+root mean squared error (RMSE):
+
 \[\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2\]
-
-We also compared the discriminative ability of the methods under study.
-We assessed discrimination using the c-for-benefit statistic described
-in {[}8{]}. Patients in each treatment arm are ranked based on their
-predicted benefit and then are matched 1:1, dropping patients in the
-larger treatment arm without a pair. We define \emph{observed}
-pair-specific treatment benefit as the difference of observed outcomes
-between the untreated and the treated patient of each pair.
-Pair-specific \emph{predicted} benefit is defined as the average of
-predicted benefits within each pair. Then, c-for-benefit is defined as
-the probability that from two randomly chosen matched patient pairs with
-unequal observed benefit, the pair with greater observed benefit also
-has a higher predicted benefit.
+We compared the discriminative ability of the methods under study using
+c-for-benefit {[}8{]}. The c-for-benefit represents the probability that
+from two randomly chosen matched patient pairs with unequal observed
+benefit, the pair with greater observed benefit also has a higher
+predicted benefit. To be able to calculate observed benefit, patients in
+each treatment arm are ranked based on their predicted benefit and then
+matched 1:1 across treatment arms. \emph{Observed} treatment benefit is
+defined as the difference of observed outcomes between the untreated and
+the treated patient of each matched patient pair. \emph{Predicted}
+benefit is defined as the average of predicted benefit within each
+matched patient pair.
 
 We evaluated calibration in a similar manner, using the integrated
-calibration index (ICI) for benefit {[}9{]}. After creating pairs based
-on predicted benefit, observed benefits are regressed on the predicted
-benefits using a locally weighted scatterplot smoother (loess). The ICI
-is the average absolute difference between predicted and smooth observed
-benefit. Values closer to \(0\) represent better calibration.
+calibration index (ICI) for benefit {[}9{]}. The observed benefits are
+regressed on the predicted benefits using a locally weighted scatterplot
+smoother (loess). The ICI-for-benefit is the average absolute difference
+between predicted and smooth observed benefit. Values closer to \(0\)
+represent better calibration.
 
@@
 median RMSE in scenarios with a true constant relative treatment effect
-(OR = 0.8, N = 4,250 and AUC = 0.75) or moderate relative deviations
-were considered (Figure \ref{fig:rmsebase}; Panel A). However, when we
+(OR = 0.8, N = 4,250 and AUC = 0.75) or with moderate relative
+deviations (Figure \ref{fig:rmsebase}; Panel A). However, when we
 considered strong linear and quadratic deviations from a constant
@@
 well to the linear interaction method. Increasing the number of knots in
-RCS resulted in higher error rates across all scenarios. The adaptive
-approach performed very similarly to the best performing model in each
-scenario.
+RCS resulted in higher RMSE across all scenarios. The adaptive approach
+performed very similarly to the best performing model in each scenario.
 
@@
 assumption of true constant relative treatment effects (Figure
-\ref{fig:rmsesamplesize}; Panel A). Contrary to previous results with
-smaller sample sizes, when introducing moderate and strong linear
-deviations the linear interaction model performed best, outperforming
+\ref{fig:rmsesamplesize}; Panel A). However, when introducing moderate
+and strong linear deviations the linear interaction model outperformed
 the constant treatment effect model (Figure \ref{fig:rmsesamplesize};
-Panels B and C). However, the linear interaction model was outperformed
-by the more flexible RCS models (3 knots) in the case of strong
-quadratic deviations. Again, the increased flexibility of RCS smoothing
-with higher number of knots resulted in overfitting and worse
-performance (Figure \ref{fig:rmsesamplesize}; Panel D).
+Panels B and C). Furthermore, the linear interaction model was
+outperformed by the more flexible RCS models (3 knots) in the case of
+strong quadratic deviations. Again, the increased flexibility of RCS
+smoothing with higher number of knots resulted in overfitting and worse
+predictive accuracy (Figure \ref{fig:rmsesamplesize}; Panel D).
 
@@
 
-When we increased the true prediction AUC to 0.85, models including RCS
-smoothing had the lowest RMSE in the presence of strong quadratic
-deviations from the base case of true constant relative treatment
-effects (Figure \ref{fig:rmseauc}; Panel D). However, with milder
-deviations, the linear interaction model had the lowest RMSE with the
-RCS smoothing methods (3 knots) being a close second (Figure
+When we increased the AUC of the true prediction model to 0.85, models
+including RCS smoothing had the lowest RMSE in the presence of strong
+quadratic deviations from the base case of true constant relative
+treatment effects (Figure \ref{fig:rmseauc}; Panel D). However, with
+milder deviations, the linear interaction model had the lowest RMSE with
+the RCS smoothing methods (3 knots) being a close second (Figure
 \ref{fig:rmseauc}; Panels B and C). Increasing the number of knots of
@@
 interaction model tended to present much lower variability compared to
-all other approaches (Figure \ref{fig:discrimination}). We also observed
-an increasing trend of discrimination for benefit variability and
-decreasing median values with increasing number of restricted cubic
-spline knots in all scenarios.
+all other approaches (Figure \ref{fig:discrimination}). With increasing
+number of RCS knots, we observed decreasing median values and increasing
+variability of the c-for-benefit in all scenarios.
 
@@
 
-When focusing on calibration, the linear interaction model had the
-lowest median ICI for benefit in the majority of scenarios except for
-the scenarios where no or moderate linear deviations from the base case
-were considered. In that case constant treatment effect models
-demonstrated the best performance, very comparable to the linear
-interaction model's performance, nonetheless (Figure
-\ref{fig:calibration}). However, under strong linear or quadratic
-deviations, the constant treatment effect model was very poorly
-calibrated (Figure \ref{fig:calibration}; Panels C and D).
+As for calibration, the linear interaction model had the lowest median
+ICI-for-benefit in the majority of scenarios except for the scenarios
+where no or moderate linear deviations from the base case were
+considered. In that case constant treatment effect models demonstrated
+optimal calibration, very comparable to the linear interaction model's
+performance, nonetheless (Figure \ref{fig:calibration}). With strong
+linear or quadratic deviations, the constant treatment effect model was
+poorly calibrated (Figure \ref{fig:calibration}; Panels C and D).
 

### 2021-07-12 submission/manuscript.rmd
use cross-validated performance metrics

@@
 )
+set.seed(19910930)
 ```
@@
 
-constantModel <- fitModelBasedHte(
+folds <- caret::createFolds(gusto$outcome, k = 4)
+
+evaluatePerformance <- function(data, testRows, settings) {
+  isStratified <- FALSE
+  if ("modelBased" %in% class(settings)) {
+    model <- fitModelBasedHte(data %>% dplyr::slice(-testRows), settings)
+    tmp <- data %>%
+      slice(testRows) %>%
+      dplyr::mutate(
+        predictedBenefit = predictBenefitModelBasedHte(
+          p             = plogis(.$riskLinearPredictor), 
+          modelBasedFit = model
+        )
+      )
+
+  } else if ("rcs" %in% class(settings)) {
+    model <- fitRcsHte(data %>% dplyr::slice(-testRows), settings)
+    tmp <- gusto %>%
+      slice(testRows) %>%
+      dplyr::mutate(
+        predictedBenefit = predictSmoothBenefit(
+          p          = plogis(riskLinearPredictor), 
+          smoothFit  = model
+        )
+      )
+  } else if ("stratified" %in% class(settings)) {
+    isStratified <- TRUE
+    model <- fitStratifiedHte(data %>% dplyr::slice(-testRows), settings)
+    tmp <- gusto %>%
+      slice(testRows) %>%
+      dplyr::mutate(
+        predictedBenefit = predictStratifiedBenefit(
+          p          = plogis(riskLinearPredictor), 
+          stratifiedHte  = model
+        )
+      )
+  }
+  cForBenefit <- calculateCForBenefit(tmp)
+  iciForBenefit <- calculateCalibrationForBenefit(tmp)
+  aic <- ifelse(
+    isStratified,
+    NA,
+    model$aic
+  )
+
+  return(
+    list(
+      c = cForBenefit,
+      ici = iciForBenefit$ici,
+      aic = aic
+    )
+  )
+}
+
+constant <- lapply(
+  folds,
+  evaluatePerformance,
   data = gusto,
@@
 
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictBenefitModelBasedHte(
-      p             = plogis(.$riskLinearPredictor), 
-      modelBasedFit = constantModel
-    )
-  )
+constantData <- do.call(bind_rows, constant)
+cConstant <- mean(constantData$c)
+iciConstant <- mean(constantData$ici)
+aicConstnat <- mean(constantData$aic)
 
-constantC   <- calculateCForBenefit(tmp)
-constantIci <- calculateCalibrationForBenefit(tmp)
-
-stratified <- fitStratifiedHte(
+stratified <- lapply(
+  folds,
+  evaluatePerformance,
   data = gusto,
@@
 
-linearModel <- fitModelBasedHte(
+stratifiedData <- do.call(bind_rows, stratified)
+cStratified <- mean(stratifiedData$c)
+iciStratified <- mean(stratifiedData$ici)
+
+linear <- lapply(
+  folds,
+  evaluatePerformance,
   data = gusto,
@@
 
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictBenefitModelBasedHte(
-      p             = plogis(.$riskLinearPredictor), 
-      modelBasedFit = linearModel
-    )
+linearData <- do.call(bind_rows, linear)
+clinear <- mean(linearData$c)
+iciLinear <- mean(linearData$ici)
+aicLinear <- mean(linearData$aic)
+
+rcs3 <- lapply(
+  folds,
+  evaluatePerformance,
+  data = gusto,
+  settings = createRcsSettings(nKnots = 3)
 )
 
-linearC   <- calculateCForBenefit(tmp)
-linearIci <- calculateCalibrationForBenefit(tmp)
+rcs3Data <- do.call(bind_rows, rcs3)
+cRcs3    <- mean(rcs3Data$c)
+iciRcs3  <- mean(rcs3Data$ici)
+aicRcs3  <- mean(rcs3Data$aic)
 
-rcs3Model <- fitRcsHte(
-  data     = gusto,
-  settings = createRcsSettings()
-)
-
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictSmoothBenefit(
-      p          = plogis(riskLinearPredictor), 
-      smoothFit  = rcs3Model
-    )
-  )
-
-rcs3C   <- calculateCForBenefit(tmp)
-rcs3Ici <- calculateCalibrationForBenefit(tmp)
-
-rcs4Model <- fitRcsHte(
+rcs4 <- lapply(
+  folds,
+  evaluatePerformance,
   data = gusto,
@@
 
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictSmoothBenefit(
-      p          = plogis(riskLinearPredictor), 
-      smoothFit  = rcs4Model
-    )
-  )
+rcs4Data <- do.call(bind_rows, rcs4)
+cRcs4    <- mean(rcs4Data$c)
+iciRcs4  <- mean(rcs4Data$ici)
+aicRcs4  <- mean(rcs4Data$aic)
 
-rcs4C   <- calculateCForBenefit(tmp)
-rcs4Ici <- calculateCalibrationForBenefit(tmp)
-
-rcs5Model <- fitRcsHte(
+rcs5 <- lapply(
+  folds,
+  evaluatePerformance,
   data = gusto,
@@
 
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictSmoothBenefit(
-      p          = plogis(.$riskLinearPredictor), 
-      smoothFit  = rcs5Model
-    )
-  )
-
-rcs5C   <- calculateCForBenefit(tmp)
-rcs5Ici <- calculateCalibrationForBenefit(tmp)
+rcs5Data <- do.call(bind_rows, rcs5)
+cRcs5    <- mean(rcs5Data$c)
+iciRcs5  <- mean(rcs5Data$ici)
+aicRcs5  <- mean(rcs5Data$aic)
 
@@
   discrimination = c(
-    constantC, linearC, rcs3C,
-    rcs4C, rcs5C
+    cConstant, clinear, cRcs3,
+    cRcs4, cRcs5
   ),
   calibration = c(
-    constantIci$ici, linearIci$ici, rcs3Ici$ici,
-    rcs4Ici$ici, rcs5Ici$ici
+    iciConstant, iciLinear, iciRcs3,
+    iciRcs4, iciRcs5
+  ),
+  aic = c(
+    aicConstnat, aicLinear, aicRcs3,
+    aicRcs4, aicRcs5
   ),
@@
 
-cBenefit <- gustoPerformance %>% arrange(discrimination)
+discriminationPerformance <- gustoPerformance %>%
+  slice(c(2, 3)) %>%
+  select(discrimination)
+
+calibrationPerformance <- gustoPerformance %>%
+  slice(c(2, 3)) %>%
+  select(calibration)
 
@@
 RCS smoothing with 4 knots. However, for very low baseline risk this model
-predicted decreasing benefit with increasing risk maybe somewhat too flexible.
+predicted decreasing benefit with increasing risk may be somewhat too flexible.
 The more robust models, the linear interaction model or the model with RCS
@@
 evolution of the stratified estimates very closely and may therefore be
-preferable for use in clinical practice.
+preferable for use in clinical practice. The model with RCS smoothing (3 knots)
+had somewhat lower AIC compared to the linear interaction model, slightly better
+cross-validated discrimination (c-for-benefit
+`r round(discriminationPerformance[2, ], 3)` vs
+`r round(discriminationPerformance[1, ], 3)`) and quite similar cross-validated
+calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
+`r round(calibrationPerformance[1, ], 4)`).
 

### 2021-07-12 submission/manuscript.tex
use cross-validated performance metrics

@@
 
-We used the risk linear predictor to fit the the proposed methods under
-study for individualizing absolute benefit predictions. All methods had
-quite comparable results, in the sense that we predicted increasing
-benefits for patients with higher baseline risk predictions. In terms of
-c-for-benefit (validated internally) all models had quite comparable
-performance ranging from 0.519 (RCS smoothing with 3 knots) to 0.542
-(RCS smoothing with 4 knots). Similar conclusions could be drawn in
-terms of ICI-for-benefit which ranged from 0.0039 (linear interaction
-approach) to 0.0053 (RCS smoothing with 4 knots).
-
-The adaptive approach picked the model with RCS smoothing with 4 knots,
-which had quite comparable performance to the smooth fit with 5 knots.
-However, for lower baseline risk these 2 models predicted implausible
-benefits and maybe should be avoided when applying such models in
-practice. The linear interaction model and the model with RCS smoothing
-(3 knots) made very similar predictions and also followed quite closely
-the evolution of the stratified estimates. In this case, we observed
-that even using a simple model with a constant relative treatment effect
-can result in very similar benefit predictions as other more complex
-approaches.
+We used the risk linear predictor to fit the proposed methods under
+study for individualizing absolute benefit predictions. All methods
+predicted increasing benefits for patients with higher baseline risk
+predictions, but the fitted patterns were clearly different. The
+adaptive approach selected the model with RCS smoothing with 4 knots.
+However, for very low baseline risk this model predicted decreasing
+benefit with increasing risk may be somewhat too flexible. The more
+robust models, the linear interaction model or the model with RCS
+smoothing (3 knots), gave very similar benefit predictions, followed the
+evolution of the stratified estimates very closely and may therefore be
+preferable for use in clinical practice. The model with RCS smoothing (3
+knots) had somewhat lower AIC compared to the linear interaction model,
+slightly better cross-validated discrimination (c-for-benefit 0.537 vs
+0.527) and quite similar cross-validated calibration (ICI-for benefit
+0.0114 vs 0.0116).
 

### 2021-07-12 submission/manuscript.rmd
cross-validate and replicate 100 times for gusto evaluation

--- 2021-07-12 submission/manuscript.rmd
@@
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
-gusto <- gusto %>%
-  tibble() %>%
-  rename(
-    "outcome" = "day30",
-    "treatment" = "tpa"
-  )
+gustoPerformance <- readr::read_csv(here::here("data/processed/gustoPerformanceMetrics.csv"))
 
-prediction <- lrm(
-  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
-  data = gusto,
-  maxit = 99
-)
-
-riskLinearPredictor <- predict(
-  prediction,
-  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
-)
-
-gusto <- gusto %>%
-  mutate(
-    riskLinearPredictor = riskLinearPredictor
-  )
-
-constantModel <- fitModelBasedHte(
-  data     = gusto, 
-  settings = createModelBasedSettings(type = "treatment")
-)
-
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictBenefitModelBasedHte(
-      p             = plogis(.$riskLinearPredictor), 
-      modelBasedFit = constantModel
-    )
-  )
-
-constantC   <- calculateCForBenefit(tmp)
-constantIci <- calculateCalibrationForBenefit(tmp)
-
-stratified <- fitStratifiedHte(
-  data     = gusto,
-  settings = createStratifiedSettings()
-)
-
-linearModel <- fitModelBasedHte(
-  data     = gusto,
-  settings = createModelBasedSettings()
-)
-
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictBenefitModelBasedHte(
-      p             = plogis(.$riskLinearPredictor), 
-      modelBasedFit = linearModel
-    )
-  )
-
-linearC   <- calculateCForBenefit(tmp)
-linearIci <- calculateCalibrationForBenefit(tmp)
-
-rcs3Model <- fitRcsHte(
-  data     = gusto,
-  settings = createRcsSettings()
-)
-
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictSmoothBenefit(
-      p          = plogis(riskLinearPredictor), 
-      smoothFit  = rcs3Model
-    )
-  )
-
-rcs3C   <- calculateCForBenefit(tmp)
-rcs3Ici <- calculateCalibrationForBenefit(tmp)
-
-rcs4Model <- fitRcsHte(
-  data     = gusto,
-  settings = createRcsSettings(nKnots = 4)
-)
-
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictSmoothBenefit(
-      p          = plogis(riskLinearPredictor), 
-      smoothFit  = rcs4Model
-    )
-  )
-
-rcs4C   <- calculateCForBenefit(tmp)
-rcs4Ici <- calculateCalibrationForBenefit(tmp)
-
-rcs5Model <- fitRcsHte(
-  data     = gusto,
-  settings = createRcsSettings(nKnots = 5)
-)
-
-tmp <- gusto %>%
-  mutate(
-    predictedBenefit = predictSmoothBenefit(
-      p          = plogis(.$riskLinearPredictor), 
-      smoothFit  = rcs5Model
-    )
-  )
-
-rcs5C   <- calculateCForBenefit(tmp)
-rcs5Ici <- calculateCalibrationForBenefit(tmp)
-
-gustoPerformance <- tibble(
-  discrimination = c(
-    constantC, linearC, rcs3C,
-    rcs4C, rcs5C
-  ),
-  calibration = c(
-    constantIci$ici, linearIci$ici, rcs3Ici$ici,
-    rcs4Ici$ici, rcs5Ici$ici
-  ),
-  method = c(
-    "constant treatment effect",
-    "linear interaction",
-    "RCS smoothing with 3 knots",
-    "RCS smoothing with 4 knots",
-    "RCS smoothing with 5 knots"
-  )
-)
-
-cBenefit <- gustoPerformance %>% arrange(discrimination)
-
+discriminationPerformance <- gustoPerformance %>%
+  select("discrimination")
+calibrationPerformance <- gustoPerformance %>%
+  select("calibration")
 ```
@@
 evolution of the stratified estimates very closely and may therefore be
-preferable for use in clinical practice.
+preferable for use in clinical practice. The linear interaction model
+had somewhat lower AIC compared to the model with RCS smoothing (3 knots), slightly better
+cross-validated discrimination (c-for-benefit
+`r round(discriminationPerformance[2, ], 3)` vs
+`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
+calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
+`r round(calibrationPerformance[3, ], 4)`).
 

### 2021-07-12 submission/manuscript.tex
cross-validate and replicate 100 times for gusto evaluation

@@
 evolution of the stratified estimates very closely and may therefore be
-preferable for use in clinical practice. The model with RCS smoothing (3
-knots) had somewhat lower AIC compared to the linear interaction model,
-slightly better cross-validated discrimination (c-for-benefit 0.537 vs
-0.527) and quite similar cross-validated calibration (ICI-for benefit
-0.0114 vs 0.0116).
+preferable for use in clinical practice. The linear interaction model
+had somewhat lower AIC compared to the model with RCS smoothing (3
+knots), slightly better cross-validated discrimination (c-for-benefit
+0.526 vs 0.525) and quite similar cross-validated calibration (ICI-for
+benefit 0.0115 vs 0.0117).
 

### 2021-07-12 submission/supplement.rmd
cross-validate and replicate 100 times for gusto evaluation

--- 2021-06-29 submission/supplement.rmd
@@
   - \usepackage{booktabs}
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
   - \newcommand\given[1][]{\:#1\vert\:}
-  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
-
 editor_options: 
@@
       "C",
-      "ICI"
+      "ICI",
+      "AIC"
     ),

### 2021-07-12 code/GustoPerformanceMetrics.R

@@ -12,169 +12,210 @@ library(tidyverse)
 library(rms)
 library(SmoothHte)
 
-set.seed(19910930)
 load("data/raw/gusto.rda")
-
-res <- list()
+set.seed(19910930)
 
 gusto <- gusto %>%
   tibble() %>%
-  filter(!is.na(tpa)) %>%
   rename(
     "outcome" = "day30",
     "treatment" = "tpa"
   )
 
-prediction <- lrm(
-  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
-  data = gusto,
-  maxit = 99
-)
-
-riskLinearPredictor <- predict(
-  prediction,
-  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
-)
-
-gusto <- gusto %>%
-  mutate(
-    riskLinearPredictor = riskLinearPredictor,
-    predictedRisk       = plogis(riskLinearPredictor)
+replicationExpression <- function(gusto) {
+  prediction <- rms::lrm(
+    outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+    data = gusto,
+    maxit = 99
   )
-
-constantModel <- fitModelBasedHte(
-  data     = gusto, 
-  settings = createModelBasedSettings(type = "treatment")
-)
-
-tmpData <- gusto %>%
-  mutate(
-    predictedBenefit = predictBenefitModelBasedHte(
-      p = predictedRisk,
-      modelBasedFit = constantModel
+  
+  riskLinearPredictor <- predict(
+    prediction,
+    newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
+  )
+  
+  gusto <- gusto %>%
+    mutate(
+      riskLinearPredictor = riskLinearPredictor
+    )
+  
+  folds <- caret::createFolds(gusto$outcome, k = 4)
+  
+  evaluatePerformance <- function(data, testRows, settings) {
+    isStratified <- FALSE
+    if ("modelBased" %in% class(settings)) {
+      model <- fitModelBasedHte(data %>% dplyr::slice(-testRows), settings)
+      tmp <- data %>%
+        slice(testRows) %>%
+        dplyr::mutate(
+          predictedBenefit = predictBenefitModelBasedHte(
+            p             = plogis(.$riskLinearPredictor), 
[…]

### 2021-07-13 submission/manuscript.rmd
draft to be shared with co-authors

@@
 [@Kent2010]. However, even though estimates at the risk subgroup level may be
-accurate, these estimated do not apply to individual patients, especially for
+accurate, these estimates do not apply to individual patients, especially for
 patients with predicted risk at the boundaries of the risk intervals. Hence, the
@@
 risk-based models including a prognostic index of baseline risk (PI) and
-treatment assignment [cite PATH !!]. We aimed to summarize and compare different
-risk-based models for predicting individualized treatment effects. We simulated
-RCT settings to compare the performance of these models under different
-assumptions of the relationship between baseline risk and treatment. We
-illustrated the different models by a case study of predicting individualized
+treatment assignment [@Kent2019; @PathEnE]. We aimed to summarize and compare
+different risk-based models for predicting individualized treatment effects. We
+simulated RCT settings to compare the performance of these models under
+different assumptions of the relationship between baseline risk and treatment.
+We illustrated the different models by a case study of predicting individualized
 effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
@@
 logistic regression model was $0.75$ and the event rate in the control arm was
-$20\%$ (Supplement 1).
+$20\%$.
 
@@
 with true probabilities $P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$, 
-the outcomes in the treatment arm based on true probabilities 
+the outcomes in the treatment arm were based on true probabilities 
 $\text{expit}(lp_1)$, with $$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0$$
@@
 interactions.
-These scenarios include 4 weak interactions
+These scenarios included 4 weak interactions
 ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
@@
 ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
-and 2 weak and 2 strong interactions. Combining all these different settings
-resulted in a simulation study of $66$ scenarios. The exact settings for each
-scenario are available in the supplementary material.
+and 2 weak and 2 strong interactions.
 
-The sample size of the base case scenario was set to 4,250 ($80%$ power for the
+The sample size of the base case scenario was set to 4,250 ($80\%$ power for the
 detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
-sample sizes of 1,064 and 17,000, respectively. We also evaluated the of worse
-or better discriminative ability for risk, adjusting the baseline covariate
-coefficients, such that the AUC of the regression model in the control arm
-was 0.65 and 0.85 respectively.
+sample sizes of 1,064 and 17,000, respectively. We also evaluated the effect of
+worse or better discriminative ability for risk, adjusting the baseline
+covariate coefficients, such that the AUC of the regression model in the control
+arm was 0.65 and 0.85 respectively.
 
-Combining all these settings resulted in a simualation study of 66 scenarios
+Combining all these settings resulted in a simulation study of 66 scenarios
 (exact settings in the supplementary material).
@@
 
-Second, we considered a model which assumes *constant relative treatment effect* (constant OR)
-Hence, absolute benefit is predicted from
+Second, we considered a model which assumes *constant relative treatment effect*
+(constant OR). Hence, absolute benefit is predicted from
 $\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.

### 2021-07-13 submission/manuscript.tex
draft to be shared with co-authors

@@
 treatment effect variation {[}6{]}. However, even though estimates at
-the risk subgroup level may be accurate, these estimated do not apply to
+the risk subgroup level may be accurate, these estimates do not apply to
 individual patients, especially for patients with predicted risk at the
@@
 various risk-based models including a prognostic index of baseline risk
-(PI) and treatment assignment {[}cite PATH !!{]}. We aimed to summarize
-and compare different risk-based models for predicting individualized
+(PI) and treatment assignment {[}3,4{]}. We aimed to summarize and
+compare different risk-based models for predicting individualized
 treatment effects. We simulated RCT settings to compare the performance
@@
 values were such, that the AUC of the logistic regression model was
-\(0.75\) and the event rate in the control arm was \(20\%\) (Supplement
-1).
+\(0.75\) and the event rate in the control arm was \(20\%\).
 
@@
 \(P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\), the outcomes in
-the treatment arm based on true probabilities \(\text{expit}(lp_1)\),
-with \[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]
+the treatment arm were based on true probabilities
+\(\text{expit}(lp_1)\), with
+\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]
 
@@
 (\(\gamma_2=0\)) and quadratic deviations. We also considered scenarios
-with treatment-covariate interactions. These scenarios include 4 weak
+with treatment-covariate interactions. These scenarios included 4 weak
 interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
 interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
-weak and 2 strong interactions. Combining all these different settings
-resulted in a simulation study of \(66\) scenarios. The exact settings
-for each scenario are available in the supplementary material.
+weak and 2 strong interactions.
 
-The sample size of the base case scenario was set to 4,250 (\(80%
-\) power for the detection of a marginal OR of 0.8). We evaluated the
+The sample size of the base case scenario was set to 4,250 (\(80\%\)
+power for the detection of a marginal OR of 0.8). We evaluated the
 effect of smaller or larger sample sizes of 1,064 and 17,000,
-respectively. We also evaluated the of worse or better discriminative
-ability for risk, adjusting the baseline covariate coefficients, such
-that the AUC of the regression model in the control arm was 0.65 and
-0.85 respectively.
+respectively. We also evaluated the effect of worse or better
+discriminative ability for risk, adjusting the baseline covariate
+coefficients, such that the AUC of the regression model in the control
+arm was 0.65 and 0.85 respectively.
 
-Combining all these settings resulted in a simualation study of 66
+Combining all these settings resulted in a simulation study of 66
 scenarios (exact settings in the supplementary material).
@@
 Second, we considered a model which assumes \emph{constant relative
-treatment effect} (constant OR) Hence, absolute benefit is predicted
+treatment effect} (constant OR). Hence, absolute benefit is predicted
 from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).

### 2021-07-13 submission/manuscript.rmd
comments on discussion

@@
 ---
-title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit in simulations"
+title: |
+ A model including an interaction of treatment with baseline risk was the optimal
+ risk-based approach to predicting absolute treatment benefit
 abstract: |
- **Objective:** To compare different risk-based methods predicting individualized treatment effects in RCT simulations.
-**Study Design and Setting:** We simulated data using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic).
-In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach using Akaike’s Information Criterion.
-We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit.
-Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect strength, and the PI's discriminative ability.
-**Results:** Models including a PI by treatment interaction performed better under most simulation settings.
-Flexible models required larger sample sizes and higher AUC of the PI to outperform the linear interaction model.
-The adaptive approach performed similarly to the best-performing method in most scenarios.
-**Conclusion:** Usually, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
+ **Objective:** To compare different risk-based methods predicting individualized
+ treatment effects in RCT simulations. **Study Design and Setting:** We simulated
+ data using diverse assumptions for a baseline prognostic index of risk (PI) and
+ the shape of its intereraction with treatment (none, linear or quadratic). In 
+ each sample we predicted absolute benefit using: models with the PI and a 
+ constant relative treatment effect; models including an interaction of treatment 
+ with the PI; stratification in quarters of the PI; nonlinear transformations of 
+ the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach 
+ using Akaike’s Information Criterion. We evaluated predictive performance using 
+ root mean squared error and measures of discrimination and calibration for 
+ benefit. Starting from a base case scenario (sample size 4,250, treatment odds 
+ ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect 
+ strength, and the PI's discriminative ability. **Results:** Models including a 
+ PI by treatment interaction performed better under most simulation settings. 
+ Flexible models required larger sample sizes and higher AUC of the PI to 
+ outperform the linear interaction model. The adaptive approach performed 
+ similarly to the best-performing method in most scenarios. **Conclusion:** 
+ Under most circumstances, a model with a linear interaction of the PI with 
+ treatment is the optimal risk-based approach to predicting absolute treatment
+ benefit.
 authors:
@@
 
-\linenumbers
 # Introduction
@@
 
-In conclusion, when comparing different risk-based approaches to predict
-individualizing benefit a models including a linear treatment interaction with
-the prognostic performed best in a wide range of scenarios. More flexible models
-with restricted cubic splines required larger sample sizes and higher AUC of the
-PI to outperform the linear model. An adaptive approach, selecting the model
-with optimal AIC, had comparable performance to the best performing approach in
-most of the scenarios.
+In conclusion, when comparing different risk-based approaches to predicting 
+individualized treatment benefit, a model including a linear treatment 
+interaction with the prognostic index performed best in a wide range of 
+scenarios. More flexible models with restricted cubic splines required larger
+sample sizes and higher AUC of the prognostic index to outperform the linear
+interaction model. An adaptive approach, selecting the model with the optimal 
+AIC, had comparable performance to the best performing approach in most of the 
+scenarios.
 
@@
 # References
-\nolinenumbers
 \setlength{\parindent}{-0.25in}

### 2021-07-13 submission/manuscript.tex
comments on discussion

@@
 
-\title{Models including an interaction of treatment with baseline risk
-were better able to predict absolute benefit in simulations}
+\title{A model including an interaction of treatment with baseline risk
+was the optimal risk-based approach to predicting absolute treatment
+benefit}
 
@@
 performed similarly to the best-performing method in most scenarios.
-\textbf{Conclusion:} Usually, a model with a linear interaction of the
-PI with treatment is better able to predict absolute treatment benefit.
+\textbf{Conclusion:} Under most circumstances, a model with a linear
+interaction of the PI with treatment is the optimal risk-based approach
+to predicting absolute treatment benefit.
 \end{abstract}
@@
   }
-
-\linenumbers
 
@@
 
-In conclusion, when comparing different risk-based approaches to predict
-individualizing benefit a models including a linear treatment
-interaction with the prognostic performed best in a wide range of
-scenarios. More flexible models with restricted cubic splines required
-larger sample sizes and higher AUC of the PI to outperform the linear
-model. An adaptive approach, selecting the model with optimal AIC, had
-comparable performance to the best performing approach in most of the
-scenarios.
+In conclusion, when comparing different risk-based approaches to
+predicting individualized treatment benefit, a model including a linear
+treatment interaction with the prognostic index performed best in a wide
+range of scenarios. More flexible models with restricted cubic splines
+required larger sample sizes and higher AUC of the prognostic index to
+outperform the linear interaction model. An adaptive approach, selecting
+the model with the optimal AIC, had comparable performance to the best
+performing approach in most of the scenarios.
 
@@
 
-\nolinenumbers
 \setlength{\parindent}{-0.25in}

### 2021-09-03 submission/supplement.rmd
add table with models selected with the adaptive approach

@@
     caption    = "Performance metrics of the different methods used to predict benefit of tPA on 30-day mortality using data from the GUSTO-I trial."
+  ) %>%
+   kable_styling(latex_options = "HOLD_position")
+```
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+library(tidyverse)
+library(kableExtra)
+library(flextable)
+
+analysisIds <- read_csv(
+  here("data/processed/analysisIds.csv"),
+  col_types = "iffiiddddddddddddd"
 )
+
+interactionAnalysisIds <- read_csv(
+  here::here("data/processed/analysisIdsInteractions.csv")
+)
+
+dat1 <- readr::read_csv(here::here("data/processed/adaptiveModel.csv")) %>%
+  filter(scenarioId < 64) %>%
+  group_by(scenarioId) %>%
+  nest() %>% 
+  mutate(
+    test = lapply(data, plyr::count),
+    n = lapply(data, nrow)
+  ) %>%
+  select(scenarioId, test, n) %>%
+  arrange(scenarioId) %>%
+  unnest(c(test, n)) %>%
+  mutate(
+    perc = round(freq / n, 2),
+    model = ifelse(
+      selectedAdaptiveModel == "RCS with 3 knots",
+      "rcs",
+      selectedAdaptiveModel
+    )
+  ) %>%
+  ungroup() %>%
+  select(scenarioId, model, perc) %>%
+  spread(model, perc) %>%
+  replace(is.na(.), 0) %>%
+  left_join(analysisIds, by = c("scenarioId" = "scenario")) %>%
+  select(
+    type,
+    effectSize,
+    sampleSize,
+    auc,
+    treatment,
+    risk,
+    rcs
+  )
+
+dat2 <- readr::read_csv(here::here("data/processed/adaptiveModel.csv")) %>%
+  filter(scenarioId >= 64) %>%
+  group_by(scenarioId) %>%
+  nest() %>% 
+  mutate(
+    test = lapply(data, plyr::count),
+    n = lapply(data, nrow)
+  ) %>%
+  select(scenarioId, test, n) %>%
+  arrange(scenarioId) %>%
+  unnest(c(test, n)) %>%
+  mutate(
+    perc = round(freq / n, 2),
+    model = ifelse(
+      selectedAdaptiveModel == "RCS with 3 knots",
+      "rcs",
+      selectedAdaptiveModel
+    )
+  ) %>%
+  ungroup() %>%
+  select(scenarioId, model, perc) %>%
+  spread(model, perc) %>%
+  replace(is.na(.), 0) %>%
+  left_join(interactionAnalysisIds, by = c("scenarioId" = "scenario")) %>%
+  select(
+    type,
+    effectSize,
+    sampleSize,
+    auc,
+    treatment,
+    risk,
+    rcs
+  )
+
+dat <- dat1 %>% bind_rows(dat2)
+
+colourer <- scales::col_quantile(
+  palette = c("#e5f5f9", "#2ca25f"),
+  domain = c(0, 1),
+  n = 20
+)
+
+flextable(dat) %>%
+  theme_booktabs(bold_header = TRUE) %>%
+  bg(
+    bg = colourer,
+    j = c("treatment", "risk", "rcs"),
+    part = "body"
+  ) %>%
+  set_caption("Rates at which the adaptive approach selected each model type.")
 ```

### 2021-09-06 submission/manuscript.rmd
add sample size restriction for adaptive method

@@
 
-Third, we considered a logistic regression model including treatment, the prognostic index, and their interaction. Absolute benefit is then estimated from 
+Third, we considered a logistic regression model including treatment, the
+prognostic index, and their linear interaction. Absolute benefit is then
+estimated from
 $\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
@@
 
-Finally, we used *restricted cubic splines* (RCS) to relax the linearity
+Fourth, we used *restricted cubic splines* (RCS) to relax the linearity
 assumption on the effect of the linear predictor [@Harrell1988]. We considered
@@
 flexibility.
+
+Finally, we considered an adaptive approach using Akaike's Information Criterion
+(AIC) for model selection. The candidate models were: a constant treatment
+effect model, a model with a linear interaction with treatment and RCS models
+with 3, 4 and 5 knots.
 
@@
 ```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
-\\textit{Panel A} presents the results of the scenario with true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
+Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_base.png"))
@@
 ```{r rmsesamplesize, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
@@
 ```{r rmseauc, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects, while holding true prediction AUC and sample size constant; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/rmse_auc.png"))
@@
 ```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \\textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/discrimination_base.png"))
@@
 ```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \\textit{Panel B} presents the results under moderate linear deviations from constant treatment effectst; \\textit{Panel C} presents the results for strong linear deviations from of constant relative treatment effects; \\textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
+True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
 knitr::include_graphics(here::here("figures/calibration_base.png"))
@@
 
-Including restricted cubic splines with 4 or 5 knots proved to be too flexible,
-as indicated by higher RMSE, increased variability of discrimination for benefit
-and worse calibration of benefit predictions. Even with larger sample sizes and
-strong quadratic deviations from the base case scenario of constant relative
-treatment effects, these more flexible restricted cubic splines did not
-outperform the simpler RCS with 3 knots. These more flexible approaches may only
-be helpful if we expect more extreme patterns of heterogeneous treatment
-effects.
+Models including restricted cubic splines with 4 or 5 knots proved to be too
+flexible, as indicated by higher RMSE, increased variability of discrimination
+for benefit and worse calibration of benefit predictions. Even with larger
+sample sizes and strong quadratic deviations from the base case scenario of
+constant relative treatment effects, these more flexible restricted cubic
+splines did not outperform the simpler RCS with 3 knots. These approaches may
+only be helpful if we expect more extreme patterns of heterogeneous treatment
+effects compared to the quadratic deviations considered here.
 
@@
 
-Increasing the discriminative ability of the risk model – by increasing the
-predictor coefficients of the true risk model – reduced RMSE for all methods.
+Increasing the discriminative ability of the risk model--by increasing the
+predictor coefficients of the true risk model--reduced RMSE for all methods.
 This increase in discriminative ability translates in higher variability of
@@
 [@vanKlaveren2019; @Burke2014; @Abadie2018].
+
+The adaptive approach had adequate performance, following closely the
+performance of the "true" model in most scenarios. However, with smaller sample
+sizes it tended to "miss" the treatment-risk interactions and selected simpler
+models. This resulted in increased RMSE variability in these scenarios,
+especially in the case of true strong linear deviations from the base case
+scenario. This tendency was mitigated when the discriminative ability of the
+prediction model was higher. Therefore, in the case of smaller sample sizes
+and/or poorly discriminating prediction models, the simpler linear interaction
+model is a safer choice for predicting absolute benefits.
 

### 2021-09-06 submission/manuscript.tex
add sample size restriction for adaptive method

@@
 Third, we considered a logistic regression model including treatment,
-the prognostic index, and their interaction. Absolute benefit is then
-estimated from
+the prognostic index, and their linear interaction. Absolute benefit is
+then estimated from
 \(\hat{\tau}(\bm{x})=\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
@@
 
-Finally, we used \emph{restricted cubic splines} (RCS) to relax the
+Fourth, we used \emph{restricted cubic splines} (RCS) to relax the
 linearity assumption on the effect of the linear predictor {[}7{]}. We
@@
 different levels of flexibility.
+
+Finally, we considered an adaptive approach using Akaike's Information
+Criterion (AIC) for model selection. The candidate models were: a
+constant treatment effect model, a model with a linear interaction with
+treatment and RCS models with 3, 4 and 5 knots.
 
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
-\textit{Panel A} presents the results of the scenario with true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:rmsebase}
+The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
+Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.}\label{fig:rmsebase}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_sample_size} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 17,000; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations constant relative treatment effects.}\label{fig:rmsesamplesize}
+Sample size 17,000 rather than 4250 in Figure \ref{fig:rmsebase}}\label{fig:rmsesamplesize}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_auc} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.85 and sample size of 4,250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects, while holding true prediction AUC and sample size constant; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:rmseauc}
+True prediction AUC of 0.85 and sample size of 4,250.}\label{fig:rmseauc}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/discrimination_base} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effects; \textit{Panel C} presents the results for strong linear deviations from constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:discrimination}
+True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:discrimination}
 \end{figure}
@@
 \includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/calibration_base} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-\textit{Panel A} presents the results under the base case scenario of true constant relative treatment effect, with a true prediction AUC of 0.75 and sample size of 4250; \textit{Panel B} presents the results under moderate linear deviations from constant treatment effectst; \textit{Panel C} presents the results for strong linear deviations from of constant relative treatment effects; \textit{Panel D} presents the results for strong quadratic deviations from constant relative treatment effects.}\label{fig:calibration}
+True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:calibration}
 \end{figure}
@@
 
-Including restricted cubic splines with 4 or 5 knots proved to be too
-flexible, as indicated by higher RMSE, increased variability of
+Models including restricted cubic splines with 4 or 5 knots proved to be
+too flexible, as indicated by higher RMSE, increased variability of
 discrimination for benefit and worse calibration of benefit predictions.
@@
 flexible restricted cubic splines did not outperform the simpler RCS
-with 3 knots. These more flexible approaches may only be helpful if we
-expect more extreme patterns of heterogeneous treatment effects.
+with 3 knots. These approaches may only be helpful if we expect more
+extreme patterns of heterogeneous treatment effects compared to the
+quadratic deviations considered here.
 
@@
 
-Increasing the discriminative ability of the risk model -- by increasing
-the predictor coefficients of the true risk model -- reduced RMSE for
-all methods. This increase in discriminative ability translates in
-higher variability of predicted risks, which, in turn, allows the
-considered methods to better capture absolute treatment benefits. As a
-consequence, the increase in discriminative ability of the risk model
-also led to higher values of c-for-benefit. Even though risk model
-performance is very important for the ability of risk-based methods to
-predict treatment benefit, prediction model development was outside the
-scope of this work and has already been studied extensively
-{[}5,12,13{]}.
+Increasing the discriminative ability of the risk model--by increasing
+the predictor coefficients of the true risk model--reduced RMSE for all
+methods. This increase in discriminative ability translates in higher
+variability of predicted risks, which, in turn, allows the considered
+methods to better capture absolute treatment benefits. As a consequence,
+the increase in discriminative ability of the risk model also led to
+higher values of c-for-benefit. Even though risk model performance is
+very important for the ability of risk-based methods to predict
+treatment benefit, prediction model development was outside the scope of
+this work and has already been studied extensively {[}5,12,13{]}.
+
+The adaptive approach had adequate performance, following closely the
+performance of the ``true'' model in most scenarios. However, with
+smaller sample sizes it tended to ``miss'' the treatment-risk
+interactions and selected simpler models. This resulted in increased
+RMSE variability in these scenarios, especially in the case of true
+strong linear deviations from the base case scenario. This tendency was
+mitigated when the discriminative ability of the prediction model was
+higher. Therefore, in the case of smaller sample sizes and/or poorly
+discriminating prediction models, the simpler linear interaction model
+is a safer choice for predicting absolute benefits.
 

### 2021-09-06 submission/manuscript.rmd
update title

@@
 title: |
- A model including an interaction of treatment with baseline risk was the optimal
- risk-based approach to predicting absolute treatment benefit
+ Linear interaction of treatment with baseline risk was sufficient for predicting
+ individualized benefit
 abstract: |
@@
 
-In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a logistic regression
-model with 6 baseline covariates, i.e. age, Killip class, systolic blood
-pressure, heart rate, an indicator of previous MI, and the location of MI, to
-predict 30-day mortality risk. A constant effect of treatment was included in
-the model. When deriving risk predictions for individuals we set the treatment
-indicator to 0. More information on model development can be found in the
-supplement.
+In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
+logistic regression model with 6 baseline covariates, i.e. age, Killip class,
+systolic blood pressure, heart rate, an indicator of previous MI, and the
+location of MI, to predict 30-day mortality risk. A constant effect of treatment
+was included in the model. When deriving risk predictions for individuals we set
+the treatment indicator to 0. More information on model development can be found
+in the supplement.
 

### 2021-09-14 submission/manuscript.rmd
replace moderate linear interaction with absence of treatment effect in figures

@@
  Flexible models required larger sample sizes and higher AUC of the PI to
- outperform the linear interaction model. The adaptive approach performed 
- similarly to the best-performing method in most scenarios. **Conclusion:** 
- Under most circumstances, a model with a linear interaction of the PI with 
- treatment is the optimal risk-based approach to predicting absolute treatment
- benefit.
+ outperform the linear interaction model. The adaptive approach tended to select
+ simpler models with smaller sample sizes and/or worse prediction performance.
+ **Conclusion:** Under most circumstances, a model with a linear interaction of
+ the PI with treatment is the optimal risk-based approach to predicting absolute
+ treatment benefit.
 authors:
@@
 sizes it tended to "miss" the treatment-risk interactions and selected simpler
-models. This resulted in increased RMSE variability in these scenarios,
-especially in the case of true strong linear deviations from the base case
-scenario. This tendency was mitigated when the discriminative ability of the
-prediction model was higher. Therefore, in the case of smaller sample sizes
-and/or poorly discriminating prediction models, the simpler linear interaction
-model is a safer choice for predicting absolute benefits.
+models (Supplementary Table S7). This resulted in increased RMSE variability in
+these scenarios, especially in the case of true strong linear deviations from
+the base case scenario. This tendency was mitigated when the discriminative
+ability of the prediction model was higher. Therefore, in the case of smaller
+sample sizes and/or poorly discriminating prediction models, the simpler linear
+interaction model is a safer choice for predicting absolute benefits.
 

### 2021-09-14 submission/manuscript.tex
update title
replace moderate linear interaction with absence of treatment effect in figures

@@
 
-\title{A model including an interaction of treatment with baseline risk
-was the optimal risk-based approach to predicting absolute treatment
-benefit}
+\title{Linear interaction of treatment with baseline risk was sufficient
+for predicting individualized benefit}
 
@@
 the PI to outperform the linear interaction model. The adaptive approach
-performed similarly to the best-performing method in most scenarios.
-\textbf{Conclusion:} Under most circumstances, a model with a linear
-interaction of the PI with treatment is the optimal risk-based approach
-to predicting absolute treatment benefit.
+tended to select simpler models with smaller sample sizes and/or worse
+prediction performance. \textbf{Conclusion:} Under most circumstances, a
+model with a linear interaction of the PI with treatment is the optimal
+risk-based approach to predicting absolute treatment benefit.
 \end{abstract}
@@
 smaller sample sizes it tended to ``miss'' the treatment-risk
-interactions and selected simpler models. This resulted in increased
-RMSE variability in these scenarios, especially in the case of true
-strong linear deviations from the base case scenario. This tendency was
-mitigated when the discriminative ability of the prediction model was
-higher. Therefore, in the case of smaller sample sizes and/or poorly
-discriminating prediction models, the simpler linear interaction model
-is a safer choice for predicting absolute benefits.
+interactions and selected simpler models (Supplementary Table S7). This
+resulted in increased RMSE variability in these scenarios, especially in
+the case of true strong linear deviations from the base case scenario.
+This tendency was mitigated when the discriminative ability of the
+prediction model was higher. Therefore, in the case of smaller sample
+sizes and/or poorly discriminating prediction models, the simpler linear
+interaction model is a safer choice for predicting absolute benefits.
 

### 2021-09-14 code/CalibrationBase.R

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(10, 28, 37, 55)
+scenarios <- c(1, 10, 37, 55)
 
 metric <- "calibration"
 titles <- c(
-  "**A.** Constant treatment effect",
-  "**B.** Moderate linear deviation",
+  "**A.** Absent treatment effect",
+  "**B.** Constant treatment effect",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-14 code/DiscriminationBase.R

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(10, 28, 37, 55)
+scenarios <- c(1, 10, 37, 55)
 
 metric <- "discrimination"
 titles <- c(
-  "**A.** Constant treatment effect",
-  "**B.** Moderate linear deviation",
+  "**A.** Absent treatment effect",
+  "**B.** Constant treatment effect",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-14 code/RmseAuc.R

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(12, 30, 39, 57)
+scenarios <- c(3, 12, 39, 57)
 
 metric <- "rmse"
 titles <- c(
-  "**A.** Constant treatment effect",
-  "**B.** Moderate linear deviation",
+  "**A.** Absent treatment effect",
+  "**B.** Constant treatment effect",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-14 code/RmseBase.R

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(10, 28, 37, 55)
+scenarios <- c(1, 10, 37, 55)
 
 metric <- "rmse"
 titles <- c(
-  "**A.** Constant treatment effect",
-  "**B.** Moderate linear deviation",
+  "**A.** Absent treatment effect",
+  "**B.** Constant treatment effect",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-14 code/RmseSampleSize.R

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(16, 34, 43, 61)
+scenarios <- c(7, 16, 43, 61)
 
 metric <- "rmse"
 titles <- c(
-  "**A.** Constant treatment effect",
-  "**B.** Moderate linear deviation",
+  "**C.** Absent treatment effect",
+  "**B.** Constant treatment effect",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-16 code/CalibrationBase.R
update figure creation

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(1, 10, 37, 55)
+scenarios <- c(10, 28, 37, 55)
 
 metric <- "calibration"
 titles <- c(
-  "**A.** Absent treatment effect",
-  "**B.** Constant treatment effect",
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-16 code/DiscriminationBase.R

@@ -8,12 +8,12 @@ library(tidyverse)
 
 source("code/helpers/CreateManuscriptPlots.R")
 
-scenarios <- c(1, 10, 37, 55)
+scenarios <- c(10, 28, 37, 55)
 
 metric <- "discrimination"
 titles <- c(
-  "**A.** Absent treatment effect",
-  "**B.** Constant treatment effect",
+  "**A.** Constant treatment effect",
+  "**B.** Moderate linear deviation",
   "**C.** Strong linear deviation",
   "**D.** Strong quadratic deviation"
 )

### 2021-09-16 code/PlotRmseAuc.R

@@ -1,0 +1,100 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+source("code/helpers/CreateManuscriptPlots.R")
+source("code/helpers/PlotResult.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv")
+
+scenarios <- c(12, 30, 39, 57)
+metric    <- "rmse"
+value     <- "auc"
+
+titles <- scenarioIds %>%
+  filter(scenario %in% scenarios) %>%
+  select(title) %>%
+  unlist()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:4],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- plotResult(scenarios, processed, titles)
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
+)
+
+fileName <- paste0(
+  paste(
+    metric,
+    value,
+    sep = "_"
+  ),
+  ".tiff"
+)
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    dpi = 1200,
+    width = 8.5, 
+    height = 7,
+    compression = "lzw"
+  )
+  
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    width = 8.5, 
+    height = 7
+  )
+  

### 2021-09-16 code/PlotRmseBase.R

@@ -1,0 +1,100 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+source("code/helpers/CreateManuscriptPlots.R")
+source("code/helpers/PlotResult.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv")
+
+scenarios <- c(10, 28, 37, 55)
+metric    <- "rmse"
+value     <- "base"
+
+titles <- scenarioIds %>%
+  filter(scenario %in% scenarios) %>%
+  select(title) %>%
+  unlist()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:4],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- plotResult(scenarios, processed, titles)
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
+)
+
+fileName <- paste0(
+  paste(
+    metric,
+    value,
+    sep = "_"
+  ),
+  ".tiff"
+)
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    dpi = 1200,
+    width = 8.5, 
+    height = 7,
+    compression = "lzw"
+  )
+  
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    width = 8.5, 
+    height = 7
+  )
+  

### 2021-09-16 code/PlotRmseResult.R

@@ -1,121 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-source("code/helpers/CreateManuscriptPlots.R")
-
-args = commandArgs(trailingOnly = TRUE)
-scenarioIds <- readr::read_csv("data/processed/analysisIds.csv")
-
-scenarios <- as.numeric(args[1:4])
-metric    <- args[5]
-
-titles <- scenarioIds %>%
-  filter(scenario %in% scenarios) %>%
-  select(title) %>%
-  unlist()
-
-names(titles) <- NULL
-
-titles <- paste(
-  titlePrefix,
-  titles
-)
-
-titlePrefix <- paste0(
-  "**",
-  LETTERS[1:4],
-  ".**"
-)
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-
-plotList <- list()
-
-for (i in seq_along(scenarios)) {
-  tmp <- processed %>%
-    dplyr::filter(scenarioId == scenarios[i])
-  
-  plot <- createPlot(
-    data = tmp,
-    metric = metric,
-    title = titles[i],
-    limits = c(0, 10),
-    pointSize = .5
-  ) +
-    ggplot2::theme(
-      legend.position = "none",
-      axis.title.x    = ggplot2::element_blank(),
-      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
-      axis.title.y    = ggplot2::element_blank(),
-      axis.text.y     = ggplot2::element_text(size = 9.5),
-      plot.title      = ggtext::element_markdown(size = 14)
-    )
-  
-  plotList[[i]] <- plot
-}
-
-pp <- gridExtra::grid.arrange(
-  plotList[[1]],
-  plotList[[2]],
-  plotList[[3]],
-  plotList[[4]],
-  nrow = 2,
-  ncol = 2,
-  left = grid::textGrob(
-    expression(
-      paste(
[…]

### 2021-09-16 code/PlotRmseSampleSize.R

@@ -1,0 +1,100 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+source("code/helpers/CreateManuscriptPlots.R")
+source("code/helpers/PlotResult.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv")
+
+scenarios <- c(16, 34, 43, 61)
+metric    <- "rmse"
+value     <- "sample_size"
+
+titles <- scenarioIds %>%
+  filter(scenario %in% scenarios) %>%
+  select(title) %>%
+  unlist()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:4],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- plotResult(scenarios, processed, titles)
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
+)
+
+fileName <- paste0(
+  paste(
+    metric,
+    value,
+    sep = "_"
+  ),
+  ".tiff"
+)
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    dpi = 1200,
+    width = 8.5, 
+    height = 7,
+    compression = "lzw"
+  )
+  
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    width = 8.5, 
+    height = 7
+  )
+  

### 2021-09-16 code/RmseAuc.R

@@ -1,98 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-
-source("code/helpers/CreateManuscriptPlots.R")
-
-scenarios <- c(3, 12, 39, 57)
-
-metric <- "rmse"
-titles <- c(
-  "**A.** Absent treatment effect",
-  "**B.** Constant treatment effect",
-  "**C.** Strong linear deviation",
-  "**D.** Strong quadratic deviation"
-)
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-plotList <- list()
-
-for (i in seq_along(scenarios)) {
-  tmp <- processed %>%
-    dplyr::filter(scenarioId == scenarios[i])
-  
-  plot <- createPlot(
-    data = tmp,
-    metric = metric,
-    title = titles[i],
-    limits = c(0, 11),
-    pointSize = .5
-  ) +
-    ggplot2::theme(
-      legend.position = "none",
-      axis.title.x    = ggplot2::element_blank(),
-      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
-      axis.title.y    = ggplot2::element_blank(),
-      axis.text.y     = ggplot2::element_text(size = 9.5),
-      plot.title      = ggtext::element_markdown(size = 14)
-    )
-  
-  plotList[[i]] <- plot
-}
-
-pp <- gridExtra::grid.arrange(
-  plotList[[1]],
-  plotList[[2]],
-  plotList[[3]],
-  plotList[[4]],
-  nrow = 2,
-  ncol = 2,
-  left = grid::textGrob(
-    expression(
-      paste(
-        "Root mean squared error (x", 
-        10^-2, 
-        ")"
-      )
-    ),
-    rot = 90
-  )
-)
-  ggplot2::ggsave(
-    file.path("figures", "rmse_auc.tiff"), 
-    plot = pp,
-    dpi = 1200,
-    width = 8.5, 
[…]

### 2021-09-16 code/RmseBase.R

@@ -1,99 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-
-source("code/helpers/CreateManuscriptPlots.R")
-
-scenarios <- c(1, 10, 37, 55)
-
-metric <- "rmse"
-titles <- c(
-  "**A.** Absent treatment effect",
-  "**B.** Constant treatment effect",
-  "**C.** Strong linear deviation",
-  "**D.** Strong quadratic deviation"
-)
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-
-plotList <- list()
-
-for (i in seq_along(scenarios)) {
-  tmp <- processed %>%
-    dplyr::filter(scenarioId == scenarios[i])
-  
-  plot <- createPlot(
-    data = tmp,
-    metric = metric,
-    title = titles[i],
-    limits = c(0, 10),
-    pointSize = .5
-  ) +
-    ggplot2::theme(
-      legend.position = "none",
-      axis.title.x    = ggplot2::element_blank(),
-      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
-      axis.title.y    = ggplot2::element_blank(),
-      axis.text.y     = ggplot2::element_text(size = 9.5),
-      plot.title      = ggtext::element_markdown(size = 14)
-    )
-  
-  plotList[[i]] <- plot
-}
-
-pp <- gridExtra::grid.arrange(
-  plotList[[1]],
-  plotList[[2]],
-  plotList[[3]],
-  plotList[[4]],
-  nrow = 2,
-  ncol = 2,
-  left = grid::textGrob(
-    expression(
-      paste(
-        "Root mean squared error (x", 
-        10^-2, 
-        ")"
-      )
-    ),
-    rot = 90
-  )
-)
-  ggplot2::ggsave(
-    file.path("figures", "rmse_base.tiff"), 
-    plot = pp,
-    dpi = 1200,
-    width = 8.5, 
[…]

### 2021-09-16 code/RmseSampleSize.R

@@ -1,100 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-
-source("code/helpers/CreateManuscriptPlots.R")
-
-scenarios <- c(7, 16, 43, 61)
-
-metric <- "rmse"
-titles <- c(
-  "**C.** Absent treatment effect",
-  "**B.** Constant treatment effect",
-  "**C.** Strong linear deviation",
-  "**D.** Strong quadratic deviation"
-)
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-
-plotList <- list()
-
-for (i in seq_along(scenarios)) {
-  tmp <- processed %>%
-    dplyr::filter(scenarioId == scenarios[i])
-  
-  plot <- createPlot(
-    data = tmp,
-    metric = metric,
-    title = titles[i],
-    limits = c(0, 7),
-    pointSize = .5
-  ) +
-    ggplot2::theme(
-      legend.position = "none",
-      axis.title.x    = ggplot2::element_blank(),
-      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
-      axis.title.y    = ggplot2::element_blank(),
-      axis.text.y     = ggplot2::element_text(size = 9.5),
-      plot.title      = ggtext::element_markdown(size = 14)
-    )
-  
-  
-  plotList[[i]] <- plot
-}
-
-pp <- gridExtra::grid.arrange(
-  plotList[[1]],
-  plotList[[2]],
-  plotList[[3]],
-  plotList[[4]],
-  nrow = 2,
-  ncol = 2,
-  left = grid::textGrob(
-    expression(
-      paste(
-        "Root mean squared error (x", 
-        10^-2, 
-        ")"
-      )
-    ),
-    rot = 90
-  )
-)
-  ggplot2::ggsave(
-    file.path("figures", "rmse_sample_size.tiff"), 
-    plot = pp,
-    dpi = 1200,
-    width = 8.5, 
[…]

### 2021-09-16 code/helpers/PlotResult.R

@@ -1,0 +1,31 @@
+
+plotResult <- function(scenarios, processed, titles) {
+  
+  plotList <- list()
+  
+  for (i in seq_along(scenarios)) {
+    tmp <- processed %>%
+      dplyr::filter(scenarioId == scenarios[i])
+    
+    plot <- createPlot(
+      data = tmp,
+      metric = metric,
+      title = titles[i],
+      limits = c(0, 10),
+      pointSize = .5
+    ) +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.y     = ggplot2::element_text(size = 9.5),
+        plot.title      = ggtext::element_markdown(size = 14)
+      )
+    
+    plotList[[i]] <- plot
+  }
+  
+  return(plotList)
+}
+

### 2021-09-16 submission/manuscript.rmd
rebuild

@@
   - \usepackage{booktabs}
+  - \date{}
   - \newcommand\given[1][]{\:#1\vert\:}
-  - \date{}
 editor_options: 
@@
 library(rms)
+library(here)
 
@@
 
-```{r rmsebase, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
+```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
 The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
 Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_base.png"))
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_base.tiff")))
+# knitr::include_graphics(here::here("figures/rmse_base.png"))
 ```
@@
 
-```{r rmsesamplesize, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
 ```
@@
 
-```{r rmseauc, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_auc.png"))
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_auc.tiff")))
+# knitr::include_graphics(here::here("figures/rmse_auc.png"))
 ```
@@
 
-```{r discrimination, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/discrimination_base.png"))
+grid::grid.raster(tiff::readTIFF(here("figures/discrimination_base.tiff")))
+# knitr::include_graphics(here::here("figures/discrimination_base.png"))
 ```
@@
 
-```{r calibration, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r calibration, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/calibration_base.png"))
+grid::grid.raster(tiff::readTIFF(here("figures/calibration_base.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
 ```
@@
 
-```{r gusto, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}

### 2021-09-16 submission/manuscript.tex
rebuild

@@
 \usepackage{booktabs}
+\date{}
 \newcommand\given[1][]{\:#1\vert\:}
-\date{}
 \usepackage{booktabs}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_base} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsebase-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
 The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_sample_size} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsesamplesize-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 Sample size 17,000 rather than 4250 in Figure \ref{fig:rmsebase}}\label{fig:rmsesamplesize}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/rmse_auc} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmseauc-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.85 and sample size of 4,250.}\label{fig:rmseauc}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/discrimination_base} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/discrimination-1} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:discrimination}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/calibration_base} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/calibration-1} \caption{Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:calibration}

### 2021-09-16 submission/supplement.rmd
rebuild

@@
 ```
+
+```{r rmseAbsent, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods when there is no treatment effect.
+Each row contains the result for sample sizes of 1064, 4250 and 17000.
+Each column contains the results for AUC of the baseline risk prediction model of 0.65, 0.75 and 0.85.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_absent.tiff")))
+# knitr::include_graphics(here::here("figures/rmse_base.png"))
+```

### 2021-09-23 submission/manuscript.rmd
add model selection rates for adaptive

@@
   - \doublespacing
-  - \usepackage[left]{lineno}
+  - \usepackage[pagewise, left]{lineno}
   - \usepackage{amssymb}
   - \usepackage{bm}
+  - \usepackage{caption}
   - \usepackage{booktabs}
@@
 ```
+
+\doublespacing 
+\linenumbers
 
@@
 of knots in RCS resulted in higher RMSE across all scenarios. The adaptive 
-approach performed very similarly to the best performing model in each scenario.
+approach selected a linear interaction model 35\% of the time in the case of
+true strong linear deviations, while the majority of the time (56\%) it selected
+a simpler constant treatment effect model (Supplementary Table S7). Similarly, 
+in the case of true strong quadratic deviations, the adaptive approach selected 
+the RCS (3 knots) model 29\% of the time, while the simpler linear interaction 
+model was selected 68\% of the time.
 
@@
 number of knots of RCS smoothing resulted in increased RMSE, which was less
-pronounced in the case of strong quadratic deviations.
+pronounced in the case of strong quadratic deviations. With increasing 
+prediction AUC we saw an increased tendency of the adaptive approach to select
+more complex models compared to the base case scenario. In the case of strong 
+linear deviations, the linear interaction model was selected the majority of the
+time (63\%), while the simpler constant treatment effect model was selected 26\%
+of the time (Supplementart Table S7). However, the increased RMSE of the
+constant treatment effect model inflated the the RMSE of the linear interaction
+model. In the case of strong quadratic deviations the RCS (3 knots) model was
+selected 84\% of the time.
 

### 2021-09-23 submission/manuscript.tex
add model selection rates for adaptive

@@
 \doublespacing
-\usepackage[left]{lineno}
+\usepackage[pagewise, left]{lineno}
 \usepackage{amssymb}
 \usepackage{bm}
+\usepackage{caption}
 \usepackage{booktabs}
@@
   }
+
+\doublespacing 
+\linenumbers
 
@@
 RCS resulted in higher RMSE across all scenarios. The adaptive approach
-performed very similarly to the best performing model in each scenario.
+selected a linear interaction model 35\% of the time in the case of true
+strong linear deviations, while the majority of the time (56\%) it
+selected a simpler constant treatment effect model (Supplementary Table
+S7). Similarly, in the case of true strong quadratic deviations, the
+adaptive approach selected the RCS (3 knots) model 29\% of the time,
+while the simpler linear interaction model was selected 68\% of the
+time.
 
@@
 RCS smoothing resulted in increased RMSE, which was less pronounced in
-the case of strong quadratic deviations.
+the case of strong quadratic deviations. With increasing prediction AUC
+we saw an increased tendency of the adaptive approach to select more
+complex models compared to the base case scenario. In the case of strong
+linear deviations, the linear interaction model was selected the
+majority of the time (63\%), while the simpler constant treatment effect
+model was selected 26\% of the time (Supplementart Table S7). However,
+the increased RMSE of the constant treatment effect model inflated the
+the RMSE of the linear interaction model. In the case of strong
+quadratic deviations the RCS (3 knots) model was selected 84\% of the
+time.
 

### 2021-10-26 submission/manuscript.rmd
start restructuring methods - figures are not final

@@
 abstract: |
- **Objective:** To compare different risk-based methods predicting individualized
- treatment effects in RCT simulations. **Study Design and Setting:** We simulated
  data using diverse assumptions for a baseline prognostic index of risk (PI) and
  the shape of its intereraction with treatment (none, linear or quadratic). In
- each sample we predicted absolute benefit using: models with the PI and a
  constant relative treatment effect; models including an interaction of treatment
@@
  ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
- strength, and the PI's discriminative ability. **Results:** Models including a
  PI by treatment interaction performed better under most simulation settings.
@@
  simpler models with smaller sample sizes and/or worse prediction performance.
- **Conclusion:** Under most circumstances, a model with a linear interaction of
  the PI with treatment is the optimal risk-based approach to predicting absolute
  treatment benefit.
+ **Objective:** To compare different risk-based methods predicting individualized
+ treatment effects in RCT simulations. **Study Design and Setting:** We simulated
+ each sample we predicted absolute benefit using: models with the PI and a
+ strength, and the PI's discriminative ability. **Results:** Models including a
+ **Conclusion:** Under most circumstances, a model with a linear interaction of
 authors:
@@
 generated from a logistic regression model including all baseline covariates. In
-the base case scenario coefficient values were such, that the AUC of the
+the base scenarios coefficient values were such, that the AUC of the
 logistic regression model was $0.75$ and the event rate in the control arm was
-$20\%$.
+$20\%$. Binary outcomes in the control arm were generated from Bernoulli variables
+with true probabilities 
+$P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.
 
-While binary outcomes in the control arm were generated from Bernoulli variables
-with true probabilities $P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$, 
-the outcomes in the treatment arm were based on true probabilities 
-$\text{expit}(lp_1)$, with $$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0$$
+Outcomes in the treatment arm were generated using 3 base scenarios: absent
+treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and high
+treatment effect (OR = 0.5). We started with simulating outcomes based on true
+constant relative treatment effects for the 3 base scenarios. We then simulated
+linear, quadratic and non-monotonic deviations from constant treatment effects
+using:
+$$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, $$
+where $lp_1$ is the true linear predictor in the treatment arm, so that
+$P(y=1|X, t_x=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
+constant absolute harm is applied across all treated patients. In this case we
+have $P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}$.
 
-The coefficients $\gamma_0,\gamma_1$ and $\gamma_2$ along with the
-centering constant $c$ were set for each simulation scenario to mimic a wide
-variety scenarios, ranging from true constant relative treatment effect
-($\gamma_1=\gamma_2=0$) to moderate and strong linear ($\gamma_2=0$) and
-quadratic deviations. We also considered scenarios with treatment-covariate
-interactions.
-These scenarios included 4 weak interactions
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
-4 strong interactions 
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
-and 2 weak and 2 strong interactions.
-
-The sample size of the base case scenario was set to 4,250 ($80\%$ power for the
+The sample size for the base scenarios was set to 4,250 ($80\%$ power for the
 detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
-sample sizes of 1,064 and 17,000, respectively. We also evaluated the effect of
+sample sizes of 1,063 and 17,000, respectively. We also evaluated the effect of
 worse or better discriminative ability for risk, adjusting the baseline
@@
 
-Combining all these settings resulted in a simulation study of 66 scenarios
+Combining all these settings resulted in a simulation study of 486 scenarios
 (exact settings in the supplementary material).

### 2021-10-26 submission/manuscript.tex
start restructuring methods - figures are not final

@@
 \begin{abstract}
+data using diverse assumptions for a baseline prognostic index of risk
+(PI) and the shape of its intereraction with treatment (none, linear or
+quadratic). In constant relative treatment effect; models including an
+interaction of treatment with the PI; stratification in quarters of the
+PI; nonlinear transformations of the PI (restricted cubic splines with
+3, 4 and 5 knots); an adaptive approach using Akaike's Information
+Criterion. We evaluated predictive performance using root mean squared
+error and measures of discrimination and calibration for benefit.
+Starting from a base case scenario (sample size 4,250, treatment odds
+ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment
+effect PI by treatment interaction performed better under most
+simulation settings. Flexible models required larger sample sizes and
+higher AUC of the PI to outperform the linear interaction model. The
+adaptive approach tended to select simpler models with smaller sample
+sizes and/or worse prediction performance. the PI with treatment is the
+optimal risk-based approach to predicting absolute treatment benefit.
 \textbf{Objective:} To compare different risk-based methods predicting
 individualized treatment effects in RCT simulations. \textbf{Study
-Design and Setting:} We simulated data using diverse assumptions for a
-baseline prognostic index of risk (PI) and the shape of its
-intereraction with treatment (none, linear or quadratic). In each sample
-we predicted absolute benefit using: models with the PI and a constant
-relative treatment effect; models including an interaction of treatment
-with the PI; stratification in quarters of the PI; nonlinear
-transformations of the PI (restricted cubic splines with 3, 4 and 5
-knots); an adaptive approach using Akaike's Information Criterion. We
-evaluated predictive performance using root mean squared error and
-measures of discrimination and calibration for benefit. Starting from a
-base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
-the PI 0.75), we varied the sample size, the treatment effect strength,
-and the PI's discriminative ability. \textbf{Results:} Models including
-a PI by treatment interaction performed better under most simulation
-settings. Flexible models required larger sample sizes and higher AUC of
-the PI to outperform the linear interaction model. The adaptive approach
-tended to select simpler models with smaller sample sizes and/or worse
-prediction performance. \textbf{Conclusion:} Under most circumstances, a
-model with a linear interaction of the PI with treatment is the optimal
-risk-based approach to predicting absolute treatment benefit.
+Design and Setting:} We simulated each sample we predicted absolute
+benefit using: models with the PI and a strength, and the PI's
+discriminative ability. \textbf{Results:} Models including a
+\textbf{Conclusion:} Under most circumstances, a model with a linear
+interaction of
 \end{abstract}
@@
 the control arm were generated from a logistic regression model
-including all baseline covariates. In the base case scenario coefficient
+including all baseline covariates. In the base scenarios coefficient
 values were such, that the AUC of the logistic regression model was
-\(0.75\) and the event rate in the control arm was \(20\%\).
+\(0.75\) and the event rate in the control arm was \(20\%\). Binary
+outcomes in the control arm were generated from Bernoulli variables with
+true probabilities
+\(P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\).
 
-While binary outcomes in the control arm were generated from Bernoulli
-variables with true probabilities
-\(P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\), the outcomes in
-the treatment arm were based on true probabilities
-\(\text{expit}(lp_1)\), with
-\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]
+Outcomes in the treatment arm were generated using 3 base scenarios:
+absent treatment effect (OR = 1), moderate treatment effect (OR = 0.8)
+and high treatment effect (OR = 0.5). We started with simulating
+outcomes based on true constant relative treatment effects for the 3
+base scenarios. We then simulated linear, quadratic and non-monotonic
+deviations from constant treatment effects using:
+\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, \] where \(lp_1\)
+is the true linear predictor in the treatment arm, so that
+\(P(y=1|X, t_x=1) = \text{expit}(lp_1)\). Finally, we simulated
+scenarios where a constant absolute harm is applied across all treated
+patients. In this case we have
+\(P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}\).
 
-The coefficients \(\gamma_0,\gamma_1\) and \(\gamma_2\) along with the
-centering constant \(c\) were set for each simulation scenario to mimic
-a wide variety scenarios, ranging from true constant relative treatment
-effect (\(\gamma_1=\gamma_2=0\)) to moderate and strong linear
-(\(\gamma_2=0\)) and quadratic deviations. We also considered scenarios
-with treatment-covariate interactions. These scenarios included 4 weak
-interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
-interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
-weak and 2 strong interactions.
+The sample size for the base scenarios was set to 4,250 (\(80\%\) power
+for the detection of a marginal OR of 0.8). We evaluated the effect of
+smaller or larger sample sizes of 1,063 and 17,000, respectively. We
+also evaluated the effect of worse or better discriminative ability for
+risk, adjusting the baseline covariate coefficients, such that the AUC
+of the regression model in the control arm was 0.65 and 0.85
+respectively.
 
-The sample size of the base case scenario was set to 4,250 (\(80\%\)
-power for the detection of a marginal OR of 0.8). We evaluated the
-effect of smaller or larger sample sizes of 1,064 and 17,000,
-respectively. We also evaluated the effect of worse or better
-discriminative ability for risk, adjusting the baseline covariate
-coefficients, such that the AUC of the regression model in the control
-arm was 0.65 and 0.85 respectively.
-
-Combining all these settings resulted in a simulation study of 66
+Combining all these settings resulted in a simulation study of 486
 scenarios (exact settings in the supplementary material).

### 2021-10-26 code/PlotRmseBase.R

@@ -5,6 +5,7 @@
 # Depends: 
 
 library(tidyverse)
+library(glue)
 source("code/helpers/CreateManuscriptPlots.R")
 source("code/helpers/PlotResult.R")
 
@@ -16,6 +17,7 @@ value     <- "base"
 
 titles <- scenarioIds %>%
   filter(scenario %in% scenarios) %>%
+  mutate(title = glue("{base} treatment effect")) %>%
   select(title) %>%
   unlist()
 
@@ -53,7 +55,7 @@ processed <- readr::read_csv(
     f
   )
 
-plotList <- plotResult(scenarios, processed, titles)
+plotList <- plotResult(scenarios, processed, titles, metric = metric)
 
 pp <- gridExtra::grid.arrange(
   plotList[[1]],

### 2021-11-03 code/PlotRmseBaseNew.R
setup figure creation script for manuscript plots

@@ -6,6 +6,7 @@
 
 library(tidyverse)
 library(glue)
+library(ggtext)
 
 source("code/helpers/PlotGammas.R")                # Used to generate the absolute plots
 source("code/helpers/PlotDeviationsFunctions.R")   # Generates the absolute plot
@@ -39,7 +40,7 @@ names(titles) <- NULL
 
 titlePrefix <- paste0(
   "**",
-  LETTERS[1:4],
+  LETTERS[1:5],
   ".**"
 )
 
@@ -71,7 +72,11 @@ processed <- readr::read_csv(
   )
 
 
-plotAbsoluteBenefit <- function(data, label) {
+plotAbsoluteBenefit <- function(data) {
+  calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
+    ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
+    return(ret)
+  }
   calcAbsoluteBenefit <- function(p, g0 = 0, g1 = 0, g2 = 0, l = 0, harm) {
     x <- log(p / (1 - p))
     sq <- calcSquare(x, g0, g1, g2, l)
@@ -82,10 +87,10 @@ plotAbsoluteBenefit <- function(data, label) {
   res <- ggplot()
   for (i in seq_along(harmSettings)) {
     harm <- case_when(
-      data$harm == "moderate-positive" ~ data$averageTrueBenefit / 4, 
-      data$harm == "strong-positive" ~ data$averageTrueBenefit / 4, 
-      data$harm == "negative" ~ -data$averageTrueBenefit / 4, 
-      TRUE                          ~ 0
+      harmSettings[i] == "moderate-positive" ~ data$averageTrueBenefit[1] / 4, 
+      harmSettings[i] == "strong-positive"   ~ data$averageTrueBenefit[1] / 2, 
+      harmSettings[i] == "negative"          ~ -data$averageTrueBenefit[1] / 4, 
+      TRUE                                   ~ 0
     )
     args <- list(
       g0 = data$g0,
@@ -94,18 +99,19 @@ plotAbsoluteBenefit <- function(data, label) {
       l = data$c,
       harm = harm
     )
+    label <- harmSettings[i]
     res <- res + ggplot2::stat_function(
-      data = data.frame(x = c(.05, .95)),
-      ggplot2::aes(x = x, color = label),
+      data = data.frame(x = c(.05, .95), label = label),
[…]

### 2021-11-04 submission/manuscript.rmd
create script for all rmse plots

@@
 
-For each patient we generated 8 baseline covariates here
+For each patient we generated 8 baseline covariates
 $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
@@
 
-$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$
+$$\text{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2}$$
 We compared the discriminative ability of the methods under study using

### 2021-11-04 code/PlotRmse.R

@@ -1,8 +1,22 @@
 #!/usr/bin/env Rscript
 
-# Description: 
-# Output: 
-# Depends: 
+# --------------------------------------------------
+# Description:
+#   Generates the RMSE plots for the base case
+# Output:
+#   figures/rmse_base.tiff
+# Depends:
+#   data/processed/analysisIds.csv
+#   data/processed/rmse.csv
+#   code/helpers/CreateManuscriptPlots.R
+#   code/helpers/PlotResult.R
+#   code/helpers/Absolute.R
+# --------------------------------------------------
+
+args <- commandArgs(trailingOnly = TRUE)
+args_sampleSize <- as.numeric(args[1])
+args_auc <- as.numeric(args[2])
+args_value <- as.character(args[3])
 
 library(tidyverse)
 library(glue)
@@ -10,21 +24,19 @@ library(ggtext)
 library(gridExtra)
 library(grid)
 
-source("code/helpers/PlotGammas.R")                # Used to generate the absolute plots
-source("code/helpers/PlotDeviationsFunctions.R")   # Generates the absolute plot
 source("code/helpers/CreateManuscriptPlots.R")     # Geneartes single boxplot
 source("code/helpers/PlotResult.R")                # Generates the boxplot list
+source("code/helpers/Absolute.R")                  # Generates the absolute plots
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(
     base == "moderate",
     type != "quadratic-moderate",
-    sampleSize == 4250,
-    auc == .75
+    sampleSize == args_sampleSize,
+    auc == args_auc
   ) 
 
 metric    <- "rmse"
-value     <- "base"
 
 titles <- scenarioIds %>%
   mutate(
@@ -59,7 +71,6 @@ f <- function(x) x * 100
 processed <- readr::read_csv(
   file = file.path("data/processed", metricFile)
 ) %>% 
-  select(-one_of("locfit")) %>%
   mutate_at(
     c(
       "constant_treatment_effect",
@@ -73,45 +84,9 @@ processed <- readr::read_csv(
     f
   )
 
-
-plotAbsoluteBenefit <- function(data) {
-  calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
-    ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
-    return(ret)
-  }
[…]

### 2021-11-04 code/PlotRmseAuc.R

@@ -1,113 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-library(glue)
-library(patchwork)
-source("code/helpers/CreateManuscriptPlots.R")
-source("code/helpers/PlotResult.R")
-
-scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
-  filter(
-    base == "moderate",
-    !(type %in% c("moderate-linear", "moderate-quadratic")),
-    sampleSize == 4250,
-    auc == .85
-  )
-metric    <- "rmse"
-value     <- "auc"
-
-titles <- scenarioIds %>%
-  mutate(
-    title = ifelse(
-      type == "constant",
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
-    )
-  ) %>%
-  select(title) %>%
-  unlist() %>%
-  unique()
-
-names(titles) <- NULL
-
-titlePrefix <- paste0(
-  "**",
-  LETTERS[1:4],
-  ".**"
-)
-
-titles <- paste(
-  titlePrefix,
-  titles
-)
-
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  select(-one_of("locfit")) %>%
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-scenarios <- scenarioIds %>%
-  filter(harm == "absent") %>%
-  select(scenario) %>%
-  unlist()
-names(scenarios) <- NULL
-
-plotList <- plotResult(scenarios, processed, titles, metric = metric)
-
-pp <- gridExtra::grid.arrange(
-  plotList[[1]] + theme(axis.text.x = element_blank()),
-  plotList[[2]] + theme(legend.position = "none", axis.text.x = element_blank()),
-  plotList[[3]] + theme(legend.position = "none"),
-  plotList[[4]] + theme(legend.position = "none"),
-  heights = c(1, 1.3),
[…]

### 2021-11-04 code/PlotRmseBase.R

@@ -1,125 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-library(glue)
-
-source("code/helpers/PlotGammas.R")
-source("code/helpers/PlotDeviationsFunctions.R")
-source("code/helpers/CreateManuscriptPlots.R")
-source("code/helpers/PlotResult.R")
-
-scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
-  filter(
-    base == "moderate",
-    type != "quadratic-moderate",
-    sampleSize == 4250,
-    auc == .75
-  ) 
-
-metric    <- "rmse"
-value     <- "base"
-
-titles <- scenarioIds %>%
-  mutate(
-    title = ifelse(
-      type == "constant",
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
-    )
-  ) %>%
-  select(title) %>%
-  unlist() %>%
-  unique()
-
-names(titles) <- NULL
-
-titlePrefix <- paste0(
-  "**",
-  LETTERS[1:4],
-  ".**"
-)
-
-titles <- paste(
-  titlePrefix,
-  titles
-)
-
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  select(-one_of("locfit")) %>%
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-scenarios <- scenarioIds %>%
-  filter(harm == "absent") %>%
-  select(scenario) %>%
-  unlist()
-names(scenarios) <- NULL
-
-plotList <- plotResult(scenarios, processed, titles, metric = metric)
-
-absolutePlot <- generateAbsolutePlot(
-  analysisIds = scenarioIds,
-  scenarios = scenarios,
-  legend.position = "top",
-  legend.title = element_blank(),
-  legend.text = element_text(size = 14),
-  axis.title = element_text(size = 12)
-)
-
-pp <- gridExtra::grid.arrange(
-  absolutePlot,
[…]

### 2021-11-04 code/PlotRmseSampleSize.R

@@ -1,112 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Description: 
-# Output: 
-# Depends: 
-
-library(tidyverse)
-library(glue)
-source("code/helpers/CreateManuscriptPlots.R")
-source("code/helpers/PlotResult.R")
-
-scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
-  filter(
-    base == "moderate",
-    !(type %in% c("moderate-linear", "moderate-quadratic")),
-    sampleSize == 17000,
-    auc == .75
-  )
-metric    <- "rmse"
-value     <- "sample_size"
-
-titles <- scenarioIds %>%
-  mutate(
-    title = ifelse(
-      type == "constant",
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
-    )
-  ) %>%
-  select(title) %>%
-  unlist() %>%
-  unique()
-
-names(titles) <- NULL
-
-titlePrefix <- paste0(
-  "**",
-  LETTERS[1:4],
-  ".**"
-)
-
-titles <- paste(
-  titlePrefix,
-  titles
-)
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-processed <- readr::read_csv(
-  file = file.path("data/processed", metricFile)
-) %>% 
-  select(-one_of("locfit")) %>%
-  mutate_at(
-    c(
-      "constant_treatment_effect",
-      "stratified",
-      "linear_predictor",
-      "rcs_3_knots",
-      "rcs_4_knots",
-      "rcs_5_knots",
-      "adaptive"
-    ),
-    f
-  )
-
-scenarios <- scenarioIds %>%
-  filter(harm == "absent") %>%
-  select(scenario) %>%
-  unlist()
-names(scenarios) <- NULL
-
-plotList <- plotResult(scenarios, processed, titles, metric = metric)
-
-pp <- gridExtra::grid.arrange(
-  plotList[[1]] + theme(axis.text.x = element_blank()),
-  plotList[[2]] + theme(legend.position = "none", axis.text.x = element_blank()),
-  plotList[[3]] + theme(legend.position = "none"),
-  plotList[[4]] + theme(legend.position = "none"),
-  heights = c(1, 1.3),
-  nrow = 2,
[…]

### 2021-11-04 code/helpers/Absolute.R

@@ -1,0 +1,37 @@
+plotAbsoluteBenefit <- function(data) {
+  calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
+    ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
+    return(ret)
+  }
+  calcAbsoluteBenefit <- function(p, g0 = 0, g1 = 0, g2 = 0, l = 0, harm) {
+    x <- log(p / (1 - p))
+    sq <- calcSquare(x, g0, g1, g2, l)
+    benefit <- plogis(x) - plogis(sq) - harm
+    return(benefit)
+  }
+  harmSettings <- unique(data$harm)
+  res <- ggplot()
+  for (i in seq_along(harmSettings)) {
+    harm <- case_when(
+      harmSettings[i] == "moderate-positive" ~ data$averageTrueBenefit[1] / 4, 
+      harmSettings[i] == "strong-positive"   ~ data$averageTrueBenefit[1] / 2, 
+      harmSettings[i] == "negative"          ~ -data$averageTrueBenefit[1] / 4, 
+      TRUE                                   ~ 0
+    )
+    args <- list(
+      g0 = data$g0,
+      g1 = data$g1,
+      g2 = data$g2,
+      l = data$c,
+      harm = harm
+    )
+    label <- harmSettings[i]
+    res <- res + ggplot2::stat_function(
+      data = data.frame(x = c(.05, .95), label = label),
+      ggplot2::aes(x = x, color = label, group = label),
+      fun = calcAbsoluteBenefit,
+      args = args
+    )
+  }
+  return(res)
+}

### 2021-11-04 code/helpers/CreateManuscriptPlots.R

@@ -16,7 +16,7 @@ createPlot <- function(
     mutate(
       harm = factor(
         harm,
-        levels = c("absent", "positive", "negative", ordered = TRUE)
+        levels = c(unique(.data$harm, ordered = TRUE))
       )
     ) %>%
     dplyr::select(
@@ -67,9 +67,14 @@ createPlot <- function(
       ),
       breaks = c(
         "absent",
-        "positive",
-        "negative"
-        )
+        "moderate_positive",
+        "strong_positive"
+      ),
+      labels = c(
+        "absent",
+        "moderate positive",
+        "strong positive"
+      )
     ) +
     ggplot2::ylab(yAxis) +
     ggplot2::ylim(limits[1], limits[2]) +

### 2021-11-04 code/helpers/PlotResult.R

@@ -5,9 +5,9 @@ plotResult <- function(scenarios, processed, titles, metric, limits = c(0, 10)) 
     tmpList <- list()
     tmpList$absent <- processed %>%
       dplyr::filter(scenarioId == scenarios[i])
-    tmpList$positive <- processed %>%
+    tmpList$moderate_positive <- processed %>%
       dplyr::filter(scenarioId == scenarios[i] + 1)
-    tmpList$negative <- processed %>%
+    tmpList$strong_positive <- processed %>%
       dplyr::filter(scenarioId == scenarios[i] + 2)
     
     tmpData <- bind_rows(tmpList, .id = "harm")
@@ -18,18 +18,7 @@ plotResult <- function(scenarios, processed, titles, metric, limits = c(0, 10)) 
       title = titles[i],
       limits = limits,
       pointSize = .2
-    ) +
-      ggplot2::theme(
-        legend.position = c(.405, .923),
-        legend.title    = ggplot2::element_text(size = 10),
-        legend.text     = ggplot2::element_text(size = 7),
-        legend.direction = "horizontal",
-        axis.title.x    = ggplot2::element_blank(),
-        axis.text.x     = ggplot2::element_text(size = 9),
-        axis.title.y    = ggplot2::element_blank(),
-        axis.text.y     = ggplot2::element_text(size = 8.5),
-        plot.title      = ggtext::element_markdown(size = 14)
-      )
+    )
     
     plotList[[i]] <- plot
   }

### 2021-11-08 code/PlotRmse.R
fix absolute benefit distributions

@@ -164,7 +164,7 @@ gridList <- list(
       ggside.rect = element_blank(),
       ggside.axis.text = element_blank(),
       ggside.axis.ticks.length = unit(0, "pt"),
-      ggside.panel.scale = .04,
+      ggside.panel.scale = .07,
       plot.title = element_markdown(size = 9),
       legend.position = "none"
     ),
@@ -178,7 +178,7 @@ gridList <- list(
       ggside.rect = element_blank(),
       ggside.axis.text = element_blank(),
       ggside.axis.ticks.length = unit(0, "pt"),
-      ggside.panel.scale = .04,
+      ggside.panel.scale = .07,
       legend.position = "none"
     ),
   absolutePlots$plot[[2]] +
@@ -207,7 +207,7 @@ gridList <- list(
       ggside.rect = element_blank(),
       ggside.axis.text = element_blank(),
       ggside.axis.ticks.length = unit(0, "pt"),
-      ggside.panel.scale = .04,
+      ggside.panel.scale = .07,
       plot.title = element_markdown(size = 9, color = "white"),
       legend.position = "none"
     ),
@@ -244,7 +244,7 @@ gridList <- list(
       ggside.rect = element_blank(),
       ggside.axis.text = element_blank(),
       ggside.axis.ticks.length = unit(0, "pt"),
-      ggside.panel.scale = .04,
+      ggside.panel.scale = .07,
       plot.title = element_markdown(size = 9, color = "white"),
       legend.position = "none"
     ),
@@ -282,7 +282,7 @@ gridList <- list(
       ggside.rect = element_blank(),
       ggside.axis.text = element_blank(),
       ggside.axis.ticks.length = unit(0, "pt"),
-      ggside.panel.scale = .04,
+      ggside.panel.scale = .07,
       plot.title = element_markdown(size = 9, color = "white"),
       legend.position = "none"
     )

### 2021-11-08 code/helpers/Absolute.R

@@ -11,6 +11,7 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL) {
   }
   harmSettings <- unique(data$harm)
   res <- ggplot()
+  validationDataset <- list()
   for (i in seq_along(harmSettings)) {
     scenarioSettings <- data %>%
       dplyr::filter(harm == harmSettings[i]) %>%
@@ -22,21 +23,11 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL) {
           TRUE                        ~ 0
         )
       )
-
-    # harm <- case_when(
-    #   harmSettings[i] == "moderate-positive" ~ data$averageTrueBenefit[1] / 4, 
-    #   harmSettings[i] == "strong-positive"   ~ data$averageTrueBenefit[1] / 2, 
-    #   harmSettings[i] == "negative"          ~ -data$averageTrueBenefit[1] / 4, 
-    #   TRUE                                   ~ 0
-    # )
-
     scenarioDir <- paste(
       "scenario",
       as.character(scenarioSettings$scenario),
       sep = "_"
     )
-
-
     args <- list(
       g0   = scenarioSettings$g0,
       g1   = scenarioSettings$g1,
@@ -52,43 +43,53 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL) {
         fun = calcAbsoluteBenefit,
         args = args
       )
-  }
-
-  if (!is.null(projectDir)) {
-    settings <- readRDS(
-      file.path(
-        projectDir,
-        "data",
-        "raw",
-        scenarioDir,
-        "settings.rds"
+    if (!is.null(projectDir)) {
+      settings <- readRDS(
+        file.path(
+          projectDir,
+          "data",
+          "raw",
+          scenarioDir,
+          "settings.rds"
+        )
       )
-    )
-    validationDatabaseSettings <- settings$simulationSettings$databaseSettings
-    validationDatabaseSettings$numberOfObservations <- 1e3
-    validationDataset <- SimulateHte::runDataGeneration(
-      databaseSettings        = validationDatabaseSettings,
-      propensitySettings      = settings$simulationSettings$propensitySettings,
-      baselineRiskSettings    = settings$simulationSettings$baselineRiskSettings,
-      treatmentEffectSettings = settings$simulationSettings$treatmentEffectSettings
-    )
+      validationDatabaseSettings <- settings$simulationSettings$databaseSettings
+      validationDatabaseSettings$numberOfObservations <- 1e5
[…]

### 2021-11-08 submission/manuscript.rmd
start working on results section

@@
  **Objective:** To compare different risk-based methods predicting individualized
- treatment effects in RCT simulations. **Study Design and Setting:** We simulated
+ treatment effects in GRIT simulations. **Study Design and Setting:** We simulated
  data using diverse assumptions for a baseline prognostic index of risk (PI) and
- the shape of its intereraction with treatment (none, linear or quadratic). In
+ the shape of its interaction with treatment (none, linear or quadratic). In
  each sample we predicted absolute benefit using: models with the PI and a
@@
  ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
- strength, and the PI's discriminative ability. **Results:** Models including a
+ strength, and the Ti's discriminative ability. **Results:** Models including a
  PI by treatment interaction performed better under most simulation settings.
@@
 
-The model including a constant relative treatment effect had the lowest median
-RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
-`r 4250` and AUC = 0.75) or with moderate relative deviations (Figure
-\ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
-quadratic deviations from a constant relative treatment effect the linear
-interaction model performed best (Figure \ref{fig:rmsebase}; Panels B and C).
-Only in the case of strong quadratic deviations models including RCS (3 knots)
-performed equally well to the linear interaction method. Increasing the number
-of knots in RCS resulted in higher RMSE across all scenarios. The adaptive 
-approach selected a linear interaction model 35\% of the time in the case of
-true strong linear deviations, while the majority of the time (56\%) it selected
-a simpler constant treatment effect model (Supplementary Table S7). Similarly, 
-in the case of true strong quadratic deviations, the adaptive approach selected 
-the RCS (3 knots) model 29\% of the time, while the simpler linear interaction 
-model was selected 68\% of the time.
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveModel.csv"))
+adaptiveSelections <- adaptiveSelections %>% 
+  group_by(scenarioId) %>% 
+  nest() %>% 
+  mutate(
+    counts = lapply(data, plyr::count),
+    n = lapply(data, nrow)
+  ) %>%
+  select(
+    scenarioId, counts, n
+  ) %>%
+  unnest(c(counts, n)) %>%
+  mutate(
+    perc = round(freq / n, 2),
+    model = case_when(
+      selectedAdaptiveModel == "RCS with 3 knots" ~ "rcs_3",
+      selectedAdaptiveModel == "RCS with 4 knots" ~ "rcs_4",
+      selectedAdaptiveModel == "RCS with 5 knots" ~ "rcs_5",
+      TRUE                                        ~ selectedAdaptiveModel
+    )
+  ) %>%
+  ungroup() %>%
+  select(scenarioId, model, perc) %>%
+  spread(model, perc) %>%
+  relocate(c(treatment, risk), .before = rcs_3)
 
-```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
+```
+
+The model including a constant relative treatment effect had the
+lowest median RMSE in scenarios with a true constant relative
+treatment effect (OR = 0.8, N = `r 4250` and AUC = 0.75) or with
+moderate linear deviations (Supplement, Figure XX). However, when we
+considered deviations from constant relative treatment effects, the
+linear interaction model performed best most of the time, with RCS-3
+having very comparable performance, outperforming the former in the
+case of strong treatment-related harms and non-monotonic
+deviations. The adaptive approach had comparable performance to the
+best-performing method in each scenario, however, we observed
+increased variability in the case of linear and non-monotonic
+deviations, especially for moderate or strong treatment-related
+harms.
+
+The adaptive approach selected the constant treatment effect model
+`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\% 
+of the time, when no constant treatment-related harms were
+considered, which dropped to
+`r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
+in the case of strong treatment-related harms. When strong linear
+deviations were considered, the adaptive approach favored the constant
+treatment effect model
+(`r adaptiveSelections %>% filter(scenarioId == 289) %>% pull(treatment) * 100`\%
+of the time) in the absence of treatment-related harms, which was
+replaced mainly by the linear interaction model (from
+`r adaptiveSelections %>% filter(scenarioId == 289) %>% pull(risk) * 100`\%
+to 
+`r adaptiveSelections %>% filter(scenarioId == 291) %>% pull(risk) * 100`\%
+of the time) with increasing treatment-related harms. In the case of
+strong quadratic deviations, the adaptive approach selected the
+linear interaction model the majority of the time which also increased
+with stronger treatment-related harms (from
+`r adaptiveSelections %>% filter(scenarioId == 361) %>% pull(risk) * 100`\%
+with no harms to
+`r adaptiveSelections %>% filter(scenarioId == 363) %>% pull(risk) * 100`\%
+with strong harms ). RCS-3 selection frequency decreased with stronger
+treatment-related harms (from
+`r adaptiveSelections %>% filter(scenarioId == 361) %>% pull(rcs_3) * 100`\%
+with no harms to
+`r adaptiveSelections %>% filter(scenarioId == 363) %>% pull(rcs_3) * 100`\%
+with strong harms ). Finally, with true non-monotonic deviations the
+adaptive approach mainly selected the constant treatment effect model
+(from
+`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(treatment) * 100`\%
+with no harms to
+`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(treatment) * 100`\%
+with strong harms), the linear interaction model (from
+`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
+with no harms to
+`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
+with strong harms) and the RCS-3 model
+(from
+`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
+with no harms to
+`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
+with strong harms). For more details, see Supplement Table XX).
+
+```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
 The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
@@
 
-When we increased the sample size (N = `r 17000`), the model including a
-constant relative treatment effect had the lowest RMSE under the assumption of
-true constant relative treatment effects (Figure \ref{fig:rmsesamplesize};
-Panel A). However, when introducing moderate and strong linear deviations the
-linear interaction model outperformed the constant treatment
-effect model (Figure \ref{fig:rmsesamplesize}; Panels B and C). Furthermore, the
-linear interaction model was outperformed by the more flexible RCS models (3
-knots) in the case of strong quadratic deviations. Again, the increased
-flexibility of RCS smoothing with higher number of knots resulted in overfitting
-and worse predictive accuracy (Figure \ref{fig:rmsesamplesize}; Panel D).
+When we increased the sample size (N = `r 17000`), the model including
+a constant relative treatment effect had the lowest RMSE under the
+assumption of true constant relative treatment effects (Figure
+\ref{fig:rmsesamplesize}; Panel A). However, when introducing strong
+linear deviations the linear interaction model outperformed the
+constant treatment effect model (Figure \ref{fig:rmsesamplesize};
+Panel B). Furthermore, the linear interaction model was outperformed
+by the more flexible RCS models (3 knots) in the case of strong
+quadratic and non-monotonic deviations. Again, the increased
+flexibility of RCS smoothing with higher number of knots resulted in
+overfitting and worse predictive accuracy (Figure
+\ref{fig:rmsesamplesize}; Panels C and D). The adaptive approach again
+followed the performance of the best performing models in each
+scenario. In addition, the increased variability in the adaptive
+method's errors in the base scenario disappeared with large sample
+size.
+
+The adaptive approach tended to increasingly favor smoother methods
+(especially RCS-3) with increasing treatment-related harms (see
+Supplement, Table XX). However, in the case of true strong quadratic
+deviations the opposite was observed: selection frequency of the
+linear interaction model increased from
+`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(risk) * 100`\% (no harms)
+to
+`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(risk) * 100`\% (strong harms)
+whereas for RCS-3 decreased from
+`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(rcs_3) * 100`\% (no harms)
+to
+`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(rcs_3) * 100`\% (strong harms).
 
@@
 time (63\%), while the simpler constant treatment effect model was selected 26\%
-of the time (Supplementart Table S7). However, the increased RMSE of the
+of the time (Supplementary Table S7). However, the increased RMSE of the
 constant treatment effect model inflated the the RMSE of the linear interaction
@@
 
-```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}

### 2021-11-09 submission/manuscript.rmd
fix plots and continue results section

@@
 
-The model including a constant relative treatment effect had the
-lowest median RMSE in scenarios with a true constant relative
-treatment effect (OR = 0.8, N = `r 4250` and AUC = 0.75) or with
-moderate linear deviations (Supplement, Figure XX). However, when we
-considered deviations from constant relative treatment effects, the
-linear interaction model performed best most of the time, with RCS-3
-having very comparable performance, outperforming the former in the
-case of strong treatment-related harms and non-monotonic
-deviations. The adaptive approach had comparable performance to the
-best-performing method in each scenario, however, we observed
-increased variability in the case of linear and non-monotonic
-deviations, especially for moderate or strong treatment-related
-harms.
+The linear interaction model outperformed all RCS methods in terms of
+RMSE in scenarios with true constant relative treatment effect (OR =
+0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
+quadratic deviations (Figure \ref{fig:rmsebase}). However, with
+non-monotonic deviations errors of the linear interaction model
+increased substantially, especially in the presence of
+treatment-related harms. In these scenarios, RCS-3 proved to be quite
+robust outperforming all other methods. The constant treatment effect
+approach had overall best performance under true constant treatment
+effect settings, but was sensitive to all considered deviations,
+resulting in increased error rates. Finally, the adaptive approach had
+comparable performance to the best-performing method in each
+scenario. However, we observed increased variability in the case of
+linear and non-monotonic deviations, especially for moderate or strong
+treatment-related harms.
 
-The adaptive approach selected the constant treatment effect model
-`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\% 
-of the time, when no constant treatment-related harms were
-considered, which dropped to
+In the case of true constant relative treatment effects, the adaptive
+approach selected the constnant effect model the majority of the time,
+even in the presence of treatment-related harms
+(`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\%
+of the time with no harms to
 `r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
-in the case of strong treatment-related harms. When strong linear
-deviations were considered, the adaptive approach favored the constant
-treatment effect model
-(`r adaptiveSelections %>% filter(scenarioId == 289) %>% pull(treatment) * 100`\%
-of the time) in the absence of treatment-related harms, which was
-replaced mainly by the linear interaction model (from
-`r adaptiveSelections %>% filter(scenarioId == 289) %>% pull(risk) * 100`\%
-to 
-`r adaptiveSelections %>% filter(scenarioId == 291) %>% pull(risk) * 100`\%
-of the time) with increasing treatment-related harms. In the case of
-strong quadratic deviations, the adaptive approach selected the
-linear interaction model the majority of the time which also increased
-with stronger treatment-related harms (from
-`r adaptiveSelections %>% filter(scenarioId == 361) %>% pull(risk) * 100`\%
+with strong harms). With strong linear or quadratic deviations and
+increasing treatment-related harms the adaptive approach increasingly
+favored the linear interaction model. In the case of true
+non-monotonic deviations, we observed increasing selection frequencies
+for RCS-3 with increasing treatment harms (from
+`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
 with no harms to
-`r adaptiveSelections %>% filter(scenarioId == 363) %>% pull(risk) * 100`\%
-with strong harms ). RCS-3 selection frequency decreased with stronger
-treatment-related harms (from
-`r adaptiveSelections %>% filter(scenarioId == 361) %>% pull(rcs_3) * 100`\%
-with no harms to
-`r adaptiveSelections %>% filter(scenarioId == 363) %>% pull(rcs_3) * 100`\%
-with strong harms ). Finally, with true non-monotonic deviations the
-adaptive approach mainly selected the constant treatment effect model
-(from
-`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(treatment) * 100`\%
-with no harms to
-`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(treatment) * 100`\%
-with strong harms), the linear interaction model (from
+`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
+with strong harms), whereas selection frequencies for the linear
+interaction model dropped from
 `r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
@@
 `r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
-with strong harms) and the RCS-3 model
-(from
-`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
-with no harms to
-`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
-with strong harms). For more details, see Supplement Table XX).
+with strong harms. Finally, in the case of strong linear and
+non-monotonic deviations selection frequencies of the constant
+treatment effect model remained high despite increasing treatment
+harms (Supplement, Table XX).
 
@@
 
-When we increased the sample size (N = `r 17000`), the model including
-a constant relative treatment effect had the lowest RMSE under the
-assumption of true constant relative treatment effects (Figure
-\ref{fig:rmsesamplesize}; Panel A). However, when introducing strong
-linear deviations the linear interaction model outperformed the
-constant treatment effect model (Figure \ref{fig:rmsesamplesize};
-Panel B). Furthermore, the linear interaction model was outperformed
-by the more flexible RCS models (3 knots) in the case of strong
-quadratic and non-monotonic deviations. Again, the increased
-flexibility of RCS smoothing with higher number of knots resulted in
-overfitting and worse predictive accuracy (Figure
-\ref{fig:rmsesamplesize}; Panels C and D). The adaptive approach again
-followed the performance of the best performing models in each
-scenario. In addition, the increased variability in the adaptive
-method's errors in the base scenario disappeared with large sample
-size.
+Increasing the sample size to `r 17000` favored the RCS-3 approach the
+most, as it achieved lowest or close to lowest RMSE across all
+scenarios (Figure \ref{fig:rmsesamplesize}). Especially in cases of
+strong quadratic and non-monotonic deviations RCS-3 had lower error
+rates compared to the linear interaction approach, regardless of
+treatment-related harms strength. The issues with the large error
+variability of the adaptive approach improved with larger sample
+sizes.
 

### 2021-11-09 code/CalibrationBase.R

@@ -6,13 +6,14 @@
 
 library(tidyverse)
 library(glue)
+library(ggtext)
 source("code/helpers/CreateManuscriptPlots.R")
 source("code/helpers/PlotResult.R")
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(
     base == "moderate",
-    !(type %in% c("moderate-linear", "moderate-quadratic")),
+    !(type %in% c("linear-moderate", "quadratic-moderate")),
     sampleSize == 4250,
     auc == .75
   )
@@ -74,11 +75,36 @@ names(scenarios) <- NULL
 plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 6))
 
 pp <- gridExtra::grid.arrange(
-  plotList[[1]] + theme(axis.text.x = element_blank()),
-  plotList[[2]] + theme(legend.position = "none", axis.text.x = element_blank()),
-  plotList[[3]] + theme(legend.position = "none"),
-  plotList[[4]] + theme(legend.position = "none"),
-  heights = c(1, 1.3),
+  plotList[[1]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.direction = "horizontal",
+      legend.position = c(.432, .93),
+      legend.text = element_text(size = 7),
+      legend.title = element_text(size = 8),
+      axis.text.x = element_blank()
+    ),
+  plotList[[2]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.position = "none",
+      axis.text.x = element_blank()
+    ),
+  plotList[[3]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.position = "none"
+    ),
+  plotList[[4]] +
+    theme(
+      axis.title = element_blank(),
+      plot.title = element_markdown(),
+      legend.position = "none"
+    ),
+  heights = c(1, 1.05),
   nrow = 2,
   ncol = 2,
   left = grid::textGrob(

### 2021-11-09 code/DiscriminationBase.R

@@ -6,13 +6,14 @@
 
 library(tidyverse)
 library(glue)
+library(ggtext)
 source("code/helpers/CreateManuscriptPlots.R")
 source("code/helpers/PlotResult.R")
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(
     base == "moderate",
-    !(type %in% c("moderate-linear", "moderate-quadratic")),
+    !(type %in% c("linear-moderate", "quadratic-moderate")),
     sampleSize == 4250,
     auc == .75
   )
@@ -59,11 +60,36 @@ names(scenarios) <- NULL
 plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(.5, .57))
 
 pp <- gridExtra::grid.arrange(
-  plotList[[1]] + theme(axis.text.x = element_blank()),
-  plotList[[2]] + theme(legend.position = "none", axis.text.x = element_blank()),
-  plotList[[3]] + theme(legend.position = "none"),
-  plotList[[4]] + theme(legend.position = "none"),
-  heights = c(1, 1.3),
+  plotList[[1]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.direction = "horizontal",
+      legend.position = c(.444, .93),
+      legend.text = element_text(size = 7),
+      legend.title = element_text(size = 8),
+      axis.text.x = element_blank()
+    ),
+  plotList[[2]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.position = "none",
+      axis.text.x = element_blank()
+    ),
+  plotList[[3]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.position = "none"
+    ),
+  plotList[[4]] +
+    theme(
+      plot.title = element_markdown(),
+      axis.title = element_blank(),
+      legend.position = "none"
+    ),
+  heights = c(1, 1.05),
   nrow = 2,
   ncol = 2,
   left = "C-statistic for benefit"

### 2021-11-09 code/PlotRmse.R

@@ -143,7 +143,7 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.05, .15),
+      limits = c(-.058, .15),
       breaks = seq(-.05, .15, .05)
     ) +
   scale_color_manual(
@@ -186,7 +186,7 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.05, .15),
+      limits = c(-.058, .15),
       breaks = seq(-.05, .15, .05)
     ) +
     scale_color_manual(
@@ -223,7 +223,7 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.05, .15),
+      limits = c(-.058, .15),
       breaks = seq(-.05, .15, .05)
     ) +
     scale_color_manual(
@@ -261,7 +261,7 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.05, .15),
+      limits = c(-.058, .15),
       breaks = seq(-.05, .15, .05)
     ) +
     scale_color_manual(

### 2021-11-10 submission/manuscript.rmd
first revision of results section

@@
  **Objective:** To compare different risk-based methods predicting individualized
- treatment effects in GRIT simulations. **Study Design and Setting:** We simulated
+ treatment effects in RCT simulations. **Study Design and Setting:** We simulated
  data using diverse assumptions for a baseline prognostic index of risk (PI) and
@@
 
-Combining all these settings resulted in a simulation study of 486 scenarios
+Combining all these settings resulted in a simulation study of 648 scenarios
 (exact settings in the supplementary material).
@@
 Second, we considered a model which assumes *constant relative treatment effect*
-(constant OR). Hence, absolute benefit is predicted from
+(constant odds ratio). Hence, absolute benefit is predicted from
 $\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveModel.csv"))
-adaptiveSelections <- adaptiveSelections %>% 
-  group_by(scenarioId) %>% 
-  nest() %>% 
-  mutate(
-    counts = lapply(data, plyr::count),
-    n = lapply(data, nrow)
-  ) %>%
-  select(
-    scenarioId, counts, n
-  ) %>%
-  unnest(c(counts, n)) %>%
-  mutate(
-    perc = round(freq / n, 2),
-    model = case_when(
-      selectedAdaptiveModel == "RCS with 3 knots" ~ "rcs_3",
-      selectedAdaptiveModel == "RCS with 4 knots" ~ "rcs_4",
-      selectedAdaptiveModel == "RCS with 5 knots" ~ "rcs_5",
-      TRUE                                        ~ selectedAdaptiveModel
-    )
-  ) %>%
-  ungroup() %>%
-  select(scenarioId, model, perc) %>%
-  spread(model, perc) %>%
-  relocate(c(treatment, risk), .before = rcs_3)
-
+```{r adaptive, echo=FALSE, warning=FALSE, message=FALSE}
+adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveSelections.csv"))
+rmseDistribution <- readr::read_csv(here::here("data/processed/rmseDistribution.csv"))
+aucDistribution <- readr::read_csv(here::here("data/processed/discriminationDistribution.csv"))
 ```
-
 The linear interaction model outperformed all RCS methods in terms of
@@
 comparable performance to the best-performing method in each
-scenario. However, we observed increased variability in the case of
-linear and non-monotonic deviations, especially for moderate or strong
-treatment-related harms.
+scenario. However, we observed increased error variability in the case
+of linear and non-monotonic deviations, especially for moderate or
+strong treatment-related harms.
 
@@
 strong quadratic and non-monotonic deviations RCS-3 had lower error
-rates compared to the linear interaction approach, regardless of
-treatment-related harms strength. The issues with the large error
-variability of the adaptive approach improved with larger sample
-sizes.
+rates
+(median RMSE
+`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
+for strong quadratic deviations and 
+`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
+for non-monotonic deviations with no treatment-related harms)
+compared to the linear interaction approach
+(median RMSE
+`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
+`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
+respectively), regardless of treatment-related harms strength. The
+issues with the large error variability of the adaptive approach
+improved with larger sample sizes.
 
@@
 
-When we increased the AUC of the true prediction model to 0.85, models including
-RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
-from the base case of true constant relative treatment effects (Figure
-\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
-interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
-being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
-number of knots of RCS smoothing resulted in increased RMSE, which was less
-pronounced in the case of strong quadratic deviations. With increasing 
-prediction AUC we saw an increased tendency of the adaptive approach to select
-more complex models compared to the base case scenario. In the case of strong 
-linear deviations, the linear interaction model was selected the majority of the
-time (63\%), while the simpler constant treatment effect model was selected 26\%
-of the time (Supplementary Table S7). However, the increased RMSE of the
-constant treatment effect model inflated the the RMSE of the linear interaction
-model. In the case of strong quadratic deviations the RCS (3 knots) model was
-selected 84\% of the time.
+When we increased the AUC of the true prediction model to 0.85 (OR =
+0.8 and N = `r 4250`) the constant effect model outperformed all other
+methods in the case of true constant treatment effects, but proved to
+be the least robust to deviations. Again, RCS-3 had the lowest
+error rates in the case of strong quadratic or monotonic deviations
+and very comparable performance to the best-performing linear
+interaction model in the case of strong linear deviations
+(median RMSE
+`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
+`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
+for the linear interaction model). The adaptive approach, though it
+performed similar to the best performing method in each scenario, on
+average, there was increased variability in error rates in the case of
+strong linear and non-monotonic deviations. In these scenarios, the adaptive approach often selected the constant treatment effect model
+(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
+and
+`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
+in the strong linear and non-monotonic deviation scenarios without
+treatment-related harms, respectively).
 
@@
 
-The constant effects model, the linear interaction model and models with RCS
-smoothing (3 knots) had the highest median c-for-benefit in the base case
-scenario and the scenarios where linear and quadratic deviations were
-considered. The constant treatment effect model and the linear interaction model
-tended to present much lower variability compared to all other approaches
-(Figure \ref{fig:discrimination}). With increasing number of RCS knots, we
-observed decreasing median values and increasing variability of the
-c-for-benefit in all scenarios.
+In terms of discrimination for benefit (OR = 0.8, N = `r 4250` and AUC
+= 0.75), all methods performed similarly, on average (Figure
+\ref{fig:discrimination}). In the case of non-monotonic deviations,
+the constant effect model had much lower discriminative performance
+compared to the rest of the methods (median AUC of
+`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
+for the constant effects model compared to the best-performing RCS-3
+with
+`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
+The linear interaction model was most stable in compared to the other
+methods in terms of error variability. With increasing number of RCS
+knots, we observed decreasing median values and increasing variability
+of the c-for-benefit in all scenarios.
 
@@
 
-As for calibration, the linear interaction model had the lowest median
-ICI-for-benefit in the majority of scenarios except for the scenarios where no
-or moderate linear deviations from the base case were considered. In that case
-constant treatment effect models demonstrated optimal calibration, very
-comparable to the linear interaction model's performance, nonetheless (Figure
-\ref{fig:calibration}). With strong linear or quadratic deviations, the
-constant treatment effect model was poorly calibrated (Figure
-\ref{fig:calibration}; Panels C and D).
+As for calibration, the constant effects model outperformed all other
+models in the case of true constant treatment effects, but it was
+miscalibrated for all deviation scenarios (Figure
+\ref{fig:calibration}). The linear interaction model showed best or
+close to best calibration across all scenarios and only showed worse
+calibration compared to RCS-3 in the case of non-monotonic deviations
+and treatment-related harms. The adaptive approach was worse
+calibrated in scenarios with strong linear and non-monotonic
+deviations compared to the linear interaction model and RCS-3.
 

### 2021-11-10 code/PlotRmse.R
add bottom labels

@@ -317,8 +317,26 @@ right.grob <- grid::textGrob(
     rot = 270
 )
 
+bottom.left.grob <- grid::textGrob(
+  "Method",
+  just = "center",
+  gp = gpar(fontsize = 10)
+)
 
-res <- grid.arrange(arrangeGrob(pp, left = left.grob, right = right.grob))
+bottom.right.grob <- grid::textGrob(
+  "Baseline risk",
+  just = "center",
+  gp = gpar(fontsize = 10)
+)
+
+bottom.grob <- grid.arrange(
+  bottom.left.grob,
+  bottom.right.grob,
+  nrow = 1,
+  widths = c(2, 1)
+)
+
+res <- grid.arrange(arrangeGrob(pp, left = left.grob, right = right.grob, bottom = bottom.grob))
 
 fileName <- paste0(
   paste(

### 2021-11-10 code/CreateAdaptiveSelections.R
modularize

@@ -1,0 +1,39 @@
+#!/usr/bin/env Rscript
+
+# --------------------------------------------------
+# Description: 
+#   Creates a csv file with the models selected
+#   in each run of the simulation
+# Output: 
+#   - data/processed/adaptiveSelections.csv
+# Depends:
+#   - data/processed/adaptiveModel.csv
+# --------------------------------------------------
+
+library(tidyverse)
+
+readr::read_csv("data/processed/adaptiveModel.csv") %>%
+  group_by(scenarioId) %>% 
+  nest() %>% 
+  mutate(
+    counts = lapply(data, plyr::count),
+    n = lapply(data, nrow)
+  ) %>%
+  select(
+    scenarioId, counts, n
+  ) %>%
+  unnest(c(counts, n)) %>%
+  mutate(
+    perc = round(freq / n, 2),
+    model = case_when(
+      selectedAdaptiveModel == "RCS with 3 knots" ~ "rcs_3",
+      selectedAdaptiveModel == "RCS with 4 knots" ~ "rcs_4",
+      selectedAdaptiveModel == "RCS with 5 knots" ~ "rcs_5",
+      TRUE                                        ~ selectedAdaptiveModel
+    )
+  ) %>%
+  ungroup() %>%
+  select(scenarioId, model, perc) %>%
+  spread(model, perc) %>%
+  relocate(c(treatment, risk), .before = rcs_3) %>%
+  readr::write_csv("data/processed/adaptiveSelections.csv")

### 2021-11-10 code/CreateDistribution.R

@@ -1,0 +1,63 @@
+#!/usr/bin/env Rscript
+
+# -----------------------------------------------------------
+# Description:
+#   Creates a csv file with all median and the
+#   0.025 and 0.975 quantiles of the
+#   RMSE/AUC/calibration sampling distribution
+# Input:
+#   The metric we are interested in:
+#     - rmse
+#     - auc
+#     - calibration
+# Output:
+#   - data/processed/(rmse|auc|calibration)Distribution.csv
+# Depends:
+#   - code/helpers/GetMedians.R
+# -----------------------------------------------------------
+
+args <- commandArgs(trailingOnly = TRUE)
+args_metric <- as.character(args[1])
+saveFile <- paste0(
+  args_metric,
+  "Distribution.csv"
+)
+
+library(tidyverse)
+library(readr)
+
+source("code/helpers/GetMedians.R")
+
+read_csv("data/processed/rmse.csv") %>%
+  group_by(scenarioId) %>%
+  arrange(scenarioId) %>%
+  nest() %>%
+  mutate(
+    median = map(
+      data,
+      ~ getMedian(
+        data = .x
+      )
+    ),
+    lower = map(
+      data,
+      ~ getLower(
+        data = .x
+      )
+    ),
+    upper = map(
+      data,
+      ~ getUpper(
+        data = .x
+      )
+    )
+  ) %>%
+  ungroup() %>%
+  select(-data) %>%
+  unnest(c("median", "lower", "upper")) %>%
+  write_csv(
+    file.path(
+      "data/processed",
+      saveFile
+    )
+  )

### 2021-11-11 submission/manuscript.rmd
update discussion

@@
 
-Fourth, we used *restricted cubic splines* (RCS) to relax the linearity
-assumption on the effect of the linear predictor [@Harrell1988]. We considered
-splines with 3, 4 and 5 knots to compare models with different levels of
-flexibility.
+Fourth, we used *restricted cubic splines* (RCS) to relax the
+linearity assumption on the effect of the linear predictor
+[@Harrell1988]. We considered splines with 3 (RCS-3), 4 (RCS-4) and 5
+(RCS-5) knots to compare models with different levels of flexibility.
 
@@
 
-Increasing the sample size to `r 17000` favored the RCS-3 approach the
-most, as it achieved lowest or close to lowest RMSE across all
-scenarios (Figure \ref{fig:rmsesamplesize}). Especially in cases of
-strong quadratic and non-monotonic deviations RCS-3 had lower error
-rates
-(median RMSE
+Increasing the sample size to `r 17000` favored RCS-3 the most, as it
+achieved lowest or close to lowest RMSE across all scenarios (Figure
+\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
+non-monotonic deviations RCS-3 had lower error rates (median RMSE
 `r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
@@
 `r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
-for non-monotonic deviations with no treatment-related harms)
-compared to the linear interaction approach
-(median RMSE
+for non-monotonic deviations with no treatment-related harms) compared
+to the linear interaction approach (median RMSE
 `r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
@@
 performed similar to the best performing method in each scenario, on
-average, there was increased variability in error rates in the case of
-strong linear and non-monotonic deviations. In these scenarios, the adaptive approach often selected the constant treatment effect model
+average, had increased variability in error rates in the case of
+strong linear and non-monotonic deviations. In these scenarios, the
+adaptive approach often selected the constant treatment effect model
 (`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
@@
 `r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
-The linear interaction model was most stable in compared to the other
+The linear interaction model was the most stable compared to the other
 methods in terms of error variability. With increasing number of RCS
@@
 
-As for calibration, the constant effects model outperformed all other
-models in the case of true constant treatment effects, but it was
-miscalibrated for all deviation scenarios (Figure
+In terms of calibration for benefit, the constant effects model
+outperformed all other models in the case of true constant treatment
+effects, but was miscalibrated for all deviation scenarios (Figure
 \ref{fig:calibration}). The linear interaction model showed best or
@@
 
-The linear interaction model proved to be superior to alternative approaches
-when predicting risk-based treatment benefit under a wide range of scenarios. It
-generally had lower mean squared error, lower e-for-benefit and higher
-c-for-benefit with lower variability across simulation replications. Models
-including restricted cubic splines with 3 knots only outperformed the linear
-interaction model in the presence of strong quadratic deviations from a constant
-relative treatment effect.
+The linear interaction model displayed very good performance overall
+under many of the considered simulation scenarios. Especially in cases
+with smaller sample sizes and moderately performing baseline risk
+prediction models it had lower RMSE, was better calibrated for benefit
+and had better discrimination for benefit, even in scenarios with
+strong quadratic deviations. However, in scenarios with true
+non-monotonic deviations, the linear interaction model was
+outperformed by RCS-3, especially in the presence of true
+treatment-related harms. Increasing the sample size or the prediction
+model's discrimination favored RCS-3 which had better or very
+comparable performance to the linear interaction model, but was more
+robust to non-monotonic deviations and the presence of
+treatment-related harms.
 
-Models including restricted cubic splines with 4 or 5 knots proved to be too
-flexible, as indicated by higher RMSE, increased variability of discrimination
-for benefit and worse calibration of benefit predictions. Even with larger
-sample sizes and strong quadratic deviations from the base case scenario of
-constant relative treatment effects, these more flexible restricted cubic
-splines did not outperform the simpler RCS with 3 knots. These approaches may
-only be helpful if we expect more extreme patterns of heterogeneous treatment
-effects compared to the quadratic deviations considered here.
+RCS-4 and RCS-5 proved to be too flexible, as indicated by higher
+RMSE, increased variability of discrimination for benefit and worse
+calibration of benefit predictions. Even with larger sample sizes and
+strong quadratic or non-monotonic deviations from the base case
+scenario of constant relative treatment effects, these more flexible
+restricted cubic splines did not outperform the simpler RCS-3 These
+approaches may only be helpful if we expect more extreme patterns of
+heterogeneous treatment effects compared to the quadratic deviations
+considered here.
 
@@
 
-The adaptive approach had adequate performance, following closely the
-performance of the "true" model in most scenarios. However, with smaller sample
-sizes it tended to "miss" the treatment-risk interactions and selected simpler
-models (Supplementary Table S7). This resulted in increased RMSE variability in
-these scenarios, especially in the case of true strong linear deviations from
-the base case scenario. This tendency was mitigated when the discriminative
-ability of the prediction model was higher. Therefore, in the case of smaller
-sample sizes and/or poorly discriminating prediction models, the simpler linear
-interaction model is a safer choice for predicting absolute benefits.
+The adaptive approach had adequate performance, following closely on
+average the performance of the "true" model in most
+scenarios. However, with smaller sample sizes it tended to "miss" the
+treatment-risk interactions and selected simpler models (Supplementary
+Table S7). This resulted in increased RMSE variability in these
+scenarios, especially in the case of true strong linear or
+non-monotonic deviations from the base case scenario.  Therefore, in
+the case of smaller sample sizes the simpler linear interaction model
+is a safer choice for predicting absolute benefits.
 
-Risk-based approaches to predictive HTE estimate treatment benefit as a function
-of baseline risk. A limitation of our study is that we assumed treatment benefit
-to be a function of baseline risk in the majority of the simulation scenarios.
-Nevertheless, our main conclusions did not change when we generated individual
-treatment-covariate interactions (supplementary table/figure x). Future
-simulation studies could explore the effect of more extensive deviations from
+Risk-based approaches to predictive HTE estimate treatment benefit as
+a function of baseline risk. A limitation of our study is that we
+assumed treatment benefit to be a function of baseline risk in the
+majority of the simulation scenarios. We attempted to address that by
+introducing constant moderate and strong treatment-related harms,
+applied on the absolute scale. Also, we considered a small number of
+scenarios with true treatment-covariate interactions, in which our
+main conclusions remained the same (Supplement, XX). Future simulation
+studies could explore the effect of more extensive deviations from
 risk-based treatment effects.
 
-Recent years have seen an increased interest in predictive HTE approaches
-focusing on individualized benefit predictions. In our simulations we only
-focused on risk-based methods, using baseline risk as a reference in a two-stage
-approach to individualizing benefit predictions. However, there is a plethora of
-different methods, ranging from treatment effect modeling to tree-based
-approaches available in more recent literature [@Athey2019; @Lu2018;
-@Wager2018]. Simulations are also needed to assess relative performance and
-define the settings where these break down or outperform each other.
+Recent years have seen an increased interest in predictive HTE
+approaches focusing on individualized benefit predictions. In our
+simulations we only focused on risk-based methods, using baseline risk
+as a reference in a two-stage approach to individualizing benefit
+predictions. However, there is a plethora of different methods,
+ranging from treatment effect modeling to tree-based approaches
+available in more recent literature [@Athey2019; @Lu2018;
+@Wager2018]. Simulations are also needed to assess relative
+performance and define the settings where these break down or
+outperform each other.
 
-In conclusion, when comparing different risk-based approaches to predicting 
-individualized treatment benefit, a model including a linear treatment 
-interaction with the prognostic index performed best in a wide range of 
-scenarios. More flexible models with restricted cubic splines required larger
-sample sizes and higher AUC of the prognostic index to outperform the linear
-interaction model. An adaptive approach, selecting the model with the optimal 
-AIC, had comparable performance to the best performing approach in most of the 
-scenarios.
+In conclusion, the best option for predicting individualized treatment
+benefit using a risk-based approach depends on the setting. With
+smaller sample sizes and/or moderately performing risk prediction
+models the linear interaction approach is a viable option. When those
+constraints are not present or when we anticipate non-negligible
+treatment-related harms, RCS-3 is a better option in terms of error
+rates, discrimination and calibration for benefit. With larger sample
+size, an adaptive approach based on AIC can also be considered as a
+more automated alternative.
 

### 2021-11-11 code/CreateAdaptiveSelections.R

@@ -1,6 +1,6 @@
 #!/usr/bin/env Rscript
 
-# --------------------------------------------------
+# ===================================================
 # Description: 
 #   Creates a csv file with the models selected
 #   in each run of the simulation
@@ -8,7 +8,7 @@
 #   - data/processed/adaptiveSelections.csv
 # Depends:
 #   - data/processed/adaptiveModel.csv
-# --------------------------------------------------
+# ===================================================
 
 library(tidyverse)
 

### 2021-11-11 code/CreateDistribution.R

@@ -1,6 +1,6 @@
 #!/usr/bin/env Rscript
 
-# -----------------------------------------------------------
+# =============================================================
 # Description:
 #   Creates a csv file with all median and the
 #   0.025 and 0.975 quantiles of the
@@ -14,7 +14,7 @@
 #   - data/processed/(rmse|auc|calibration)Distribution.csv
 # Depends:
 #   - code/helpers/GetMedians.R
-# -----------------------------------------------------------
+# =============================================================
 
 args <- commandArgs(trailingOnly = TRUE)
 args_metric <- as.character(args[1])

### 2021-11-11 code/PlotRmse.R

@@ -1,6 +1,6 @@
 #!/usr/bin/env Rscript
 
-# --------------------------------------------------
+# ===================================================
 # Description:
 #   Generates the RMSE plots for the base case
 # Input:
@@ -15,7 +15,7 @@
 #   - code/helpers/CreateManuscriptPlots.R
 #   - code/helpers/PlotResult.R
 #   - code/helpers/Absolute.R
-# --------------------------------------------------
+# ===================================================
 
 args <- commandArgs(trailingOnly = TRUE)
 args_sampleSize <- as.numeric(args[1])

### 2021-11-12 submission/manuscript.rmd
update abstract

@@
 abstract: |
- **Objective:** To compare different risk-based methods predicting individualized
- treatment effects in RCT simulations. **Study Design and Setting:** We simulated
- data using diverse assumptions for a baseline prognostic index of risk (PI) and
- the shape of its interaction with treatment (none, linear or quadratic). In
- each sample we predicted absolute benefit using: models with the PI and a
- constant relative treatment effect; models including an interaction of treatment
- with the PI; stratification in quarters of the PI; nonlinear transformations of
- the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
- using Akaike’s Information Criterion. We evaluated predictive performance using
- root mean squared error and measures of discrimination and calibration for
- benefit. Starting from a base case scenario (sample size 4,250, treatment odds
- ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
- strength, and the Ti's discriminative ability. **Results:** Models including a
- PI by treatment interaction performed better under most simulation settings.
- Flexible models required larger sample sizes and higher AUC of the PI to
- outperform the linear interaction model. The adaptive approach tended to select
- simpler models with smaller sample sizes and/or worse prediction performance.
- **Conclusion:** Under most circumstances, a model with a linear interaction of
- the PI with treatment is the optimal risk-based approach to predicting absolute
- treatment benefit.
+ **Objective:** To compare different risk-based methods predicting
+ individualized treatment effects in RCTs. **Study Design and
+ Setting:** We simulated data using diverse assumptions for a baseline
+ prognostic index of risk (PI), the shape of its interaction with
+ treatment (none, linear, quadratic or non-monotonic) and the magnitude
+ of treatment-related harms. In each sample we predicted absolute
+ benefit using: models with a constant relative treatment effect;
+ stratification in quarters of the PI; models including a linear
+ interaction of treatment with the PI; models including an interaction
+ of treatment with a restricted cubic spline (RCS) transformation of
+ the PI; an adaptive approach using Akaike’s Information Criterion. We
+ evaluated predictive performance using root mean squared error and
+ measures of discrimination and calibration for benefit. Starting from
+ a base scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
+ the PI 0.75), we varied the sample size, the treatment effect size,
+ and the PI's AUC. **Results:** Models including a PI by treatment
+ interaction performed best with smaller sample sizes and/or lower AUC
+ of the PI. Otherwise, RCS (3 knots) models proved more robust for the
+ majority of the deviation settings. The adaptive approach was unstable
+ with smaller sample sizes. **Conclusion:** Depending on the setting,
+ the linear interaction or the RCS (3 knots) model should be preferred.
+
 authors:
@@
 
+The results from all individual scenarios can be explored online at
+[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).
+
 ```{r calibration, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
@@
 
-We demonstrate the different methods for individualizing treatment benefits
-using data from `r nrow(gusto)` patients with an acute myocardial infarction
-(MI) included in the GUSTO-I trial.
+We demonstrate the different methods for individualizing treatment
+benefits using data from `r nrow(gusto)` patients with an acute
+myocardial infarction (MI) included in the GUSTO-I trial.
 `r treatmentArms %>% filter(tpa == 1) %>% select(n)` patients were randomized to
@@
 `r treatmentArms %>% filter(tpa == 0) %>% select(n)` were randomized to
-streptokinase. The outcome of interest was 30-day mortality, recorded for all
-patients.
+streptokinase. The outcome of interest was 30-day mortality, recorded
+for all patients.
 
-In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
-logistic regression model with 6 baseline covariates, i.e. age, Killip class,
-systolic blood pressure, heart rate, an indicator of previous MI, and the
-location of MI, to predict 30-day mortality risk. A constant effect of treatment
-was included in the model. When deriving risk predictions for individuals we set
-the treatment indicator to 0. More information on model development can be found
-in the supplement.
+In line with previous analyses [@Califf1997; @Steyerberg2000], we
+fitted a logistic regression model with 6 baseline covariates,
+i.e. age, Killip class, systolic blood pressure, heart rate, an
+indicator of previous MI, and the location of MI, to predict 30-day
+mortality risk. A constant effect of treatment was included in the
+model. When deriving risk predictions for individuals we set the
+treatment indicator to 0. More information on model development can be
+found in the supplement.
 
@@
 
-We used the risk linear predictor to fit the proposed methods under study for
-individualizing absolute benefit predictions. All methods predicted increasing
-benefits for patients with higher baseline risk predictions, but the fitted
-patterns were clearly different. The adaptive approach selected the model with
-RCS smoothing with 4 knots. However, for very low baseline risk this model
-predicted decreasing benefit with increasing risk may be somewhat too flexible.
-The more robust models, the linear interaction model or the model with RCS
-smoothing (3 knots), gave very similar benefit predictions, followed the
-evolution of the stratified estimates very closely and may therefore be
-preferable for use in clinical practice. The linear interaction model
-had somewhat lower AIC compared to the model with RCS smoothing (3 knots), slightly better
-cross-validated discrimination (c-for-benefit
+We used the risk linear predictor to fit the proposed methods under
+study for individualizing absolute benefit predictions. All methods
+predicted increasing benefits for patients with higher baseline risk
+predictions, but the fitted patterns were clearly different. The
+adaptive approach selected the model with RCS smoothing with 4
+knots. However, for very low baseline risk this model predicted
+decreasing benefit with increasing risk may be somewhat too flexible.
+The more robust models, the linear interaction model or the model with
+RCS smoothing (3 knots), gave very similar benefit predictions,
+followed the evolution of the stratified estimates very closely and
+may therefore be preferable for use in clinical practice. The linear
+interaction model had somewhat lower AIC compared to the model with
+RCS smoothing (3 knots), slightly better cross-validated
+discrimination (c-for-benefit
 `r round(discriminationPerformance[2, ], 3)` vs
@@
 
-The constant treatment effect model, despite having adequate performance in the
-presence of weak treatment effect heterogeneity on the relative scale, quickly
-broke down with stronger deviations from constant relative treatment effects. In
-these cases, the stratified approach generally had lower error rates compared to
-the constant treatment effect model. Stepwise treatment benefit estimates are
-very useful for demonstrating treatment effect heterogeneity--because
-estimating treatment effect requires groups of patients rather than individual
-patients--but are not helpful for making individualized absolute benefit
-predictions.
+The constant treatment effect model, despite having adequate
+performance in the presence of weak treatment effect heterogeneity on
+the relative scale, quickly broke down with stronger deviations from
+constant relative treatment effects. In these cases, the stratified
+approach generally had lower error rates compared to the constant
+treatment effect model. Stepwise treatment benefit estimates are very
+useful for demonstrating treatment effect heterogeneity--because
+estimating treatment effect requires groups of patients rather than
+individual patients--but are not helpful for making individualized
+absolute benefit predictions.
 
-Increasing the discriminative ability of the risk model--by increasing the
-predictor coefficients of the true risk model--reduced RMSE for all methods.
-This increase in discriminative ability translates in higher variability of
-predicted risks, which, in turn, allows the considered methods to better capture
-absolute treatment benefits. As a consequence, the increase in discriminative
-ability of the risk model also led to higher values of c-for-benefit. Even
-though risk model performance is very important for the ability of risk-based
-methods to predict treatment benefit, prediction model development was outside
+Increasing the discriminative ability of the risk model--by increasing
+the predictor coefficients of the true risk model--reduced RMSE for
+all methods.  This increase in discriminative ability translates in
+higher variability of predicted risks, which, in turn, allows the
+considered methods to better capture absolute treatment benefits. As a
+consequence, the increase in discriminative ability of the risk model
+also led to higher values of c-for-benefit. Even though risk model
+performance is very important for the ability of risk-based methods to
+predict treatment benefit, prediction model development was outside
 the scope of this work and has already been studied extensively

### 2021-12-08 submission/manuscript.tex
new version

@@
 \begin{abstract}
-data using diverse assumptions for a baseline prognostic index of risk
-(PI) and the shape of its intereraction with treatment (none, linear or
-quadratic). In constant relative treatment effect; models including an
-interaction of treatment with the PI; stratification in quarters of the
-PI; nonlinear transformations of the PI (restricted cubic splines with
-3, 4 and 5 knots); an adaptive approach using Akaike's Information
-Criterion. We evaluated predictive performance using root mean squared
-error and measures of discrimination and calibration for benefit.
-Starting from a base case scenario (sample size 4,250, treatment odds
-ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment
-effect PI by treatment interaction performed better under most
-simulation settings. Flexible models required larger sample sizes and
-higher AUC of the PI to outperform the linear interaction model. The
-adaptive approach tended to select simpler models with smaller sample
-sizes and/or worse prediction performance. the PI with treatment is the
-optimal risk-based approach to predicting absolute treatment benefit.
 \textbf{Objective:} To compare different risk-based methods predicting
-individualized treatment effects in RCT simulations. \textbf{Study
-Design and Setting:} We simulated each sample we predicted absolute
-benefit using: models with the PI and a strength, and the PI's
-discriminative ability. \textbf{Results:} Models including a
-\textbf{Conclusion:} Under most circumstances, a model with a linear
-interaction of
+individualized treatment effects in RCTs. \textbf{Study Design and
+Setting:} We simulated data using diverse assumptions for a baseline
+prognostic index of risk (PI), the shape of its interaction with
+treatment (none, linear, quadratic or non-monotonic) and the magnitude
+of treatment-related harms. In each sample we predicted absolute benefit
+using: models with a constant relative treatment effect; stratification
+in quarters of the PI; models including a linear interaction of
+treatment with the PI; models including an interaction of treatment with
+a restricted cubic spline (RCS) transformation of the PI; an adaptive
+approach using Akaike's Information Criterion. We evaluated predictive
+performance using root mean squared error and measures of discrimination
+and calibration for benefit. Starting from a base scenario (sample size
+4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the
+sample size, the treatment effect size, and the PI's AUC.
+\textbf{Results:} Models including a PI by treatment interaction
+performed best with smaller sample sizes and/or lower AUC of the PI.
+Otherwise, RCS (3 knots) models proved more robust for the majority of
+the deviation settings. The adaptive approach was unstable with smaller
+sample sizes. \textbf{Conclusion:} Depending on the setting, the linear
+interaction or the RCS (3 knots) model should be preferred.
 \end{abstract}
@@
 
-For each patient we generated 8 baseline covariates here
+For each patient we generated 8 baseline covariates
 \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
@@
 
-Combining all these settings resulted in a simulation study of 486
+Combining all these settings resulted in a simulation study of 648
 scenarios (exact settings in the supplementary material).
@@
 Second, we considered a model which assumes \emph{constant relative
-treatment effect} (constant OR). Hence, absolute benefit is predicted
-from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).
+treatment effect} (constant odds ratio). Hence, absolute benefit is
+predicted from
+\(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).
 
@@
 linearity assumption on the effect of the linear predictor {[}7{]}. We
-considered splines with 3, 4 and 5 knots to compare models with
-different levels of flexibility.
+considered splines with 3 (RCS-3), 4 (RCS-4) and 5 (RCS-5) knots to
+compare models with different levels of flexibility.
 
@@
 
-\[\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2\]
+\[\text{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2}\]
 We compared the discriminative ability of the methods under study using
@@
 
-The model including a constant relative treatment effect had the lowest
-median RMSE in scenarios with a true constant relative treatment effect
-(OR = 0.8, N = 4,250 and AUC = 0.75) or with moderate relative
-deviations (Figure \ref{fig:rmsebase}; Panel A). However, when we
-considered strong linear and quadratic deviations from a constant
-relative treatment effect the linear interaction model performed best
-(Figure \ref{fig:rmsebase}; Panels B and C). Only in the case of strong
-quadratic deviations models including RCS (3 knots) performed equally
-well to the linear interaction method. Increasing the number of knots in
-RCS resulted in higher RMSE across all scenarios. The adaptive approach
-selected a linear interaction model 35\% of the time in the case of true
-strong linear deviations, while the majority of the time (56\%) it
-selected a simpler constant treatment effect model (Supplementary Table
-S7). Similarly, in the case of true strong quadratic deviations, the
-adaptive approach selected the RCS (3 knots) model 29\% of the time,
-while the simpler linear interaction model was selected 68\% of the
-time.
+The linear interaction model outperformed all RCS methods in terms of
+RMSE in scenarios with true constant relative treatment effect (OR =
+0.8, N = 4,250 and AUC = 0.75), strong linear and even strong quadratic
+deviations (Figure \ref{fig:rmsebase}). However, with non-monotonic
+deviations errors of the linear interaction model increased
+substantially, especially in the presence of treatment-related harms. In
+these scenarios, RCS-3 proved to be quite robust outperforming all other
+methods. The constant treatment effect approach had overall best
+performance under true constant treatment effect settings, but was
+sensitive to all considered deviations, resulting in increased error
+rates. Finally, the adaptive approach had comparable performance to the
+best-performing method in each scenario. However, we observed increased
+error variability in the case of linear and non-monotonic deviations,
+especially for moderate or strong treatment-related harms.
+
+In the case of true constant relative treatment effects, the adaptive
+approach selected the constnant effect model the majority of the time,
+even in the presence of treatment-related harms (96\% of the time with
+no harms to 88\% with strong harms). With strong linear or quadratic
+deviations and increasing treatment-related harms the adaptive approach
+increasingly favored the linear interaction model. In the case of true
+non-monotonic deviations, we observed increasing selection frequencies
+for RCS-3 with increasing treatment harms (from 32\% with no harms to
+53\% with strong harms), whereas selection frequencies for the linear
+interaction model dropped from 40\% with no harms to 7\% with strong
+harms. Finally, in the case of strong linear and non-monotonic
+deviations selection frequencies of the constant treatment effect model
+remained high despite increasing treatment harms (Supplement, Table XX).
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsebase-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated superpopulation of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsebase-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
 The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
@@
 
-When we increased the sample size (N = 17,000), the model including a
-constant relative treatment effect had the lowest RMSE under the
-assumption of true constant relative treatment effects (Figure
-\ref{fig:rmsesamplesize}; Panel A). However, when introducing moderate
-and strong linear deviations the linear interaction model outperformed
-the constant treatment effect model (Figure \ref{fig:rmsesamplesize};
-Panels B and C). Furthermore, the linear interaction model was
-outperformed by the more flexible RCS models (3 knots) in the case of
-strong quadratic deviations. Again, the increased flexibility of RCS
-smoothing with higher number of knots resulted in overfitting and worse
-predictive accuracy (Figure \ref{fig:rmsesamplesize}; Panel D).
+Increasing the sample size to 17,000 favored RCS-3 the most, as it
+achieved lowest or close to lowest RMSE across all scenarios (Figure
+\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
+non-monotonic deviations RCS-3 had lower error rates (median RMSE 0.011
+for strong quadratic deviations and 0.010 for non-monotonic deviations
+with no treatment-related harms) compared to the linear interaction
+approach (median RMSE 0.013 and 0.014, respectively), regardless of
+treatment-related harms strength. The issues with the large error
+variability of the adaptive approach improved with larger sample sizes.
+
+The adaptive approach tended to increasingly favor smoother methods
+(especially RCS-3) with increasing treatment-related harms (see
+Supplement, Table XX). However, in the case of true strong quadratic
+deviations the opposite was observed: selection frequency of the linear
+interaction model increased from 31\% (no harms) to 50\% (strong harms)
+whereas for RCS-3 decreased from 52\% (no harms) to 34\% (strong harms).
 
@@
 
-When we increased the AUC of the true prediction model to 0.85, models
-including RCS smoothing had the lowest RMSE in the presence of strong
-quadratic deviations from the base case of true constant relative
-treatment effects (Figure \ref{fig:rmseauc}; Panel D). However, with
-milder deviations, the linear interaction model had the lowest RMSE with
-the RCS smoothing methods (3 knots) being a close second (Figure
-\ref{fig:rmseauc}; Panels B and C). Increasing the number of knots of
-RCS smoothing resulted in increased RMSE, which was less pronounced in
-the case of strong quadratic deviations. With increasing prediction AUC
-we saw an increased tendency of the adaptive approach to select more
-complex models compared to the base case scenario. In the case of strong
-linear deviations, the linear interaction model was selected the
-majority of the time (63\%), while the simpler constant treatment effect
-model was selected 26\% of the time (Supplementart Table S7). However,
-the increased RMSE of the constant treatment effect model inflated the
-the RMSE of the linear interaction model. In the case of strong
-quadratic deviations the RCS (3 knots) model was selected 84\% of the
-time.
+When we increased the AUC of the true prediction model to 0.85 (OR = 0.8
+and N = 4,250) the constant effect model outperformed all other methods
+in the case of true constant treatment effects, but proved to be the
+least robust to deviations. Again, RCS-3 had the lowest error rates in
+the case of strong quadratic or monotonic deviations and very comparable
+performance to the best-performing linear interaction model in the case
+of strong linear deviations (median RMSE 0.016 for RCS-3 compared to
+0.014 for the linear interaction model). The adaptive approach, though
+it performed similar to the best performing method in each scenario, on
+average, had increased variability in error rates in the case of strong
+linear and non-monotonic deviations. In these scenarios, the adaptive
+approach often selected the constant treatment effect model (23\% and
+25\% in the strong linear and non-monotonic deviation scenarios without
+treatment-related harms, respectively).
 
@@
 
-The constant effects model, the linear interaction model and models with
-RCS smoothing (3 knots) had the highest median c-for-benefit in the base
-case scenario and the scenarios where linear and quadratic deviations
-were considered. The constant treatment effect model and the linear
-interaction model tended to present much lower variability compared to
-all other approaches (Figure \ref{fig:discrimination}). With increasing
-number of RCS knots, we observed decreasing median values and increasing
-variability of the c-for-benefit in all scenarios.
+In terms of discrimination for benefit (OR = 0.8, N = 4,250 and AUC =
+0.75), all methods performed similarly, on average (Figure
+\ref{fig:discrimination}). In the case of non-monotonic deviations, the
+constant effect model had much lower discriminative performance compared
+to the rest of the methods (median AUC of 0.04 for the constant effects
+model compared to the best-performing RCS-3 with 0.02). The linear
+interaction model was the most stable compared to the other methods in
+terms of error variability. With increasing number of RCS knots, we
+observed decreasing median values and increasing variability of the
+c-for-benefit in all scenarios.
 
@@
 
-As for calibration, the linear interaction model had the lowest median
-ICI-for-benefit in the majority of scenarios except for the scenarios
-where no or moderate linear deviations from the base case were
-considered. In that case constant treatment effect models demonstrated
-optimal calibration, very comparable to the linear interaction model's
-performance, nonetheless (Figure \ref{fig:calibration}). With strong
-linear or quadratic deviations, the constant treatment effect model was
-poorly calibrated (Figure \ref{fig:calibration}; Panels C and D).
+In terms of calibration for benefit, the constant effects model
+outperformed all other models in the case of true constant treatment
+effects, but was miscalibrated for all deviation scenarios (Figure
+\ref{fig:calibration}). The linear interaction model showed best or
+close to best calibration across all scenarios and only showed worse
+calibration compared to RCS-3 in the case of non-monotonic deviations
+and treatment-related harms. The adaptive approach was worse calibrated
+in scenarios with strong linear and non-monotonic deviations compared to
+the linear interaction model and RCS-3.
+
+The results from all individual scenarios can be explored online at
+\url{https://arekkas.shinyapps.io/simulation_viewer/}.
 
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/gusto} \caption{Individualized absolute benefit predictions based on baseline risk when using a constantn treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/gusto} \caption{Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.}\label{fig:gusto}
@@
 
-The linear interaction model proved to be superior to alternative
-approaches when predicting risk-based treatment benefit under a wide
-range of scenarios. It generally had lower mean squared error, lower
-e-for-benefit and higher c-for-benefit with lower variability across
-simulation replications. Models including restricted cubic splines with
-3 knots only outperformed the linear interaction model in the presence
-of strong quadratic deviations from a constant relative treatment
-effect.
+The linear interaction model displayed very good performance overall
+under many of the considered simulation scenarios. Especially in cases
+with smaller sample sizes and moderately performing baseline risk
+prediction models it had lower RMSE, was better calibrated for benefit
+and had better discrimination for benefit, even in scenarios with strong
+quadratic deviations. However, in scenarios with true non-monotonic
+deviations, the linear interaction model was outperformed by RCS-3,
+especially in the presence of true treatment-related harms. Increasing
+the sample size or the prediction model's discrimination favored RCS-3
+which had better or very comparable performance to the linear
+interaction model, but was more robust to non-monotonic deviations and
+the presence of treatment-related harms.
 
-Models including restricted cubic splines with 4 or 5 knots proved to be
-too flexible, as indicated by higher RMSE, increased variability of
-discrimination for benefit and worse calibration of benefit predictions.
-Even with larger sample sizes and strong quadratic deviations from the
-base case scenario of constant relative treatment effects, these more
-flexible restricted cubic splines did not outperform the simpler RCS
-with 3 knots. These approaches may only be helpful if we expect more
-extreme patterns of heterogeneous treatment effects compared to the
-quadratic deviations considered here.
+RCS-4 and RCS-5 proved to be too flexible, as indicated by higher RMSE,
+increased variability of discrimination for benefit and worse
+calibration of benefit predictions. Even with larger sample sizes and
+strong quadratic or non-monotonic deviations from the base case scenario
+of constant relative treatment effects, these more flexible restricted
+cubic splines did not outperform the simpler RCS-3 These approaches may
+only be helpful if we expect more extreme patterns of heterogeneous
+treatment effects compared to the quadratic deviations considered here.
 
@@
 
-The adaptive approach had adequate performance, following closely the
-performance of the ``true'' model in most scenarios. However, with
-smaller sample sizes it tended to ``miss'' the treatment-risk
-interactions and selected simpler models (Supplementary Table S7). This
-resulted in increased RMSE variability in these scenarios, especially in
-the case of true strong linear deviations from the base case scenario.
-This tendency was mitigated when the discriminative ability of the
-prediction model was higher. Therefore, in the case of smaller sample
-sizes and/or poorly discriminating prediction models, the simpler linear
-interaction model is a safer choice for predicting absolute benefits.
+The adaptive approach had adequate performance, following closely on
+average the performance of the ``true'' model in most scenarios.
+However, with smaller sample sizes it tended to ``miss'' the
+treatment-risk interactions and selected simpler models (Supplementary
+Table S7). This resulted in increased RMSE variability in these
+scenarios, especially in the case of true strong linear or non-monotonic
+deviations from the base case scenario. Therefore, in the case of
+smaller sample sizes the simpler linear interaction model is a safer
+choice for predicting absolute benefits.
 
@@
 treatment benefit to be a function of baseline risk in the majority of
-the simulation scenarios. Nevertheless, our main conclusions did not
-change when we generated individual treatment-covariate interactions
-(supplementary table/figure x). Future simulation studies could explore
-the effect of more extensive deviations from risk-based treatment
-effects.
+the simulation scenarios. We attempted to address that by introducing
+constant moderate and strong treatment-related harms, applied on the
+absolute scale. Also, we considered a small number of scenarios with
+true treatment-covariate interactions, in which our main conclusions
+remained the same (Supplement, XX). Future simulation studies could
+explore the effect of more extensive deviations from risk-based
+treatment effects.
 
@@
 
-In conclusion, when comparing different risk-based approaches to
-predicting individualized treatment benefit, a model including a linear
-treatment interaction with the prognostic index performed best in a wide
-range of scenarios. More flexible models with restricted cubic splines
-required larger sample sizes and higher AUC of the prognostic index to
-outperform the linear interaction model. An adaptive approach, selecting
-the model with the optimal AIC, had comparable performance to the best
-performing approach in most of the scenarios.
+In conclusion, the best option for predicting individualized treatment
+benefit using a risk-based approach depends on the setting. With smaller
+sample sizes and/or moderately performing risk prediction models the
+linear interaction approach is a viable option. When those constraints
+are not present or when we anticipate non-negligible treatment-related
+harms, RCS-3 is a better option in terms of error rates, discrimination
+and calibration for benefit. With larger sample size, an adaptive
+approach based on AIC can also be considered as a more automated
+alternative.
 

### 2021-12-08 submission/supplement.rmd
new version

--- 2021-07-12 submission/supplement.rmd
@@
 ## Base-case scenario
-The base-case scenario assumes a constant odds ratio of $0.8$ in favor of treatment.
-The simulated datasets are of size $n=4250$, where treatment is allocated at random using a 50/50 split (80% power for the detection of an unadjusted OR of 0.8, assuming an event rate of 20% in the untreated arm).
-Outcome incidence in the untreated population is set at $20\%$.
-For the development of the prediction model we use the model defined in \@ref(eq:p0) including a constant treatment effect.
-When doing predictions, $t_x$ is set to $0$.
-The value of the true $\bm{\beta}$ is such that the above prediction model has an AUC of $0.75$.
+The base-case scenario assumes a constant odds ratio of $0.8$ in favor
+of treatment. The simulated datasets are of size $n=4250$, where
+treatment is allocated at random using a 50/50 split (80% power for
+the detection of an unadjusted OR of 0.8, assuming an event rate of
+20% in the untreated arm). Outcome incidence in the untreated
+population is set at $20\%$. For the development of the prediction
+model we use the model defined in \@ref(eq:p0) including a constant
+treatment effect. When doing predictions, $t_x$ is set to $0$. The
+value of the true $\bm{\beta}$ is such that the above prediction model
+has an AUC of $0.75$.
 
-The previously defined targets are achieved when $\bm{\beta}=(-2.08, 0.49,\dots,0.49)^t$. For the derivations in the treatment arm we use $\bm{\gamma}=(\log(0.8), 1, 0)^t$.
+The previously defined targets are achieved when $\bm{\beta}=(-2.08,
+0.49,\dots,0.49)^t$. For the derivations in the treatment arm we use
+$\bm{\gamma}=(\log(0.8), 1, 0)^t$.
 
 ## Deviations from base-case
-We deviate from the base-case scenario in two ways.
-First, we alter the overall target settings of sample size, overall treatment effect and prediction model AUC.
-In a second stage, we consider settings that violate the assumption of a constant relative treatment effect, using a model-based approach.
+We deviate from the base-case scenario in two ways. First, we alter
+the overall target settings of sample size, overall treatment effect
+and prediction model AUC. In a second stage, we consider settings that
+violate the assumption of a constant relative treatment effect, using
+a model-based approach.
 
@@
 
-For the second part linear and quadratic deviations from the assumption of
-constant relative effect are considered. We also consider different intensity
-levels of these deviations. The settings for these deviations are defined in
-Table \@ref(tab:tab1) and result in the effects of Figure
-\@ref(fig:fig1).
+For the second part linear and quadratic deviations from the
+assumption of constant relative effect are considered. We also
+consider different intensity levels of these deviations. The settings
+for these deviations are defined in Table \@ref(tab:tab1) and result
+in the effects of Figure \@ref(fig:fig1).
 
@@
 
-Finally, we consider 3 additional scenarios of interaction of individual
-covariates with treatment. These scenarios include a 4 weak interactions
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2 weak and 2 strong interactions (Table \@ref(tab:tab2)).
+Finally, we consider 3 additional scenarios of interaction of
+individual covariates with treatment. These scenarios include a 4 weak
+interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong
+interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2
+weak and 2 strong interactions (Table \@ref(tab:tab2)).
 
@@
 ### Risk stratification
-Derive a prediction model using the same approach as above and divide the
-population in equally sized risk-based subgroups. Estimate subgroup-specific
-absolute benefit from the observed absolute differences. Subject-specific
-benefit predictions are made by attributing to individuals their corresponding
-subgroup-specific estimate.
+Derive a prediction model using the same approach as above and divide
+the population in equally sized risk-based subgroups. Estimate
+subgroup-specific absolute benefit from the observed absolute
+differences. Subject-specific benefit predictions are made by
+attributing to individuals their corresponding subgroup-specific
+estimate.
 
 ### Constant treatment effect
-Assuming a constant relative treatment effect, fit the adjusted model in
-\@ref(eq:risk). Then, an estimate of absolute benefit can be derived from$$\hat{f}_{\text{benefit}}(lp\given \bm{x}, \hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
+Assuming a constant relative treatment effect, fit the adjusted model
+in \@ref(eq:risk). Then, an estimate of absolute benefit can be
+derived from$$\hat{f}_{\text{benefit}}(lp\given \bm{x},
+\hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
 
 ### Linear interaction
-The assumption of constant relative treatment effect is relaxed modeling a
-linear interaction of treatment with the risk linear predictor:
-$$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} = g\big(lp+(\gamma_0+\gamma_1lp)t_x\big)$$
-We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) = g(lp) - g\big(\gamma_0+(1+\gamma_1)lp\big)$$
+The assumption of constant relative treatment effect is relaxed
+modeling a linear interaction of treatment with the risk linear
+predictor: $$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} =
+g\big(lp+(\gamma_0+\gamma_1lp)t_x\big)$$ We predict absolute benefit
+from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) =
+g(lp) - g\big(\gamma_0+(1+\gamma_1)lp\big)$$
 
@@
 
-```{r tab1, echo=FALSE, eval=TRUE, warning=FALSE}
+```{r tab1, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
  asd <- read_csv(
-  here("data/processed/analysisIds.csv"),
-  col_types = "iffiiddddddddddddd"
+  here("data/processed/analysisIds.csv")
 ) %>%
   select(-"type")
-
 kableExtra::kable(
@@
     "Scenario",
-    "Effect",
+    "Base",
     "N",
     "AUC",
+    "Treatment-related harm",
     paste0("b", 0:8),
     paste0("g", 0:2),
-    "c"
+    "c",
+    "Average benefit"
   ),
@@
 ) %>%
-  kableExtra::group_rows(
-    group_label = "Constant treatment effect",
-    start_row   = 10,
-    italic      = TRUE,
-    end_row     = 27
-  ) %>%
-  kableExtra::group_rows(
-    group_label = "Linear deviation",
-    start_row   = 28,
-    italic      = TRUE,
-    end_row     = 45
-  ) %>%
-  kableExtra::group_rows(
-    group_label = "Quadratic deviation",
-    start_row   = 46,
-    italic      = TRUE,
-    end_row     = 63
-  ) %>%
   kableExtra::landscape() %>%
@@
     c(
-      "Analysis ID"           = 4,
+      "Analysis ID"           = 5,
       "Baseline risk"         = 9,
@@
       )
-  )
+  ) %>%
+  kableExtra::kable_styling(font_size = 7)
 ```
-
-\newpage
-```{r tab2, echo=FALSE, eval=TRUE, warning=FALSE}
-asd <- read_csv(
-  here("data/processed/analysisIdsInteractions.csv"),
-  col_types = "iffiidddddddddddddd"
-) %>%
-  select(-c("type", "g0"))
-
-  knitr::kable(
-    asd,
-    format      = "latex",
-    longtable   = TRUE,
-    align       = "r",
-    booktabs    = TRUE,
-    col.names   = c(
-      "Scenario",
-      "Effect",
-      "N",
-      "AUC",
-      paste0("b", 0:8),
-      "g1", "g2", "g5", "g6"
-    ),
-    digits      = 2,
-    format.args = list(
-      big.mar      = ",",
-      decimal.mark = "."
-    ),
-    caption    = ""
-  ) %>%
-  kableExtra::landscape() %>%
-  kableExtra::add_header_above(
-    c(
-      "Analysis ID"           = 4,
-      "Baseline risk"         = 9,
-      "Coefficient in\ntreatment arm" = 4
-      )
-  )
-```
-
-# Results in scenarios with interactions
-
-When we considered a set of 4 true linear treatment-covariate interactions the
-model containing a linear interaction with the prognostic index had the lowest
-median RMSE. We observed an increasing trend in prediction errors with
-increasing interaction intensity (Figure \@ref(fig:rmseInteractions)). The
-model with restricted cubic spline smoothing (3 knots) had very comparable
-performance to the linear interaction model. Increasing the flexibility of the
-smooth methods resulted in increasing median RMSE. These results may be
-explained by the fact that the interactions considered were linear, thus
-favoring the linear interaction model. More flexible approaches may be better
-suited for higher-order treatment-covariate interactions. Finally, the adaptive
-approach had adequate performance under all scenarios, resembling the
-performance of the best-performing approach every time.
-
-```{r rmseInteractions, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/rmse_interactions.png"))
-```
-
-The constant treatment effects model, the linear interaction model and the model
-with RCS smoothing (3 knots) had the highest c-for-benefit across all scenarios
-(Figrue \@ref(fig:discriminationInteractions)). RCS smoothing with 4 or 5 knots
-did not improve performance. On the contrary, we observed an increasing trend in
-c-for-benefit variability, as was the case in the main text. The adaptive
-apprach again had statisfactory performance.
-
-```{r discriminationInteractions, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/discrimination_interactions.png"))
-```
-
-Despite the very similar performance in terms of prediction errors, the linear
-interaction model resulted in better-calibrated benefit predictions compared to
-the rest of the methods (Figure \@ref(fig:calibrationInteractions)). The constant
-treatment effects model had the highest median ICI-for-benefit across all
-scenario settings, which became more pornounced with increasing
-treatment-covariate interaction intensity.
-
-```{r calibrationInteractions, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/calibration_interactions.png"))
-```
-
-# Case study
-
-We fitted a logistic regression model to predict 30-day mortality after a
-myocardial infarction using data from the GUSTO-I trial. The risk factors
-modeled were age, Killip class, systolic blood pressure heart rate, primary MI
-and MI location. We restricted systolic blood pressure values to 120 mmHg and
-modeled heart rate using linear spline transformation with a knot at 50 beats
-per minute. A brief overview of the distribution of baseline characteristics is
-presented in Table \@ref(tab:table1). The coefficients of the model can be seen
-in Table \@ref(tab:modelCoefficients). We used this model in the manuscript to
-derive baseline risk predictions to all patients of the study. For all patients
-we set the treatment indicator to 0, assuming a constant relative treatment
-effect. The model displayed very good internal validity (Table
-\@ref(tab:internalValidity)).
-
-```{r table1, echo=FALSE, warning=FALSE, message=FALSE}
-load(here::here("data/raw/gusto.rda"))
-gusto <- gusto %>%
-  tibble() %>%
-  filter(!is.na(tpa))
-
-gustoTable1 <- gusto %>%
-  mutate(
-    tpa = factor(
-      tpa,
-      levels = c(0, 1),
-      labels = c(
-        "Tissue plasminogen activator",
-        "Streptokinase"
-      )
-    ),
-    Age = age,
-    Outcome = factor(
-      day30,
-      levels = c(0, 1),
-      labels = c(
-        "Alive",
-        "Dead"
-      )
-    )
-  )
-
-res <- table1::table1(
-  ~ sex + Age + Killip + sysbp + + miloc + pulse + pmi + Outcome | tpa,
-  data = gustoTable1,
-  caption = "Baseline characteristics of the patients included in GUSTO-I trial."
-)
-
-table1::t1kable(res)
-```
-
-```{r modelCoefficients, echo=FALSE, warning=FALSE, message=FALSE}
-gustoModelData <- gusto %>%
-  rename(
-    "outcome" = "day30",
-    "treatment" = "tpa"
-  )
-
-prediction <- glm(
-  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + rms::lsp(pulse, 50) + pmi + miloc,
-  data = gustoModelData,
-  maxit = 99,
-  family = "binomial"
-)
-
-predictionSummary <- summary(prediction)
-table <- predictionSummary$coefficients
-
-rownames(table)[2] <- "Treatment"
-rownames(table)[3] <- "Age"
-rownames(table)[4] <- "KILLIP (II)"
-rownames(table)[5] <- "KILLIP (III)"
-rownames(table)[6] <- "KILLIP (IV)"
-rownames(table)[7] <- "Systolic blood pressure"
-rownames(table)[8] <- "Pulse rate"
-rownames(table)[9] <- "Pulse rate'"
-rownames(table)[10] <- "Prior MI (yes)"
-rownames(table)[11] <- "MI location (other)"
-rownames(table)[12] <- "MI location (anterior)"
-
- knitr::kable(
-    table,
-    format      = "latex",
-    align       = "r",
-    booktabs    = TRUE,
-    digits      = 4,
-    format.args = list(
-      big.mar      = ",",
-      decimal.mark = "."
-    ),
-    caption    = "Coefficients of the prediction model for 30-day mortality after myocardial infarction (MI)."
-  )
-
-```
-
-```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Smooth calibration of the derived prediction model."}
-predictionStats <- rms::val.prob(
-  y = gustoModelData$outcome, 
-  logit = prediction$linear.predictors,
-  logistic.cal = FALSE,
-  statloc = FALSE
-)
-```
-
-```{r internalValidity, echo=FALSE, warning=FALSE, message=FALSE}
-table <- tibble(
-  Metric = c(
-    "AUC",
-    "Brier score",
-    "Emax",
-    "E90",
-    "Eavg"
-  ),
-  Value = c(
-    predictionStats[2],
-    predictionStats["Brier"],
-    predictionStats["Emax"],
-    predictionStats["E90"],
-    predictionStats["Eavg"]
-  )
-)
-
- knitr::kable(
-    table,
-    format      = "latex",
-    align       = "l",
-    booktabs    = TRUE,
-    digits      = 4,
-    format.args = list(
-      big.mar      = ",",
-      decimal.mark = "."
-    ),
-    caption    = "Evaluation statistics of the prediction model for 30-day mortality in GUSTO-I."
-  ) 
-```
-
-We used the derived prediction model to predict benefit of tPA to 30-day
-mortality using the data at hand. The models are presented in the main
-manuscript. All approaches displayed comaprable discrimination for benefit, with
-the constant treatment effect model, the linear interaction model and the RCS (5
-knots) model displaying the highest performance (Table
-\@ref(tab:gustoPerformanceMetrics)). However, the linear interaction model and
-the RCS (5 knots) model were worse calibrated. The constant treatment effect
-model also had the best performance among the considered methods in terms of
-calibration for benefit.
-
-```{r gustoPerformanceMetrics, echo=FALSE, warning=FALSE, message=FALSE}
-metrics <- readr::read_csv(
-  here::here("data/processed/gustoPerformanceMetrics.csv")
-)
-
- knitr::kable(
-    metrics,
-    format      = "latex",
-    align       = "l",
-    booktabs    = TRUE,
-    digits      = 4,
-    format.args = list(
-      big.mar      = ",",
-      decimal.mark = "."
-    ),
-    col.names = c(
-      "Model",
-      "C",
-      "ICI",
-      "AIC"
-    ),
-    caption    = "Performance metrics of the different methods used to predict benefit of tPA on 30-day mortality using data from the GUSTO-I trial."
-  ) 
-```

### 2021-12-08 code/SimulationScript.R

@@ -37,21 +37,23 @@ selectedScenario <- as.numeric(
   stringr::str_extract(args[1], "[0-9]+")
 )
 
-if (selectedScenario <= 486) {
+if (selectedScenario <= 648) {
   idSettings <- analysisIds %>%
     filter(scenario == selectedScenario)
   
   if (idSettings$base != "absent") {
     harm <- case_when(
-      idSettings$harm == "positive" ~ idSettings$averageTrueBenefit / 4, 
-      idSettings$harm == "negative" ~ -idSettings$averageTrueBenefit / 4, 
-      TRUE                          ~ 0
+      idSettings$harm == "moderate-positive" ~ idSettings$averageTrueBenefit / 4, 
+      idSettings$harm == "strong-positive"   ~ idSettings$averageTrueBenefit / 2, 
+      idSettings$harm == "negative"          ~ -idSettings$averageTrueBenefit / 4, 
+      TRUE                                   ~ 0
     )
   } else {
     harm <- case_when(
-      idSettings$harm == "positive" ~ .01,
-      idSettings$harm == "negative" ~ -.01,
-      TRUE                          ~ 0
+      idSettings$harm == "moderate-positive" ~ .01,
+      idSettings$harm == "strong-positive"   ~ .02,
+      idSettings$harm == "negative"          ~ -.01,
+      TRUE                                   ~ 0
     )
   }
   
@@ -193,7 +195,9 @@ smoothSettings <- list(
   adaptive = createHteSettings(
     settings = createAdaptiveSettings(
       list(
-        nonLinearSettings = SmoothHte::createRcsSettings(),
+        rcs3 = SmoothHte::createRcsSettings(),
+        rcs4 = SmoothHte::createRcsSettings(nKnots = 4),
+        rcs5 = SmoothHte::createRcsSettings(nKnots = 5),
         linearSettings    = SmoothHte::createModelBasedSettings(),
         constantSettings  = SmoothHte::createModelBasedSettings(
           type                 = "treatment",

### 2021-12-08 submission/manuscript.rmd
incorporate comments and update rmse figures

@@
 abstract: |
- **Objective:** To compare different risk-based methods predicting
+ **Objective**: To compare different risk-based methods predicting
  individualized treatment effects in RCTs. **Study Design and
- Setting:** We simulated data using diverse assumptions for a baseline
- prognostic index of risk (PI), the shape of its interaction with
- treatment (none, linear, quadratic or non-monotonic) and the magnitude
- of treatment-related harms. In each sample we predicted absolute
- benefit using: models with a constant relative treatment effect;
- stratification in quarters of the PI; models including a linear
- interaction of treatment with the PI; models including an interaction
- of treatment with a restricted cubic spline (RCS) transformation of
- the PI; an adaptive approach using Akaike’s Information Criterion. We
- evaluated predictive performance using root mean squared error and
- measures of discrimination and calibration for benefit. Starting from
- a base scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
- the PI 0.75), we varied the sample size, the treatment effect size,
- and the PI's AUC. **Results:** Models including a PI by treatment
- interaction performed best with smaller sample sizes and/or lower AUC
- of the PI. Otherwise, RCS (3 knots) models proved more robust for the
- majority of the deviation settings. The adaptive approach was unstable
- with smaller sample sizes. **Conclusion:** Depending on the setting,
- the linear interaction or the RCS (3 knots) model should be preferred.
-
-authors:
+ Setting**: We simulated RCT data using diverse assumptions for the
+ average treatment effect, a baseline prognostic index of risk (PI),
+ the shape of its interaction with treatment (none, linear, quadratic
+ or non-monotonic) and the magnitude of treatment-related harms. In
+ each sample we predicted absolute benefit using: models with a
+ constant relative treatment effect; stratification in quarters of the
+ PI; models including a linear interaction of treatment with the PI;
+ models including an interaction of treatment with a restricted cubic
+ spline (RCS) transformation of the PI; an adaptive approach using
+ Akaike’s Information Criterion. We evaluated predictive performance
+ using root mean squared error and measures of discrimination and
+ calibration for benefit. **Results**: The linear-interaction model
+ and the RCS-interaction model outperformed the constant treatment
+ effect model in many simulation scenarios. The RCS-model was optimal
+ when quadratic or non-monotonic deviations from a constant treatment
+ effect were stronger, and when sample size was larger. Larger sample
+ size also supported the adaptive approach. **Conclusion**: An
+ interaction between risk and treatment assignment generally improved
+ treatment effect predictions. Non-linear RCS interactions should be
+ considered for larger sample size.
+author:
   - name: Alexandros Rekkas
-    department: Department of Medical Informatics
-    affiliation: Erasmus Medical Center
-    location: Rotterdam, The Netherlands
+    affiliation: 1
   - name: Peter R. Rijnbeek
-    department: Department of Medical Informatics
-    affiliation: Erasmus Medical Center
-    location: Rotterdam, The Netherlands
+    affiliation: 1
   - name: Ewout W. Steyerberg
-    department: Department of Biomedical Data Sciences
-    affiliation: Leiden University Medical Center
-    location: Leiden, The Netherlands
+    affiliation: 2
   - name: David van Klaveren
-    department: Department of Public Health
-    affiliation: Erasmus Medical Center
-    location: Rotterdam, The Netherlands
+    affiliation: 3
+address:
+  - code: 1
+    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
+  - code: 2
+    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
+  - code: 3
+    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
+numbersections: true
+linenumbers: false
+journal: arXiv.org
 keywords:
@@
 output: 
-  rticles::arxiv_article:
+  rticles::elsevier_article:
     keep_tex: true
@@
   - \doublespacing
-  - \usepackage[pagewise, left]{lineno}
   - \usepackage{amssymb}
@@
 
-d <- function(x, decimals = 2) {
+d <- function(x, decimals = 4) {
   sprintf(paste0("%1.", decimals, "f"), x) 
@@
 this prediction model to evaluate risk-based treatment effect variation
-[@Kent2010]. However, even though estimates at the risk subgroup level may be
-accurate, these estimates do not apply to individual patients, especially for
-patients with predicted risk at the boundaries of the risk intervals. Hence, the
-risk-stratified approach is useful for exploring and presenting HTE, but is not
-useful for supporting treatment decisions for individual patients.
+[@Kent2010]. This two-step approach identified substantial absolute treatment
+effect differences between low-risk and high-risk patients in a re-analysis of
+32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
+level may be accurate, these estimates do not apply to individual patients,
+especially for patients with predicted risk at the boundaries of the risk
+intervals. Hence, the risk-stratified approach is useful for exploring and
+presenting HTE, but is not useful for supporting treatment decisions for
+individual patients.
 
@@
 
-For each patient we generated 8 baseline covariates
-$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
-was allocated using a 50:50 split. Outcomes for patients in the control arm were
-generated from a logistic regression model including all baseline covariates. In
-the base scenarios coefficient values were such, that the AUC of the
-logistic regression model was $0.75$ and the event rate in the control arm was
-$20\%$. Binary outcomes in the control arm were generated from Bernoulli variables
-with true probabilities 
+We simulated a typical RCT that is undertaken to compare a binary outcome
+(e.g. death) between a group of treated patients in the treatment arm and a
+group of untreated patients in the control arm. For each patient we generated 8
+baseline covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1,
+0.2)$. Treatment was allocated using a 50:50 split. Outcomes for patients in the
+control arm were generated from a logistic regression model including all
+baseline covariates. In the base scenarios coefficient values were such, that
+the AUC of the logistic regression model was $0.75$ and the event rate in the
+control arm was $20\%$. Binary outcomes in the control arm were generated from
+Bernoulli variables with true probabilities
 $P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.
@@
 Outcomes in the treatment arm were generated using 3 base scenarios: absent
-treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and high
+treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and strong
 treatment effect (OR = 0.5). We started with simulating outcomes based on true
@@
 detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
-sample sizes of 1,063 and 17,000, respectively. We also evaluated the effect of
-worse or better discriminative ability for risk, adjusting the baseline
-covariate coefficients, such that the AUC of the regression model in the control
-arm was 0.65 and 0.85 respectively.
+sample sizes of 1,063 (4,250 divided by 4) and 17,000 (4250 multiplied by 4),
+respectively. We also evaluated the effect of worse or better discriminative
+ability for risk, adjusting the baseline covariate coefficients, such that the
+AUC of the regression model in the control arm was 0.65 and 0.85 respectively.
 
 Combining all these settings resulted in a simulation study of 648 scenarios
-(exact settings in the supplementary material).
+(exact settings in the supplementary material). With these scenarios we were
+able to cover the observed treatment effect heterogeneity in 32 large trials as
+well as many other potential variations of risk-based treatment effect
+[@Kent2016].
 
@@
 
-All methods assume that a risk prediction model is available to assign risk
-predictions to individual patients. For the simulations we developed a
-prediction model internally, using logistic regression including main effects
+All risk-based methods assume that a risk prediction model is available to
+assign risk predictions to individual patients. For the simulations we developed
+a prediction model internally, using logistic regression including main effects
 for all baseline covariates and treatment assignment. Risk predictions for
@@
 
-Fourth, we used *restricted cubic splines* (RCS) to relax the
-linearity assumption on the effect of the linear predictor
-[@Harrell1988]. We considered splines with 3 (RCS-3), 4 (RCS-4) and 5
-(RCS-5) knots to compare models with different levels of flexibility.
+Fourth, we used *restricted cubic splines* (RCS) to relax the linearity
+assumption on the effect of the linear predictor [@Harrell1988]. We considered
+splines with 3 (RCS-3), 4 (RCS-4) and 5 (RCS-5) knots to compare models with
+different levels of flexibility.
 
@@
 0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
-quadratic deviations (Figure \ref{fig:rmsebase}). However, with
-non-monotonic deviations errors of the linear interaction model
-increased substantially, especially in the presence of
-treatment-related harms. In these scenarios, RCS-3 proved to be quite
-robust outperforming all other methods. The constant treatment effect
+quadratic deviations from a constant relative treatment effect
+(Figure \ref{fig:rmsebase}; panels A-C). However, with
+non-monotonic deviations from a constant relative treatment effect,
+the RMSE of the linear interaction model increased substantially, 
+especially in the presence of treatment-related harms (Figure 
+\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
+methods in terms of RMSE. The constant treatment effect
 approach had overall best performance under true constant treatment
 effect settings, but was sensitive to all considered deviations,
-resulting in increased error rates. Finally, the adaptive approach had
+resulting in increased RMSE. Finally, the adaptive approach had
 comparable performance to the best-performing method in each
-scenario. However, we observed increased error variability in the case
-of linear and non-monotonic deviations, especially for moderate or
-strong treatment-related harms.
+scenario. However, in comparison with the best-performing approach, its RMSE was
+more variable across the 1000 replications in the scenarios with linear and
+non-monotonic deviations, especially when also including moderate or strong
+treatment-related harms. This is caused by wrongly selecting the constant
+treatment effect model in a substantial proportion of the replications
+(Supplement, Table XX).
 
-In the case of true constant relative treatment effects, the adaptive
-approach selected the constnant effect model the majority of the time,
-even in the presence of treatment-related harms
+In the scenario with a true constant relative treatment effects, the adaptive
+approach selected the constant effect model for the majority of the replcations, even in
+the presence of treatment-related harms
 (`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\%
-of the time with no harms to
+of the replications without harms and 
 `r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
-with strong harms). With strong linear or quadratic deviations and
-increasing treatment-related harms the adaptive approach increasingly
-favored the linear interaction model. In the case of true
-non-monotonic deviations, we observed increasing selection frequencies
-for RCS-3 with increasing treatment harms (from
+with strong harms). With strong linear or quadratic deviations and increasing
+treatment-related harms the adaptive approach increasingly favored the linear
+interaction model. With non-monotonic deviations, we observed increasing
+selection frequencies for RCS-3 with increasing treatment harms (from
 `r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
@@
 `r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
-with strong harms), whereas selection frequencies for the linear
-interaction model dropped from
+with strong harms), whereas selection frequencies for the linear interaction
+model dropped from
 `r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
@@
 `r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
-with strong harms. Finally, in the case of strong linear and
-non-monotonic deviations selection frequencies of the constant
-treatment effect model remained high despite increasing treatment
-harms (Supplement, Table XX).
+with strong harms. Finally, in the case of strong linear and non-monotonic
+deviations selection frequencies of the constant treatment effect model remained
+high despite increasing treatment harms (Supplement, Table XX).
 
 ```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
-The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
-Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/rmse_base.tiff")))
+The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250.
+The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects.
+Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis).
+The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
 # knitr::include_graphics(here::here("figures/rmse_base.png"))
@@
 \ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
-non-monotonic deviations RCS-3 had lower error rates (median RMSE
+non-monotonic deviations RCS-3 had lower RMSE (median 
 `r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
@@
 for non-monotonic deviations with no treatment-related harms) compared
-to the linear interaction approach (median RMSE
+to the linear interaction approach (median 
 `r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
 `r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
-respectively), regardless of treatment-related harms strength. The
-issues with the large error variability of the adaptive approach
-improved with larger sample sizes.
-
-The adaptive approach tended to increasingly favor smoother methods
-(especially RCS-3) with increasing treatment-related harms (see
-Supplement, Table XX). However, in the case of true strong quadratic
-deviations the opposite was observed: selection frequency of the
-linear interaction model increased from
-`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(risk) * 100`\% (no harms)
-to
-`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(risk) * 100`\% (strong harms)
-whereas for RCS-3 decreased from
-`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(rcs_3) * 100`\% (no harms)
-to
-`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(rcs_3) * 100`\% (strong harms).
+respectively), regardless of the strength of treatment-related
+harms. Due to the large sample size, the RMSE of the adaptive approach
+was even more similar to the best-performing method, and the constant
+relative treatment effect model was less often wrongly selected
+(***numbers????***).
 
@@
 Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/rmse_sample_size.tiff")))
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
 # knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
@@
 When we increased the AUC of the true prediction model to 0.85 (OR =
-0.8 and N = `r 4250`) the constant effect model outperformed all other
-methods in the case of true constant treatment effects, but proved to
-be the least robust to deviations. Again, RCS-3 had the lowest
-error rates in the case of strong quadratic or monotonic deviations
-and very comparable performance to the best-performing linear
-interaction model in the case of strong linear deviations
-(median RMSE
+0.8 and N = `r 4250`). RCS-3 had the lowest RMSE in the case of strong
+quadratic or non-monotonic deviations and very comparable performance to
+the -- optimal -- linear interaction model in the case of strong
+linear deviations (median RMSE
 `r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
 `r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
-for the linear interaction model). The adaptive approach, though it
-performed similar to the best performing method in each scenario, on
-average, had increased variability in error rates in the case of
-strong linear and non-monotonic deviations. In these scenarios, the
-adaptive approach often selected the constant treatment effect model
+for the linear interaction model). As observed in the base case
+scenario the adaptive approach wrongly selected the constant treatment
+effect model
 (`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
@@
 `r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
-in the strong linear and non-monotonic deviation scenarios without
-treatment-related harms, respectively).
+of the replications in the strong linear and non-monotonic deviation
+scenarios without treatment-related harms, respectively), leading to
+more variability of the RMSE.
 
@@
 True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/rmse_auc.tiff")))
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
 # knitr::include_graphics(here::here("figures/rmse_auc.png"))
@@
 
-In terms of discrimination for benefit (OR = 0.8, N = `r 4250` and AUC
-= 0.75), all methods performed similarly, on average (Figure
-\ref{fig:discrimination}). In the case of non-monotonic deviations,
-the constant effect model had much lower discriminative performance
-compared to the rest of the methods (median AUC of
+In comparison with the true approach, dsicrimination for benefit in
+the scenario with a constant relative treatment effect was only
+slightly lower for the linear interaction model, but more
+substantially lower for the non-linear RCS approaches (Figure
+\ref{fig:discrimination}; panel A). With strong linear or quadratic
+deviations from a constant relative treatment effect, all methods
+discriminated quite similarly (Figure \ref{fig:discrimination}; panels
+B-C). In the scenario with non-monotonic deviations, the constant
+effect model had much lower discriminative ability compared to
+non-linear RCS methods (median AUC of
 `r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
-for the constant effects model compared to the best-performing RCS-3
-with
-`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
-The linear interaction model was the most stable compared to the other
-methods in terms of error variability. With increasing number of RCS
-knots, we observed decreasing median values and increasing variability
-of the c-for-benefit in all scenarios.
+for the constant effects model,
+`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
+for the linear interaction model and
+`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
+for the best-performing RCS-3; Figure \ref{fig:rmseauc}; panel D).
+With increasing number of RCS knots, we observed decreasing median
+values and increasing variability of the c-for-benefit in all
+scenarios. ***With increasing sample size... (Suppl Fig...)***
 
@@
 In terms of calibration for benefit, the constant effects model
-outperformed all other models in the case of true constant treatment
-effects, but was miscalibrated for all deviation scenarios (Figure
-\ref{fig:calibration}). The linear interaction model showed best or
-close to best calibration across all scenarios and only showed worse
-calibration compared to RCS-3 in the case of non-monotonic deviations
-and treatment-related harms. The adaptive approach was worse
-calibrated in scenarios with strong linear and non-monotonic
-deviations compared to the linear interaction model and RCS-3.
+outperformed all other models in the scenario with true constant
+treatment effects, but was miscalibrated for all deviation scenarios
+(Figure \ref{fig:calibration}). The linear interaction model showed
+best or close to best calibration across all scenarios and only showed
+worse calibration compared to RCS-3 in case of non-monotonic
+deviations and treatment-related harms (Figure \ref{fig:calibration};
+panel D). The adaptive approach was worse calibrated in scenarios with
+strong linear and non-monotonic deviations compared to the linear
+interaction model and RCS-3. ***With increasing sample size... (Supp
+Fig...)***
 
@@
 adaptive approach selected the model with RCS smoothing with 4
-knots. However, for very low baseline risk this model predicted
-decreasing benefit with increasing risk may be somewhat too flexible.
-The more robust models, the linear interaction model or the model with
-RCS smoothing (3 knots), gave very similar benefit predictions,
-followed the evolution of the stratified estimates very closely and
-may therefore be preferable for use in clinical practice. The linear
+knots. However, for very low baseline risk the decreasing predicted
+benefit with increasing risk may be somewhat too flexible. The more
+robust models, the linear interaction model or the model with RCS
+smoothing (3 knots), gave very similar benefit predictions, followed
+the evolution of the stratified estimates very closely and may
+therefore be preferable for use in clinical practice. The linear
 interaction model had somewhat lower AIC compared to the model with
@@
 
-The linear interaction model displayed very good performance overall
-under many of the considered simulation scenarios. Especially in cases
-with smaller sample sizes and moderately performing baseline risk
-prediction models it had lower RMSE, was better calibrated for benefit
-and had better discrimination for benefit, even in scenarios with
-strong quadratic deviations. However, in scenarios with true
-non-monotonic deviations, the linear interaction model was
-outperformed by RCS-3, especially in the presence of true
-treatment-related harms. Increasing the sample size or the prediction
-model's discrimination favored RCS-3 which had better or very
-comparable performance to the linear interaction model, but was more
-robust to non-monotonic deviations and the presence of
-treatment-related harms.
+The linear interaction model and the RCS-3 model both displayed very
+good performance under many of the considered simulation scenarios, in
+contrast with the constant relative treatment effect model. The linear
+interaction model was optimal in cases with smaller sample sizes and
+moderately performing baseline risk prediction models, that is, it had
+lower RMSE, was better calibrated for benefit and had better
+discrimination for benefit, even in scenarios with strong quadratic
+deviations. In scenarios with true non-monotonic deviations, the
+linear interaction model was outperformed by RCS-3, especially in the
+presence of true treatment-related harms. Increasing the sample size
+or the prediction model’s discriminative ability favored RCS-3,
+especially in scenarios with non-monotonic deviations and in the
+presence of treatment-related harms.
 
-RCS-4 and RCS-5 proved to be too flexible, as indicated by higher
-RMSE, increased variability of discrimination for benefit and worse
-calibration of benefit predictions. Even with larger sample sizes and
-strong quadratic or non-monotonic deviations from the base case
-scenario of constant relative treatment effects, these more flexible
-restricted cubic splines did not outperform the simpler RCS-3 These
-approaches may only be helpful if we expect more extreme patterns of
-heterogeneous treatment effects compared to the quadratic deviations
-considered here.
+RCS-4 and RCS-5 proved to be too flexible in all considered scenarios,
+as indicated by higher RMSE, increased variability of discrimination
+for benefit and worse calibration of benefit predictions. Even with
+larger sample sizes and strong quadratic or non-monotonic deviations
+from the base case scenario of constant relative treatment effects,
+these more flexible restricted cubic splines did not outperform the
+simpler RCS-3 These approaches may only be helpful if we expect more
+extreme patterns of heterogeneous treatment effects compared to the
+quadratic deviations considered here.
 
@@
 treatment effect model. Stepwise treatment benefit estimates are very
-useful for demonstrating treatment effect heterogeneity--because
+useful for demonstrating treatment effect heterogeneity–because
 estimating treatment effect requires groups of patients rather than
-individual patients--but are not helpful for making individualized
+individual patients–but are not helpful for making individualized
 absolute benefit predictions.
 
-Increasing the discriminative ability of the risk model--by increasing
-the predictor coefficients of the true risk model--reduced RMSE for
-all methods.  This increase in discriminative ability translates in
-higher variability of predicted risks, which, in turn, allows the
-considered methods to better capture absolute treatment benefits. As a
+Increasing the discriminative ability of the risk model–by increasing
+the predictor coefficients of the true risk model–reduced RMSE for all
+methods. This increase in discriminative ability translates in higher
+variability of predicted risks, which, in turn, allows the considered
+methods to better capture absolute treatment benefits. As a
 consequence, the increase in discriminative ability of the risk model
@@
 The adaptive approach had adequate performance, following closely on
-average the performance of the "true" model in most
-scenarios. However, with smaller sample sizes it tended to "miss" the
+average the performance of the “true” model in most
+scenarios. However, with smaller sample sizes it tended to “miss” the
 treatment-risk interactions and selected simpler models (Supplementary
@@
 ranging from treatment effect modeling to tree-based approaches
-available in more recent literature [@Athey2019; @Lu2018;
-@Wager2018]. Simulations are also needed to assess relative
-performance and define the settings where these break down or
-outperform each other.
+available in more recent literature [@Athey2019; @Lu2018;@Wager2018].
+Simulations are also needed to assess relative performance and define
+the settings where these break down or outperform each other.
 
-In conclusion, the best option for predicting individualized treatment
-benefit using a risk-based approach depends on the setting. With
+In conclusion, the linear interaction approach is a viable option with
 smaller sample sizes and/or moderately performing risk prediction
-models the linear interaction approach is a viable option. When those
-constraints are not present or when we anticipate non-negligible
-treatment-related harms, RCS-3 is a better option in terms of error
-rates, discrimination and calibration for benefit. With larger sample
-size, an adaptive approach based on AIC can also be considered as a
-more automated alternative.
+modelsRCS-3 is a better option when non-monotonic deviations from a
+constant relative treatment effect and/or substantial
+treatment-related harms are anticipated. With larger sample size, an
+adaptive approach -- selecting the method with optimal AIC -- can also
+be considered as an automated alternative.
 

### 2021-12-08 submission/manuscript.tex
incorporate comments and update rmse figures

@@
-\documentclass{article}
+\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
+%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+\usepackage[hyphens]{url}
 
-\usepackage{arxiv}
+  \journal{arXiv.org} % Sets Journal name
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{lipsum}
+\usepackage{lineno} % add
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \usepackage{graphicx}
+\usepackage{booktabs} % book-quality tables
+%%%%%%%%%%%%%%%% end my additions to header
 
-\title{Linear interaction of treatment with baseline risk was sufficient
-for predicting individualized benefit}
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \usepackage{fontspec}
+  \ifxetex
+    \usepackage{xltxtra,xunicode}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage[margin=1.0in]{geometry}
+\bibliographystyle{elsarticle-harv}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={Linear interaction of treatment with baseline risk was sufficient for predicting individualized benefit},
+            colorlinks=false,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
 
-\author{
-    Alexandros Rekkas
-   \\
-    Department of Medical Informatics \\
-    Erasmus Medical Center \\
-  Rotterdam, The Netherlands \\
-  \texttt{} \\
-   \And
-    Peter R. Rijnbeek
-   \\
-    Department of Medical Informatics \\
-    Erasmus Medical Center \\
-  Rotterdam, The Netherlands \\
-  \texttt{} \\
-   \And
-    Ewout W. Steyerberg
-   \\
-    Department of Biomedical Data Sciences \\
-    Leiden University Medical Center \\
-  Leiden, The Netherlands \\
-  \texttt{} \\
-   \And
-    David van Klaveren
-   \\
-    Department of Public Health \\
-    Erasmus Medical Center \\
-  Rotterdam, The Netherlands \\
-  \texttt{} \\
-  }
+\setcounter{secnumdepth}{5}
+% Pandoc toggle for numbering sections (defaults to be off)
 
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
 \newlength{\csllabelwidth}
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
 % for Pandoc 2.8 to 2.10.1
@@
  {}
-\usepackage{calc} % for calculating minipage widths
+\usepackage{calc}
 \newcommand{\CSLBlock}[1]{#1\hfill\break}
@@
 
+% Pandoc header
 \renewcommand*\familydefault{\sfdefault}
@@
 \doublespacing
-\usepackage[pagewise, left]{lineno}
 \usepackage{amssymb}
@@
 \begin{document}
-\maketitle
+\begin{frontmatter}
 
-\def\tightlist{}
+  \title{Linear interaction of treatment with baseline risk was
+sufficient for predicting individualized benefit}
+    \author[1]{Alexandros Rekkas}
+
+    \author[1]{Peter R. Rijnbeek}
+
+    \author[2]{Ewout W. Steyerberg}
+
+    \author[3]{David van Klaveren}
+
+      \address[1]{Department of Medical Informatics, Erasmus Medical
+Center, Rotterdam, The Netherlands}
+    \address[2]{Department of Biomedical Data Sciences, Leiden
+University Medical Center, Leiden, The Netherlands}
+    \address[3]{Department of Public Health, Erasmus Medical Center,
+Rotterdam, The Netherlands}
 
   \begin{abstract}
-\textbf{Objective:} To compare different risk-based methods predicting
+  \textbf{Objective}: To compare different risk-based methods predicting
   individualized treatment effects in RCTs. \textbf{Study Design and
-Setting:} We simulated data using diverse assumptions for a baseline
-prognostic index of risk (PI), the shape of its interaction with
-treatment (none, linear, quadratic or non-monotonic) and the magnitude
-of treatment-related harms. In each sample we predicted absolute benefit
-using: models with a constant relative treatment effect; stratification
-in quarters of the PI; models including a linear interaction of
-treatment with the PI; models including an interaction of treatment with
-a restricted cubic spline (RCS) transformation of the PI; an adaptive
-approach using Akaike's Information Criterion. We evaluated predictive
-performance using root mean squared error and measures of discrimination
-and calibration for benefit. Starting from a base scenario (sample size
-4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the
-sample size, the treatment effect size, and the PI's AUC.
-\textbf{Results:} Models including a PI by treatment interaction
-performed best with smaller sample sizes and/or lower AUC of the PI.
-Otherwise, RCS (3 knots) models proved more robust for the majority of
-the deviation settings. The adaptive approach was unstable with smaller
-sample sizes. \textbf{Conclusion:} Depending on the setting, the linear
-interaction or the RCS (3 knots) model should be preferred.
+  Setting}: We simulated RCT data using diverse assumptions for the
+  average treatment effect, a baseline prognostic index of risk (PI),
+  the shape of its interaction with treatment (none, linear, quadratic
+  or non-monotonic) and the magnitude of treatment-related harms. In
+  each sample we predicted absolute benefit using: models with a
+  constant relative treatment effect; stratification in quarters of the
+  PI; models including a linear interaction of treatment with the PI;
+  models including an interaction of treatment with a restricted cubic
+  spline (RCS) transformation of the PI; an adaptive approach using
+  Akaike's Information Criterion. We evaluated predictive performance
+  using root mean squared error and measures of discrimination and
+  calibration for benefit. \textbf{Results}: The linear-interaction
+  model and the RCS-interaction model outperformed the constant
+  treatment effect model in many simulation scenarios. The RCS-model was
+  optimal when quadratic or non-monotonic deviations from a constant
+  treatment effect were stronger, and when sample size was larger.
+  Larger sample size also supported the adaptive approach.
+  \textbf{Conclusion}: An interaction between risk and treatment
+  assignment generally improved treatment effect predictions. Non-linear
+  RCS interactions should be considered for larger sample size.
   \end{abstract}
-
-\keywords{
-    treatment effect heterogeneity
-   \and
-    absolute benefit
-   \and
-    prediction models
-  }
+   \begin{keyword} treatment effect heterogeneity absolute
+benefit prediction models\end{keyword}
+ \end{frontmatter}
 
@@
 is stratified using this prediction model to evaluate risk-based
-treatment effect variation {[}6{]}. However, even though estimates at
-the risk subgroup level may be accurate, these estimates do not apply to
-individual patients, especially for patients with predicted risk at the
-boundaries of the risk intervals. Hence, the risk-stratified approach is
-useful for exploring and presenting HTE, but is not useful for
-supporting treatment decisions for individual patients.
+treatment effect variation {[}6{]}. This two-step approach identified
+substantial absolute treatment effect differences between low-risk and
+high-risk patients in a re-analysis of 32 large trials {[}7{]}. However,
+even though estimates at the risk subgroup level may be accurate, these
+estimates do not apply to individual patients, especially for patients
+with predicted risk at the boundaries of the risk intervals. Hence, the
+risk-stratified approach is useful for exploring and presenting HTE, but
+is not useful for supporting treatment decisions for individual
+patients.
 
@@
 
-For each patient we generated 8 baseline covariates
+We simulated a typical RCT that is undertaken to compare a binary
+outcome (e.g.~death) between a group of treated patients in the
+treatment arm and a group of untreated patients in the control arm. For
+each patient we generated 8 baseline covariates
 \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
@@
 absent treatment effect (OR = 1), moderate treatment effect (OR = 0.8)
-and high treatment effect (OR = 0.5). We started with simulating
+and strong treatment effect (OR = 0.5). We started with simulating
 outcomes based on true constant relative treatment effects for the 3
@@
 for the detection of a marginal OR of 0.8). We evaluated the effect of
-smaller or larger sample sizes of 1,063 and 17,000, respectively. We
-also evaluated the effect of worse or better discriminative ability for
-risk, adjusting the baseline covariate coefficients, such that the AUC
-of the regression model in the control arm was 0.65 and 0.85
-respectively.
+smaller or larger sample sizes of 1,063 (4,250 divided by 4) and 17,000
+(4250 multiplied by 4), respectively. We also evaluated the effect of
+worse or better discriminative ability for risk, adjusting the baseline
+covariate coefficients, such that the AUC of the regression model in the
+control arm was 0.65 and 0.85 respectively.
 
 Combining all these settings resulted in a simulation study of 648
-scenarios (exact settings in the supplementary material).
+scenarios (exact settings in the supplementary material). With these
+scenarios we were able to cover the observed treatment effect
+heterogeneity in 32 large trials as well as many other potential
+variations of risk-based treatment effect {[}7{]}.
 
@@
 
-All methods assume that a risk prediction model is available to assign
-risk predictions to individual patients. For the simulations we
-developed a prediction model internally, using logistic regression
+All risk-based methods assume that a risk prediction model is available
+to assign risk predictions to individual patients. For the simulations
+we developed a prediction model internally, using logistic regression
 including main effects for all baseline covariates and treatment
@@
 Fourth, we used \emph{restricted cubic splines} (RCS) to relax the
-linearity assumption on the effect of the linear predictor {[}7{]}. We
+linearity assumption on the effect of the linear predictor {[}8{]}. We
 considered splines with 3 (RCS-3), 4 (RCS-4) and 5 (RCS-5) knots to
@@
 We compared the discriminative ability of the methods under study using
-c-for-benefit {[}8{]}. The c-for-benefit represents the probability that
+c-for-benefit {[}9{]}. The c-for-benefit represents the probability that
 from two randomly chosen matched patient pairs with unequal observed
@@
 We evaluated calibration in a similar manner, using the integrated
-calibration index (ICI) for benefit {[}9{]}. The observed benefits are
+calibration index (ICI) for benefit {[}10{]}. The observed benefits are
 regressed on the predicted benefits using a locally weighted scatterplot
@@
 0.8, N = 4,250 and AUC = 0.75), strong linear and even strong quadratic
-deviations (Figure \ref{fig:rmsebase}). However, with non-monotonic
-deviations errors of the linear interaction model increased
-substantially, especially in the presence of treatment-related harms. In
-these scenarios, RCS-3 proved to be quite robust outperforming all other
-methods. The constant treatment effect approach had overall best
-performance under true constant treatment effect settings, but was
-sensitive to all considered deviations, resulting in increased error
-rates. Finally, the adaptive approach had comparable performance to the
-best-performing method in each scenario. However, we observed increased
-error variability in the case of linear and non-monotonic deviations,
-especially for moderate or strong treatment-related harms.
+deviations from a constant relative treatment effect (Figure
+\ref{fig:rmsebase}; panels A-C). However, with non-monotonic deviations
+from a constant relative treatment effect, the RMSE of the linear
+interaction model increased substantially, especially in the presence of
+treatment-related harms (Figure \ref{fig:rmsebase}; panel D). In these
+scenarios, RCS-3 outperformed all other methods in terms of RMSE. The
+constant treatment effect approach had overall best performance under
+true constant treatment effect settings, but was sensitive to all
+considered deviations, resulting in increased RMSE. Finally, the
+adaptive approach had comparable performance to the best-performing
+method in each scenario. However, in comparison with the best-performing
+approach, its RMSE was more variable across the 1000 replications in the
+scenarios with linear and non-monotonic deviations, especially when also
+including moderate or strong treatment-related harms. This is caused by
+wrongly selecting the constant treatment effect model in a substantial
+proportion of the replications (Supplement, Table XX).
 
-In the case of true constant relative treatment effects, the adaptive
-approach selected the constnant effect model the majority of the time,
-even in the presence of treatment-related harms (96\% of the time with
-no harms to 88\% with strong harms). With strong linear or quadratic
-deviations and increasing treatment-related harms the adaptive approach
-increasingly favored the linear interaction model. In the case of true
-non-monotonic deviations, we observed increasing selection frequencies
-for RCS-3 with increasing treatment harms (from 32\% with no harms to
-53\% with strong harms), whereas selection frequencies for the linear
-interaction model dropped from 40\% with no harms to 7\% with strong
-harms. Finally, in the case of strong linear and non-monotonic
+In the scenario with a true constant relative treatment effects, the
+adaptive approach selected the constant effect model for the majority of
+the replcations, even in the presence of treatment-related harms (96\%
+of the replications without harms and 88\% with strong harms). With
+strong linear or quadratic deviations and increasing treatment-related
+harms the adaptive approach increasingly favored the linear interaction
+model. With non-monotonic deviations, we observed increasing selection
+frequencies for RCS-3 with increasing treatment harms (from 32\% with no
+harms to 53\% with strong harms), whereas selection frequencies for the
+linear interaction model dropped from 40\% with no harms to 7\% with
+strong harms. Finally, in the case of strong linear and non-monotonic
 deviations selection frequencies of the constant treatment effect model
@@
 \includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsebase-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
-The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250.
-Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.}\label{fig:rmsebase}
+The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250.
+The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects.
+Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis).
+The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.}\label{fig:rmsebase}
 \end{figure}
@@
 \ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
-non-monotonic deviations RCS-3 had lower error rates (median RMSE 0.011
-for strong quadratic deviations and 0.010 for non-monotonic deviations
-with no treatment-related harms) compared to the linear interaction
-approach (median RMSE 0.013 and 0.014, respectively), regardless of
-treatment-related harms strength. The issues with the large error
-variability of the adaptive approach improved with larger sample sizes.
-
-The adaptive approach tended to increasingly favor smoother methods
-(especially RCS-3) with increasing treatment-related harms (see
-Supplement, Table XX). However, in the case of true strong quadratic
-deviations the opposite was observed: selection frequency of the linear
-interaction model increased from 31\% (no harms) to 50\% (strong harms)
-whereas for RCS-3 decreased from 52\% (no harms) to 34\% (strong harms).
+non-monotonic deviations RCS-3 had lower RMSE (median 0.011 for strong
+quadratic deviations and 0.010 for non-monotonic deviations with no
+treatment-related harms) compared to the linear interaction approach
+(median 0.013 and 0.014, respectively), regardless of the strength of
+treatment-related harms. Due to the large sample size, the RMSE of the
+adaptive approach was even more similar to the best-performing method,
+and the constant relative treatment effect model was less often wrongly
+selected (\textbf{\emph{numbers????}}).
 
@@
 When we increased the AUC of the true prediction model to 0.85 (OR = 0.8
-and N = 4,250) the constant effect model outperformed all other methods
-in the case of true constant treatment effects, but proved to be the
-least robust to deviations. Again, RCS-3 had the lowest error rates in
-the case of strong quadratic or monotonic deviations and very comparable
-performance to the best-performing linear interaction model in the case
-of strong linear deviations (median RMSE 0.016 for RCS-3 compared to
-0.014 for the linear interaction model). The adaptive approach, though
-it performed similar to the best performing method in each scenario, on
-average, had increased variability in error rates in the case of strong
-linear and non-monotonic deviations. In these scenarios, the adaptive
-approach often selected the constant treatment effect model (23\% and
-25\% in the strong linear and non-monotonic deviation scenarios without
-treatment-related harms, respectively).
+and N = 4,250). RCS-3 had the lowest RMSE in the case of strong
+quadratic or non-monotonic deviations and very comparable performance to
+the -- optimal -- linear interaction model in the case of strong linear
+deviations (median RMSE 0.016 for RCS-3 compared to 0.014 for the linear
+interaction model). As observed in the base case scenario the adaptive
+approach wrongly selected the constant treatment effect model (23\% and
+25\% of the replications in the strong linear and non-monotonic
+deviation scenarios without treatment-related harms, respectively),
+leading to more variability of the RMSE.
 
@@
 
-In terms of discrimination for benefit (OR = 0.8, N = 4,250 and AUC =
-0.75), all methods performed similarly, on average (Figure
-\ref{fig:discrimination}). In the case of non-monotonic deviations, the
-constant effect model had much lower discriminative performance compared
-to the rest of the methods (median AUC of 0.04 for the constant effects
-model compared to the best-performing RCS-3 with 0.02). The linear
-interaction model was the most stable compared to the other methods in
-terms of error variability. With increasing number of RCS knots, we
+In comparison with the true approach, dsicrimination for benefit in the
+scenario with a constant relative treatment effect was only slightly
+lower for the linear interaction model, but more substantially lower for
+the non-linear RCS approaches (Figure \ref{fig:discrimination}; panel
+A). With strong linear or quadratic deviations from a constant relative
+treatment effect, all methods discriminated quite similarly (Figure
+\ref{fig:discrimination}; panels B-C). In the scenario with
+non-monotonic deviations, the constant effect model had much lower
+discriminative ability compared to non-linear RCS methods (median AUC of
+0.4971 for the constant effects model, 0.5285 for the linear interaction
+model and 0.5304 for the best-performing RCS-3; Figure
+\ref{fig:rmseauc}; panel D). With increasing number of RCS knots, we
 observed decreasing median values and increasing variability of the
-c-for-benefit in all scenarios.
+c-for-benefit in all scenarios. \textbf{\emph{With increasing sample
+size\ldots{} (Suppl Fig\ldots)}}
 
@@
 In terms of calibration for benefit, the constant effects model
-outperformed all other models in the case of true constant treatment
-effects, but was miscalibrated for all deviation scenarios (Figure
-\ref{fig:calibration}). The linear interaction model showed best or
-close to best calibration across all scenarios and only showed worse
-calibration compared to RCS-3 in the case of non-monotonic deviations
-and treatment-related harms. The adaptive approach was worse calibrated
-in scenarios with strong linear and non-monotonic deviations compared to
-the linear interaction model and RCS-3.
+outperformed all other models in the scenario with true constant
+treatment effects, but was miscalibrated for all deviation scenarios
+(Figure \ref{fig:calibration}). The linear interaction model showed best
+or close to best calibration across all scenarios and only showed worse
+calibration compared to RCS-3 in case of non-monotonic deviations and
+treatment-related harms (Figure \ref{fig:calibration}; panel D). The
+adaptive approach was worse calibrated in scenarios with strong linear
+and non-monotonic deviations compared to the linear interaction model
+and RCS-3. \textbf{\emph{With increasing sample size\ldots{} (Supp
+Fig\ldots)}}
 
@@
 
-In line with previous analyses {[}10,11{]}, we fitted a logistic
+In line with previous analyses {[}11,12{]}, we fitted a logistic
 regression model with 6 baseline covariates, i.e.~age, Killip class,
@@
 adaptive approach selected the model with RCS smoothing with 4 knots.
-However, for very low baseline risk this model predicted decreasing
-benefit with increasing risk may be somewhat too flexible. The more
-robust models, the linear interaction model or the model with RCS
-smoothing (3 knots), gave very similar benefit predictions, followed the
-evolution of the stratified estimates very closely and may therefore be
-preferable for use in clinical practice. The linear interaction model
-had somewhat lower AIC compared to the model with RCS smoothing (3
-knots), slightly better cross-validated discrimination (c-for-benefit
-0.526 vs 0.525) and quite similar cross-validated calibration (ICI-for
-benefit 0.0115 vs 0.0117).
+However, for very low baseline risk the decreasing predicted benefit
+with increasing risk may be somewhat too flexible. The more robust
+models, the linear interaction model or the model with RCS smoothing (3
+knots), gave very similar benefit predictions, followed the evolution of
+the stratified estimates very closely and may therefore be preferable
+for use in clinical practice. The linear interaction model had somewhat
+lower AIC compared to the model with RCS smoothing (3 knots), slightly
+better cross-validated discrimination (c-for-benefit 0.526 vs 0.525) and
+quite similar cross-validated calibration (ICI-for benefit 0.0115 vs
+0.0117).
 
@@
 
-The linear interaction model displayed very good performance overall
-under many of the considered simulation scenarios. Especially in cases
-with smaller sample sizes and moderately performing baseline risk
-prediction models it had lower RMSE, was better calibrated for benefit
-and had better discrimination for benefit, even in scenarios with strong
-quadratic deviations. However, in scenarios with true non-monotonic
-deviations, the linear interaction model was outperformed by RCS-3,
-especially in the presence of true treatment-related harms. Increasing
-the sample size or the prediction model's discrimination favored RCS-3
-which had better or very comparable performance to the linear
-interaction model, but was more robust to non-monotonic deviations and
-the presence of treatment-related harms.
+The linear interaction model and the RCS-3 model both displayed very
+good performance under many of the considered simulation scenarios, in
+contrast with the constant relative treatment effect model. The linear
+interaction model was optimal in cases with smaller sample sizes and
+moderately performing baseline risk prediction models, that is, it had
+lower RMSE, was better calibrated for benefit and had better
+discrimination for benefit, even in scenarios with strong quadratic
+deviations. In scenarios with true non-monotonic deviations, the linear
+interaction model was outperformed by RCS-3, especially in the presence
+of true treatment-related harms. Increasing the sample size or the
+prediction model's discriminative ability favored RCS-3, especially in
+scenarios with non-monotonic deviations and in the presence of
+treatment-related harms.
 
-RCS-4 and RCS-5 proved to be too flexible, as indicated by higher RMSE,
-increased variability of discrimination for benefit and worse
-calibration of benefit predictions. Even with larger sample sizes and
-strong quadratic or non-monotonic deviations from the base case scenario
-of constant relative treatment effects, these more flexible restricted
-cubic splines did not outperform the simpler RCS-3 These approaches may
-only be helpful if we expect more extreme patterns of heterogeneous
-treatment effects compared to the quadratic deviations considered here.
+RCS-4 and RCS-5 proved to be too flexible in all considered scenarios,
+as indicated by higher RMSE, increased variability of discrimination for
+benefit and worse calibration of benefit predictions. Even with larger
+sample sizes and strong quadratic or non-monotonic deviations from the
+base case scenario of constant relative treatment effects, these more
+flexible restricted cubic splines did not outperform the simpler RCS-3
+These approaches may only be helpful if we expect more extreme patterns
+of heterogeneous treatment effects compared to the quadratic deviations
+considered here.
 
@@
 treatment benefit, prediction model development was outside the scope of
-this work and has already been studied extensively {[}5,12,13{]}.
+this work and has already been studied extensively {[}5,13,14{]}.
 
@@
 from treatment effect modeling to tree-based approaches available in
-more recent literature {[}14--16{]}. Simulations are also needed to
+more recent literature {[}15--17{]}. Simulations are also needed to
 assess relative performance and define the settings where these break
@@
 
-In conclusion, the best option for predicting individualized treatment
-benefit using a risk-based approach depends on the setting. With smaller
-sample sizes and/or moderately performing risk prediction models the
-linear interaction approach is a viable option. When those constraints
-are not present or when we anticipate non-negligible treatment-related
-harms, RCS-3 is a better option in terms of error rates, discrimination
-and calibration for benefit. With larger sample size, an adaptive
-approach based on AIC can also be considered as a more automated
-alternative.
+In conclusion, the linear interaction approach is a viable option with
+smaller sample sizes and/or moderately performing risk prediction
+modelsRCS-3 is a better option when non-monotonic deviations from a
+constant relative treatment effect and/or substantial treatment-related
+harms are anticipated. With larger sample size, an adaptive approach --
+selecting the method with optimal AIC -- can also be considered as an
+automated alternative.
 
@@
 
+\leavevmode\hypertarget{ref-Kent2016}{}%
+{[}7{]} Kent DM, Nelson J, Dahabreh IJ, Rothwell PM, Altman DG, Hayward
+RA. Risk and treatment effect heterogeneity: Re-analysis of individual
+participant data from 32 large clinical trials. International Journal of
+Epidemiology 2016:dyw118. \url{https://doi.org/10.1093/ije/dyw118}.
+
 \leavevmode\hypertarget{ref-Harrell1988}{}%
-{[}7{]} Harrell FE, Lee KL, Pollock BG. Regression models in clinical
+{[}8{]} Harrell FE, Lee KL, Pollock BG. Regression models in clinical
 studies: Determining relationships between predictors and response. JNCI
@@
 \leavevmode\hypertarget{ref-vanKlaveren2018}{}%
-{[}8{]} Klaveren D van, Steyerberg EW, Serruys PW, Kent DM. The proposed
+{[}9{]} Klaveren D van, Steyerberg EW, Serruys PW, Kent DM. The proposed
 ``concordance-statistic for benefit'' provided a useful metric when
@@
 \leavevmode\hypertarget{ref-Austin2019}{}%
-{[}9{]} Austin PC, Steyerberg EW. The integrated calibration index (ICI)
-and related metrics for quantifying the calibration of logistic
+{[}10{]} Austin PC, Steyerberg EW. The integrated calibration index
+(ICI) and related metrics for quantifying the calibration of logistic
 regression models. Statistics in Medicine 2019;38:4051--65.
@@
 \leavevmode\hypertarget{ref-Califf1997}{}%
-{[}10{]} Califf RM, Woodlief LH, Harrell FE, Lee KL, White HD, Guerci A,
+{[}11{]} Califf RM, Woodlief LH, Harrell FE, Lee KL, White HD, Guerci A,
 et al. Selection of thrombolytic therapy for individual patients:
@@
 \leavevmode\hypertarget{ref-Steyerberg2000}{}%
-{[}11{]} Steyerberg EW, Bossuyt PMM, Lee KL. Clinical trials in acute
+{[}12{]} Steyerberg EW, Bossuyt PMM, Lee KL. Clinical trials in acute
 myocardial infarction: Should we adjust for baseline characteristics?
@@
 \leavevmode\hypertarget{ref-Burke2014}{}%
-{[}12{]} Burke JF, Hayward RA, Nelson JP, Kent DM. Using internally
+{[}13{]} Burke JF, Hayward RA, Nelson JP, Kent DM. Using internally
 developed risk models to assess heterogeneity in treatment effects in
@@
 \leavevmode\hypertarget{ref-Abadie2018}{}%
-{[}13{]} Abadie A, Chingos MM, West MR. Endogenous stratification in
+{[}14{]} Abadie A, Chingos MM, West MR. Endogenous stratification in
 randomized experiments. The Review of Economics and Statistics
@@
 \leavevmode\hypertarget{ref-Athey2019}{}%
-{[}14{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
+{[}15{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
 Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.
@@
 \leavevmode\hypertarget{ref-Lu2018}{}%
-{[}15{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
+{[}16{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
 treatment effect in observational data using random forest methods.
@@
 \leavevmode\hypertarget{ref-Wager2018}{}%
-{[}16{]} Wager S, Athey S. Estimation and inference of heterogeneous
+{[}17{]} Wager S, Athey S. Estimation and inference of heterogeneous
 treatment effects using random forests. Journal of the American
@@
 
-\bibliographystyle{unsrt}
-\bibliography{references.bib}
-
 \end{document}

### 2021-12-08 code/CreateDistribution.R

@@ -11,7 +11,7 @@
 #     - auc
 #     - calibration
 # Output:
-#   - data/processed/(rmse|auc|calibration)Distribution.csv
+#   - data/processed/(rmse|discrimination|calibration)Distribution.csv
 # Depends:
 #   - code/helpers/GetMedians.R
 # =============================================================
@@ -23,12 +23,18 @@ saveFile <- paste0(
   "Distribution.csv"
 )
 
+
 library(tidyverse)
 library(readr)
 
 source("code/helpers/GetMedians.R")
 
-read_csv("data/processed/rmse.csv") %>%
+read_csv(
+  file.path(
+    "data/processed",
+    paste0(args_metric, ".csv")
+  )
+) %>%
   group_by(scenarioId) %>%
   arrange(scenarioId) %>%
   nest() %>%

### 2021-12-08 code/PlotRmse.R

@@ -18,9 +18,10 @@
 # ===================================================
 
 args <- commandArgs(trailingOnly = TRUE)
-args_sampleSize <- as.numeric(args[1])
-args_auc <- as.numeric(args[2])
-args_value <- as.character(args[3])
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
 
 library(tidyverse)
 library(glue)
@@ -35,7 +36,7 @@ source("code/helpers/Absolute.R")                  # Generates the absolute plot
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(
-    base == "moderate",
+    base == args_base,
     !(type %in% c("quadratic-moderate", "linear-moderate")),
     sampleSize == args_sampleSize,
     auc == args_auc
@@ -102,7 +103,8 @@ absolutePlots <- scenarioIds %>%
       type,
       ~ plotAbsoluteBenefit(
       data = .x,
-      projectDir = "~/Documents/Projects/arekkas_HteSimulation_XXXX_2021"
+      projectDir = "~/Documents/Projects/arekkas_HteSimulation_XXXX_2021",
+      type = .y
       )
     )
   )
@@ -341,6 +343,7 @@ res <- grid.arrange(arrangeGrob(pp, left = left.grob, right = right.grob, bottom
 fileName <- paste0(
   paste(
     metric,
+    args_base,
     args_value,
     sep = "_"
   ),

### 2021-12-08 code/helpers/Absolute.R

@@ -1,4 +1,4 @@
-plotAbsoluteBenefit <- function(data, projectDir = NULL) {
+plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
   calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
     ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
     return(ret)
@@ -68,19 +68,23 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL) {
       bind_rows(.id = "harmSetting")
     res <- res +
       # ggside::ggside(x.pos = "bottom") +
-      ggside::ggside(y.pos = "left") +
+      ggside::ggside(y.pos = "left")
+    if (type == "constant") {
+      res <- res +
       ggside::geom_xsideboxplot(
         aes(x = probs),
         data = validationDataset[[1]] %>% mutate(probs = plogis(untreatedRiskLinearPredictor)),
         orientation = "y",
         outlier.shape = NA
-      ) +
-      ggside::geom_ysideboxplot(
-        aes(y = trueBenefit, fill = harmSetting),
-        data = fullValidationDataset,
-        orientation = "x",
-        outlier.shape = NA
-        ) +
+      )
+    }
+      # ggside::geom_ysideboxplot(
+      #   aes(y = trueBenefit, fill = harmSetting),
+      #   data = fullValidationDataset,
+      #   orientation = "x",
+      #   outlier.shape = NA
+      #   ) +
+    res <- res +
       scale_ysidex_continuous() +
       scale_xsidey_continuous() +
       scale_fill_manual(

### 2021-12-09 submission/manuscript.rmd
add model selections and rmse for high effects to supplement

@@
 scenario. However, in comparison with the best-performing approach, its RMSE was
-more variable across the 1000 replications in the scenarios with linear and
+more variable across the 500 replications in the scenarios with linear and
 non-monotonic deviations, especially when also including moderate or strong
@@
 treatment effect model in a substantial proportion of the replications
-(Supplement, Table XX).
+(Supplement, Figure S3).
 
@@
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
-knitr::include_graphics(here::here("figures/gusto.png"))
+grid::grid.raster(tiff::readTIFF(here("figures/gusto.tiff")))
+# knitr::include_graphics(here::here("figures/gusto.png"))
 ```

### 2021-12-09 submission/manuscript.tex
add model selections and rmse for high effects to supplement

@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{/home/arekkas/Documents/Projects/arekkas_HteSimulation_XXXX_2021/figures/gusto} \caption{Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/gusto-1} \caption{Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.}\label{fig:gusto}

### 2021-12-09 submission/supplement.rmd
add model selections and rmse for high effects to supplement

@@
   select(-"type")
-kableExtra::kable(
+kableExtra::kbl(
   asd,
@@
     "c",
-    "Average benefit"
+    "Before harms",
+    "After harms"
   ),
+  linesep = c('', '', '', '\\addlinespace'),
   digits      = 2,
@@
       "Baseline risk"         = 9,
-      "True treatment effect" = 4
+      "True treatment effect" = 4,
+      "Benefit"               = 2
       )
@@
 ```
+
+\newpage
+\begin{landscape}
+```{r adaptiveBase, cache=FALSE, echo=FALSE, fig.cap="Adaptive model selection rates for the base case scenario true prediction AUC of 0.75, sample size of 4250", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_base.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+\end{landscape}
+
+\newpage
+\begin{landscape}
+```{r adaptiveSampleSize, cache=FALSE, echo=FALSE, fig.cap="Adaptive model selection rates for sample size of 17000.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+\end{landscape}
+
+\newpage
+\begin{landscape}
+```{r adaptiveAuc, cache=FALSE, echo=FALSE, fig.cap="Adaptive model selection rates for true prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_auc.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+\end{landscape}
+
+\newpage
+```{r rmseHighBase, cache=FALSE, echo=FALSE, fig.cap="RMSE high effect base", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_base.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+\newpage
+```{r rmseHighSampleSize, cache=FALSE, echo=FALSE, fig.cap="RMSE high effect sample size", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+\newpage
+```{r rmseHighAuc, cache=FALSE, echo=FALSE, fig.cap="RMSE high effect auc", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_auc.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```

### 2021-12-09 code/GustoPlot.R

@@ -64,20 +64,6 @@ rcs5Model <- fitRcsHte(
 
 
 plot <- ggplot() +
-  stat_function(
-    data = data.frame(
-      x     = c(0, maxRisk), 
-      label = "Constant treatment effect"
-    ),
-    aes(
-      x     = x, 
-      color = label
-    ),
-    fun  = predictBenefitModelBasedHte,
-    args = list(
-      modelBasedFit = constantModel
-    )
-  ) +
   geom_pointrange(
     data = stratified$data,
     aes(
@@ -91,13 +77,32 @@ plot <- ggplot() +
   ) +
   stat_function(
     data = data.frame(
+      x     = c(0, maxRisk), 
+      label = "Constant treatment effect"
+    ),
+    aes(
+      x     = x, 
+      color = label,
+      linetype = label,
+      alpha = .8
+    ),
+    size = .8,
+    fun  = predictBenefitModelBasedHte,
+    args = list(
+      modelBasedFit = constantModel
+    )
+  ) +
+  stat_function(
+    data = data.frame(
       x = c(0, maxRisk), 
       label = "Linear interaction"
     ),
     aes(
       x     = x,
+      linetype = label,
       color = label
     ),
+    size = .65,
     fun  = predictBenefitModelBasedHte,
     args = list(
       modelBasedFit = linearModel
@@ -110,8 +115,10 @@ plot <- ggplot() +
     ),
     aes(
       x     = x,
-      color = label
+      color = label,
+      linetype = label
     ),
+    size = .65,
     fun  = predictSmoothBenefit,
     args = list(
       smoothFit = rcs3Model
@@ -124,7 +131,9 @@ plot <- ggplot() +
     ),
     aes(
       x     = x,
-      color = label
+      color = label,
+      linetype = label,
+      alpha = .8
     ),
     fun  = predictSmoothBenefit,
     args = list(
@@ -138,21 +147,9 @@ plot <- ggplot() +
     ),
     aes(
       x     = x,
-      color = label
-    ),
-    fun  = predictSmoothBenefit,
-    args = list(
-      smoothFit = rcs5Model
-    )
-  ) +
-  stat_function(
-    data = data.frame(
-      x = c(0, maxRisk), 
-      label = "RCS 5 knots"
-    ),
-    aes(
-      x     = x,
-      color = label
+      color = label,
[…]

### 2021-12-09 code/PlotRmse.R

@@ -74,6 +74,12 @@ metricFile <- paste(metric, "csv", sep = ".")
 
 f <- function(x) x * 100
 
+limitsHigh <- ifelse(
+  args_base == "high",
+  yes = .2,
+  no = .15
+)
+
 processed <- readr::read_csv(
   file = file.path("data/processed", metricFile)
 ) %>% 
@@ -145,8 +151,8 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.058, .15),
-      breaks = seq(-.05, .15, .05)
+      limits = c(-.058, limitsHigh),
+      breaks = seq(-.05, limitsHigh, .05)
     ) +
   scale_color_manual(
       name = "Constant treatment-\n related harm",
@@ -188,8 +194,8 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.058, .15),
-      breaks = seq(-.05, .15, .05)
+      limits = c(-.058, limitsHigh),
+      breaks = seq(-.05, limitsHigh, .05)
     ) +
     scale_color_manual(
       name = "Constant treatment-\n related harm",
@@ -225,8 +231,8 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.058, .15),
-      breaks = seq(-.05, .15, .05)
+      limits = c(-.058, limitsHigh),
+      breaks = seq(-.05, limitsHigh, .05)
     ) +
     scale_color_manual(
       name = "Constant treatment-\n related harm",
@@ -263,8 +269,8 @@ gridList <- list(
     xlim(c(0, .5)) +
     scale_y_continuous(
       position = "right",
-      limits = c(-.058, .15),
-      breaks = seq(-.05, .15, .05)
+      limits = c(-.058, limitsHigh),
+      breaks = seq(-.05, limitsHigh, .05)
     ) +
     scale_color_manual(
       name = "Constant treatment-\n related harm",

### 2021-12-09 code/WriteAnalysisIds.R

@@ -148,6 +148,17 @@ expand_grid(base, type, sampleSize, auc, harm) %>%
       base == "high" & auc == .75 ~ .079,
       base == "high" & auc == .65 ~ .089,
       base == "high" & auc == .85 ~ .069,
+    ),
+    averageBenefit = case_when(
+      base == "absent" & harm == "absent" ~ 0,
+      base == "absent" & harm == "moderate-positive" ~ -.01,
+      base == "absent" & harm == "strong-positive" ~ -.02,
+      base == "absent" & harm == "negative" ~ .01,
+      base != "absent" & harm == "absent" ~ averageTrueBenefit,
+      base != "absent" & harm == "moderate-positive" ~ averageTrueBenefit - averageTrueBenefit / 4,
+      base != "absent" & harm == "strong-positive" ~ averageTrueBenefit - averageTrueBenefit / 4,
+      base != "absent" & harm == "moderate-positive" ~ averageTrueBenefit - averageTrueBenefit / 2,
+      base != "absent" & harm == "negative" ~ averageTrueBenefit + averageTrueBenefit / 4,
     )
   ) %>%
   relocate(scenario) %>%

### 2021-12-13 submission/supplement.rmd
add caption

@@
   booktabs    = TRUE,
-  caption     = "",
+  caption     = "Scenario settings of the entire simulation study.",
   col.names   = c(

### 2021-12-13 submission/supplement.rmd
add sections

--- 2021-12-08 submission/supplement.rmd
@@
 
-For the second part linear and quadratic deviations from the
-assumption of constant relative effect are considered. We also
-consider different intensity levels of these deviations. The settings
-for these deviations are defined in Table \@ref(tab:tab1) and result
-in the effects of Figure \@ref(fig:fig1).
-
-```{r fig1, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '50%'}
-knitr::include_graphics(here::here("figures/deviate_linear_08.png"))
-knitr::include_graphics(here::here("figures/deviate_quadratic_08.png"))
-```
-
-In Figure \@ref(fig:fig2) the absolute benefits observed based on different
-settings are presented. The base-case scenario is also presented as a reference.
-
-```{r fig2, echo=FALSE, fig.cap="Linear and quadratic deviations from the base-case scenario of constant relative effect (OR=0.8)", fig.show="hold", out.width = '50%'}
-knitr::include_graphics(here::here("figures/deviate_linear_absolute_08.png"))
-knitr::include_graphics(here::here("figures/deviate_quadratic_absolute_08.png"))
-```
-
-Finally, we consider 3 additional scenarios of interaction of
-individual covariates with treatment. These scenarios include a 4 weak
-interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong
-interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2
-weak and 2 strong interactions (Table \@ref(tab:tab2)).
+For the second part linear, quadratic and non-monotonic deviations
+from the assumption of constant relative effect are considered. We
+also consider different intensity levels of these deviations. Finally,
+constant absolute treatment-related harms are introduced,
+i.e. positive ($0.25\times\text{true average benefit}$), strong
+positive ($0.50\times\text{true average benefit}$) and negative
+($-0.25\times\text{true average benefit}$; i.e. constant absolute
+treatment-related benefit). In case of true absent treatment effects,
+treatment-related harms are set to $1\%, 2\%$ and $-1\%$ for
+positive, strong positive and negative setting, respectively. The
+settings for these deviations are defined in Table \@ref(tab:tab1).
 
@@
 ```
+
+# Adaptive model selection frequencies
+
+```{r adaptiveBase, cache=TRUE, echo=FALSE, fig.cap="Adaptive model selection rates for the base case scenario true prediction AUC of 0.75, sample size of 4250", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_base.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r adaptiveSampleSize, cache=TRUE, echo=FALSE, fig.cap="Adaptive model selection rates for sample size of 17000.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r adaptiveAuc, cache=TRUE, echo=FALSE, fig.cap="Adaptive model selection rates for true prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_auc.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+\newpage
+
+# Strong relative treatment effect
+
+Here we present the root mean squared error of the considered methods
+using strong constant relative treatment effect ($\text{OR}=0.5$) as
+the reference. Again, the same sample size and prediction performance
+settings were considered along with the same settings for linear,
+quadratic and non-monotonic deviations from the base case scenario of
+constant relative treatment effects are considered. All results can be found at [https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).
+
+```{r rmseHighBase, cache=TRUE, echo=FALSE, fig.cap="RMSE high effect base", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_base.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r rmseHighSampleSize, cache=TRUE, echo=FALSE, fig.cap="RMSE high effect sample size", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r rmseHighAuc, cache=TRUE, echo=FALSE, fig.cap="RMSE high effect auc", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_auc.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```

### 2021-12-13 code/GustoPlot.R

@@ -161,8 +161,8 @@ plot <- ggplot() +
       "#2274A5", 
       "#969A97",
       "#273043",
-      "#DB3A34",
-      "#F75C03",
+      "#F15152",
+      "#F4D8CD",
       "#FFC857"
     ),
     breaks = c(

### 2021-12-13 submission/manuscript.rmd
include discrimination and calibration plots to supplement

@@
 ```
+
 The linear interaction model outperformed all RCS methods in terms of
@@
 0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
-quadratic deviations from a constant relative treatment effect
-(Figure \ref{fig:rmsebase}; panels A-C). However, with
-non-monotonic deviations from a constant relative treatment effect,
-the RMSE of the linear interaction model increased substantially, 
-especially in the presence of treatment-related harms (Figure 
-\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
-methods in terms of RMSE. The constant treatment effect
-approach had overall best performance under true constant treatment
-effect settings, but was sensitive to all considered deviations,
-resulting in increased RMSE. Finally, the adaptive approach had
-comparable performance to the best-performing method in each
-scenario. However, in comparison with the best-performing approach, its RMSE was
-more variable across the 500 replications in the scenarios with linear and
-non-monotonic deviations, especially when also including moderate or strong
-treatment-related harms. This is caused by wrongly selecting the constant
-treatment effect model in a substantial proportion of the replications
-(Supplement, Figure S3).
-
-In the scenario with a true constant relative treatment effects, the adaptive
-approach selected the constant effect model for the majority of the replcations, even in
-the presence of treatment-related harms
-(`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\%
-of the replications without harms and 
-`r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
-with strong harms). With strong linear or quadratic deviations and increasing
-treatment-related harms the adaptive approach increasingly favored the linear
-interaction model. With non-monotonic deviations, we observed increasing
-selection frequencies for RCS-3 with increasing treatment harms (from
-`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
-with no harms to
-`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
-with strong harms), whereas selection frequencies for the linear interaction
-model dropped from
-`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
-with no harms to
-`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
-with strong harms. Finally, in the case of strong linear and non-monotonic
-deviations selection frequencies of the constant treatment effect model remained
-high despite increasing treatment harms (Supplement, Table XX).
+quadratic deviations from a constant relative treatment effect (Figure
+\ref{fig:rmsebase}; panels A-C). However, with non-monotonic
+deviations from a constant relative treatment effect, the RMSE of the
+linear interaction model increased substantially, especially in the
+presence of treatment-related harms (Figure \ref{fig:rmsebase}; panel
+D). In these scenarios, RCS-3 outperformed all other methods in terms
+of RMSE. The constant treatment effect approach had overall best
+performance under true constant treatment effect settings, but was
+sensitive to all considered deviations, resulting in increased
+RMSE. Finally, the adaptive approach had comparable performance to the
+best-performing method in each scenario. However, in comparison with
+the best-performing approach, its RMSE was more variable across the
+500 replications in the scenarios with linear and non-monotonic
+deviations, especially when also including moderate or strong
+treatment-related harms. This is caused by wrongly selecting the
+constant treatment effect model in a substantial proportion of the
+replications (Supplement, Figure S1).
 
@@
 relative treatment effect model was less often wrongly selected
-(***numbers????***).
+(Supplement, Figure S2).
 
@@
 scenarios without treatment-related harms, respectively), leading to
-more variability of the RMSE.
+more variability of the RMSE (Supplement, Figure S3).
 
@@
 True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/discrimination_base.tiff")))
+grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
 # knitr::include_graphics(here::here("figures/discrimination_base.png"))
@@
 True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/calibration_base.tiff")))
+grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_base.tiff")))
 # knitr::include_graphics(here::here("figures/calibration_base.png"))

### 2021-12-13 submission/manuscript.tex
include discrimination and calibration plots to supplement

@@
 method in each scenario. However, in comparison with the best-performing
-approach, its RMSE was more variable across the 1000 replications in the
+approach, its RMSE was more variable across the 500 replications in the
 scenarios with linear and non-monotonic deviations, especially when also
@@
 wrongly selecting the constant treatment effect model in a substantial
-proportion of the replications (Supplement, Table XX).
-
-In the scenario with a true constant relative treatment effects, the
-adaptive approach selected the constant effect model for the majority of
-the replcations, even in the presence of treatment-related harms (96\%
-of the replications without harms and 88\% with strong harms). With
-strong linear or quadratic deviations and increasing treatment-related
-harms the adaptive approach increasingly favored the linear interaction
-model. With non-monotonic deviations, we observed increasing selection
-frequencies for RCS-3 with increasing treatment harms (from 32\% with no
-harms to 53\% with strong harms), whereas selection frequencies for the
-linear interaction model dropped from 40\% with no harms to 7\% with
-strong harms. Finally, in the case of strong linear and non-monotonic
-deviations selection frequencies of the constant treatment effect model
-remained high despite increasing treatment harms (Supplement, Table XX).
+proportion of the replications (Supplement, Figure S1).
 
@@
 and the constant relative treatment effect model was less often wrongly
-selected (\textbf{\emph{numbers????}}).
+selected (Supplement, Figure S2).
 
@@
 deviation scenarios without treatment-related harms, respectively),
-leading to more variability of the RMSE.
+leading to more variability of the RMSE (Supplement, Figure S3).
 

### 2021-12-13 submission/supplement.rmd
include discrimination and calibration plots to supplement

@@
 
-```{r adaptiveBase, cache=TRUE, echo=FALSE, fig.cap="Adaptive model selection rates for the base case scenario true prediction AUC of 0.75, sample size of 4250", fig.show="hold", out.width = '100%'}
+```{r adaptiveBase, cache=TRUE, echo=FALSE, fig.cap="Model selection frequencies of the adaptive approach based on Akaike's Information Criterion across 500 replications.
+The scenario with the true constant relative treatment effect (first panel) had a true prediction AUC of 0.75 and sample size of 4,250.
+", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_base.tiff")))
@@
 
-```{r adaptiveSampleSize, cache=TRUE, echo=FALSE, fig.cap="Adaptive model selection rates for sample size of 17000.", fig.show="hold", out.width = '100%'}
+```{r adaptiveSampleSize, cache=TRUE, echo=FALSE, fig.cap="Model selection frequencies of the adaptive approach based on Akaike's Information Criterion across 500 replications.
+Sample size is 17,000 rather than 4,250 in Figure \\ref{fig:adaptiveBase}", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_sample_size.tiff")))
@@
 
-```{r adaptiveAuc, cache=TRUE, echo=FALSE, fig.cap="Adaptive model selection rates for true prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
+```{r adaptiveAuc, cache=TRUE, echo=FALSE, fig.cap="Model selection frequencies of the adaptive approach based on Akaike's Information Criterion across 500 replications.
+AUC is 0.85 rather than 0.75 in Figure \\ref{fig:adaptiveBase}", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_auc.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+\newpage
+
+# Discrimination and calibration for benefit
+
+```{r discriminationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Sample size", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r discriminationAuc, cache=FALSE, echo=FALSE, fig.cap="Auc", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_auc.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r calibrationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Calibration Sample size", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_sample_size.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r calibrationAuc, cache=FALSE, echo=FALSE, fig.cap="Calibration Auc", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_auc.tiff")))
 # knitr::include_graphics(here::here("figures/calibration_base.png"))
@@
 
-```{r rmseHighBase, cache=TRUE, echo=FALSE, fig.cap="RMSE high effect base", fig.show="hold", out.width = '100%'}
+```{r rmseHighBase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
+The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4,250.
+The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) deviations from constant relative treatment effects.
+Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis).
+The 2.5, 25, 75 and 97.5 percentiles of the risk distribution are expressed in the boxplot.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_base.tiff")))
@@
 
-```{r rmseHighSampleSize, cache=TRUE, echo=FALSE, fig.cap="RMSE high effect sample size", fig.show="hold", out.width = '100%'}
+```{r rmseHighSampleSize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
+Sample size is 17,000 rather than 4,250 in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_sample_size.tiff")))
@@
 
-```{r rmseHighAuc, cache=TRUE, echo=FALSE, fig.cap="RMSE high effect auc", fig.show="hold", out.width = '100%'}
+```{r rmseHighAuc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
+AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_auc.tiff")))

### 2021-12-13 code/CalibrationBase.R

@@ -1,8 +1,26 @@
 #!/usr/bin/env Rscript
 
-# Description: 
-# Output: 
-# Depends: 
+# ===================================================
+# Description:
+#   Generates the calibration plots
+# Input:
+#   - sample size
+#   - auc
+#   - value
+# Output:
+#   - figures/calibration_*.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/calibration.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+# ===================================================
+
+args <- commandArgs(trailingOnly = TRUE)
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
 
 library(tidyverse)
 library(glue)
@@ -12,13 +30,14 @@ source("code/helpers/PlotResult.R")
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(
-    base == "moderate",
-    !(type %in% c("linear-moderate", "quadratic-moderate")),
-    sampleSize == 4250,
-    auc == .75
-  )
+    base == args_base,
+    harm != "negative",
+    !(type %in% c("quadratic-moderate", "linear-moderate")),
+    sampleSize == args_sampleSize,
+    auc == args_auc
+  ) 
+
 metric    <- "calibration"
-value     <- "base"
 
 titles <- scenarioIds %>%
   mutate(
@@ -53,7 +72,6 @@ f <- function(x) x * 100
 processed <- readr::read_csv(
   file = file.path("data/processed", metricFile)
 ) %>% 
-  select(-one_of("locfit")) %>%
   mutate_at(
     c(
       "constant_treatment_effect",
@@ -72,11 +90,12 @@ scenarios <- scenarioIds %>%
   unlist()
 names(scenarios) <- NULL
 
-plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 6))
+plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 6, 1))
 
-pp <- gridExtra::grid.arrange(
+res <- gridExtra::grid.arrange(
   plotList[[1]] +
     theme(
+      panel.grid.minor = element_blank(),
       plot.title = element_markdown(),
       axis.title = element_blank(),
       legend.direction = "horizontal",
[…]

### 2021-12-13 code/DiscriminationBase.R

@@ -1,8 +1,27 @@
 #!/usr/bin/env Rscript
 
-# Description: 
-# Output: 
-# Depends: 
+# ===================================================
+# Description:
+#   Generates the discrimination plots
+# Input:
+#   - sample size
+#   - auc
+#   - value
+# Output:
+#   - figures/rmse_*.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/discrimination.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+# ===================================================
+
+args <- commandArgs(trailingOnly = TRUE)
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
+
 
 library(tidyverse)
 library(glue)
@@ -12,13 +31,14 @@ source("code/helpers/PlotResult.R")
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(
-    base == "moderate",
-    !(type %in% c("linear-moderate", "quadratic-moderate")),
-    sampleSize == 4250,
-    auc == .75
-  )
+    base == args_base,
+    harm != "negative",
+    !(type %in% c("quadratic-moderate", "linear-moderate")),
+    sampleSize == args_sampleSize,
+    auc == args_auc
+  ) 
+
 metric <- "discrimination"
-value  <- "base"
 
 titles <- scenarioIds %>%
   mutate(
@@ -57,9 +77,9 @@ scenarios <- scenarioIds %>%
   unlist()
 names(scenarios) <- NULL
 
-plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(.5, .57))
+plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(.49, .6, .01))
 
-pp <- gridExtra::grid.arrange(
+res <- gridExtra::grid.arrange(
   plotList[[1]] +
     theme(
       plot.title = element_markdown(),
@@ -68,6 +88,7 @@ pp <- gridExtra::grid.arrange(
       legend.position = c(.444, .93),
       legend.text = element_text(size = 7),
       legend.title = element_text(size = 8),
+      panel.grid.minor = element_blank(),
       axis.text.x = element_blank()
     ),
   plotList[[2]] +
@@ -75,18 +96,21 @@ pp <- gridExtra::grid.arrange(
       plot.title = element_markdown(),
[…]

### 2021-12-13 code/GustoPlot.R

@@ -162,8 +162,8 @@ plot <- ggplot() +
       "#969A97",
       "#273043",
       "#F15152",
-      "#F4D8CD",
-      "#FFC857"
+      "#FFC857",
+      "#F4D8CD"
     ),
     breaks = c(
       "Constant treatment effect",

### 2021-12-13 code/PlotRmse.R

@@ -125,7 +125,7 @@ scenarios <- scenarioIds %>%
 
 names(scenarios) <- NULL
 
-plotList <- plotResult(scenarios, processed, titles, metric = metric)
+plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 10, 2.5))
 
 
 # ------------------------------------
@@ -136,6 +136,7 @@ plotList <- plotResult(scenarios, processed, titles, metric = metric)
 gridList <- list(
   plotList[[1]] +
     theme(
+      panel.grid.minor = element_blank(),
       plot.title = element_markdown(size = 9),
       axis.title.x = ggplot2::element_blank(),
       axis.title.y = ggplot2::element_blank(),
@@ -164,6 +165,7 @@ gridList <- list(
     ) +
     theme_bw() +
     theme(
+      panel.grid.minor = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.x = element_blank(),
@@ -178,6 +180,7 @@ gridList <- list(
     ),
   plotList[[2]] + 
     theme(
+      panel.grid.minor = element_blank(),
       plot.title = element_markdown(size = 9),
       axis.title = element_blank(),
       axis.text.x = element_blank(),
@@ -207,6 +210,7 @@ gridList <- list(
     ) +
     theme_bw() +
     theme(
+      panel.grid.minor = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.x = element_blank(),
@@ -221,6 +225,7 @@ gridList <- list(
     ),
   plotList[[3]] +
     theme(
+      panel.grid.minor = element_blank(),
       plot.title = element_markdown(size = 9),
       axis.title = element_blank(),
       axis.text.x = element_blank(),
@@ -244,6 +249,7 @@ gridList <- list(
     ) +
     theme_bw() +
     theme(
+      panel.grid.minor = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.y = element_text(size = 8),
@@ -258,6 +264,7 @@ gridList <- list(
     ),
   plotList[[4]] +
     theme(
+      panel.grid.minor = element_blank(),
       axis.title = element_blank(),
       axis.text.x = element_text(size = 8),
       axis.text.y = element_text(size = 8),
@@ -282,6 +289,7 @@ gridList <- list(
     ) +
     theme_bw() +
     theme(
[…]

### 2021-12-13 code/helpers/CreateManuscriptPlots.R

@@ -77,7 +77,11 @@ createPlot <- function(
       )
     ) +
     ggplot2::ylab(yAxis) +
-    ggplot2::ylim(limits[1], limits[2]) +
+    ggplot2::scale_y_continuous(
+               breaks = seq(limits[1], limits[2], limits[3]),
+               limits = c(limits[1], limits[2])
+             ) +
+    # ggplot2::ylim(limits[1], limits[2]) +
     ggplot2::theme_bw()
   
   if (!missing(title)) {

### 2021-12-15 submission/manuscript.rmd
shared version

@@
 title: |
- Linear interaction of treatment with baseline risk was sufficient for predicting
- individualized benefit
+  An interaction between baseline risk and treatment assignment 
+  generally imporoved treatment effect predictions
 abstract: |
@@
     affiliation: 1
+  - name: David M. Kent
+    affiliation: 2
   - name: Ewout W. Steyerberg
-    affiliation: 2
+    affiliation: 3
   - name: David van Klaveren
-    affiliation: 3
+    affiliation: 4
 address:
@@
   - code: 2
+    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
+  - code: 3
     address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
-  - code: 3
+  - code: 4
     address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
@@
 the scenario with a constant relative treatment effect was only
-slightly lower for the linear interaction model, but more
-substantially lower for the non-linear RCS approaches (Figure
+slightly lower for the linear interaction model, but substantially
+lower for the non-linear RCS approaches (Figure
 \ref{fig:discrimination}; panel A). With strong linear or quadratic
@@
 effect model had much lower discriminative ability compared to
-non-linear RCS methods (median AUC of
+all other methods (median AUC of
 `r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
@@
 `r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
-for the best-performing RCS-3; Figure \ref{fig:rmseauc}; panel D).
-With increasing number of RCS knots, we observed decreasing median
-values and increasing variability of the c-for-benefit in all
-scenarios. ***With increasing sample size... (Suppl Fig...)***
+for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel
+D). The adaptive approach was unstable in terms of discrimination for
+benefit, especially in the presence of treamtnt-related harms. With
+increasing number of RCS knots, we observed decreasing median values
+and increasing variability of the c-for-benefit in all scenarios. When
+we increased the sample size to 17,000 we observed similar trends,
+however the performance of all methods was more stable (Supplement,
+Figure S4). Finally, when we increased the true prediction AUC to 0.85
+the adaptive approach in the case of non-monotonic deviations was,
+again, quite unstable, especially with null or moderate
+treatment-related harms (Supplement, Figure S5). In these scenarios,
+the adaptive approach tended to select more often the much inferior
+constant treatment effect method (Supplement, Figure S3)
 
@@
 strong linear and non-monotonic deviations compared to the linear
-interaction model and RCS-3. ***With increasing sample size... (Supp
-Fig...)***
+interaction model and RCS-3. When we increased sample size to 17,000
+similar conclusions on calibration for benefit could be drawn. As
+expected, all methods displayed more stable calibration performance
+due to the larger number of patients (Supplement, Figure S6). When we
+increased the true prediction AUC to 0.85, the linear interaction
+model was worse calibrated, on average, than RCS-3 in the case of
+strong quadratic deviations from constant relative treatment effects
+(Supplement, Figure S7).
 

### 2021-12-15 submission/manuscript.tex
shared version

@@
             pdfauthor={},
-            pdftitle={Linear interaction of treatment with baseline risk was sufficient for predicting individualized benefit},
+            pdftitle={An interaction between baseline risk and treatment assignment generally imporoved treatment effect predictions},
             colorlinks=false,
@@
 
-  \title{Linear interaction of treatment with baseline risk was
-sufficient for predicting individualized benefit}
+  \title{An interaction between baseline risk and treatment assignment
+generally imporoved treatment effect predictions}
     \author[1]{Alexandros Rekkas}
@@
 
-    \author[2]{Ewout W. Steyerberg}
+    \author[2]{David M. Kent}
 
-    \author[3]{David van Klaveren}
+    \author[3]{Ewout W. Steyerberg}
+
+    \author[4]{David van Klaveren}
 
@@
 Center, Rotterdam, The Netherlands}
-    \address[2]{Department of Biomedical Data Sciences, Leiden
+    \address[2]{Predictive Analytics and Comparative Effectiveness
+Center, Institute for Clinical Research and Health Policy Studies, Tufts
+Medical Center, Boston, Massachusetts, USA}
+    \address[3]{Department of Biomedical Data Sciences, Leiden
 University Medical Center, Leiden, The Netherlands}
-    \address[3]{Department of Public Health, Erasmus Medical Center,
+    \address[4]{Department of Public Health, Erasmus Medical Center,
 Rotterdam, The Netherlands}
@@
 scenario with a constant relative treatment effect was only slightly
-lower for the linear interaction model, but more substantially lower for
-the non-linear RCS approaches (Figure \ref{fig:discrimination}; panel
-A). With strong linear or quadratic deviations from a constant relative
+lower for the linear interaction model, but substantially lower for the
+non-linear RCS approaches (Figure \ref{fig:discrimination}; panel A).
+With strong linear or quadratic deviations from a constant relative
 treatment effect, all methods discriminated quite similarly (Figure
@@
 non-monotonic deviations, the constant effect model had much lower
-discriminative ability compared to non-linear RCS methods (median AUC of
+discriminative ability compared to all other methods (median AUC of
 0.4971 for the constant effects model, 0.5285 for the linear interaction
 model and 0.5304 for the best-performing RCS-3; Figure
-\ref{fig:rmseauc}; panel D). With increasing number of RCS knots, we
-observed decreasing median values and increasing variability of the
-c-for-benefit in all scenarios. \textbf{\emph{With increasing sample
-size\ldots{} (Suppl Fig\ldots)}}
+\ref{fig:discrimination}; panel D). The adaptive approach was unstable
+in terms of discrimination for benefit, especially in the presence of
+treamtnt-related harms. With increasing number of RCS knots, we observed
+decreasing median values and increasing variability of the c-for-benefit
+in all scenarios. When we increased the sample size to 17,000 we
+observed similar trends, however the performance of all methods was more
+stable (Supplement, Figure S4). Finally, when we increased the true
+prediction AUC to 0.85 the adaptive approach in the case of
+non-monotonic deviations was, again, quite unstable, especially with
+null or moderate treatment-related harms (Supplement, Figure S5). In
+these scenarios, the adaptive approach tended to select more often the
+much inferior constant treatment effect method (Supplement, Figure S3)
 
@@
 and non-monotonic deviations compared to the linear interaction model
-and RCS-3. \textbf{\emph{With increasing sample size\ldots{} (Supp
-Fig\ldots)}}
+and RCS-3. When we increased sample size to 17,000 similar conclusions
+on calibration for benefit could be drawn. As expected, all methods
+displayed more stable calibration performance due to the larger number
+of patients (Supplement, Figure S6). When we increased the true
+prediction AUC to 0.85, the linear interaction model was worse
+calibrated, on average, than RCS-3 in the case of strong quadratic
+deviations from constant relative treatment effects (Supplement, Figure
+S7).
 

### 2021-12-15 submission/supplement.rmd
shared version

@@
 
-```{r discriminationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Sample size", fig.show="hold", out.width = '100%'}
+```{r discriminationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_sample_size.tiff")))
@@
 
-```{r discriminationAuc, cache=FALSE, echo=FALSE, fig.cap="Auc", fig.show="hold", out.width = '100%'}
+```{r discriminationAuc, cache=FALSE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_auc.tiff")))
@@
 
-```{r calibrationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Calibration Sample size", fig.show="hold", out.width = '100%'}
+```{r calibrationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_sample_size.tiff")))
@@
 
-```{r calibrationAuc, cache=FALSE, echo=FALSE, fig.cap="Calibration Auc", fig.show="hold", out.width = '100%'}
+```{r calibrationAuc, cache=FALSE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_auc.tiff")))

### 2022-01-26 submission/manuscript.rmd
add treatment interaction scenarios

@@
 title: |
-  An interaction between baseline risk and treatment assignment 
-  generally imporoved treatment effect predictions
+ Individualized treatment effect was predicted best by modeling 
+ baseline risk in interaction with treatment assignment
 abstract: |
- **Objective**: To compare different risk-based methods predicting
- individualized treatment effects in RCTs. **Study Design and
- Setting**: We simulated RCT data using diverse assumptions for the
- average treatment effect, a baseline prognostic index of risk (PI),
- the shape of its interaction with treatment (none, linear, quadratic
- or non-monotonic) and the magnitude of treatment-related harms. In
- each sample we predicted absolute benefit using: models with a
- constant relative treatment effect; stratification in quarters of the
- PI; models including a linear interaction of treatment with the PI;
- models including an interaction of treatment with a restricted cubic
- spline (RCS) transformation of the PI; an adaptive approach using
- Akaike’s Information Criterion. We evaluated predictive performance
- using root mean squared error and measures of discrimination and
- calibration for benefit. **Results**: The linear-interaction model
- and the RCS-interaction model outperformed the constant treatment
- effect model in many simulation scenarios. The RCS-model was optimal
- when quadratic or non-monotonic deviations from a constant treatment
+ **Objective**: To compare different risk-based methods for optimal
+ prediction of individualized treatment effects from RCTs. **Study
+ Design and Setting**: We simulated RCT data using diverse assumptions
+ for the average treatment effect, a baseline prognostic index of risk
+ (PI), the shape of its interaction with treatment (none, linear,
+ quadratic or non-monotonic) and the magnitude of treatment-related
+ harms (none or constant independent of the PI). In each sample we
+ predicted absolute benefit using: models with a constant relative
+ treatment effect; stratification in quarters of the PI; models
+ including a linear interaction of treatment with the PI; models
+ including an interaction of treatment with a restricted cubic spline
+ (RCS) transformation of the PI; an adaptive approach using Akaike’s
+ Information Criterion. We evaluated predictive performance using root
+ mean squared error and measures of discrimination and calibration for
+ benefit. **Results**: The linear-interaction model and the
+ RCS-interaction model outperformed the constant treatment effect
+ model in many simulation scenarios. The RCS-model was optimal when
+ quadratic or non-monotonic deviations from a constant treatment
  effect were stronger, and when sample size was larger. Larger sample
- size also supported the adaptive approach. **Conclusion**: An
- interaction between risk and treatment assignment generally improved
- treatment effect predictions. Non-linear RCS interactions should be
- considered for larger sample size.
+ size also supported the adaptive approach. Illustrations in the
+ GUSTO-I trial confirmed these findings. **Conclusion**: An
+ interaction between baseline risk and treatment assignment should be
+ considered to improve treatment effect predictions. Non-linear RCS
+ interactions should be considered only in larger sample sizes.
 author:
@@
 reference class used for defining patient similarity when making individualized
-predictions or recommendations [@Rekkas2020]. Risk-modeling approaches use
-prediction of baseline risk as the reference; treatment effect modeling
-approaches also model treatment-covariate interactions, in addition to risk
-factors; optimal treatment regime approaches focus on developing treatment
-assignment rules and therefore rely heavily on modeling treatment effect
-modifiers.
+predictions or recommendations [@Rekkas2020]. First, risk-modeling approaches
+use prediction of baseline risk as the reference; second, treatment effect
+modeling approaches also model treatment-covariate interactions, in addition to
+risk factors; third, optimal treatment regime approaches focus on developing
+treatment assignment rules and rely heavily on modeling treatment effect
+modifiers. A key difference between these approaches is their parsimony in
+dealing the treatment effect modifiers, with no interaction considered (risk
+modeling), a limited number of interactions (effect modeling), or a larger set
+of interactions (optimal treatment regime approaches).
 
@@
 32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
-level may be accurate, these estimates do not apply to individual patients,
-especially for patients with predicted risk at the boundaries of the risk
-intervals. Hence, the risk-stratified approach is useful for exploring and
-presenting HTE, but is not useful for supporting treatment decisions for
-individual patients.
+level may be accurate, these estimates may need further refinement for
+individual patients, especially for patients with predicted risk at the
+boundaries of the risk intervals. Hence, the risk-stratified approach is useful
+for exploring and presenting HTE, but is not sufficient for supporting treatment
+decisions for individual patients.
 
@@
 simulated RCT settings to compare the performance of these models under
-different assumptions of the relationship between baseline risk and treatment.
-We illustrated the different models by a case study of predicting individualized
-effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
-patients with an acute myocardial infarction (MI).
+different assumptions of the relation between baseline risk and treatment. We
+illustrated the different models by a case study of predicting individualized
+effects of treatments for acute myocardial infarction (MI) in a large randomized
+controlled trial (RCT).
 
@@
 We simulated a typical RCT that is undertaken to compare a binary outcome
-(e.g. death) between a group of treated patients in the treatment arm and a
+(e.g. death) between a group of patients in the treatment arm and a
 group of untreated patients in the control arm. For each patient we generated 8
@@
 
-The sample size for the base scenarios was set to 4,250 ($80\%$ power for the
-detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
-sample sizes of 1,063 (4,250 divided by 4) and 17,000 (4250 multiplied by 4),
+The sample size for the base scenarios was set to 4,250, since this sample size
+provides $80\%$ power for the detection of a marginal OR of 0.8 with the
+standard alpha of 0.5\%. We evaluated the effect of smaller or larger sample
+sizes of 1,063 (4,250 divided by 4) and 17,000 (4250 multiplied by 4),
 respectively. We also evaluated the effect of worse or better discriminative
@@
 arm and patients in the treated arm. We considered this approach as a reference,
-expecting it to perform worse than the other candidates, as its objective is not
-to individualize benefit prediction.
+expecting it to perform worse than the other candidates, as its objective is to
+provide an illustration of HTE rather than to optimize individualized benefit
+predictions.
 
@@
 effect model, a model with a linear interaction with treatment and RCS models
-with 3, 4 and 5 knots.
+with 3, 4 and 5 knots. The extra degrees of freedom were 1 (linear interaction),
+2, 3 and 4 (RCS models) for these increasingly complex interactions with the
+treatment effect.
 
@@
 
-The linear interaction model outperformed all RCS methods in terms of
-RMSE in scenarios with true constant relative treatment effect (OR =
-0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
-quadratic deviations from a constant relative treatment effect (Figure
-\ref{fig:rmsebase}; panels A-C). However, with non-monotonic
-deviations from a constant relative treatment effect, the RMSE of the
-linear interaction model increased substantially, especially in the
-presence of treatment-related harms (Figure \ref{fig:rmsebase}; panel
-D). In these scenarios, RCS-3 outperformed all other methods in terms
-of RMSE. The constant treatment effect approach had overall best
-performance under true constant treatment effect settings, but was
-sensitive to all considered deviations, resulting in increased
+The linear interaction model outperformed all RCS methods in terms of RMSE in
+scenarios with true constant relative treatment effect (OR = 0.8, N = `r 4250`
+and AUC = 0.75), strong linear and even strong quadratic deviations from a
+constant relative treatment effect (Figure \ref{fig:rmsebase}; panels
+A-C). However, with non-monotonic deviations from a constant relative treatment
+effect, the RMSE of the linear interaction model increased substantially,
+especially in the presence of treatment-related harms (Figure
+\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
+methods in terms of RMSE. As might be expected the constant treatment effect
+approach had overall best performance under true constant treatment effect
+settings. It was sensitive to all considered deviations, resulting in increased
 RMSE. Finally, the adaptive approach had comparable performance to the
-best-performing method in each scenario. However, in comparison with
-the best-performing approach, its RMSE was more variable across the
-500 replications in the scenarios with linear and non-monotonic
-deviations, especially when also including moderate or strong
-treatment-related harms. This is caused by wrongly selecting the
-constant treatment effect model in a substantial proportion of the
-replications (Supplement, Figure S1).
+best-performing method in each scenario. However, in comparison with the
+best-performing approach, its RMSE was more variable in the scenarios with
+linear and non-monotonic deviations, especially when also including moderate or
+strong treatment-related harms. On closer inspection, we found that this
+behavior was caused by wrongly selecting the constant
+treatment effect model in a substantial proportion of the replications
+(Supplement, Figure S1). This problematic behavior was less with larger sample
+sizes (see below).
 
-```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
+```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000.
 The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250.
 The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects.
-Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis).
+Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis).
 The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.", fig.show="hold", out.width = '100%'}
@@
 
-Increasing the sample size to `r 17000` favored RCS-3 the most, as it
-achieved lowest or close to lowest RMSE across all scenarios (Figure
+Increasing the sample size to `r 17000` favored RCS-3 the most, It achieved
+lowest or close to lowest RMSE across all scenarios (Figure
 \ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
@@
 
-```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
+```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
@@
 
-When we increased the AUC of the true prediction model to 0.85 (OR =
-0.8 and N = `r 4250`). RCS-3 had the lowest RMSE in the case of strong
-quadratic or non-monotonic deviations and very comparable performance to
-the -- optimal -- linear interaction model in the case of strong
-linear deviations (median RMSE
+When we increased the AUC of the true prediction model to 0.85 (OR = 0.8 and N =
+`r 4250`). RCS-3 had the lowest RMSE in the case of strong quadratic or
+non-monotonic deviations and very comparable performance to the -- optimal --
+linear interaction model in the case of strong linear deviations (median RMSE
 `r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
 `r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
-for the linear interaction model). As observed in the base case
-scenario the adaptive approach wrongly selected the constant treatment
-effect model
+for the linear interaction model). As observed in the base case scenario the
+adaptive approach wrongly selected the constant treatment effect model
 (`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
@@
 `r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
-of the replications in the strong linear and non-monotonic deviation
-scenarios without treatment-related harms, respectively), leading to
-more variability of the RMSE (Supplement, Figure S3).
+of the replications in the strong linear and non-monotonic deviation scenarios
+without treatment-related harms, respectively), leading to more variability of
+the RMSE (Supplement, Figure S3).
 
-```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
+```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated samples 4,250.
+True prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
@@
 
-In comparison with the true approach, dsicrimination for benefit in
-the scenario with a constant relative treatment effect was only
-slightly lower for the linear interaction model, but substantially
-lower for the non-linear RCS approaches (Figure
-\ref{fig:discrimination}; panel A). With strong linear or quadratic
-deviations from a constant relative treatment effect, all methods
-discriminated quite similarly (Figure \ref{fig:discrimination}; panels
-B-C). In the scenario with non-monotonic deviations, the constant
-effect model had much lower discriminative ability compared to
-all other methods (median AUC of
+In comparison with the true approach, dsicrimination for benefit in the scenario
+with a constant relative treatment effect was only slightly lower for the linear
+interaction model, but substantially lower for the non-linear RCS approaches
+(Figure \ref{fig:discrimination}; panel A). With strong linear or quadratic
+deviations from a constant relative treatment effect, all methods discriminated
+quite similarly (Figure \ref{fig:discrimination}; panels B-C). In the scenario
+with non-monotonic deviations, the constant effect model had much lower
+discriminative ability compared to all other methods (median AUC of
 `r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
@@
 `r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
-for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel
-D). The adaptive approach was unstable in terms of discrimination for
-benefit, especially in the presence of treamtnt-related harms. With
-increasing number of RCS knots, we observed decreasing median values
-and increasing variability of the c-for-benefit in all scenarios. When
-we increased the sample size to 17,000 we observed similar trends,
-however the performance of all methods was more stable (Supplement,
-Figure S4). Finally, when we increased the true prediction AUC to 0.85
-the adaptive approach in the case of non-monotonic deviations was,
-again, quite unstable, especially with null or moderate
-treatment-related harms (Supplement, Figure S5). In these scenarios,
-the adaptive approach tended to select more often the much inferior
-constant treatment effect method (Supplement, Figure S3)
+for the best-performing RCS-3; Figure \ref{fig:discrimination}; panel D). The
+adaptive approach was unstable in terms of discrimination for benefit,
+especially in the presence of treatment-related harms. With increasing number of
+RCS knots, we observed decreasing median values and increasing variability of
+the c-for-benefit in all scenarios. When we increased the sample size to 17,000
+we observed similar trends, however the performance of all methods was more
+stable (Supplement, Figure S4). Finally, when we increased the true prediction
+AUC to 0.85 the adaptive approach in the case of non-monotonic deviations was,
+again, quite unstable, especially with null or moderate treatment-related harms
+(Supplement, Figure S5). In these scenarios, the adaptive approach tended to
+select more often the inferior constant treatment effect method (Supplement,
+Figure S3)
 
-```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
+```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250.
+True prediction AUC of 0.75.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
@@
 
-In terms of calibration for benefit, the constant effects model
-outperformed all other models in the scenario with true constant
-treatment effects, but was miscalibrated for all deviation scenarios
-(Figure \ref{fig:calibration}). The linear interaction model showed
-best or close to best calibration across all scenarios and only showed
-worse calibration compared to RCS-3 in case of non-monotonic
-deviations and treatment-related harms (Figure \ref{fig:calibration};
-panel D). The adaptive approach was worse calibrated in scenarios with
-strong linear and non-monotonic deviations compared to the linear
-interaction model and RCS-3. When we increased sample size to 17,000
-similar conclusions on calibration for benefit could be drawn. As
-expected, all methods displayed more stable calibration performance
-due to the larger number of patients (Supplement, Figure S6). When we
-increased the true prediction AUC to 0.85, the linear interaction
-model was worse calibrated, on average, than RCS-3 in the case of
-strong quadratic deviations from constant relative treatment effects
-(Supplement, Figure S7).
+In terms of calibration for benefit, the constant effects model outperformed all
+other models in the scenario with true constant treatment effects, but was
+miscalibrated for all deviation scenarios (Figure \ref{fig:calibration}). The
+linear interaction model showed best or close to best calibration across all
+scenarios and only showed worse calibration compared to RCS-3 in case of
+non-monotonic deviations and treatment-related harms (Figure
+\ref{fig:calibration}; panel D). The adaptive approach was worse calibrated in
+scenarios with strong linear and non-monotonic deviations compared to the linear
+interaction model and RCS-3. When we increased sample size to 17,000 similar
+conclusions on calibration for benefit could be drawn. As expected, all methods
+displayed more stable calibration performance due to the larger number of
+patients (Supplement, Figure S6). When we increased the true prediction AUC to
+0.85, the linear interaction model was worse calibrated, on average, than RCS-3
+in the case of strong quadratic deviations from constant relative treatment
+effects (Supplement, Figure S7).
 
@@
 
+gusto <- gusto %>%
+  tibble() %>%
+  filter(tx != "SK+tPA") %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
 treatmentArms <- gusto %>%
-  group_by(tpa) %>%
+  group_by(treatment) %>%
   summarise(n = n())
@@
 
-We demonstrate the different methods for individualizing treatment
-benefits using data from `r nrow(gusto)` patients with an acute
-myocardial infarction (MI) included in the GUSTO-I trial.
-`r treatmentArms %>% filter(tpa == 1) %>% select(n)` patients were randomized to
-tissue plasminogen activator (tPA) treatment and
-`r treatmentArms %>% filter(tpa == 0) %>% select(n)` were randomized to
-streptokinase. The outcome of interest was 30-day mortality, recorded
-for all patients.
+We demonstrate the different methods for individualizing treatment benefits
+using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
+included in the GUSTO-I trial.
+`r treatmentArms %>% filter(treatment == 1) %>% select(n)` 
+patients were randomized to tissue plasminogen activator (tPA) treatment and
+`r treatmentArms %>% filter(treatment == 0) %>% select(n)`
+were randomized to streptokinase. The outcome of interest was 30-day mortality
+(total of `r sum(gusto$outcome)` events), recorded for all patients.
 
-In line with previous analyses [@Califf1997; @Steyerberg2000], we
-fitted a logistic regression model with 6 baseline covariates,
-i.e. age, Killip class, systolic blood pressure, heart rate, an
-indicator of previous MI, and the location of MI, to predict 30-day
-mortality risk. A constant effect of treatment was included in the
-model. When deriving risk predictions for individuals we set the
-treatment indicator to 0. More information on model development can be
-found in the supplement.
+In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
+logistic regression model with 6 baseline covariates, i.e. age, Killip class,
+systolic blood pressure, heart rate, an indicator of previous MI, and the
+location of MI, to predict 30-day mortality risk. A constant effect of treatment
+was included in the model. When deriving risk predictions for individuals we set
+the treatment indicator to 0. More information on model development can be found
+in the supplement.
 
@@
 
-We used the risk linear predictor to fit the proposed methods under
-study for individualizing absolute benefit predictions. All methods
-predicted increasing benefits for patients with higher baseline risk
-predictions, but the fitted patterns were clearly different. The
-adaptive approach selected the model with RCS smoothing with 4
-knots. However, for very low baseline risk the decreasing predicted
-benefit with increasing risk may be somewhat too flexible. The more
-robust models, the linear interaction model or the model with RCS
-smoothing (3 knots), gave very similar benefit predictions, followed
-the evolution of the stratified estimates very closely and may
-therefore be preferable for use in clinical practice. The linear
-interaction model had somewhat lower AIC compared to the model with
-RCS smoothing (3 knots), slightly better cross-validated
-discrimination (c-for-benefit
+We used the risk linear predictor to fit the proposed methods under study for
+individualizing absolute benefit predictions. All methods predicted increasing
+benefits for patients with higher baseline risk predictions, but the fitted
+patterns were somewhat different. The adaptive approach selected the model with
+RCS smoothing with 4 knots. However, for very low baseline risk the decreasing
+predicted benefit with increasing risk may be somewhat too flexible. The more
+robust models, such as the linear interaction model and the model with RCS
+smoothing (3 knots), gave very similar benefit predictions. These models
+followed the evolution of the stratified estimates very closely and may
+therefore be preferable for use in clinical practice. The linear interaction
+model had somewhat lower AIC compared to the model with RCS smoothing (3 knots),
+slightly better cross-validated discrimination (c-for-benefit
 `r round(discriminationPerformance[2, ], 3)` vs
@@
 
-The linear interaction model and the RCS-3 model both displayed very
-good performance under many of the considered simulation scenarios, in
-contrast with the constant relative treatment effect model. The linear
-interaction model was optimal in cases with smaller sample sizes and
-moderately performing baseline risk prediction models, that is, it had
-lower RMSE, was better calibrated for benefit and had better
-discrimination for benefit, even in scenarios with strong quadratic
-deviations. In scenarios with true non-monotonic deviations, the
-linear interaction model was outperformed by RCS-3, especially in the
-presence of true treatment-related harms. Increasing the sample size
-or the prediction model’s discriminative ability favored RCS-3,
-especially in scenarios with non-monotonic deviations and in the
-presence of treatment-related harms.
+The linear interaction model and the RCS-3 model both displayed very good
+performance under many of the considered simulation scenarios, in contrast with
+the constant relative treatment effect model. The linear interaction model was
+optimal in cases with smaller sample sizes and moderately performing baseline
+risk prediction models, that is, it had lower RMSE, was better calibrated for
+benefit and had better discrimination for benefit, even in scenarios with strong
+quadratic deviations. In scenarios with true non-monotonic deviations, the
+linear interaction model was outperformed by RCS-3, especially in the presence
+of true treatment-related harms. Increasing the sample size or the prediction
+model’s discriminative ability favored RCS-3, especially in scenarios with
+non-monotonic deviations and in the presence of treatment-related harms.
 
-RCS-4 and RCS-5 proved to be too flexible in all considered scenarios,
-as indicated by higher RMSE, increased variability of discrimination
-for benefit and worse calibration of benefit predictions. Even with
-larger sample sizes and strong quadratic or non-monotonic deviations
-from the base case scenario of constant relative treatment effects,
-these more flexible restricted cubic splines did not outperform the
-simpler RCS-3 These approaches may only be helpful if we expect more
-extreme patterns of heterogeneous treatment effects compared to the
-quadratic deviations considered here.
+RCS-4 and RCS-5 proved to be too flexible in all considered scenarios, as
+indicated by higher RMSE, increased variability of discrimination for benefit
+and worse calibration of benefit predictions. Even with larger sample sizes and
+strong quadratic or non-monotonic deviations from the base case scenario of
+constant relative treatment effects, these more flexible restricted cubic
+splines did not outperform the simpler RCS-3 These approaches may only be
+helpful if we expect more extreme patterns of heterogeneous treatment effects
+compared to the quadratic deviations considered here. Considering interactions
+in RCS-3 models as the most complex approach often may be reasonable.
 
-The constant treatment effect model, despite having adequate
-performance in the presence of weak treatment effect heterogeneity on
-the relative scale, quickly broke down with stronger deviations from
-constant relative treatment effects. In these cases, the stratified
-approach generally had lower error rates compared to the constant
-treatment effect model. Stepwise treatment benefit estimates are very
-useful for demonstrating treatment effect heterogeneity–because
-estimating treatment effect requires groups of patients rather than
-individual patients–but are not helpful for making individualized
-absolute benefit predictions.
+The constant treatment effect model, despite having adequate performance in the
+presence of weak treatment effect heterogeneity on the relative scale, quickly
+broke down with stronger deviations from constant relative treatment effects. In
+these cases, the stratified approach generally had lower error rates compared to
+the constant treatment effect model. Such stepwise treatment benefit estimates
+are useful for visually demonstrating treatment effect heterogeneity but may be
+considered insufficient for making individualized benefit predictions.
 
-Increasing the discriminative ability of the risk model–by increasing
-the predictor coefficients of the true risk model–reduced RMSE for all
-methods. This increase in discriminative ability translates in higher
-variability of predicted risks, which, in turn, allows the considered
-methods to better capture absolute treatment benefits. As a
-consequence, the increase in discriminative ability of the risk model
-also led to higher values of c-for-benefit. Even though risk model
-performance is very important for the ability of risk-based methods to
-predict treatment benefit, prediction model development was outside
-the scope of this work and has already been studied extensively
-[@vanKlaveren2019; @Burke2014; @Abadie2018].
+Increasing the discriminative ability of the risk model–by increasing the
+predictor coefficients of the true risk model–reduced RMSE for all methods. This
+increase in discriminative ability translates in higher variability of predicted
+risks, which, in turn, allows the considered methods to better capture absolute
+treatment benefits. As a consequence, the increase in discriminative ability of
+the risk model also led to higher discrimination between those with low or high
+benefit (as reflected in values of c-for-benefit). Even though risk model
+performance is very important for the ability of risk-based methods to predict
+treatment benefit, prediction model development was outside the scope of this
+work and has already been studied extensively [@vanKlaveren2019; @Burke2014;
+@Abadie2018].
 
-The adaptive approach had adequate performance, following closely on
-average the performance of the “true” model in most
-scenarios. However, with smaller sample sizes it tended to “miss” the
-treatment-risk interactions and selected simpler models (Supplementary
-Table S7). This resulted in increased RMSE variability in these
-scenarios, especially in the case of true strong linear or
-non-monotonic deviations from the base case scenario. Therefore, in
-the case of smaller sample sizes the simpler linear interaction model
-is a safer choice for predicting absolute benefits.
+The adaptive approach had adequate performance, following closely on average the
+performance of the “true” model in most scenarios. With smaller sample sizes it
+tended to miss the treatment-risk interactions and selected simpler models
+(Supplementary Table S7). This conservative behavior resulted in increased RMSE
+variability in these scenarios, especially in the case of true strong linear or
+non-monotonic deviations from the base case scenario. Therefore, in the case of
+smaller sample sizes the simpler linear interaction model may be a safer choice
+for predicting absolute benefits, if a non-constant treatment effect is
+suspected.
 
-Risk-based approaches to predictive HTE estimate treatment benefit as
-a function of baseline risk. A limitation of our study is that we
-assumed treatment benefit to be a function of baseline risk in the
-majority of the simulation scenarios. We attempted to address that by
-introducing constant moderate and strong treatment-related harms,
-applied on the absolute scale. Also, we considered a small number of
-scenarios with true treatment-covariate interactions, in which our
-main conclusions remained the same (Supplement, XX). Future simulation
-studies could explore the effect of more extensive deviations from
-risk-based treatment effects.
+Risk-based approaches to predictive HTE estimate treatment benefit as a function
+of baseline risk. A limitation of our study is that we assumed treatment benefit
+to be a function of baseline risk in the majority of the simulation
+scenarios. We also considered constant moderate and strong treatment-related
+harms, applied on the absolute scale to expand the range of scenarios in line
+with previous work (REF!!). Also, we considered a small number of scenarios with
+true treatment-covariate interactions, in which our main conclusions remained
+the same (Supplement, XX). Future simulation studies could explore the effect of
+more extensive deviations from risk-based treatment effects.
 
-Recent years have seen an increased interest in predictive HTE
-approaches focusing on individualized benefit predictions. In our
-simulations we only focused on risk-based methods, using baseline risk
-as a reference in a two-stage approach to individualizing benefit
-predictions. However, there is a plethora of different methods,
-ranging from treatment effect modeling to tree-based approaches
-available in more recent literature [@Athey2019; @Lu2018;@Wager2018].
-Simulations are also needed to assess relative performance and define
-the settings where these break down or outperform each other.
+Recent years have seen an increased interest in predictive HTE approaches
+focusing on individualized benefit predictions. In our simulations we only
+focused on risk-based methods, using baseline risk as a reference in a two-stage
+approach to individualizing benefit predictions. However, there is a plethora of
+different methods, ranging from treatment effect modeling to tree-based
+approaches available in more recent literature [@Athey2019; @Lu2018;@Wager2018].
+Simulations are also needed to assess relative performance and define the
+settings where these break down or outperform each other.
 
-In conclusion, the linear interaction approach is a viable option with
-smaller sample sizes and/or moderately performing risk prediction
-modelsRCS-3 is a better option when non-monotonic deviations from a
-constant relative treatment effect and/or substantial
-treatment-related harms are anticipated. With larger sample size, an
-adaptive approach -- selecting the method with optimal AIC -- can also
-be considered as an automated alternative.
+In conclusion, the linear interaction approach is a viable option with smaller
+sample sizes and/or moderately performing risk prediction models if we consider
+a non-constant relative treatment effect plausible. RCS-3 is a better option
+when non-monotonic deviations from a constant relative treatment effect and/or
+substantial treatment-related harms are anticipated. With larger sample size, an
+adaptive approach -- selecting the method with optimal AIC -- can also be
+considered as an automated alternative.
 

### 2022-01-26 submission/manuscript.tex
add treatment interaction scenarios

@@
             pdfauthor={},
-            pdftitle={An interaction between baseline risk and treatment assignment generally imporoved treatment effect predictions},
+            pdftitle={Individualized treatment effect was predicted best by modeling baseline risk in interaction with treatment assignment},
             colorlinks=false,
@@
 
-  \title{An interaction between baseline risk and treatment assignment
-generally imporoved treatment effect predictions}
+  \title{Individualized treatment effect was predicted best by modeling
+baseline risk in interaction with treatment assignment}
     \author[1]{Alexandros Rekkas}
@@
   \begin{abstract}
-  \textbf{Objective}: To compare different risk-based methods predicting
-  individualized treatment effects in RCTs. \textbf{Study Design and
-  Setting}: We simulated RCT data using diverse assumptions for the
-  average treatment effect, a baseline prognostic index of risk (PI),
-  the shape of its interaction with treatment (none, linear, quadratic
-  or non-monotonic) and the magnitude of treatment-related harms. In
+  \textbf{Objective}: To compare different risk-based methods for
+  optimal prediction of individualized treatment effects from RCTs.
+  \textbf{Study Design and Setting}: We simulated RCT data using diverse
+  assumptions for the average treatment effect, a baseline prognostic
+  index of risk (PI), the shape of its interaction with treatment (none,
+  linear, quadratic or non-monotonic) and the magnitude of
+  treatment-related harms (none or constant independent of the PI). In
   each sample we predicted absolute benefit using: models with a
@@
   treatment effect were stronger, and when sample size was larger.
-  Larger sample size also supported the adaptive approach.
-  \textbf{Conclusion}: An interaction between risk and treatment
-  assignment generally improved treatment effect predictions. Non-linear
-  RCS interactions should be considered for larger sample size.
+  Larger sample size also supported the adaptive approach. Illustrations
+  in the GUSTO-I trial confirmed these findings. \textbf{Conclusion}: An
+  interaction between baseline risk and treatment assignment should be
+  considered to improve treatment effect predictions. Non-linear RCS
+  interactions should be considered only in larger sample sizes.
   \end{abstract}
@@
 similarity when making individualized predictions or recommendations
-{[}2{]}. Risk-modeling approaches use prediction of baseline risk as the
-reference; treatment effect modeling approaches also model
-treatment-covariate interactions, in addition to risk factors; optimal
-treatment regime approaches focus on developing treatment assignment
-rules and therefore rely heavily on modeling treatment effect modifiers.
+{[}2{]}. First, risk-modeling approaches use prediction of baseline risk
+as the reference; second, treatment effect modeling approaches also
+model treatment-covariate interactions, in addition to risk factors;
+third, optimal treatment regime approaches focus on developing treatment
+assignment rules and rely heavily on modeling treatment effect
+modifiers. A key difference between these approaches is their parsimony
+in dealing the treatment effect modifiers, with no interaction
+considered (risk modeling), a limited number of interactions (effect
+modeling), or a larger set of interactions (optimal treatment regime
+approaches).
 
@@
 even though estimates at the risk subgroup level may be accurate, these
-estimates do not apply to individual patients, especially for patients
-with predicted risk at the boundaries of the risk intervals. Hence, the
-risk-stratified approach is useful for exploring and presenting HTE, but
-is not useful for supporting treatment decisions for individual
-patients.
+estimates may need further refinement for individual patients,
+especially for patients with predicted risk at the boundaries of the
+risk intervals. Hence, the risk-stratified approach is useful for
+exploring and presenting HTE, but is not sufficient for supporting
+treatment decisions for individual patients.
 
@@
 treatment effects. We simulated RCT settings to compare the performance
-of these models under different assumptions of the relationship between
+of these models under different assumptions of the relation between
 baseline risk and treatment. We illustrated the different models by a
-case study of predicting individualized effects of tissue plasminogen
-activator (tPA) versus streptokinase treatment in patients with an acute
-myocardial infarction (MI).
+case study of predicting individualized effects of treatments for acute
+myocardial infarction (MI) in a large randomized controlled trial (RCT).
 
@@
 We simulated a typical RCT that is undertaken to compare a binary
-outcome (e.g.~death) between a group of treated patients in the
-treatment arm and a group of untreated patients in the control arm. For
-each patient we generated 8 baseline covariates
-\(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
-Treatment was allocated using a 50:50 split. Outcomes for patients in
-the control arm were generated from a logistic regression model
-including all baseline covariates. In the base scenarios coefficient
-values were such, that the AUC of the logistic regression model was
-\(0.75\) and the event rate in the control arm was \(20\%\). Binary
-outcomes in the control arm were generated from Bernoulli variables with
-true probabilities
+outcome (e.g.~death) between a group of patients in the treatment arm
+and a group of untreated patients in the control arm. For each patient
+we generated 8 baseline covariates \(x_1,\dots,x_4\sim N(0, 1)\) and
+\(x_5,\dots,x_8\sim B(1, 0.2)\). Treatment was allocated using a 50:50
+split. Outcomes for patients in the control arm were generated from a
+logistic regression model including all baseline covariates. In the base
+scenarios coefficient values were such, that the AUC of the logistic
+regression model was \(0.75\) and the event rate in the control arm was
+\(20\%\). Binary outcomes in the control arm were generated from
+Bernoulli variables with true probabilities
 \(P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\).
@@
 
-The sample size for the base scenarios was set to 4,250 (\(80\%\) power
-for the detection of a marginal OR of 0.8). We evaluated the effect of
+The sample size for the base scenarios was set to 4,250, since this
+sample size provides \(80\%\) power for the detection of a marginal OR
+of 0.8 with the standard alpha of 0.5\%. We evaluated the effect of
 smaller or larger sample sizes of 1,063 (4,250 divided by 4) and 17,000
@@
 expecting it to perform worse than the other candidates, as its
-objective is not to individualize benefit prediction.
+objective is to provide an illustration of HTE rather than to optimize
+individualized benefit predictions.
 
@@
 constant treatment effect model, a model with a linear interaction with
-treatment and RCS models with 3, 4 and 5 knots.
+treatment and RCS models with 3, 4 and 5 knots. The extra degrees of
+freedom were 1 (linear interaction), 2, 3 and 4 (RCS models) for these
+increasingly complex interactions with the treatment effect.
 
@@
 treatment-related harms (Figure \ref{fig:rmsebase}; panel D). In these
-scenarios, RCS-3 outperformed all other methods in terms of RMSE. The
-constant treatment effect approach had overall best performance under
-true constant treatment effect settings, but was sensitive to all
-considered deviations, resulting in increased RMSE. Finally, the
-adaptive approach had comparable performance to the best-performing
-method in each scenario. However, in comparison with the best-performing
-approach, its RMSE was more variable across the 500 replications in the
-scenarios with linear and non-monotonic deviations, especially when also
-including moderate or strong treatment-related harms. This is caused by
-wrongly selecting the constant treatment effect model in a substantial
-proportion of the replications (Supplement, Figure S1).
+scenarios, RCS-3 outperformed all other methods in terms of RMSE. As
+might be expected the constant treatment effect approach had overall
+best performance under true constant treatment effect settings. It was
+sensitive to all considered deviations, resulting in increased RMSE.
+Finally, the adaptive approach had comparable performance to the
+best-performing method in each scenario. However, in comparison with the
+best-performing approach, its RMSE was more variable in the scenarios
+with linear and non-monotonic deviations, especially when also including
+moderate or strong treatment-related harms. On closer inspection, we
+found that this behavior was caused by wrongly selecting the constant
+treatment effect model in a substantial proportion of the replications
+(Supplement, Figure S1). This problematic behavior was less with larger
+sample sizes (see below).
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsebase-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsebase-1} \caption{RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000.
 The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250.
 The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects.
-Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis).
+Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis).
 The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.}\label{fig:rmsebase}
@@
 
-Increasing the sample size to 17,000 favored RCS-3 the most, as it
-achieved lowest or close to lowest RMSE across all scenarios (Figure
+Increasing the sample size to 17,000 favored RCS-3 the most, It achieved
+lowest or close to lowest RMSE across all scenarios (Figure
 \ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsesamplesize-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-Sample size 17,000 rather than 4250 in Figure \ref{fig:rmsebase}}\label{fig:rmsesamplesize}
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsesamplesize-1} \caption{RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \ref{fig:rmsebase}}\label{fig:rmsesamplesize}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmseauc-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-True prediction AUC of 0.85 and sample size of 4,250.}\label{fig:rmseauc}
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmseauc-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated samples 4,250.
+True prediction AUC of 0.85.}\label{fig:rmseauc}
 \end{figure}
@@
 in terms of discrimination for benefit, especially in the presence of
-treamtnt-related harms. With increasing number of RCS knots, we observed
-decreasing median values and increasing variability of the c-for-benefit
-in all scenarios. When we increased the sample size to 17,000 we
-observed similar trends, however the performance of all methods was more
-stable (Supplement, Figure S4). Finally, when we increased the true
-prediction AUC to 0.85 the adaptive approach in the case of
-non-monotonic deviations was, again, quite unstable, especially with
-null or moderate treatment-related harms (Supplement, Figure S5). In
-these scenarios, the adaptive approach tended to select more often the
-much inferior constant treatment effect method (Supplement, Figure S3)
+treatment-related harms. With increasing number of RCS knots, we
+observed decreasing median values and increasing variability of the
+c-for-benefit in all scenarios. When we increased the sample size to
+17,000 we observed similar trends, however the performance of all
+methods was more stable (Supplement, Figure S4). Finally, when we
+increased the true prediction AUC to 0.85 the adaptive approach in the
+case of non-monotonic deviations was, again, quite unstable, especially
+with null or moderate treatment-related harms (Supplement, Figure S5).
+In these scenarios, the adaptive approach tended to select more often
+the inferior constant treatment effect method (Supplement, Figure S3)
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/discrimination-1} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
-True prediction AUC of 0.75 and sample size of 4,250.}\label{fig:discrimination}
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/discrimination-1} \caption{Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250.
+True prediction AUC of 0.75.}\label{fig:discrimination}
 \end{figure}
@@
 We demonstrate the different methods for individualizing treatment
-benefits using data from 30,510 patients with an acute myocardial
+benefits using data from 30,510 patients with acute myocardial
 infarction (MI) included in the GUSTO-I trial. 10,348 patients were
@@
 were randomized to streptokinase. The outcome of interest was 30-day
-mortality, recorded for all patients.
+mortality (total of 2,128 events), recorded for all patients.
 
@@
 predicted increasing benefits for patients with higher baseline risk
-predictions, but the fitted patterns were clearly different. The
+predictions, but the fitted patterns were somewhat different. The
 adaptive approach selected the model with RCS smoothing with 4 knots.
@@
 with increasing risk may be somewhat too flexible. The more robust
-models, the linear interaction model or the model with RCS smoothing (3
-knots), gave very similar benefit predictions, followed the evolution of
-the stratified estimates very closely and may therefore be preferable
-for use in clinical practice. The linear interaction model had somewhat
-lower AIC compared to the model with RCS smoothing (3 knots), slightly
-better cross-validated discrimination (c-for-benefit 0.526 vs 0.525) and
-quite similar cross-validated calibration (ICI-for benefit 0.0115 vs
-0.0117).
+models, such as the linear interaction model and the model with RCS
+smoothing (3 knots), gave very similar benefit predictions. These models
+followed the evolution of the stratified estimates very closely and may
+therefore be preferable for use in clinical practice. The linear
+interaction model had somewhat lower AIC compared to the model with RCS
+smoothing (3 knots), slightly better cross-validated discrimination
+(c-for-benefit 0.526 vs 0.525) and quite similar cross-validated
+calibration (ICI-for benefit 0.0115 vs 0.0117).
 
@@
 of heterogeneous treatment effects compared to the quadratic deviations
-considered here.
+considered here. Considering interactions in RCS-3 models as the most
+complex approach often may be reasonable.
 
@@
 generally had lower error rates compared to the constant treatment
-effect model. Stepwise treatment benefit estimates are very useful for
-demonstrating treatment effect heterogeneity--because estimating
-treatment effect requires groups of patients rather than individual
-patients--but are not helpful for making individualized absolute benefit
-predictions.
+effect model. Such stepwise treatment benefit estimates are useful for
+visually demonstrating treatment effect heterogeneity but may be
+considered insufficient for making individualized benefit predictions.
 
@@
 the increase in discriminative ability of the risk model also led to
-higher values of c-for-benefit. Even though risk model performance is
-very important for the ability of risk-based methods to predict
-treatment benefit, prediction model development was outside the scope of
-this work and has already been studied extensively {[}5,13,14{]}.
+higher discrimination between those with low or high benefit (as
+reflected in values of c-for-benefit). Even though risk model
+performance is very important for the ability of risk-based methods to
+predict treatment benefit, prediction model development was outside the
+scope of this work and has already been studied extensively
+{[}5,13,14{]}.
 
 The adaptive approach had adequate performance, following closely on
-average the performance of the ``true'' model in most scenarios.
-However, with smaller sample sizes it tended to ``miss'' the
-treatment-risk interactions and selected simpler models (Supplementary
-Table S7). This resulted in increased RMSE variability in these
-scenarios, especially in the case of true strong linear or non-monotonic
-deviations from the base case scenario. Therefore, in the case of
-smaller sample sizes the simpler linear interaction model is a safer
-choice for predicting absolute benefits.
+average the performance of the ``true'' model in most scenarios. With
+smaller sample sizes it tended to miss the treatment-risk interactions
+and selected simpler models (Supplementary Table S7). This conservative
+behavior resulted in increased RMSE variability in these scenarios,
+especially in the case of true strong linear or non-monotonic deviations
+from the base case scenario. Therefore, in the case of smaller sample
+sizes the simpler linear interaction model may be a safer choice for
+predicting absolute benefits, if a non-constant treatment effect is
+suspected.
 
@@
 treatment benefit to be a function of baseline risk in the majority of
-the simulation scenarios. We attempted to address that by introducing
-constant moderate and strong treatment-related harms, applied on the
-absolute scale. Also, we considered a small number of scenarios with
-true treatment-covariate interactions, in which our main conclusions
-remained the same (Supplement, XX). Future simulation studies could
-explore the effect of more extensive deviations from risk-based
-treatment effects.
+the simulation scenarios. We also considered constant moderate and
+strong treatment-related harms, applied on the absolute scale to expand
+the range of scenarios in line with previous work (REF!!). Also, we
+considered a small number of scenarios with true treatment-covariate
+interactions, in which our main conclusions remained the same
+(Supplement, XX). Future simulation studies could explore the effect of
+more extensive deviations from risk-based treatment effects.
 
@@
 In conclusion, the linear interaction approach is a viable option with
-smaller sample sizes and/or moderately performing risk prediction
-modelsRCS-3 is a better option when non-monotonic deviations from a
-constant relative treatment effect and/or substantial treatment-related
-harms are anticipated. With larger sample size, an adaptive approach --
-selecting the method with optimal AIC -- can also be considered as an
-automated alternative.
+smaller sample sizes and/or moderately performing risk prediction models
+if we consider a non-constant relative treatment effect plausible. RCS-3
+is a better option when non-monotonic deviations from a constant
+relative treatment effect and/or substantial treatment-related harms are
+anticipated. With larger sample size, an adaptive approach -- selecting
+the method with optimal AIC -- can also be considered as an automated
+alternative.
 

### 2022-01-26 submission/supplement.rmd
add treatment interaction scenarios

@@
 geometry: margin=1.0in
-toc: false
+toc: true
 font-size: 12pt
@@
 
-```{r discriminationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r discriminationSampleSize, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
@@
 
-```{r discriminationAuc, cache=FALSE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r discriminationAuc, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
@@
 
-```{r calibrationSampleSize, cache=FALSE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r calibrationSampleSize, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
@@
 
-```{r calibrationAuc, cache=FALSE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
+```{r calibrationAuc, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000.
 True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
@@
 ```
+
+# Treatment interactions
+
+Here we present the root mean squared error of the considered methods
+when true interactions of the covariates with treatment are
+assumed. We ran our analyses on a limited set of scenarios. More
+specifically, true prediction AUC was set to 0.75 and sample size was
+set to 4,250. The simulation scenarios included 4 weak interactions
+($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong interactions
+($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2 weak and 2
+strong interactions.
+
+```{r rmseInteraction, cache=FALSE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
+AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```

### 2022-01-26 code/PlotRmseInteractions.R

@@ -1,0 +1,160 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description:
+#   Generates the RMSE plots for the base case
+# Input:
+#   - type
+#   - value
+# Output:
+#   - figures/rmse_base.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/rmse.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+#   - code/helpers/Absolute.R
+# ===================================================
+
+library(tidyverse)
+library(glue)
+library(ggtext)
+library(gridExtra)
+library(grid)
+library(ggside)
+
+source("code/helpers/CreateManuscriptPlots.R")     # Geneartes single boxplot
+source("code/helpers/PlotResult.R")                # Generates the boxplot list
+source("code/helpers/Absolute.R")                  # Generates the absolute plots
+
+scenarioIds <- readr::read_csv("data/processed/analysisIdsInteractions.csv")
+
+metric    <- "rmse"
+titles <- scenarioIds %>%
+  mutate(
+    title = str_to_sentence(glue("{str_replace_all(type, '-', ' ')} interactions")
+    )
+  ) %>%
+  select(title) %>%
+  unlist() %>%
+  unique()
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:3],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+limitsHigh <- .15
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  filter(scenarioId > 648) %>%
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+scenarios <- scenarioIds %>%
+  filter(harm == "absent") %>%
+  select(scenario) %>%
+  unlist()
+
+names(scenarios) <- NULL
+
+plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 10, 2.5))
+
+gridList <- list(
+  plotList[[1]] +
+    theme(
+      panel.grid.minor = element_blank(),
[…]

### 2022-01-26 code/WriteAnalysisIdsInteractions.R

@@ -3,19 +3,20 @@
 library(tidyverse)
 library(readr)
 
-effect <- c("weak", "strong", "mixed")
-typeOfEffect <- "interaction"
-n <- 4250
-auc <- 75
+base <- "interaction"
+type <- c("weak", "mixed", "strong")
+sampleSize <- 4250
+auc <- .75
+harm <- c("absent", "moderate-positive", "strong-positive", "negative")
 
-table <- tibble(
-  type       = typeOfEffect,
-  effectSize = effect,
-  sampleSize = n,
-  auc        = auc
-) %>%
+table <- expand_grid(base, type, sampleSize, auc, harm) %>%
   mutate(
-    b0 = -2.08,
+    # b0 = -2.08,
+    b0 = case_when(
+      type == "weak" ~ -3.05,
+      type == "mixed" ~ -3.32,
+      TRUE ~ -3.50
+    ),
     b1 = .49,
     b2 = .49,
     b3 = .49,
@@ -26,25 +27,25 @@ table <- tibble(
     b8 = .49,
     g0 = log(.8),
     g1 = case_when(
-      (effect == "weak" | effect == "mixed") ~ -.19,
-      (effect == "strong"                  ) ~ -.49
+      (type == "weak" | type == "mixed"  ) ~ -.19,
+      (type == "strong"                  ) ~ -.49
     ),
     g2 = case_when(
-      (effect == "weak"                      ) ~ -.19,
-      (effect == "mixed" | effect == "strong") ~ -.49
+      (type == "weak"                    ) ~ -.19,
+      (type == "mixed" | type == "strong") ~ -.49
     ),
     g5 = case_when(
-      (effect == "weak" | effect == "mixed") ~ -.19,
-      (effect == "strong"                  ) ~ -.49
+      (type == "weak" | type == "mixed"  ) ~ -.19,
+      (type == "strong"                  ) ~ -.49
     ),
     g6 = case_when(
-      (effect == "weak"                      ) ~ -.19,
-      (effect == "mixed" | effect == "strong") ~ -.49
+      (type == "weak"                    ) ~ -.19,
+      (type == "mixed" | type == "strong") ~ -.49
     ),
-    scenario = 63 + 1:n()
+    averageTrueBenefit = .029,
+    scenario = 648 + 1:n()
   ) %>%
   select(scenario, everything())
 
 # saveRDS(table, "data/processed/analysisIds.rds")
 readr::write_csv(table, "data/processed/analysisIdsInteractions.csv")
-

### 2022-01-31 submission/manuscript.rmd
working on updates following co-authors' comments

@@
 assign risk predictions to individual patients. For the simulations we developed
-a prediction model internally, using logistic regression including main effects
-for all baseline covariates and treatment assignment. Risk predictions for
-individual patients were based on treatment assignment to the control arm, that
-is setting treatment assignment to 0.
+a prediction model internally on the entire population, using a logistic
+regression model with main effects for all baseline covariates and treatment
+assignment. Risk predictions for individual patients were derived by setting
+treatment assignment to 0. Another common approach is to derive the prediction
+model solely on the control patients, however this approach has been shown to
+lead to biased benefit predictions [@vanKlaveren2019; @Burke2014; @Abadie2018].
 
@@
 
-Finally, we considered an adaptive approach using Akaike's Information Criterion
-(AIC) for model selection. The candidate models were: a constant treatment
-effect model, a model with a linear interaction with treatment and RCS models
-with 3, 4 and 5 knots. The extra degrees of freedom were 1 (linear interaction),
-2, 3 and 4 (RCS models) for these increasingly complex interactions with the
-treatment effect.
+Finally, we considered an *adaptive approach* using Akaike's Information
+Criterion (AIC) for model selection. More specifically, for the adaptive
+approach we ranked the constant relative treatment effect model, the linear
+interaction model, and the RCS models with 3, 4, and 5 knots based on their AIC
+and selected the one with the lowest value. The extra degrees of freedom were 1
+(linear interaction), 2, 3 and 4 (RCS models) for these increasingly complex
+interactions with the treatment effect.
 
@@
 smooth observed benefit. Values closer to $0$ represent better calibration.
+
+## Empirical illustration
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+load(here::here("data/raw/gusto.rda"))
+gusto <- gusto %>%
+  tibble() %>%
+  filter(!is.na(tpa))
+
+gusto <- gusto %>%
+  tibble() %>%
+  filter(tx != "SK+tPA") %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+treatmentArms <- gusto %>%
+  group_by(treatment) %>%
+  summarise(n = n())
+```
+
+We demonstrate the different methods for individualizing treatment benefits
+using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
+included in the GUSTO-I trial.
+`r treatmentArms %>% filter(treatment == 1) %>% select(n)` 
+patients were randomized to tissue plasminogen activator (tPA) treatment and
+`r treatmentArms %>% filter(treatment == 0) %>% select(n)`
+were randomized to streptokinase. The outcome of interest was 30-day mortality
+(total of `r sum(gusto$outcome)` events), recorded for all patients.
 
@@
 AUC to 0.85 the adaptive approach in the case of non-monotonic deviations was,
+
 again, quite unstable, especially with null or moderate treatment-related harms
@@
 
-## Case study
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-load(here::here("data/raw/gusto.rda"))
-gusto <- gusto %>%
-  tibble() %>%
-  filter(!is.na(tpa))
-
-gusto <- gusto %>%
-  tibble() %>%
-  filter(tx != "SK+tPA") %>%
-  rename(
-    "outcome" = "day30",
-    "treatment" = "tpa"
-  )
-
-treatmentArms <- gusto %>%
-  group_by(treatment) %>%
-  summarise(n = n())
-```
-
-We demonstrate the different methods for individualizing treatment benefits
-using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
-included in the GUSTO-I trial.
-`r treatmentArms %>% filter(treatment == 1) %>% select(n)` 
-patients were randomized to tissue plasminogen activator (tPA) treatment and
-`r treatmentArms %>% filter(treatment == 0) %>% select(n)`
-were randomized to streptokinase. The outcome of interest was 30-day mortality
-(total of `r sum(gusto$outcome)` events), recorded for all patients.
+## Empirical illustration
 
@@
 constant relative treatment effects, these more flexible restricted cubic
-splines did not outperform the simpler RCS-3 These approaches may only be
+splines did not outperform the simpler RCS-3. These approaches may only be
 helpful if we expect more extreme patterns of heterogeneous treatment effects
@@
 harms, applied on the absolute scale to expand the range of scenarios in line
-with previous work (REF!!). Also, we considered a small number of scenarios with
-true treatment-covariate interactions, in which our main conclusions remained
-the same (Supplement, XX). Future simulation studies could explore the effect of
-more extensive deviations from risk-based treatment effects.
+with previous work [@Glasziou1995]. Also, we considered a small number of
+scenarios with true treatment-covariate interactions, in which our main
+conclusions remained the same (Supplement, XX). Future simulation studies could
+explore the effect of more extensive deviations from risk-based treatment
+effects.
 
-Recent years have seen an increased interest in predictive HTE approaches
-focusing on individualized benefit predictions. In our simulations we only
-focused on risk-based methods, using baseline risk as a reference in a two-stage
-approach to individualizing benefit predictions. However, there is a plethora of
-different methods, ranging from treatment effect modeling to tree-based
-approaches available in more recent literature [@Athey2019; @Lu2018;@Wager2018].
-Simulations are also needed to assess relative performance and define the
-settings where these break down or outperform each other.
+In our simulations we only focused on risk-based methods, using baseline risk as
+a reference in a two-stage approach to individualizing benefit
+predictions. However, there is a plethora of different methods, ranging from
+treatment effect modeling to tree-based approaches available in more recent
+literature [@Athey2019; @Lu2018;@Wager2018;@powers2018some]. Many of these
+methods rely on incorporating treatment-covariate interactions in the prediction
+of benefit. An important caveat of such approaches is that they may be prone to
+overfitting, thus exaggerating the magnitude of the predicted benefits. In a
+wide range of simulation settings, a simpler risk modeling approach was
+consistently better calibrated for benefit compared to more complex treatment
+effect modelling approaches [@vanKlaveren2019]. Similarly, when SYNTAX score II,
+a model developed for identifying patients with complex coronary artery disease
+that benefit more from percutaneous coronary intervention or from coronary
+artery bypass grafting was redeveloped using fewer treatment-covariate
+interactions had better external performance compared to its
+predecessor[@farooq2013anatomical; @takahashi2020redevelopment].
 

### 2022-01-31 submission/manuscript.tex
working on updates following co-authors' comments

@@
 to assign risk predictions to individual patients. For the simulations
-we developed a prediction model internally, using logistic regression
-including main effects for all baseline covariates and treatment
-assignment. Risk predictions for individual patients were based on
-treatment assignment to the control arm, that is setting treatment
-assignment to 0.
+we developed a prediction model internally on the entire population,
+using a logistic regression model with main effects for all baseline
+covariates and treatment assignment. Risk predictions for individual
+patients were derived by setting treatment assignment to 0. Another
+common approach is to derive the prediction model solely on the control
+patients, however this approach has been shown to lead to biased benefit
+predictions {[}5,8,9{]}.
 
@@
 Fourth, we used \emph{restricted cubic splines} (RCS) to relax the
-linearity assumption on the effect of the linear predictor {[}8{]}. We
+linearity assumption on the effect of the linear predictor {[}10{]}. We
 considered splines with 3 (RCS-3), 4 (RCS-4) and 5 (RCS-5) knots to
@@
 
-Finally, we considered an adaptive approach using Akaike's Information
-Criterion (AIC) for model selection. The candidate models were: a
-constant treatment effect model, a model with a linear interaction with
-treatment and RCS models with 3, 4 and 5 knots. The extra degrees of
-freedom were 1 (linear interaction), 2, 3 and 4 (RCS models) for these
-increasingly complex interactions with the treatment effect.
+Finally, we considered an \emph{adaptive approach} using Akaike's
+Information Criterion (AIC) for model selection. More specifically, for
+the adaptive approach we ranked the constant relative treatment effect
+model, the linear interaction model, and the RCS models with 3, 4, and 5
+knots based on their AIC and selected the one with the lowest value. The
+extra degrees of freedom were 1 (linear interaction), 2, 3 and 4 (RCS
+models) for these increasingly complex interactions with the treatment
+effect.
 
@@
 We compared the discriminative ability of the methods under study using
-c-for-benefit {[}9{]}. The c-for-benefit represents the probability that
-from two randomly chosen matched patient pairs with unequal observed
-benefit, the pair with greater observed benefit also has a higher
-predicted benefit. To be able to calculate observed benefit, patients in
-each treatment arm are ranked based on their predicted benefit and then
-matched 1:1 across treatment arms. \emph{Observed} treatment benefit is
-defined as the difference of observed outcomes between the untreated and
-the treated patient of each matched patient pair. \emph{Predicted}
-benefit is defined as the average of predicted benefit within each
-matched patient pair.
+c-for-benefit {[}11{]}. The c-for-benefit represents the probability
+that from two randomly chosen matched patient pairs with unequal
+observed benefit, the pair with greater observed benefit also has a
+higher predicted benefit. To be able to calculate observed benefit,
+patients in each treatment arm are ranked based on their predicted
+benefit and then matched 1:1 across treatment arms. \emph{Observed}
+treatment benefit is defined as the difference of observed outcomes
+between the untreated and the treated patient of each matched patient
+pair. \emph{Predicted} benefit is defined as the average of predicted
+benefit within each matched patient pair.
 
 We evaluated calibration in a similar manner, using the integrated
-calibration index (ICI) for benefit {[}10{]}. The observed benefits are
+calibration index (ICI) for benefit {[}12{]}. The observed benefits are
 regressed on the predicted benefits using a locally weighted scatterplot
@@
 represent better calibration.
+
+\hypertarget{empirical-illustration}{%
+\subsection{Empirical illustration}\label{empirical-illustration}}
+
+We demonstrate the different methods for individualizing treatment
+benefits using data from 30,510 patients with acute myocardial
+infarction (MI) included in the GUSTO-I trial. 10,348 patients were
+randomized to tissue plasminogen activator (tPA) treatment and 20,162
+were randomized to streptokinase. The outcome of interest was 30-day
+mortality (total of 2,128 events), recorded for all patients.
 
@@
 increased the true prediction AUC to 0.85 the adaptive approach in the
-case of non-monotonic deviations was, again, quite unstable, especially
-with null or moderate treatment-related harms (Supplement, Figure S5).
-In these scenarios, the adaptive approach tended to select more often
-the inferior constant treatment effect method (Supplement, Figure S3)
+case of non-monotonic deviations was,
+
+again, quite unstable, especially with null or moderate
+treatment-related harms (Supplement, Figure S5). In these scenarios, the
+adaptive approach tended to select more often the inferior constant
+treatment effect method (Supplement, Figure S3)
 
@@
 
-\hypertarget{case-study}{%
-\subsection{Case study}\label{case-study}}
+\hypertarget{empirical-illustration-1}{%
+\subsection{Empirical illustration}\label{empirical-illustration-1}}
 
-We demonstrate the different methods for individualizing treatment
-benefits using data from 30,510 patients with acute myocardial
-infarction (MI) included in the GUSTO-I trial. 10,348 patients were
-randomized to tissue plasminogen activator (tPA) treatment and 20,162
-were randomized to streptokinase. The outcome of interest was 30-day
-mortality (total of 2,128 events), recorded for all patients.
-
-In line with previous analyses {[}11,12{]}, we fitted a logistic
+In line with previous analyses {[}13,14{]}, we fitted a logistic
 regression model with 6 baseline covariates, i.e.~age, Killip class,
@@
 base case scenario of constant relative treatment effects, these more
-flexible restricted cubic splines did not outperform the simpler RCS-3
+flexible restricted cubic splines did not outperform the simpler RCS-3.
 These approaches may only be helpful if we expect more extreme patterns
@@
 predict treatment benefit, prediction model development was outside the
-scope of this work and has already been studied extensively
-{[}5,13,14{]}.
+scope of this work and has already been studied extensively {[}5,8,9{]}.
 
@@
 strong treatment-related harms, applied on the absolute scale to expand
-the range of scenarios in line with previous work (REF!!). Also, we
-considered a small number of scenarios with true treatment-covariate
-interactions, in which our main conclusions remained the same
-(Supplement, XX). Future simulation studies could explore the effect of
-more extensive deviations from risk-based treatment effects.
+the range of scenarios in line with previous work {[}{\textbf{???}}{]}.
+Also, we considered a small number of scenarios with true
+treatment-covariate interactions, in which our main conclusions remained
+the same (Supplement, XX). Future simulation studies could explore the
+effect of more extensive deviations from risk-based treatment effects.
 
-Recent years have seen an increased interest in predictive HTE
-approaches focusing on individualized benefit predictions. In our
-simulations we only focused on risk-based methods, using baseline risk
-as a reference in a two-stage approach to individualizing benefit
+In our simulations we only focused on risk-based methods, using baseline
+risk as a reference in a two-stage approach to individualizing benefit
 predictions. However, there is a plethora of different methods, ranging
 from treatment effect modeling to tree-based approaches available in
-more recent literature {[}15--17{]}. Simulations are also needed to
-assess relative performance and define the settings where these break
-down or outperform each other.
+more recent literature {[}15--18{]}. Many of these methods rely on
+incorporating treatment-covariate interactions in the prediction of
+benefit. An important caveat of such approaches is that they may be
+prone to overfitting, thus exaggerating the magnitude of the predicted
+benefits. In a wide range of simulation settings, a simpler risk
+modeling approach was consistently better calibrated for benefit
+compared to more complex treatment effect modelling approaches {[}5{]}.
+Similarly, when SYNTAX score II, a model developed for identifying
+patients with complex coronary artery disease that benefit more from
+percutaneous coronary intervention or from coronary artery bypass
+grafting was redeveloped using fewer treatment-covariate interactions
+had better external performance compared to its predecessor{[}19,20{]}.
 
@@
 
+\leavevmode\hypertarget{ref-Burke2014}{}%
+{[}8{]} Burke JF, Hayward RA, Nelson JP, Kent DM. Using internally
+developed risk models to assess heterogeneity in treatment effects in
+clinical trials. Circulation: Cardiovascular Quality and Outcomes
+2014;7:163--9. \url{https://doi.org/10.1161/circoutcomes.113.000497}.
+
+\leavevmode\hypertarget{ref-Abadie2018}{}%
+{[}9{]} Abadie A, Chingos MM, West MR. Endogenous stratification in
+randomized experiments. The Review of Economics and Statistics
+2018;100:567--80. \url{https://doi.org/10.1162/rest_a_00732}.
+
 \leavevmode\hypertarget{ref-Harrell1988}{}%
-{[}8{]} Harrell FE, Lee KL, Pollock BG. Regression models in clinical
+{[}10{]} Harrell FE, Lee KL, Pollock BG. Regression models in clinical
 studies: Determining relationships between predictors and response. JNCI
@@
 \leavevmode\hypertarget{ref-vanKlaveren2018}{}%
-{[}9{]} Klaveren D van, Steyerberg EW, Serruys PW, Kent DM. The proposed
-``concordance-statistic for benefit'' provided a useful metric when
-modeling heterogeneous treatment effects. Journal of Clinical
+{[}11{]} Klaveren D van, Steyerberg EW, Serruys PW, Kent DM. The
+proposed ``concordance-statistic for benefit'' provided a useful metric
+when modeling heterogeneous treatment effects. Journal of Clinical
 Epidemiology 2018;94:59--68.
@@
 \leavevmode\hypertarget{ref-Austin2019}{}%
-{[}10{]} Austin PC, Steyerberg EW. The integrated calibration index
+{[}12{]} Austin PC, Steyerberg EW. The integrated calibration index
 (ICI) and related metrics for quantifying the calibration of logistic
@@
 \leavevmode\hypertarget{ref-Califf1997}{}%
-{[}11{]} Califf RM, Woodlief LH, Harrell FE, Lee KL, White HD, Guerci A,
+{[}13{]} Califf RM, Woodlief LH, Harrell FE, Lee KL, White HD, Guerci A,
 et al. Selection of thrombolytic therapy for individual patients:
@@
 \leavevmode\hypertarget{ref-Steyerberg2000}{}%
-{[}12{]} Steyerberg EW, Bossuyt PMM, Lee KL. Clinical trials in acute
+{[}14{]} Steyerberg EW, Bossuyt PMM, Lee KL. Clinical trials in acute
 myocardial infarction: Should we adjust for baseline characteristics?
@@
 \url{https://doi.org/10.1016/s0002-8703(00)90001-2}.
-
-\leavevmode\hypertarget{ref-Burke2014}{}%
-{[}13{]} Burke JF, Hayward RA, Nelson JP, Kent DM. Using internally
-developed risk models to assess heterogeneity in treatment effects in
-clinical trials. Circulation: Cardiovascular Quality and Outcomes
-2014;7:163--9. \url{https://doi.org/10.1161/circoutcomes.113.000497}.
-
-\leavevmode\hypertarget{ref-Abadie2018}{}%
-{[}14{]} Abadie A, Chingos MM, West MR. Endogenous stratification in
-randomized experiments. The Review of Economics and Statistics
-2018;100:567--80. \url{https://doi.org/10.1162/rest_a_00732}.
 
@@
 \url{https://doi.org/10.1080/01621459.2017.1319839}.
+
+\leavevmode\hypertarget{ref-powers2018some}{}%
+{[}18{]} Powers S, Qian J, Jung K, Schuler A, Shah NH, Hastie T, et al.
+Some methods for heterogeneous treatment effect estimation in high
+dimensions. Statistics in Medicine 2018;37:1767--87.
+
+\leavevmode\hypertarget{ref-farooq2013anatomical}{}%
+{[}19{]} Farooq V, Van Klaveren D, Steyerberg EW, Meliga E, Vergouwe Y,
+Chieffo A, et al. Anatomical and clinical characteristics to guide
+decision making between coronary artery bypass surgery and percutaneous
+coronary intervention for individual patients: Development and
+validation of syntax score ii. The Lancet 2013;381:639--50.
+
+\leavevmode\hypertarget{ref-takahashi2020redevelopment}{}%
+{[}20{]} Takahashi K, Serruys PW, Fuster V, Farkouh ME, Spertus JA,
+Cohen DJ, et al. Redevelopment and validation of the syntax score ii to
+individualise decision making between percutaneous and surgical
+revascularisation in patients with complex coronary artery disease:
+Secondary analysis of the multicentre randomised controlled syntaxes
+trial with external cohort validation. The Lancet 2020;396:1399--412.
 \end{cslreferences}

### 2022-01-31 submission/supplement.rmd
working on updates following co-authors' comments

@@
   linesep = c('', '', '', '\\addlinespace'),
-  digits      = 2,
+  digits      = 3,
   format.args = list(
@@
 
-Here we present the root mean squared error of the considered methods
-using strong constant relative treatment effect ($\text{OR}=0.5$) as
-the reference. Again, the same sample size and prediction performance
-settings were considered along with the same settings for linear,
-quadratic and non-monotonic deviations from the base case scenario of
-constant relative treatment effects are considered. All results can be found at [https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).
+Here we present the root mean squared error of the considered methods using
+strong constant relative treatment effect ($\text{OR}=0.5$) as the
+reference. Again, the same sample size and prediction performance settings were
+considered along with the same settings for linear, quadratic and non-monotonic
+deviations from the base case scenario of constant relative treatment effects
+are considered. All results can be found at
+[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).
 
@@
 
-Here we present the root mean squared error of the considered methods
-when true interactions of the covariates with treatment are
-assumed. We ran our analyses on a limited set of scenarios. More
-specifically, true prediction AUC was set to 0.75 and sample size was
-set to 4,250. The simulation scenarios included 4 weak interactions
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 4 strong interactions
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), and 2 weak and 2
-strong interactions.
+We carried out a smaller set of simulations, in which we assumed true
+treatment-covariate interactions. Sample size was set to 4,250 and the AUC of
+the true prediction model was set to 0.75. The following scenarios were
+considered: 1) 4 true weak positive interactions
+($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.83$); 2) 4 strong positive
+interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$); 3) 2 weak and 2
+strong positive interactions; 4) 4 weak negative interactions 
+($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=1.17$); 5) 4 strong negative interactions 
+($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=1.39$); 6) 2 weak and 2 strong negative
+interactions; 7) combined positive and negative strong interactions. We also
+considered constant treatment-related harms applied on the absolute scale to all
+treated patients. The exact settings were: 1) absent treatment-related harms; 2)
+moderate treatment-related harms, defined as 25\% of the average true benefit of the
+scenario without treatment-related harms; 3) strong treatment-related harms
+defined as 50\% of the true average benefit of the scenario without
+treatment-related harms; 4) negative treatment-related harms (benefit), defined
+as an absolute risk reduction for treated patients of 50\% of the true average
+benefit of the scenario without treatment-related harms. The exact settings can
+be found in Table \@ref(tab:tab2).
 
-```{r rmseInteraction, cache=FALSE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
+```{r tab2, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
+asd <- read_csv(
+  here("data/processed/analysisIdsInteractions.csv")
+) %>%
+  select(-"type")
+kableExtra::kbl(
+  asd,
+  format = "latex",
+  escape = FALSE,
+  longtable   = TRUE,
+  align       = "r",
+  booktabs    = TRUE,
+  caption     = "Scenario settings of the entire simulation study.",
+  col.names   = c(
+    "Scenario",
+    "Base",
+    "N",
+    "AUC",
+    "Treatment-related harm",
+    paste0("b", 0:8),
+    "g0", "g1", "g2", "g5", "g6",
+    "Before harms",
+    "After harms"
+  ),
+  linesep = c('', '', '', '\\addlinespace'),
+  digits      = 2,
+  format.args = list(
+    big.mar      = ",",
+    decimal.mark = "."
+  )
+) %>%
+  kableExtra::landscape() %>%
+  kableExtra::add_header_above(
+    c(
+      "Analysis ID"           = 5,
+      "Baseline risk"         = 9,
+      "True treatment effect" = 5,
+      "Benefit"               = 2
+      )
+  ) %>%
+  kableExtra::kable_styling(font_size = 6)
+```
+
+```{r rmseInteractionPositive, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
 AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
-grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction.tiff")))
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_positive.tiff")))
 # knitr::include_graphics(here::here("figures/calibration_base.png"))
 ```
+```{r rmseInteractionNegative, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
+AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_negative.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+```{r rmseInteractionCombined, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
+AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_combined.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+# Empirical illustration
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+load(here::here("data/raw/gusto.rda"))
+gusto <- gusto %>%
+  tibble() %>%
+  filter(!is.na(tpa))
+
+gusto <- gusto %>%
+  tibble() %>%
+  filter(tx != "SK+tPA") %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+treatmentArms <- gusto %>%
+  group_by(treatment) %>%
+  summarise(n = n())
+```
+
+\begin{equation}
+P(outcome=1|X=x) = \text{expit}(lp(x)),
+(\#eq:gusto1)
+\end{equation}
+where 
+\begin{equation}
+\begin{aligned}
+lp(x)=& \beta_0 + \beta_1 \text{age} + \beta_2 I(\text{Killip}=II) + \beta_3I(\text{Killip}=III) +\\
+&\beta_4 I(\text{Killip}=IV) + \beta_5ceil(\text{sysbp}, 120) + lsp(\text{pulse}, 50) +\\
+&\beta_6 I(\text{pmi}=yes)+\beta_8 I(\text{miloc}=Anterior) + \\
+&\beta_9 I(\text{miloc}=Other) + \gamma t_x
+\end{aligned}
+(\#eq:gusto2)
+\end{equation}

### 2022-01-31 code/PlotRmseInteractions.R

@@ -23,11 +23,24 @@ library(gridExtra)
 library(grid)
 library(ggside)
 
+args <- commandArgs(trailingOnly = TRUE)
+interactionType <- as.character(args[1])
+
+if (interactionType == "positive") {
+  scenarioLimits <- c(649, 660)
+} else if (interactionType == "negative") {
+  scenarioLimits <- c(661, 672)
+} else {
+  scenarioLimits <- c(673, 676)
+}
+
+
 source("code/helpers/CreateManuscriptPlots.R")     # Geneartes single boxplot
 source("code/helpers/PlotResult.R")                # Generates the boxplot list
 source("code/helpers/Absolute.R")                  # Generates the absolute plots
 
-scenarioIds <- readr::read_csv("data/processed/analysisIdsInteractions.csv")
+scenarioIds <- readr::read_csv("data/processed/analysisIdsInteractions.csv") %>%
+  filter(scenario >= scenarioLimits[1] & scenario <= scenarioLimits[2])
 
 metric    <- "rmse"
 titles <- scenarioIds %>%
@@ -59,7 +72,7 @@ limitsHigh <- .15
 processed <- readr::read_csv(
   file = file.path("data/processed", metricFile)
 ) %>% 
-  filter(scenarioId > 648) %>%
+  filter(scenarioId >= scenarioLimits[1] & scenarioId <= scenarioLimits[2]) %>%
   mutate_at(
     c(
       "constant_treatment_effect",
@@ -82,69 +95,85 @@ names(scenarios) <- NULL
 
 plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 10, 2.5))
 
-gridList <- list(
-  plotList[[1]] +
+if (interactionType == "combined") {
+  res <- plotList[[1]] +
     theme(
       panel.grid.minor = element_blank(),
       plot.title = element_markdown(size = 9),
       axis.title.x = ggplot2::element_blank(),
       axis.title.y = ggplot2::element_blank(),
-      axis.text.x = element_blank(),
-      axis.text.y = element_text(size = 8),
-      legend.direction = "horizontal",
-      legend.title = element_text(size = 7.5),
-      legend.text = element_text(size = 7),
-      legend.position = c(.211, .910)
-    ),
-  plotList[[2]] + 
-    theme(
-      panel.grid.minor = element_blank(),
-      plot.title = element_markdown(size = 9),
-      axis.title = element_blank(),
-      axis.text.x = element_blank(),
-      axis.text.y = element_text(size = 8),
[…]

### 2022-01-31 code/WriteAnalysisIdsInteractions.R

@@ -4,19 +4,27 @@ library(tidyverse)
 library(readr)
 
 base <- "interaction"
-type <- c("weak", "mixed", "strong")
+type <- c(
+  "weak", 
+  "mixed", 
+  "strong",
+  "negative-weak", 
+  "negative-mixed", 
+  "negative-strong",
+  "combined"
+)
 sampleSize <- 4250
 auc <- .75
 harm <- c("absent", "moderate-positive", "strong-positive", "negative")
 
 table <- expand_grid(base, type, sampleSize, auc, harm) %>%
   mutate(
-    # b0 = -2.08,
-    b0 = case_when(
-      type == "weak" ~ -3.05,
-      type == "mixed" ~ -3.32,
-      TRUE ~ -3.50
-    ),
+    b0 = -2.08,
+    # b0 = case_when(
+    #   type == "weak" ~ -3.05,
+    #   type == "mixed" ~ -3.32,
+    #   TRUE ~ -3.50
+    # ),
     b1 = .49,
     b2 = .49,
     b3 = .49,
@@ -25,24 +33,49 @@ table <- expand_grid(base, type, sampleSize, auc, harm) %>%
     b6 = .49,
     b7 = .49,
     b8 = .49,
-    g0 = log(.8),
+    g0 = log(.5),
     g1 = case_when(
-      (type == "weak" | type == "mixed"  ) ~ -.19,
-      (type == "strong"                  ) ~ -.49
+      (type == "weak" | type == "mixed"                   ) ~ -.19,
+      (type == "strong" | type == "combined"              ) ~ -.49,
+      (type == "negative-weak" | type == "negative-mixed" ) ~ .16,
+      (type == "negative-strong"                          ) ~ .33
     ),
     g2 = case_when(
-      (type == "weak"                    ) ~ -.19,
-      (type == "mixed" | type == "strong") ~ -.49
+      (type == "weak"                                                           ) ~ -.19,
+      (type == "mixed" | type == "strong"                                       ) ~ -.49,
+      (type == "negative-weak"                                                  ) ~ .16,
+      (type == "negative-mixed" | type == "negative-strong" | type == "combined") ~ .33
     ),
     g5 = case_when(
-      (type == "weak" | type == "mixed"  ) ~ -.19,
-      (type == "strong"                  ) ~ -.49
+      (type == "weak" | type == "mixed"                   ) ~ -.19,
+      (type == "strong" | type =="combined"               ) ~ -.49,
[…]

### 2022-02-02 submission/manuscript.rmd
co-author comments

@@
 the treatment indicator to 0. More information on model development can be found
-in the supplement.
+in the supplement (Section XX).
 
@@
 
-We used the risk linear predictor to fit the proposed methods under study for
-individualizing absolute benefit predictions. All methods predicted increasing
-benefits for patients with higher baseline risk predictions, but the fitted
-patterns were somewhat different. The adaptive approach selected the model with
-RCS smoothing with 4 knots. However, for very low baseline risk the decreasing
-predicted benefit with increasing risk may be somewhat too flexible. The more
-robust models, such as the linear interaction model and the model with RCS
-smoothing (3 knots), gave very similar benefit predictions. These models
-followed the evolution of the stratified estimates very closely and may
-therefore be preferable for use in clinical practice. The linear interaction
-model had somewhat lower AIC compared to the model with RCS smoothing (3 knots),
-slightly better cross-validated discrimination (c-for-benefit
-`r round(discriminationPerformance[2, ], 3)` vs
-`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
-calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
-`r round(calibrationPerformance[3, ], 4)`).
+We used the risk linear predictor to fit a constant treatment effect model, a
+linear interaction model and a RCS-3 model individualizing absolute benefit
+predictions. RCS-4 and RCS-5 models were excluded. In our simulations these
+methods were always outperformed by the simpler approaches and were often
+overfitted. Finally, an adaptive approach with only the 3 candidate models was
+also applied.
+
+All considered methods provided similar fits, predicting increasing benefits for
+patients with higher baseline risk predictions. All models followed the
+evolution of the stratified estimates very closely. The adaptive approach based
+on AIC selected the constant treatment effect model. The constant treatment
+effect model had somewhat lower AIC compared to the linear interaction model
+slightly worse cross-validated discrimination (c-for-benefit
+`r round(discriminationPerformance[1, ], 3)` vs
+`r round(discriminationPerformance[2, ], 3)`) and better cross-validated
+calibration (ICI-for benefit `r round(calibrationPerformance[1, ], 4)` vs
+`r round(calibrationPerformance[2, ], 4)`). In conclusion, a simpler constant
+treatment effect model is adequate for predicting absolute 30-day mortality
+benefits of treatment with tPA in patients with with acute MI.
 
@@
 harms, applied on the absolute scale to expand the range of scenarios in line
-with previous work [@Glasziou1995]. Also, we considered a small number of
-scenarios with true treatment-covariate interactions, in which our main
-conclusions remained the same (Supplement, XX). Future simulation studies could
-explore the effect of more extensive deviations from risk-based treatment
-effects.
+with previous work [@Glasziou1995]. In a limited set of scenarios where we
+assumed the existence of true treatment-covariate interactions, our conclusions
+remained unchanged. Even though the average error rates increased for all the
+considered methods, due to the miss-specification of the outcome model, the
+linear interaction model had the lowest error rates. RCS-3 had very comparable
+performance. The constant treatment effect model often gave biased results,
+espcially in the presence of moderate or strong treatment-related harms. All the
+results of these simulations can be found in Supplement, Section XX. Future
+simulation studies could explore the effect of more extensive deviations from
+risk-based treatment effects.
 
@@
 when non-monotonic deviations from a constant relative treatment effect and/or
-substantial treatment-related harms are anticipated. With larger sample size, an
-adaptive approach -- selecting the method with optimal AIC -- can also be
-considered as an automated alternative.
+substantial treatment-related harms are anticipated. Increasing the complexity
+of the RCS models by increasing the number of knots does not translate to
+improved benefit prediction. Using AIC for model selection among the constant
+treatment effect, the linear interaction and RCS-3 model is a viable option,
+especially with larger sample size.
 

### 2022-02-02 submission/manuscript.tex
co-author comments

@@
 for individuals we set the treatment indicator to 0. More information on
-model development can be found in the supplement.
+model development can be found in the supplement (Section XX).
 
-We used the risk linear predictor to fit the proposed methods under
-study for individualizing absolute benefit predictions. All methods
-predicted increasing benefits for patients with higher baseline risk
-predictions, but the fitted patterns were somewhat different. The
-adaptive approach selected the model with RCS smoothing with 4 knots.
-However, for very low baseline risk the decreasing predicted benefit
-with increasing risk may be somewhat too flexible. The more robust
-models, such as the linear interaction model and the model with RCS
-smoothing (3 knots), gave very similar benefit predictions. These models
-followed the evolution of the stratified estimates very closely and may
-therefore be preferable for use in clinical practice. The linear
-interaction model had somewhat lower AIC compared to the model with RCS
-smoothing (3 knots), slightly better cross-validated discrimination
-(c-for-benefit 0.526 vs 0.525) and quite similar cross-validated
-calibration (ICI-for benefit 0.0115 vs 0.0117).
+We used the risk linear predictor to fit a constant treatment effect
+model, a linear interaction model and a RCS-3 model individualizing
+absolute benefit predictions. RCS-4 and RCS-5 models were excluded. In
+our simulations these methods were always outperformed by the simpler
+approaches and were often overfitted. Finally, an adaptive approach with
+only the 3 candidate models was also applied.
+
+All considered methods provided similar fits, predicting increasing
+benefits for patients with higher baseline risk predictions. All models
+followed the evolution of the stratified estimates very closely. The
+adaptive approach based on AIC selected the constant treatment effect
+model. The constant treatment effect model had somewhat lower AIC
+compared to the linear interaction model slightly worse cross-validated
+discrimination (c-for-benefit 0.525 vs 0.526) and better cross-validated
+calibration (ICI-for benefit 0.0104 vs 0.0115). In conclusion, a simpler
+constant treatment effect model is adequate for predicting absolute
+30-day mortality benefits of treatment with tPA in patients with with
+acute MI.
 
@@
 strong treatment-related harms, applied on the absolute scale to expand
-the range of scenarios in line with previous work {[}{\textbf{???}}{]}.
-Also, we considered a small number of scenarios with true
-treatment-covariate interactions, in which our main conclusions remained
-the same (Supplement, XX). Future simulation studies could explore the
-effect of more extensive deviations from risk-based treatment effects.
+the range of scenarios in line with previous work {[}15{]}. In a limited
+set of scenarios where we assumed the existence of true
+treatment-covariate interactions, our conclusions remained unchanged.
+Even though the average error rates increased for all the considered
+methods, due to the miss-specification of the outcome model, the linear
+interaction model had the lowest error rates. RCS-3 had very comparable
+performance. The constant treatment effect model often gave biased
+results, espcially in the presence of moderate or strong
+treatment-related harms. All the results of these simulations can be
+found in Supplement, Section XX. Future simulation studies could explore
+the effect of more extensive deviations from risk-based treatment
+effects.
 
@@
 from treatment effect modeling to tree-based approaches available in
-more recent literature {[}15--18{]}. Many of these methods rely on
+more recent literature {[}16--19{]}. Many of these methods rely on
 incorporating treatment-covariate interactions in the prediction of
@@
 grafting was redeveloped using fewer treatment-covariate interactions
-had better external performance compared to its predecessor{[}19,20{]}.
+had better external performance compared to its predecessor{[}20,21{]}.
 
@@
 relative treatment effect and/or substantial treatment-related harms are
-anticipated. With larger sample size, an adaptive approach -- selecting
-the method with optimal AIC -- can also be considered as an automated
-alternative.
+anticipated. Increasing the complexity of the RCS models by increasing
+the number of knots does not translate to improved benefit prediction.
+Using AIC for model selection among the constant treatment effect, the
+linear interaction and RCS-3 model is a viable option, especially with
+larger sample size.
 
@@
 
+\leavevmode\hypertarget{ref-Glasziou1995}{}%
+{[}15{]} Glasziou PP, Irwig LM. An evidence based approach to
+individualising treatment. BMJ 1995;311:1356--9.
+\url{https://doi.org/10.1136/bmj.311.7016.1356}.
+
 \leavevmode\hypertarget{ref-Athey2019}{}%
-{[}15{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
+{[}16{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
 Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.
@@
 \leavevmode\hypertarget{ref-Lu2018}{}%
-{[}16{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
+{[}17{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
 treatment effect in observational data using random forest methods.
@@
 \leavevmode\hypertarget{ref-Wager2018}{}%
-{[}17{]} Wager S, Athey S. Estimation and inference of heterogeneous
+{[}18{]} Wager S, Athey S. Estimation and inference of heterogeneous
 treatment effects using random forests. Journal of the American
@@
 \leavevmode\hypertarget{ref-powers2018some}{}%
-{[}18{]} Powers S, Qian J, Jung K, Schuler A, Shah NH, Hastie T, et al.
+{[}19{]} Powers S, Qian J, Jung K, Schuler A, Shah NH, Hastie T, et al.
 Some methods for heterogeneous treatment effect estimation in high
@@
 \leavevmode\hypertarget{ref-farooq2013anatomical}{}%
-{[}19{]} Farooq V, Van Klaveren D, Steyerberg EW, Meliga E, Vergouwe Y,
+{[}20{]} Farooq V, Van Klaveren D, Steyerberg EW, Meliga E, Vergouwe Y,
 Chieffo A, et al. Anatomical and clinical characteristics to guide
@@
 \leavevmode\hypertarget{ref-takahashi2020redevelopment}{}%
-{[}20{]} Takahashi K, Serruys PW, Fuster V, Farkouh ME, Spertus JA,
+{[}21{]} Takahashi K, Serruys PW, Fuster V, Farkouh ME, Spertus JA,
 Cohen DJ, et al. Redevelopment and validation of the syntax score ii to

### 2022-02-02 submission/supplement.rmd
co-author comments

@@
 library(here)
+library(rms)
 
@@
 
+\newpage
 # Adaptive model selection frequencies
@@
 \newpage
-
 # Discrimination and calibration for benefit
@@
 \newpage
-
 # Strong relative treatment effect
@@
 
+\newpage
 # Treatment interactions
@@
 
-```{r rmseInteractionPositive, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
-AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
+```{r rmseInteractionPositive, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions all favoring treatment were considered.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_positive.tiff")))
@@
 ```
-```{r rmseInteractionNegative, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
-AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
+```{r rmseInteractionNegative, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions all favoring the control were considered.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_negative.tiff")))
@@
 ```
-```{r rmseInteractionCombined, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000.
-AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
+```{r rmseInteractionCombined, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions 2 favoring treatment and 2 favoring the control were considered.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_combined.tiff")))
@@
 
+\newpage
 # Empirical illustration
@@
   summarise(n = n())
+
+prediction <- glm(
+   outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+   data = gusto,
+   family = "binomial",
+   maxit = 99
+)
+
+predictionCoefficients <- coef(prediction)
+predictionCoefficients <- predictionCoefficients[c(1, 3:12, 2)]
 ```
@@
 lp(x)=& \beta_0 + \beta_1 \text{age} + \beta_2 I(\text{Killip}=II) + \beta_3I(\text{Killip}=III) +\\
-&\beta_4 I(\text{Killip}=IV) + \beta_5ceil(\text{sysbp}, 120) + lsp(\text{pulse}, 50) +\\
-&\beta_6 I(\text{pmi}=yes)+\beta_8 I(\text{miloc}=Anterior) + \\
-&\beta_9 I(\text{miloc}=Other) + \gamma t_x
+&\beta_4 I(\text{Killip}=IV) + \beta_5ceil(\text{sysbp}, 120) + \beta_6 \text{pulse}+\\
+&\beta_7 max(\text{pulse - 50, 0}) + \beta_8 I(\text{pmi}=yes)+\\
+&\beta_9 I(\text{miloc}=Anterior) + \beta_9 I(\text{miloc}=Other) +\\
+&\gamma\times\text{treatment}
 \end{aligned}
@@
 \end{equation}
+
+and $expit(x) = \frac{e^x}{1+e^x}$
+
+```{r, echo=FALSE}
+predictionSummary <- summary(prediction)
+# predictionSummaryCoef <- tibble(predictionSummary$coefficients) %>%
+#   rownames_to_column("Variable")
+
+predictionCoefficientsTable <- predictionSummary$coefficients %>%
+  as.data.frame() %>%
+  mutate(
+    rownames = c(
+      "Intercept",
+      "Age",
+      "Killip class = II",
+      "Killip class = III",
+      "Killip class = IV",
+      "Systolic blood pressure",
+      "Pulse rate (1)",
+      "Pulse rate (2)",
+      "Previous MI (yes)",
+      "MI location (Other)",
+      "MI location (Anterior)",
+      "Treatment"
+    )
+  ) %>%
+  rownames_to_column("Variable") %>%
+  select(-Variable) %>%
+  select(rownames, everything())
+
+colnames(predictionCoefficientsTable) <- c(
+  "Variable",
+  "Estimate",
+  "stderror",
+  "zvalue",
+  "pvalue"
+)
+
+kableExtra::kbl(
+  predictionCoefficientsTable,
+  format = "latex",
+  escape = FALSE,
+  longtable = TRUE,
+  align = c("l", rep("r", 4)),
+  booktabs = TRUE,
+  caption = "Coefficients of the prediction model for 30-day mortality, based on the data from GUSTO-I trial.",
+  digits = 3,
+  linesep = "",
+  format.args = list(
+    big.mar      = ",",
+    decimal.mark = "."
+  )
+) %>%
+  kableExtra::kable_styling(font_size = 7)
+```

### 2022-02-02 code/DiscriminationBase.R

@@ -85,7 +85,7 @@ res <- gridExtra::grid.arrange(
       plot.title = element_markdown(),
       axis.title = element_blank(),
       legend.direction = "horizontal",
-      legend.position = c(.444, .93),
+      legend.position = c(.364, .94),
       legend.text = element_text(size = 7),
       legend.title = element_text(size = 8),
       panel.grid.minor = element_blank(),

### 2022-02-02 code/PlotRmse.R

@@ -145,7 +145,7 @@ gridList <- list(
       legend.direction = "horizontal",
       legend.title = element_text(size = 7.5),
       legend.text = element_text(size = 7),
-      legend.position = c(.333, .87)
+      legend.position = c(.273, .87)
     ),
   absolutePlots$plot[[1]] +
     ggtitle("Simulated absolute benefit in treated patients") +

### 2022-02-02 code/helpers/Absolute.R

@@ -67,7 +67,7 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
     fullValidationDataset <- validationDataset %>%
       bind_rows(.id = "harmSetting")
     res <- res +
-      # ggside::ggside(x.pos = "bottom") +
+      ggside::ggside(x.pos = "bottom") +
       ggside::ggside(y.pos = "left")
     if (type == "constant") {
       res <- res +
@@ -78,12 +78,12 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
         outlier.shape = NA
       )
     }
-      # ggside::geom_ysideboxplot(
-      #   aes(y = trueBenefit, fill = harmSetting),
-      #   data = fullValidationDataset,
-      #   orientation = "x",
-      #   outlier.shape = NA
-      #   ) +
+      res <- res + ggside::geom_ysideboxplot(
+        aes(y = trueBenefit, fill = harmSetting),
+        data = fullValidationDataset,
+        orientation = "x",
+        outlier.shape = NA
+        )
     res <- res +
       scale_ysidex_continuous() +
       scale_xsidey_continuous() +

### 2022-02-02 code/helpers/CreateManuscriptPlots.R

@@ -72,8 +72,8 @@ createPlot <- function(
       ),
       labels = c(
         "absent",
-        "moderate positive",
-        "strong positive"
+        "moderate",
+        "strong"
       )
     ) +
     ggplot2::ylab(yAxis) +

### 2022-02-03 submission/manuscript.rmd
new updates to manuscript

@@
 Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis).
-The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.", fig.show="hold", out.width = '100%'}
+The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot on the top.
+The 2.5, 25, 50, 75, and 97.5 percentiles of the true benefit distributions are expressed by the boxplots on the side of the right-handside panel.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))

### 2022-02-03 submission/manuscript.tex
new updates to manuscript

@@
 Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis).
-The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.}\label{fig:rmsebase}
+The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot on the top.
+The 2.5, 25, 50, 75, and 97.5 percentiles of the true benefit distributions are expressed by the boxplots on the side of the right-handside panel.}\label{fig:rmsebase}
 \end{figure}

### 2022-02-03 submission/supplement.rmd
new updates to manuscript

@@
 \beginsupplement
+\newpage
 
@@
 
-We first, generate the binary outcomes $y$ for the untreated patients ($t_x=0$), based on
+We first, generate the binary outcomes $y$ for the untreated patients ($t_x=0$),
+based on
 
@@
 ## Base-case scenario
-The base-case scenario assumes a constant odds ratio of $0.8$ in favor
-of treatment. The simulated datasets are of size $n=4250$, where
-treatment is allocated at random using a 50/50 split (80% power for
-the detection of an unadjusted OR of 0.8, assuming an event rate of
-20% in the untreated arm). Outcome incidence in the untreated
-population is set at $20\%$. For the development of the prediction
-model we use the model defined in \@ref(eq:p0) including a constant
-treatment effect. When doing predictions, $t_x$ is set to $0$. The
-value of the true $\bm{\beta}$ is such that the above prediction model
+The base-case scenario assumes a constant odds ratio of $0.8$ in favor of
+treatment. The simulated datasets are of size $n=4250$, where treatment is
+allocated at random using a 50/50 split (80% power for the detection of an
+unadjusted OR of 0.8, assuming an event rate of 20% in the untreated
+arm). Outcome incidence in the untreated population is set at $20\%$. For the
+development of the prediction model we use the model defined in \@ref(eq:p0)
+including a constant treatment effect. When doing predictions, $t_x$ is set to
+$0$. The value of the true $\bm{\beta}$ is such that the above prediction model
 has an AUC of $0.75$.
@@
 ## Deviations from base-case
-We deviate from the base-case scenario in two ways. First, we alter
-the overall target settings of sample size, overall treatment effect
-and prediction model AUC. In a second stage, we consider settings that
-violate the assumption of a constant relative treatment effect, using
-a model-based approach.
+We deviate from the base-case scenario in two ways. First, we alter the overall
+target settings of sample size, overall treatment effect and prediction model
+AUC. In a second stage, we consider settings that violate the assumption of a
+constant relative treatment effect, using a model-based approach.
 
@@
 
-We set the true risk model coefficients to be $\bm{\beta} = \big(-1.63, 0.26,\dots,0.26\big)^t$ for $AUC=0.65$ and $\bm{\beta} = \big(-2.7, 0.82,\dots,0.82\big)^t$ for $AUC=0.85$.
-In both cases, $\beta_0$ is selected so that an event rate of $20\%$ is maintained in the control arm.
+We set the true risk model coefficients to be
+$\bm{\beta} = \big(-1.63,0.26,\dots,0.26\big)^t$ for $AUC=0.65$ and $\bm{\beta} = \big(-2.7,0.82,\dots,0.82\big)^t$ 
+for $AUC=0.85$. In both cases, $\beta_0$ is selected so that an event rate of
+$20\%$ is maintained in the control arm.
 
-For the second part linear, quadratic and non-monotonic deviations
-from the assumption of constant relative effect are considered. We
-also consider different intensity levels of these deviations. Finally,
-constant absolute treatment-related harms are introduced,
-i.e. positive ($0.25\times\text{true average benefit}$), strong
-positive ($0.50\times\text{true average benefit}$) and negative
+For the second part linear, quadratic and non-monotonic deviations from the
+assumption of constant relative effect are considered. We also consider
+different intensity levels of these deviations. Finally, constant absolute
+treatment-related harms are introduced, i.e. positive 
+($0.25\times\text{true average benefit}$),
+strong positive ($0.50\times\text{true average benefit}$) and negative
 ($-0.25\times\text{true average benefit}$; i.e. constant absolute
 treatment-related benefit). In case of true absent treatment effects,
-treatment-related harms are set to $1\%, 2\%$ and $-1\%$ for
-positive, strong positive and negative setting, respectively. The
-settings for these deviations are defined in Table \@ref(tab:tab1).
+treatment-related harms are set to $1\%, 2\%$ and $-1\%$ for positive, strong
+positive and negative setting, respectively. The settings for these deviations
+are defined in Table \@ref(tab:tab1).
 
@@
 ### Risk stratification
-Derive a prediction model using the same approach as above and divide
-the population in equally sized risk-based subgroups. Estimate
-subgroup-specific absolute benefit from the observed absolute
-differences. Subject-specific benefit predictions are made by
-attributing to individuals their corresponding subgroup-specific
-estimate.
+Derive a prediction model using the same approach as above and divide the
+population in equally sized risk-based subgroups. Estimate subgroup-specific
+absolute benefit from the observed absolute differences. Subject-specific
+benefit predictions are made by attributing to individuals their corresponding
+subgroup-specific estimate.
 
 ### Constant treatment effect
-Assuming a constant relative treatment effect, fit the adjusted model
-in \@ref(eq:risk). Then, an estimate of absolute benefit can be
-derived from$$\hat{f}_{\text{benefit}}(lp\given \bm{x},
-\hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
+Assuming a constant relative treatment effect, fit the adjusted model in
+\@ref(eq:risk). Then, an estimate of absolute benefit can be derived
+from
+$$\hat{f}_{\text{benefit}}(lp\given \bm{x}, \hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
 
 ### Linear interaction
-The assumption of constant relative treatment effect is relaxed
-modeling a linear interaction of treatment with the risk linear
-predictor: $$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} =
-g\big(lp+(\gamma_0+\gamma_1lp)t_x\big)$$ We predict absolute benefit
-from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) =
-g(lp) - g\big(\gamma_0+(1+\gamma_1)lp\big)$$
+The assumption of constant relative treatment effect is relaxed modeling a
+linear interaction of treatment with the risk linear predictor: $$E\{y\given
+\bm{x}, t_x, \hat{\bm{\beta}}\} = g\big(lp+(\gamma_0+\gamma_1lp)t_x\big)$$ We
+predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp\given\bm{x},
+\hat{\bm{\beta}}) = g(lp) - g\big(\gamma_0+(1+\gamma_1)lp\big)$$
 
@@
 Finally, we drop the linearity assumption and predict absolute benefit using
-smoothing with restricted cubic splines with 3, 4, and 5 knots. More
+smoothing with restricted cubic splines with $k=3, 4$ and $5$ knots. More
 specifically, we fit the model:
@@
 where
-$$f_{RCS}(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$ with
-$$h_{j+1}(x)= (x-t_j)^3-(x-t_{k-1})_+\frac{t_k-t_j}{t_k-t_{k-1}}+(x-t_k)^3_+\frac{t_{k-1}-t_j}{t_k-t_{k-   1}}$$
-and $t_1,\dots,t_k$ are the selected knots. We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp \given \bm{x},\hat{\bm{\beta}})=P(y=1\given lp, t_x=0)-P(y=1\given lp,t_x=1)$$.
+$$f_{RCS}(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$
+with $h_1(x)=x$ and for $j=2,\dots,k-2$
+$$h_{j+1}(x)= (x-t_j)^3-(x-t_{k-1})_+^3 \frac{t_k-t_j}{t_k-t_{k-1}}+(x-t_k)^3_+\frac{t_{k-1}-t_j}{t_k-t_{k-1}}$$
+where 
+$t_1,\dots,t_k$ are the selected knots. We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp \given \bm{x},\hat{\bm{\beta}})=P(y=1\given lp, t_x=0)-P(y=1\given lp,t_x=1)$$.
 
@@
 ```
+
+\newpage
+# Plausible scenario settings
+
+In this section we present specific scenarios from our simulation settings
+in which evolution of benefit followed similar patterns to (REF!!). In this case
+patients were stratified into risk quarters based on their true baseline
+risk. Within each risk quarter we constructed boxplots of true benefit.
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/scenario_397.tiff")))
+```
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/scenario_251.tiff")))
+```
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/scenario_406.tiff")))
+```
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/scenario_324.tiff")))
+```
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/scenario_422.tiff")))
+```

### 2022-02-03 code/PlotRmse.R

@@ -30,9 +30,15 @@ library(gridExtra)
 library(grid)
 library(ggside)
 
-source("code/helpers/CreateManuscriptPlots.R")     # Geneartes single boxplot
-source("code/helpers/PlotResult.R")                # Generates the boxplot list
-source("code/helpers/Absolute.R")                  # Generates the absolute plots
+# --------------------------------------
+# Sourcing helper files:
+#   1. Generates single boxplot
+#   2. Generates the boxplot list
+#   3. Generates the absolute plots
+# --------------------------------------
+source("code/helpers/CreateManuscriptPlots.R") 
+source("code/helpers/PlotResult.R")
+source("code/helpers/Absolute.R")
 
 scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
   filter(

### 2022-02-03 code/PlotRmseInteractions.R

@@ -35,9 +35,15 @@ if (interactionType == "positive") {
 }
 
 
-source("code/helpers/CreateManuscriptPlots.R")     # Geneartes single boxplot
-source("code/helpers/PlotResult.R")                # Generates the boxplot list
-source("code/helpers/Absolute.R")                  # Generates the absolute plots
+# --------------------------------------
+# Sourcing helper files:
+#   1. Generates single boxplot
+#   2. Generates the boxplot list
+#   3. Generates the absolute plots
+# --------------------------------------
+source("code/helpers/CreateManuscriptPlots.R")     
+source("code/helpers/PlotResult.R")                
+source("code/helpers/Absolute.R")                  
 
 scenarioIds <- readr::read_csv("data/processed/analysisIdsInteractions.csv") %>%
   filter(scenario >= scenarioLimits[1] & scenario <= scenarioLimits[2])
@@ -52,11 +58,15 @@ titles <- scenarioIds %>%
   unlist() %>%
   unique()
 
-titlePrefix <- paste0(
-  "**",
-  LETTERS[1:3],
-  ".**"
-)
+if (interactionType != "combined") {
+  titlePrefix <- paste0(
+    "**",
+    LETTERS[1:3],
+    ".**"
+  )
+} else {
+  titlePrefix <- ""
+}
 
 titles <- paste(
   titlePrefix,
@@ -86,6 +96,26 @@ processed <- readr::read_csv(
     f
   )
 
+# ------------------------------------
+# Create the absolute benefit plots
+# ------------------------------------
+absolutePlots <- scenarioIds %>%
+  filter(harm != "negative") %>%
+  group_by(type) %>%
+  nest() %>%
+  mutate(
+    plot = map2(
+      data,
+      type,
+      ~ plotAbsoluteBenefit(
+      data = .x,
+      projectDir = "~/Documents/Projects/arekkas_HteSimulation_XXXX_2021",
+      type = .y
+      )
+    )
+  )
+
+
 scenarios <- scenarioIds %>%
   filter(harm == "absent") %>%
   select(scenario) %>%
@@ -93,21 +123,22 @@ scenarios <- scenarioIds %>%
 
 names(scenarios) <- NULL
 
-plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(0, 10, 2.5))
+plotList <- plotResult(
+  scenarios,
+  processed,
+  titles,
+  metric = metric,
[…]

### 2022-02-03 code/helpers/Absolute.R

@@ -1,4 +1,4 @@
-plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
+plotAbsoluteBenefit <- function(data, projectDir, type) {
   calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
     ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
     return(ret)
@@ -9,6 +9,7 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
     benefit <- plogis(x) - plogis(sq) - harm
     return(benefit)
   }
+  base <- data$base[1]
   harmSettings <- unique(data$harm)
   res <- ggplot()
   validationDataset <- list()
@@ -36,14 +37,16 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
       harm = scenarioSettings$harmValue
     )
     label <- harmSettings[i]
-    res <- res +
-      ggplot2::stat_function(
-        data = data.frame(x = c(.05, .95), label = label),
-        ggplot2::aes(x = x, color = label, group = label),
-        fun = calcAbsoluteBenefit,
-        args = args
-      )
-    if (!is.null(projectDir)) {
+    if (base != "interaction") {
+      res <- res +
+        ggplot2::stat_function(
+          data = data.frame(x = c(.05, .95), label = label),
+          ggplot2::aes(x = x, color = label, group = label),
+          fun = calcAbsoluteBenefit,
+          args = args
+        )
+    }
+    if (!missing(projectDir)) {
       settings <- readRDS(
         file.path(
           projectDir,
@@ -56,16 +59,36 @@ plotAbsoluteBenefit <- function(data, projectDir = NULL, type) {
       validationDatabaseSettings <- settings$simulationSettings$databaseSettings
       validationDatabaseSettings$numberOfObservations <- 1e5
       validationDataset[[i]] <- SimulateHte::runDataGeneration(
-        databaseSettings        = validationDatabaseSettings,
-        propensitySettings      = settings$simulationSettings$propensitySettings,
-        baselineRiskSettings    = settings$simulationSettings$baselineRiskSettings,
+        databaseSettings = validationDatabaseSettings,
+        propensitySettings = settings$simulationSettings$propensitySettings,
+        baselineRiskSetting = settings$simulationSettings$baselineRiskSettings,
         treatmentEffectSettings = settings$simulationSettings$treatmentEffectSettings
       )
[…]

### 2022-02-08 submission/manuscript.rmd
almost final version after comments

@@
  interaction between baseline risk and treatment assignment should be
- considered to improve treatment effect predictions. Non-linear RCS
+ considered to improve treatment effect predictions. Non-linear 
  interactions should be considered only in larger sample sizes.
@@
 A *stratified HTE method* has been suggested as an alternative to traditional
-subgroup analyses. Patients are stratified into equally-sized risk strata---in
-this case based on risk quartiles. Absolute treatment effects within risk strata
-are estimated by the difference in event rate between patients in the control
-arm and patients in the treated arm. We considered this approach as a reference,
-expecting it to perform worse than the other candidates, as its objective is to
-provide an illustration of HTE rather than to optimize individualized benefit
-predictions.
+subgroup analyses [@Kent2019; @PathEnE]. Patients are stratified into
+equally-sized risk strata---in this case based on risk quartiles. Absolute
+treatment effects within risk strata are estimated by the difference in event
+rate between patients in the control arm and patients in the treated arm. We
+considered this approach as a reference, expecting it to perform worse than the
+other candidates, as its objective is to provide an illustration of HTE rather
+than to optimize individualized benefit predictions.
 
@@
 
+For each scenario setting we performed 500 replications, within which all the
+considered models were fitted. For comparing between models we simulated a
+super-population of size 500,000 for each scenario. We calculated RMSE and
+discrimination and calibration for benefit of the models derived in each
+replication of our simulation settings within this super-population. 
 ## Empirical illustration
@@
 
-We demonstrate the different methods for individualizing treatment benefits
+We demonstrated the different methods for individualizing treatment benefits
 using data from `r nrow(gusto)` patients with acute myocardial infarction (MI)
@@
 (total of `r sum(gusto$outcome)` events), recorded for all patients.
+
+In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
+logistic regression model with 6 baseline covariates, i.e. age, Killip class,
+systolic blood pressure, heart rate, an indicator of previous MI, and the
+location of MI, to predict 30-day mortality risk. A constant effect of treatment
+was included in the model. When deriving risk predictions for individuals we set
+the treatment indicator to 0. More information on model development can be found
+in the supplement (Section 6).
 
@@
 treatment effect model in a substantial proportion of the replications
-(Supplement, Figure S1). This problematic behavior was less with larger sample
+(Supplement, Figure S3). This problematic behavior was less with larger sample
 sizes (see below).
@@
 relative treatment effect model was less often wrongly selected
-(Supplement, Figure S2).
+(Supplement, Figure S4).
 
-```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
+```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}.
+RMSE was calculated on a super-population of size 500,000", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
@@
 without treatment-related harms, respectively), leading to more variability of
-the RMSE (Supplement, Figure S3).
+the RMSE (Supplement, Figure S5).
 
-```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated samples 4,250.
-True prediction AUC of 0.85.", fig.show="hold", out.width = '100%'}
+```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples 4,250.
+True prediction AUC of 0.85.
+RMSE was calculated on a super-population of size 500,000", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
@@
 
-In comparison with the true approach, dsicrimination for benefit in the scenario
-with a constant relative treatment effect was only slightly lower for the linear
-interaction model, but substantially lower for the non-linear RCS approaches
-(Figure \ref{fig:discrimination}; panel A). With strong linear or quadratic
-deviations from a constant relative treatment effect, all methods discriminated
-quite similarly (Figure \ref{fig:discrimination}; panels B-C). In the scenario
-with non-monotonic deviations, the constant effect model had much lower
+When assuming a true constant relative treatment effect, discrimination for
+benefit was only slightly lower for the linear interaction model, but
+substantially lower for the non-linear RCS approaches (Figure
+\ref{fig:discrimination}; panel A). With strong linear or quadratic deviations
+from a constant relative treatment effect, all methods discriminated quite
+similarly (Figure \ref{fig:discrimination}; panels B-C). In the scenario with
+non-monotonic deviations, the constant effect model had much lower
 discriminative ability compared to all other methods (median AUC of
@@
 we observed similar trends, however the performance of all methods was more
-stable (Supplement, Figure S4). Finally, when we increased the true prediction
+stable (Supplement, Figure S6). Finally, when we increased the true prediction
 AUC to 0.85 the adaptive approach in the case of non-monotonic deviations was,
-
-again, quite unstable, especially with null or moderate treatment-related harms
-(Supplement, Figure S5). In these scenarios, the adaptive approach tended to
-select more often the inferior constant treatment effect method (Supplement,
-Figure S3)
+again, more conservative, especially with null or moderate treatment-related
+harms (Supplement, Figure S7).
 
@@
 displayed more stable calibration performance due to the larger number of
-patients (Supplement, Figure S6). When we increased the true prediction AUC to
+patients (Supplement, Figure S8). When we increased the true prediction AUC to
 0.85, the linear interaction model was worse calibrated, on average, than RCS-3
 in the case of strong quadratic deviations from constant relative treatment
-effects (Supplement, Figure S7).
+effects (Supplement, Figure S9).
 
 The results from all individual scenarios can be explored online at
-[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).
+[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/). Additionally,
+all the code for the simulations can be found at
+[https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021](https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021)
 
@@
 ## Empirical illustration
-
-In line with previous analyses [@Califf1997; @Steyerberg2000], we fitted a
-logistic regression model with 6 baseline covariates, i.e. age, Killip class,
-systolic blood pressure, heart rate, an indicator of previous MI, and the
-location of MI, to predict 30-day mortality risk. A constant effect of treatment
-was included in the model. When deriving risk predictions for individuals we set
-the treatment indicator to 0. More information on model development can be found
-in the supplement (Section XX).
 
@@
 
-We used the risk linear predictor to fit a constant treatment effect model, a
+We used the derived prognostic index to fit a constant treatment effect model, a
 linear interaction model and a RCS-3 model individualizing absolute benefit
@@
 
-```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3 knots.
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
@@
 tended to miss the treatment-risk interactions and selected simpler models
-(Supplementary Table S7). This conservative behavior resulted in increased RMSE
+(Supplement Section 4). This conservative behavior resulted in increased RMSE
 variability in these scenarios, especially in the case of true strong linear or
@@
 espcially in the presence of moderate or strong treatment-related harms. All the
-results of these simulations can be found in Supplement, Section XX. Future
+results of these simulations can be found in Supplement, Section 6. Future
 simulation studies could explore the effect of more extensive deviations from

### 2022-02-08 submission/manuscript.tex
almost final version after comments

@@
   interaction between baseline risk and treatment assignment should be
-  considered to improve treatment effect predictions. Non-linear RCS
+  considered to improve treatment effect predictions. Non-linear
   interactions should be considered only in larger sample sizes.
@@
 A \emph{stratified HTE method} has been suggested as an alternative to
-traditional subgroup analyses. Patients are stratified into
+traditional subgroup analyses {[}3,4{]}. Patients are stratified into
 equally-sized risk strata---in this case based on risk quartiles.
@@
 
-\hypertarget{empirical-illustration}{%
-\subsection{Empirical illustration}\label{empirical-illustration}}
+For each scenario setting we performed 500 replications, within which
+all the considered models were fitted. For comparing between models we
+simulated a super-population of size 500,000 for each scenario. We
+calculated RMSE and discrimination and calibration for benefit of the
+models derived in each replication of our simulation settings within
+this super-population. \#\# Empirical illustration
 
-We demonstrate the different methods for individualizing treatment
+We demonstrated the different methods for individualizing treatment
 benefits using data from 30,510 patients with acute myocardial
@@
 mortality (total of 2,128 events), recorded for all patients.
+
+In line with previous analyses {[}13,14{]}, we fitted a logistic
+regression model with 6 baseline covariates, i.e.~age, Killip class,
+systolic blood pressure, heart rate, an indicator of previous MI, and
+the location of MI, to predict 30-day mortality risk. A constant effect
+of treatment was included in the model. When deriving risk predictions
+for individuals we set the treatment indicator to 0. More information on
+model development can be found in the supplement (Section 6).
 
@@
 treatment effect model in a substantial proportion of the replications
-(Supplement, Figure S1). This problematic behavior was less with larger
+(Supplement, Figure S3). This problematic behavior was less with larger
 sample sizes (see below).
@@
 and the constant relative treatment effect model was less often wrongly
-selected (Supplement, Figure S2).
+selected (Supplement, Figure S4).
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsesamplesize-1} \caption{RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \ref{fig:rmsebase}}\label{fig:rmsesamplesize}
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmsesamplesize-1} \caption{RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \ref{fig:rmsebase}.
+RMSE was calculated on a super-population of size 500,000}\label{fig:rmsesamplesize}
 \end{figure}
@@
 deviation scenarios without treatment-related harms, respectively),
-leading to more variability of the RMSE (Supplement, Figure S3).
+leading to more variability of the RMSE (Supplement, Figure S5).
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmseauc-1} \caption{RMSE of the considered methods across 500 replications calculated in a simulated samples 4,250.
-True prediction AUC of 0.85.}\label{fig:rmseauc}
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/rmseauc-1} \caption{RMSE of the considered methods across 500 replications calculated in simulated samples 4,250.
+True prediction AUC of 0.85.
+RMSE was calculated on a super-population of size 500,000}\label{fig:rmseauc}
 \end{figure}
 
-In comparison with the true approach, dsicrimination for benefit in the
-scenario with a constant relative treatment effect was only slightly
-lower for the linear interaction model, but substantially lower for the
-non-linear RCS approaches (Figure \ref{fig:discrimination}; panel A).
-With strong linear or quadratic deviations from a constant relative
-treatment effect, all methods discriminated quite similarly (Figure
-\ref{fig:discrimination}; panels B-C). In the scenario with
-non-monotonic deviations, the constant effect model had much lower
-discriminative ability compared to all other methods (median AUC of
-0.4971 for the constant effects model, 0.5285 for the linear interaction
-model and 0.5304 for the best-performing RCS-3; Figure
-\ref{fig:discrimination}; panel D). The adaptive approach was unstable
-in terms of discrimination for benefit, especially in the presence of
-treatment-related harms. With increasing number of RCS knots, we
-observed decreasing median values and increasing variability of the
-c-for-benefit in all scenarios. When we increased the sample size to
-17,000 we observed similar trends, however the performance of all
-methods was more stable (Supplement, Figure S4). Finally, when we
+When assuming a true constant relative treatment effect, discrimination
+for benefit was only slightly lower for the linear interaction model,
+but substantially lower for the non-linear RCS approaches (Figure
+\ref{fig:discrimination}; panel A). With strong linear or quadratic
+deviations from a constant relative treatment effect, all methods
+discriminated quite similarly (Figure \ref{fig:discrimination}; panels
+B-C). In the scenario with non-monotonic deviations, the constant effect
+model had much lower discriminative ability compared to all other
+methods (median AUC of 0.4971 for the constant effects model, 0.5285 for
+the linear interaction model and 0.5304 for the best-performing RCS-3;
+Figure \ref{fig:discrimination}; panel D). The adaptive approach was
+unstable in terms of discrimination for benefit, especially in the
+presence of treatment-related harms. With increasing number of RCS
+knots, we observed decreasing median values and increasing variability
+of the c-for-benefit in all scenarios. When we increased the sample size
+to 17,000 we observed similar trends, however the performance of all
+methods was more stable (Supplement, Figure S6). Finally, when we
 increased the true prediction AUC to 0.85 the adaptive approach in the
-case of non-monotonic deviations was,
-
-again, quite unstable, especially with null or moderate
-treatment-related harms (Supplement, Figure S5). In these scenarios, the
-adaptive approach tended to select more often the inferior constant
-treatment effect method (Supplement, Figure S3)
+case of non-monotonic deviations was, again, more conservative,
+especially with null or moderate treatment-related harms (Supplement,
+Figure S7).
 
@@
 displayed more stable calibration performance due to the larger number
-of patients (Supplement, Figure S6). When we increased the true
+of patients (Supplement, Figure S8). When we increased the true
 prediction AUC to 0.85, the linear interaction model was worse
@@
 deviations from constant relative treatment effects (Supplement, Figure
-S7).
+S9).
 
 The results from all individual scenarios can be explored online at
-\url{https://arekkas.shinyapps.io/simulation_viewer/}.
+\url{https://arekkas.shinyapps.io/simulation_viewer/}. Additionally, all
+the code for the simulations can be found at
+\url{https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021}
 
@@
 
-\hypertarget{empirical-illustration-1}{%
-\subsection{Empirical illustration}\label{empirical-illustration-1}}
+\hypertarget{empirical-illustration}{%
+\subsection{Empirical illustration}\label{empirical-illustration}}
 
-In line with previous analyses {[}13,14{]}, we fitted a logistic
-regression model with 6 baseline covariates, i.e.~age, Killip class,
-systolic blood pressure, heart rate, an indicator of previous MI, and
-the location of MI, to predict 30-day mortality risk. A constant effect
-of treatment was included in the model. When deriving risk predictions
-for individuals we set the treatment indicator to 0. More information on
-model development can be found in the supplement (Section XX).
-
-We used the risk linear predictor to fit a constant treatment effect
+We used the derived prognostic index to fit a constant treatment effect
 model, a linear interaction model and a RCS-3 model individualizing
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/gusto-1} \caption{Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3,4 and 5 knots.
+\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/gusto-1} \caption{Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3 knots.
 Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.}\label{fig:gusto}
@@
 smaller sample sizes it tended to miss the treatment-risk interactions
-and selected simpler models (Supplementary Table S7). This conservative
+and selected simpler models (Supplement Section 4). This conservative
 behavior resulted in increased RMSE variability in these scenarios,
@@
 treatment-related harms. All the results of these simulations can be
-found in Supplement, Section XX. Future simulation studies could explore
+found in Supplement, Section 6. Future simulation studies could explore
 the effect of more extensive deviations from risk-based treatment

### 2022-02-08 submission/supplement.rmd
almost final version after comments

--- 2022-02-02 submission/supplement.rmd
@@
 
-We set the true risk model coefficients to be $\bm{\beta} = \big(-1.63, 0.26,\dots,0.26\big)^t$ for $AUC=0.65$ and $\bm{\beta} = \big(-2.7, 0.82,\dots,0.82\big)^t$ for $AUC=0.85$.
-In both cases, $\beta_0$ is selected so that an event rate of $20\%$ is maintained in the control arm.
+We set the true risk model coefficients to be
+$\bm{\beta} = \big(-1.63,0.26,\dots,0.26\big)^t$ for $AUC=0.65$ and $\bm{\beta} = \big(-2.7,0.82,\dots,0.82\big)^t$ 
+for $AUC=0.85$. In both cases, $\beta_0$ is selected so that an event rate of
+$20\%$ is maintained in the control arm.
 
-For the second part linear, quadratic and non-monotonic deviations
-from the assumption of constant relative effect are considered. We
-also consider different intensity levels of these deviations. Finally,
-constant absolute treatment-related harms are introduced,
-i.e. positive ($0.25\times\text{true average benefit}$), strong
-positive ($0.50\times\text{true average benefit}$) and negative
+For the second part linear, quadratic and non-monotonic deviations from the
+assumption of constant relative effect are considered. We also consider
+different intensity levels of these deviations. Finally, constant absolute
+treatment-related harms are introduced, i.e. positive 
+($0.25\times\text{true average benefit}$),
+strong positive ($0.50\times\text{true average benefit}$) and negative
 ($-0.25\times\text{true average benefit}$; i.e. constant absolute
 treatment-related benefit). In case of true absent treatment effects,
-treatment-related harms are set to $1\%, 2\%$ and $-1\%$ for
-positive, strong positive and negative setting, respectively. The
-settings for these deviations are defined in Table \@ref(tab:tab1).
-
-## Risk modeling
-Merging treatment arms, we develop prediction models including a constant relative treatment effect:
-
-\begin{equation}
-E\{y\given x,t_x\} = P(y\given x, t_x) = g(\beta_0+\beta_1x_1+\dots+\beta_8x_8+\gamma t_x)
-\end{equation}
-(\#eq:risk)
-Individualized predictions are derived setting $t_x=0$.
-
-## Approaches to individualize benefit predictions
-
-### Risk stratification
-Derive a prediction model using the same approach as above and divide
-the population in equally sized risk-based subgroups. Estimate
-subgroup-specific absolute benefit from the observed absolute
-differences. Subject-specific benefit predictions are made by
-attributing to individuals their corresponding subgroup-specific
-estimate.
-
-### Constant treatment effect
-Assuming a constant relative treatment effect, fit the adjusted model
-in \@ref(eq:risk). Then, an estimate of absolute benefit can be
-derived from$$\hat{f}_{\text{benefit}}(lp\given \bm{x},
-\hat{\bm{\beta}}) = g(lp) - g(lp+\hat{\gamma}) $$
-
-### Linear interaction
-The assumption of constant relative treatment effect is relaxed
-modeling a linear interaction of treatment with the risk linear
-predictor: $$E\{y\given \bm{x}, t_x, \hat{\bm{\beta}}\} =
-g\big(lp+(\gamma_0+\gamma_1lp)t_x\big)$$ We predict absolute benefit
-from $$\hat{f}_{\text{benefit}}(lp\given\bm{x}, \hat{\bm{\beta}}) =
-g(lp) - g\big(\gamma_0+(1+\gamma_1)lp\big)$$
-
-### Restricted cubic splines
-Finally, we drop the linearity assumption and predict absolute benefit using
-smoothing with restricted cubic splines with 3, 4, and 5 knots. More
-specifically, we fit the model:
-
-$$P(y=1\given lp, t_x) = g\big(\beta_0+\beta_{t_x}t_x+f_{RCS}(lp) + f_{RCS}(lp)\times t_x\big)$$
-where
-$$f_{RCS}(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$ with
-$$h_{j+1}(x)= (x-t_j)^3-(x-t_{k-1})_+\frac{t_k-t_j}{t_k-t_{k-1}}+(x-t_k)^3_+\frac{t_{k-1}-t_j}{t_k-t_{k-   1}}$$
-and $t_1,\dots,t_k$ are the selected knots. We predict absolute benefit from $$\hat{f}_{\text{benefit}}(lp \given \bm{x},\hat{\bm{\beta}})=P(y=1\given lp, t_x=0)-P(y=1\given lp,t_x=1)$$.
+treatment-related harms are set to $1\%, 2\%$ and $-1\%$ for positive, strong
+positive and negative setting, respectively. The settings for these deviations
+are defined in Table \@ref(tab:tab1).
 
@@
 ```
+
+\newpage
+# Plausible scenario settings
+
+In this section we present specific scenarios from our simulation settings
+in which evolution of benefit followed similar patterns to (REF!!). In this case
+patients were stratified into risk quarters based on their true baseline
+risk. Within each risk quarter we constructed boxplots of true benefit.
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials.
+In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
+include_graphics(
+  c(
+    here("figures/scenario_251.png"),
+    here("figures/scenario_324.png")
+  )
+)
+```
+
+```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials.
+In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
+include_graphics(
+  c(
+    here("figures/scenario_406.png"),
+    here("figures/scenario_422.png")
+  )
+)
+```
+
+# Approaches to individualize benefit predictions
+
+## Risk modeling
+Merging treatment arms, we develop prediction models including a constant relative treatment effect:
+
+\begin{equation}
+P(y=1\given\bm{x}, t_x) = g(\bm{x}^t\bm{\beta} + \gamma t_x)
+\end{equation}
+(\#eq:risk)
+
+The risk linear predictor is then 
+$lp(\bm{x}; \bm{\beta}) = \bm{x}^t\bm{\beta}$. 
+We derive baseline risk predictions for patients by setting $t_x=0$ in
+\@ref(eq:risk). All methods for individualizing benefit predictions are 2-stage
+methods, that start by fitting a model for predicting baseline risk. The
+estimated linear predictor of this model is
+
+\begin{equation*}
+\hat{lp} = lp(\bm{x};\hat{\bm{\beta}}) = \bm{x}^t\hat{\bm{\beta}}
+\end{equation*}
+
+## Risk stratification
+Derive a prediction model using the same approach as above and divide the
+population in equally sized risk-based subgroups. Estimate subgroup-specific
+absolute benefit from the observed absolute differences. Subject-specific
+benefit predictions are made by attributing to individuals their corresponding
+subgroup-specific estimate.
+
+## Constant treatment effect
+Assuming a constant relative treatment effect, fit the adjusted model in
+\@ref(eq:risk). Then, an estimate of absolute benefit can be derived from
+$$\tau(lp;  \hat{\bm{\beta}}, \hat{\gamma}) = g(\hat{lp}) - g(\hat{lp}+\hat{\gamma}) $$
+
+## Linear interaction
+The assumption of constant relative treatment effect is relaxed modeling a
+linear interaction of treatment with the risk linear predictor:
+
+$$P(y=1\given lp, t_x;\hat{\bm{\beta}}) = g(\delta_0 + \delta_1\hat{lp} +
+\gamma_0t_x + \gamma_1\hat{lp}t_x)$$
+
+Then the predicted absolute benefit is 
+$$\tau(lp;\hat{\bm{\beta}},\bm{\gamma}, \bm{\delta}) = g(\delta_0 +
+\delta_1\hat{lp}) - g\big((\delta_0+\gamma_0) + (\delta_1+\gamma_1)\hat{lp}\big)$$
+
+## Restricted cubic splines
+Finally, we drop the linearity assumption and predict absolute benefit using
+smoothing with restricted cubic splines with $k=3, 4$ and $5$ knots. More
+specifically, we fit the model:
+
+$$P(y=1|lp, t_x;\hat{\bm{\beta}}, \bm{\delta}, f_{RCS}) = g\big(\delta_0 + \gamma
+t_x +f_{RCS}(\hat{lp})\times t_x\big)$$
+where
+$$f_{RCS}(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$
+with $h_1(x)=x$ and for $j=2,\dots,k-2$
+$$h_{j+1}(x)= (x-t_j)^3-(x-t_{k-1})_+^3 \frac{t_k-t_j}{t_k-t_{k-1}}+(x-t_k)^3_+\frac{t_{k-1}-t_j}{t_k-t_{k-1}}$$
+where 
+$t_1,\dots,t_k$ are the selected knots. We predict absolute benefit from 
+$$ \tau(lp; \hat{\bm{\beta}}, \delta_0, f_{RCS}) = P(y=1\given\hat{lp},
+t_x=0) - P(y=1\given \hat{lp}, t_x=1) $$
 
@@
   booktabs = TRUE,
-  caption     = "Scenario settings of the entire simulation study.",
+  caption = "Scenario settings of the treatment-covariate interaction scenarios.",
   col.names = c(
@@
 
+For predicting baseline risk of 30-day mortality we fitted a logistic regression
+model with age, Killip class (*Killip*), systolic blood pressure (*sysbp*),
+pulse rate (*pulse*), prior myocardial infarction (*pmi*), location of
+myocardial infarction (*miloc*) and treatment as the covariates. Baseline
+predictions were made setting treatment to 0.
+
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
-\begin{equation}
+\begin{equation*}
 P(outcome=1|X=x) = \text{expit}(lp(x)),
 (\#eq:gusto1)
-\end{equation}
+\end{equation*}
 where 
-\begin{equation}
+\begin{equation*}
 \begin{aligned}
 lp(x)=& \beta_0 + \beta_1 \text{age} + \beta_2 I(\text{Killip}=II) + \beta_3I(\text{Killip}=III) +\\
-&\beta_4 I(\text{Killip}=IV) + \beta_5ceil(\text{sysbp}, 120) + \beta_6 \text{pulse}+\\
+&\beta_4 I(\text{Killip}=IV) + \beta_5min(\text{sysbp}, 120) + \beta_6 \text{pulse}+\\
 &\beta_7 max(\text{pulse - 50, 0}) + \beta_8 I(\text{pmi}=yes)+\\
@@
 (\#eq:gusto2)
-\end{equation}
+\end{equation*}
 

### 2022-02-09 submission/manuscript.rmd
new updates

@@
 the treatment indicator to 0. More information on model development can be found
-in the supplement (Section 6).
+in the supplement (Supplement, Section 8).
 
@@
 again, more conservative, especially with null or moderate treatment-related
-harms (Supplement, Figure S7).
+harms (Supplement, Figure S5).
 
@@
 displayed more stable calibration performance due to the larger number of
-patients (Supplement, Figure S8). When we increased the true prediction AUC to
+patients (Supplement, Figure S6). When we increased the true prediction AUC to
 0.85, the linear interaction model was worse calibrated, on average, than RCS-3
 in the case of strong quadratic deviations from constant relative treatment
-effects (Supplement, Figure S9).
+effects (Supplement, Figure S7).
 
@@
 espcially in the presence of moderate or strong treatment-related harms. All the
-results of these simulations can be found in Supplement, Section 6. Future
+results of these simulations can be found in Supplement, Section 7. Future
 simulation studies could explore the effect of more extensive deviations from

### 2022-02-09 submission/manuscript.tex
new updates

@@
 for individuals we set the treatment indicator to 0. More information on
-model development can be found in the supplement (Section 6).
+model development can be found in the supplement (Supplement, Section
+8).
 
@@
 especially with null or moderate treatment-related harms (Supplement,
-Figure S7).
+Figure S5).
 
@@
 displayed more stable calibration performance due to the larger number
-of patients (Supplement, Figure S8). When we increased the true
+of patients (Supplement, Figure S6). When we increased the true
 prediction AUC to 0.85, the linear interaction model was worse
@@
 deviations from constant relative treatment effects (Supplement, Figure
-S9).
+S7).
 
@@
 treatment-related harms. All the results of these simulations can be
-found in Supplement, Section 6. Future simulation studies could explore
+found in Supplement, Section 7. Future simulation studies could explore
 the effect of more extensive deviations from risk-based treatment

### 2022-02-09 submission/supplement.rmd
new updates

@@
 
-In this section we present specific scenarios from our simulation settings
-in which evolution of benefit followed similar patterns to (REF!!). In this case
-patients were stratified into risk quarters based on their true baseline
+In this section we present specific scenarios from our simulation settings in
+which evolution of benefit followed similar patterns to [@Kent2016]. In this
+case patients were stratified into risk quarters based on their true baseline
 risk. Within each risk quarter we constructed boxplots of true benefit.
@@
 ```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials.
-In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
+In this case, we see increasing absolute benefits with increasing baseline risk up to the third risk quarter.
+In the fourth risk quarter this trend is interrupted and benefits are diminished.", fig.show="hold", out.width = '50%'}
 include_graphics(
@@
 ```
+
+\newpage
+# References
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+<div id="refs"></div>
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+\noindent

### 2022-02-16 submission/iscb_2022_abstract.rmd
add outline of iscb abstract

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: word_document
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(SmoothHte)
library(rms)
library(here)

d <- function(x, decimals = 4) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}

knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
  )
set.seed(19910930)
```

# Abstract
Our objective was to compare different risk-based methods for optimal prediction
of individualized treatment effects from RCTs.

We simulated RCT data using diverse assumptions for the average treatment
effect, a baseline prognostic index of risk (PI), the shape of its interaction
with treatment (none, linear, quadratic or non-monotonic) and the magnitude of
treatment-related harms (none or constant independent of the PI). The
combination of these settings resulted in the definition of 648 simulation
scenarios. In each sample we predicted absolute benefit using: models with a
constant relative treatment effect; stratification in quarters of the PI; models
including a linear interaction of treatment with the PI; models including an
interaction of treatment with a restricted cubic spline (RCS) transformation of
the PI; an adaptive approach using Akaike’s Information Criterion. We evaluated
predictive performance using root mean squared error and measures of
discrimination and calibration for benefit.

The linear-interaction model and the RCS-interaction model outperformed the
constant treatment effect model in many simulation scenarios. The RCS-model was
optimal when quadratic or non-monotonic deviations from a constant treatment
effect were stronger, and when sample size was larger. Larger sample size also
supported the adaptive approach. All the simulation results can be explored at
https://arekkas.shinyapps.io/simulation_viewer/. We also illustrated the
application of the considered methods in the GUSTO-I trial, which confirmed our
simulations' findings.

In conclusion, an interaction between baseline risk and treatment assignment
should be considered to improve treatment effect predictions. Non-linear
interactions should be considered only in larger sample sizes.

### 2022-03-17 submission/manuscript.tex
update after co-author comments, closes #8

@@
   \textbf{Objective}: To compare different risk-based methods for
-  optimal prediction of individualized treatment effects from RCTs.
-  \textbf{Study Design and Setting}: We simulated RCT data using diverse
-  assumptions for the average treatment effect, a baseline prognostic
-  index of risk (PI), the shape of its interaction with treatment (none,
-  linear, quadratic or non-monotonic) and the magnitude of
-  treatment-related harms (none or constant independent of the PI). In
-  each sample we predicted absolute benefit using: models with a
-  constant relative treatment effect; stratification in quarters of the
-  PI; models including a linear interaction of treatment with the PI;
-  models including an interaction of treatment with a restricted cubic
-  spline (RCS) transformation of the PI; an adaptive approach using
-  Akaike's Information Criterion. We evaluated predictive performance
-  using root mean squared error and measures of discrimination and
-  calibration for benefit. \textbf{Results}: The linear-interaction
-  model and the RCS-interaction model outperformed the constant
-  treatment effect model in many simulation scenarios. The RCS-model was
-  optimal when quadratic or non-monotonic deviations from a constant
-  treatment effect were stronger, and when sample size was larger.
-  Larger sample size also supported the adaptive approach. Illustrations
-  in the GUSTO-I trial confirmed these findings. \textbf{Conclusion}: An
-  interaction between baseline risk and treatment assignment should be
-  considered to improve treatment effect predictions. Non-linear
-  interactions should be considered only in larger sample sizes.
+  optimal prediction of individualized treatment effects. \textbf{Study
+  Design and Setting}: We simulated RCT data using diverse assumptions
+  for the average treatment effect, a baseline prognostic index of risk
+  (PI), the shape of its interaction with treatment (none, linear,
+  quadratic or non-monotonic), and the magnitude of treatment-related
+  harms (none or constant independent of the PI). We predicted absolute
+  benefit using: models with a constant relative treatment effect;
+  stratification in quarters of the PI; models including a linear
+  interaction of treatment with the PI; models including an interaction
+  of treatment with a restricted cubic spline (RCS) transformation of
+  the PI; an adaptive approach using Akaike's Information Criterion. We
+  evaluated predictive performance using root mean squared error and
+  measures of discrimination and calibration for benefit.
+  \textbf{Results}: The linear-interaction model and the RCS-interaction
+  displayed robust performance across many simulation scenarios. The
+  RCS-model was optimal when quadratic or non-monotonic deviations from
+  a constant treatment effect were stronger, and when sample size was
+  larger. The adaptive approach required larger sample sizes.
+  Illustrations in the GUSTO-I trial confirmed these findings.
+  \textbf{Conclusion}: An interaction between baseline risk and
+  treatment assignment should be considered to improve treatment effect
+  predictions.
   \end{abstract}
@@
 treatment-covariate interactions, often led to miscalibrated predictions
-of benefit, while risk-based methods proved quite robust {[}5{]}. Most
-often, risk-modeling approaches are carried out in two steps: first a
-risk prediction model is developed externally or internally on the
-entire RCT population, ``blinded'' to treatment; then the RCT population
-is stratified using this prediction model to evaluate risk-based
-treatment effect variation {[}6{]}. This two-step approach identified
-substantial absolute treatment effect differences between low-risk and
-high-risk patients in a re-analysis of 32 large trials {[}7{]}. However,
-even though estimates at the risk subgroup level may be accurate, these
+of absolute benefit, while risk-based methods proved quite robust in
+terms of benefit calibration, although provided weaker discrimination of
+benefit in the presence of true effect modifiers {[}5{]}. Most often,
+risk-modeling approaches are carried out in two steps: first a risk
+prediction model is developed externally or internally on the entire RCT
+population, ``blinded'' to treatment; then the RCT population is
+stratified using this prediction model to evaluate risk-based treatment
+effect variation {[}6{]}. This two-step approach identified substantial
+absolute treatment effect differences between low-risk and high-risk
+patients in a re-analysis of 32 large trials {[}7{]}. However, even
+though estimates at the risk subgroup level may be accurate, these
 estimates may need further refinement for individual patients,
@@
 various risk-based models including a prognostic index of baseline risk
-(PI) and treatment assignment {[}3,4{]}. We aimed to summarize and
-compare different risk-based models for predicting individualized
-treatment effects. We simulated RCT settings to compare the performance
-of these models under different assumptions of the relation between
-baseline risk and treatment. We illustrated the different models by a
-case study of predicting individualized effects of treatments for acute
+(PI) and treatment assignment {[}3,4{]}. In the current simulation
+study, we aim to summarize and compare different risk-based models for
+predicting individualized treatment effects. We simulate different
+relations between baseline risk and treatment effects and also consider
+potential harms of treatment. We illustrate the different models by a
+case study of predicting individualized effects of treatment for acute
 myocardial infarction (MI) in a large randomized controlled trial (RCT).
@@
 using a logistic regression model with main effects for all baseline
-covariates and treatment assignment. Risk predictions for individual
-patients were derived by setting treatment assignment to 0. Another
-common approach is to derive the prediction model solely on the control
-patients, however this approach has been shown to lead to biased benefit
-predictions {[}5,8,9{]}.
+covariates and treatment assignment. Baseline risk predictions for
+individual patients were derived by setting treatment assignment to 0.
+Another common approach is to derive the prediction model solely on the
+control patients, however this approach has been shown to lead to biased
+benefit predictions {[}5,8,9{]}.
 
@@
 models derived in each replication of our simulation settings within
-this super-population. \#\# Empirical illustration
+this super-population.
+
+\hypertarget{empirical-illustration}{%
+\subsection{Empirical illustration}\label{empirical-illustration}}
 
@@
 
-\hypertarget{empirical-illustration}{%
-\subsection{Empirical illustration}\label{empirical-illustration}}
+\hypertarget{empirical-illustration-1}{%
+\subsection{Empirical illustration}\label{empirical-illustration-1}}
 
@@
 The linear interaction model and the RCS-3 model both displayed very
-good performance under many of the considered simulation scenarios, in
-contrast with the constant relative treatment effect model. The linear
-interaction model was optimal in cases with smaller sample sizes and
-moderately performing baseline risk prediction models, that is, it had
-lower RMSE, was better calibrated for benefit and had better
+good performance under many of the considered simulation scenarios. The
+linear interaction model was optimal in cases with smaller sample sizes
+and moderately performing baseline risk prediction models, that is, it
+had lower RMSE, was better calibrated for benefit and had better
 discrimination for benefit, even in scenarios with strong quadratic
@@
 sizes the simpler linear interaction model may be a safer choice for
-predicting absolute benefits, if a non-constant treatment effect is
-suspected.
+predicting absolute benefits in the presence of any suspected
+treatment-related harms.
 
-Risk-based approaches to predictive HTE estimate treatment benefit as a
-function of baseline risk. A limitation of our study is that we assumed
-treatment benefit to be a function of baseline risk in the majority of
-the simulation scenarios. We also considered constant moderate and
-strong treatment-related harms, applied on the absolute scale to expand
-the range of scenarios in line with previous work {[}15{]}. In a limited
-set of scenarios where we assumed the existence of true
-treatment-covariate interactions, our conclusions remained unchanged.
-Even though the average error rates increased for all the considered
-methods, due to the miss-specification of the outcome model, the linear
-interaction model had the lowest error rates. RCS-3 had very comparable
-performance. The constant treatment effect model often gave biased
-results, espcially in the presence of moderate or strong
-treatment-related harms. All the results of these simulations can be
-found in Supplement, Section 7. Future simulation studies could explore
-the effect of more extensive deviations from risk-based treatment
-effects.
+A limitation of our study is that we assumed treatment benefit to be a
+function of baseline risk in the majority of the simulation scenarios.
+We also considered constant moderate and strong treatment-related harms,
+applied on the absolute scale to expand the range of scenarios in line
+with previous work {[}15{]}. In a limited set of scenarios where we
+assumed the existence of true treatment-covariate interactions, our
+conclusions remained unchanged. Even though the average error rates
+increased for all the considered methods, due to the miss-specification
+of the outcome model, the linear interaction model had the lowest error
+rates. RCS-3 had very comparable performance. The constant treatment
+effect model often gave biased results, espcially in the presence of
+moderate or strong treatment-related harms. All the results of these
+simulations can be found in Supplement, Section 7. Future simulation
+studies could explore the effect of more extensive deviations from
+risk-based treatment effects.
 
@@
 compared to more complex treatment effect modelling approaches {[}5{]}.
-Similarly, when SYNTAX score II, a model developed for identifying
-patients with complex coronary artery disease that benefit more from
-percutaneous coronary intervention or from coronary artery bypass
-grafting was redeveloped using fewer treatment-covariate interactions
-had better external performance compared to its predecessor{[}20,21{]}.
+However, whether this remains the case in a range of empirical settings
+still needs to be explored Similarly, when SYNTAX score II, a model
+developed for identifying patients with complex coronary artery disease
+that benefit more from percutaneous coronary intervention or from
+coronary artery bypass grafting was redeveloped using fewer
+treatment-covariate interactions had better external performance
+compared to its predecessor{[}20,21{]}.
 

### 2022-03-17 code/PlotDeviations.R

@@ -21,55 +21,58 @@ library(SimulationEvaluationHte)
 library(dplyr)
 
 analysisIds <- readr::read_csv(
-  "data/processed/analysisIds.csv",
-)
+  "data/processed/analysisIds.csv"
+  )
 
 analysisIdsInteractions <- readr::read_csv(
   "data/processed/analysisIdsInteractions.csv",
   col_types = "icfiidddddddddddddd"
 )
 
-generatePlot <- function(scenarios) {
+generatePlot <- function(scenarios, analysisIds) {
   
-  gammas <- list()
-  for (i in seq_along(scenarios, analysisIds)) {
-    idSettings <- analysisIds %>%
-      filter(scenario == scenarios[i])
+  gammas <- deviationPlots <- list()
+  idSettings <- analysisIds %>% filter(scenario %in% scenarios)
+  for (i in seq_along(scenarios)) {
     gammas[[i]] <- list(
-      g0 = idSettings$g0,
-      g1 = idSettings$g1,
-      g2 = idSettings$g2,
-      c  = idSettings$c
+      g0 = idSettings %>% filter(scenario == scenarios[i]) %>% pull(g0),
+      g1 = idSettings %>% filter(scenario == scenarios[i]) %>% pull(g1),
+      g2 = idSettings %>% filter(scenario == scenarios[i]) %>% pull(g2),
+      c  = idSettings %>% filter(scenario == scenarios[i]) %>% pull(c)
     )
   }
+
+  names(gammas) <- sub("^[lq].*-", "", idSettings$type)
   
-  names(gammas) <- c(
-    "gammas_moderate",
-    "gammas_high"
+  initialPlot <- constantEffectPlot(or = .8, label = "base-case") +
+    plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed")
+  types <- names(gammas)
+
+
+  deviationPlots <- map(
+    types,
+    function(type) {
+      plotSquare(
+        g0 = gammas[[type]]$g0,
+        g1 = gammas[[type]]$g1,
+        g2 = gammas[[type]]$g2,
+        l = gammas[[type]]$c,
+        label = type
+      )
+    }
   )
-  
-  p <- constantEffectPlot(or = .8, label = "base-case") +
-    plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed") +
-    plotSquare(
-      g0    = gammas$gammas_moderate$g0,
-      g1    = gammas$gammas_moderate$g1,
[…]

### 2022-03-17 submission/manuscript.rmd
update after co-author comments, closes #1

@@
 abstract: |
- **Objective**: To compare different risk-based methods for optimal
- prediction of individualized treatment effects from RCTs. **Study
- Design and Setting**: We simulated RCT data using diverse assumptions
- for the average treatment effect, a baseline prognostic index of risk
- (PI), the shape of its interaction with treatment (none, linear,
- quadratic or non-monotonic) and the magnitude of treatment-related
- harms (none or constant independent of the PI). In each sample we
- predicted absolute benefit using: models with a constant relative
- treatment effect; stratification in quarters of the PI; models
- including a linear interaction of treatment with the PI; models
- including an interaction of treatment with a restricted cubic spline
- (RCS) transformation of the PI; an adaptive approach using Akaike’s
- Information Criterion. We evaluated predictive performance using root
- mean squared error and measures of discrimination and calibration for
- benefit. **Results**: The linear-interaction model and the
- RCS-interaction model outperformed the constant treatment effect
- model in many simulation scenarios. The RCS-model was optimal when
- quadratic or non-monotonic deviations from a constant treatment
- effect were stronger, and when sample size was larger. Larger sample
- size also supported the adaptive approach. Illustrations in the
- GUSTO-I trial confirmed these findings. **Conclusion**: An
- interaction between baseline risk and treatment assignment should be
- considered to improve treatment effect predictions. Non-linear 
- interactions should be considered only in larger sample sizes.
+ **Objective**: To compare different risk-based methods for optimal prediction
+ of individualized treatment effects. **Study Design and Setting**: We simulated
+ RCT data using diverse assumptions for the average treatment effect, a baseline
+ prognostic index of risk (PI), the shape of its interaction with treatment
+ (none, linear, quadratic or non-monotonic), and the magnitude of
+ treatment-related harms (none or constant independent of the PI). We predicted
+ absolute benefit using: models with a constant relative treatment effect;
+ stratification in quarters of the PI; models including a linear interaction of
+ treatment with the PI; models including an interaction of treatment with a
+ restricted cubic spline (RCS) transformation of the PI; an adaptive approach
+ using Akaike’s Information Criterion. We evaluated predictive performance using
+ root mean squared error and measures of discrimination and calibration for
+ benefit. **Results**: The linear-interaction model and the RCS-interaction
+ displayed robust performance across many simulation scenarios. The RCS-model
+ was optimal when quadratic or non-monotonic deviations from a constant
+ treatment effect were stronger, and when sample size was larger. The adaptive
+ approach required larger sample sizes. Illustrations in the GUSTO-I trial
+ confirmed these findings. **Conclusion**: An interaction between baseline risk
+ and treatment assignment should be considered to improve treatment effect
+ predictions.
 author:
@@
 treatment-covariate interactions, often led to miscalibrated predictions of
-benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
-often, risk-modeling approaches are carried out in two steps: first a risk
-prediction model is developed externally or internally on the entire RCT
-population, “blinded” to treatment; then the RCT population is stratified using
-this prediction model to evaluate risk-based treatment effect variation
-[@Kent2010]. This two-step approach identified substantial absolute treatment
-effect differences between low-risk and high-risk patients in a re-analysis of
-32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
-level may be accurate, these estimates may need further refinement for
-individual patients, especially for patients with predicted risk at the
-boundaries of the risk intervals. Hence, the risk-stratified approach is useful
-for exploring and presenting HTE, but is not sufficient for supporting treatment
-decisions for individual patients.
+absolute benefit, while risk-based methods proved quite robust in terms of
+benefit calibration, although provided weaker discrimination of benefit in the
+presence of true effect modifiers [@vanKlaveren2019]. Most often, risk-modeling
+approaches are carried out in two steps: first a risk prediction model is
+developed externally or internally on the entire RCT population, “blinded” to
+treatment; then the RCT population is stratified using this prediction model to
+evaluate risk-based treatment effect variation [@Kent2010]. This two-step
+approach identified substantial absolute treatment effect differences between
+low-risk and high-risk patients in a re-analysis of 32 large trials
+[@Kent2016]. However, even though estimates at the risk subgroup level may be
+accurate, these estimates may need further refinement for individual patients,
+especially for patients with predicted risk at the boundaries of the risk
+intervals. Hence, the risk-stratified approach is useful for exploring and
+presenting HTE, but is not sufficient for supporting treatment decisions for
+individual patients.
 
@@
 risk-based models including a prognostic index of baseline risk (PI) and
-treatment assignment [@Kent2019; @PathEnE]. We aimed to summarize and compare
+treatment assignment [@Kent2019; @PathEnE]. In the current simulation study, we aim to summarize and compare
 different risk-based models for predicting individualized treatment effects. We
-simulated RCT settings to compare the performance of these models under
-different assumptions of the relation between baseline risk and treatment. We
-illustrated the different models by a case study of predicting individualized
-effects of treatments for acute myocardial infarction (MI) in a large randomized
+simulate different relations between baseline risk and treatment effects and also consider potential harms of treatment. We
+illustrate the different models by a case study of predicting individualized
+effects of treatment for acute myocardial infarction (MI) in a large randomized
 controlled trial (RCT).
@@
 regression model with main effects for all baseline covariates and treatment
-assignment. Risk predictions for individual patients were derived by setting
-treatment assignment to 0. Another common approach is to derive the prediction
-model solely on the control patients, however this approach has been shown to
-lead to biased benefit predictions [@vanKlaveren2019; @Burke2014; @Abadie2018].
+assignment. Baseline risk predictions for individual patients were derived by
+setting treatment assignment to 0. Another common approach is to derive the
+prediction model solely on the control patients, however this approach has been
+shown to lead to biased benefit predictions [@vanKlaveren2019; @Burke2014;
+@Abadie2018].
 
@@
 replication of our simulation settings within this super-population.
+
 ## Empirical illustration
@@
 The linear interaction model and the RCS-3 model both displayed very good
-performance under many of the considered simulation scenarios, in contrast with
-the constant relative treatment effect model. The linear interaction model was
-optimal in cases with smaller sample sizes and moderately performing baseline
-risk prediction models, that is, it had lower RMSE, was better calibrated for
-benefit and had better discrimination for benefit, even in scenarios with strong
-quadratic deviations. In scenarios with true non-monotonic deviations, the
-linear interaction model was outperformed by RCS-3, especially in the presence
-of true treatment-related harms. Increasing the sample size or the prediction
-model’s discriminative ability favored RCS-3, especially in scenarios with
-non-monotonic deviations and in the presence of treatment-related harms.
+performance under many of the considered simulation scenarios. The linear
+interaction model was optimal in cases with smaller sample sizes and moderately
+performing baseline risk prediction models, that is, it had lower RMSE, was
+better calibrated for benefit and had better discrimination for benefit, even in
+scenarios with strong quadratic deviations. In scenarios with true non-monotonic
+deviations, the linear interaction model was outperformed by RCS-3, especially
+in the presence of true treatment-related harms. Increasing the sample size or
+the prediction model’s discriminative ability favored RCS-3, especially in
+scenarios with non-monotonic deviations and in the presence of treatment-related
+harms.
 
@@
 smaller sample sizes the simpler linear interaction model may be a safer choice
-for predicting absolute benefits, if a non-constant treatment effect is
-suspected.
+for predicting absolute benefits in the presence of any suspected
+treatment-related harms.
 
-Risk-based approaches to predictive HTE estimate treatment benefit as a function
-of baseline risk. A limitation of our study is that we assumed treatment benefit
+A limitation of our study is that we assumed treatment benefit
 to be a function of baseline risk in the majority of the simulation
@@
 consistently better calibrated for benefit compared to more complex treatment
-effect modelling approaches [@vanKlaveren2019]. Similarly, when SYNTAX score II,
-a model developed for identifying patients with complex coronary artery disease
-that benefit more from percutaneous coronary intervention or from coronary
-artery bypass grafting was redeveloped using fewer treatment-covariate
-interactions had better external performance compared to its
+effect modelling approaches [@vanKlaveren2019]. However, whether this remains
+the case in a range of empirical settings still needs to be explored Similarly,
+when SYNTAX score II, a model developed for identifying patients with complex
+coronary artery disease that benefit more from percutaneous coronary
+intervention or from coronary artery bypass grafting was redeveloped using fewer
+treatment-covariate interactions had better external performance compared to its
 predecessor[@farooq2013anatomical; @takahashi2020redevelopment].

### 2022-03-22 submission/manuscript.rmd
fix shiny and rebuild manuscript + supplement

@@
     affiliation: 1
+    correspondingauthor: true
   - name: Peter R. Rijnbeek
@@
 \linenumbers
+
+# What is new? {-}
+* Models including a linear interaction of the prognostic index with treatment
+  were a good option for predicting absolute benefit with smaller sample sizes or
+  moderately performing prediction models.
+* Models including an interaction of treatment with a restricted cubic spline
+  transformation using 3 knots performed better in the presence of true
+  non-monotonic deviations from a constant relative treatment effect or with
+  treatment-related harms.
+* Using an AIC adaptive approach to select among a constant effects model, a
+  linear interaction model, and a model using restricted cubic spline
+  transformations was better suited for larger sample sizes.
+
+# What this adds to what was known? {-}
+* Using a constant relative treatment effect for making risk-based
+  individualized benefit predictions is not always the best approach.
+* Models including linear interaction of treatment with the prognostic index
+  have robust performance across a wide variety of settings.
+* Increasing the flexibility of the transformation of the prognostic index does
+  not improve performance.
+
+# What is the implication and what should change now? {-}
+* Assuming a constant relative treatment effect for all patients may lead to
+  treatment mistargeting.
+* We recommend using a linear interaction of treatment with the prognostic index
+  for making individualized risk-based benefit predictions.
+* With larger sample size, using an adaptive approach based on AIC to explore
+  non-linear interactions of the prognostic index with treatment is a viable
+  option.
 

### 2022-03-22 submission/manuscript.tex
fix shiny and rebuild manuscript + supplement

@@
 \linenumbers
+
+\hypertarget{what-is-new}{%
+\section*{What is new?}\label{what-is-new}}
+\addcontentsline{toc}{section}{What is new?}
+
+\begin{itemize}
+\tightlist
+\item
+  Models including a linear interaction of the prognostic index with
+  treatment were a good option for predicting absolute benefit with
+  smaller sample sizes or moderately performing prediction models.
+\item
+  Models including an interaction of treatment with a restricted cubic
+  spline transformation using 3 knots performed better in the presence
+  of true non-monotonic deviations from a constant relative treatment
+  effect or with treatment-related harms.
+\item
+  Using an AIC adaptive approach to select among a constant effects
+  model, a linear interaction model, and a model using restricted cubic
+  spline transformations was better suited for larger sample sizes.
+\end{itemize}
+
+\hypertarget{what-this-adds-to-what-was-known}{%
+\section*{What this adds to what was
+known?}\label{what-this-adds-to-what-was-known}}
+\addcontentsline{toc}{section}{What this adds to what was known?}
+
+\begin{itemize}
+\tightlist
+\item
+  Using a constant relative treatment effect for making risk-based
+  individualized benefit predictions is not always the best approach.
+\item
+  Models including linear interaction of treatment with the prognostic
+  index have robust performance across a wide variety of settings.
+\item
+  Increasing the flexibility of the transformation of the prognostic
+  index does not improve performance.
+\end{itemize}
+
+\hypertarget{what-is-the-implication-and-what-should-change-now}{%
+\section*{What is the implication and what should change
+now?}\label{what-is-the-implication-and-what-should-change-now}}
+\addcontentsline{toc}{section}{What is the implication and what should
+change now?}
+
+\begin{itemize}
+\tightlist
+\item
+  Assuming a constant relative treatment effect for all patients may
+  lead to treatment mistargeting.
+\item
+  We recommend using a linear interaction of treatment with the
+  prognostic index for making individualized risk-based benefit
+  predictions.
+\item
+  With larger sample size, using an adaptive approach based on AIC to
+  explore non-linear interactions of the prognostic index with treatment
+  is a viable option.
+\end{itemize}
 

### 2022-03-22 submission/supplement.rmd
fix shiny and rebuild manuscript + supplement

@@
   here("data/processed/analysisIdsInteractions.csv")
-) %>%
-  select(-"type")
+)
 kableExtra::kbl(
@@
     "Base",
+    "Type",
     "N",

### 2022-03-22 extras/shiny/results/global.R

@@ -12,16 +12,15 @@ calibration <- readr::read_csv(
   col_types = readr::cols(.default = "d")
 )
 
-analysisIds <- readr::read_csv(
+analysisIds1 <- readr::read_csv(
   "data/analysisIds.csv"
 ) %>%
   select(scenario, base, type, harm, sampleSize, auc)
 
-# analysisIds2 <- readr::read_csv("data/analysisIdsInteractions.csv") %>%
-#   select(scenario, base, type, sampleSize, auc) %>%
-#   mutate(harm = "absent")
-# 
-# analysisIds <- bind_rows(analysisIds1, analysisIds2)
+analysisIds2 <- readr::read_csv("data/analysisIdsInteractions.csv") %>%
+  select(scenario, base, type, harm, sampleSize, auc)
+
+analysisIds <- bind_rows(analysisIds1, analysisIds2)
 
 
 
@@ -56,11 +55,13 @@ createPlot2 <- function(data) {
     plotly::layout(
       yaxis = list(
         hoverformat = ".3f",
-        title = ""
+        title = "",
+        gridcolor = "ffff"
       ),
       xaxis = list(
-        title = ""
-      )
+          title = ""
+      ),
+      plot_bgcolor = '#e1e9f5'
     )
 }
 

### 2022-03-22 extras/shiny/results/server.R

@@ -128,7 +128,10 @@ shiny::shinyServer(
             pull(scenario)
         } else {
           analysisIds %>%
-            filter(effectSize == input$effect) %>%
+            filter(
+              type == input$effect,
+              harm == input$harm
+            ) %>%
             pull(scenario)
         }
       }

### 2022-03-22 extras/shiny/results/ui.R

@@ -9,9 +9,9 @@ shiny::shinyUI(
       shinydashboard::sidebarMenu(
         id = "menu1",
         shinydashboard::menuItem(
-          tabName = "abstract",
-          text    = "Abstract",
-          icon    = icon("file-alt")
+          tabName = "home",
+          text    = "Home",
+          icon    = icon("home")
         ),
         shinydashboard::menuItem(
           tabName = "settings",
@@ -37,14 +37,24 @@ shiny::shinyUI(
           shiny::uiOutput("typeInput"),
           shiny::uiOutput("toggleNPatients"),
           shiny::uiOutput("togglePredictionPerformance"),
-          shiny::uiOutput("toggleHarmInput")
+          shiny::selectInput(
+            inputId  = "harm",
+            label    = "Constant harm",
+            choices  = c(
+              "absent",
+              "moderate-positive",
+              "strong-positive",
+              "negative"
+            ),
+            selected = "absent"
+          )
         )
       )
     ),
     shinydashboard::dashboardBody(
       shinydashboard::tabItems(
         shinydashboard::tabItem(
-          tabName = "abstract",
+          tabName = "home",
           shiny::includeHTML(
             "html/abstract.html"
           )

### 2022-04-08 submission/manuscript.rmd
update manuscript and supplement

@@
   - \newcommand\given[1][]{\:#1\vert\:}
+  - \newcommand{\indep}{\perp \!\!\! \perp}
 editor_options: 
@@
 risk-based models including a prognostic index of baseline risk (PI) and
-treatment assignment [@Kent2019; @PathEnE]. In the current simulation study, we aim to summarize and compare
-different risk-based models for predicting individualized treatment effects. We
-simulate different relations between baseline risk and treatment effects and also consider potential harms of treatment. We
-illustrate the different models by a case study of predicting individualized
-effects of treatment for acute myocardial infarction (MI) in a large randomized
-controlled trial (RCT).
+treatment assignment [@Kent2019; @PathEnE]. In the current simulation study, we
+aim to summarize and compare different risk-based models for predicting
+individualized treatment effects. We simulate different relations between
+baseline risk and treatment effects and also consider potential harms of
+treatment. We illustrate the different models by a case study of predicting
+individualized effects of treatment for acute myocardial infarction (MI) in a
+large randomized controlled trial (RCT).
 
 # Methods
+
+## Background
+
+We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
+treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
+covariates measured. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
+outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
+predicting the conditional average treatment effect (CATE), 
+$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
+Assuming that $(Z, X, Y)$ is a random sample from the target population and that
+$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
+CATE from
+\begin{align*}
+\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
+&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
+\end{align*}
+
+Based on an estimate of baseline risk
+$$
+E\{Y\given X=x, Z=0\}=g\big(lp(x)\big)
+$$
+with $u=lp(x)=\hat{\beta}^tx$ the linear predictor and $g$ the link function,
+we predict CATE from
+$$
+\hat{\tau}(x) = g\big(f(u, 0)\big) - g\big(f(u, 1)\big)
+$$
+where $f(u,z)$ describes interactions of the baseline risk linear predictor with
+treatment. In the rest of the paper we use the linear predictor $lp(x)$ as the
+prognostic index (PI) for the outcome $Y$.
 
@@
 where $lp_1$ is the true linear predictor in the treatment arm, so that
-$P(y=1|X, t_x=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
+$P(y=1|X, Z=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
 constant absolute harm is applied across all treated patients. In this case we
-have $P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}$.
+have $P(y=1|X,Z=1) = \text{expit}(lp_1) + \text{harm}$.
 
@@
 (constant odds ratio). Hence, absolute benefit is predicted from
-$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.
+$\hat{\tau}(\bm{x}) = \text{expit}(PI) - \text{expit}(PI +\log(\text{OR}))$.
 

### 2022-04-08 submission/manuscript.tex
update manuscript and supplement

@@
 \newcommand\given[1][]{\:#1\vert\:}
+\newcommand{\indep}{\perp \!\!\! \perp}
 \usepackage{booktabs}
@@
 
+\hypertarget{background}{%
+\subsection{Background}\label{background}}
+
+We observe RCT data \((Z, X, Y)\), where for each patient \(Z_i= 0, 1\)
+is the treatment status, \(Y_i = 0, 1\) is the observed outcome and
+\(X_i\) is a set of covariates measured. Let \(\{Y_i(z), z=0, 1\}\)
+denote the unobservable potential outcomes. We observe
+\(Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)\). We are interested in predicting
+the conditional average treatment effect (CATE),
+\[\tau(x) = E\{Y(0) - Y(1)|X=x\}\] Assuming that \((Z, X, Y)\) is a
+random sample from the target population and that
+\(\big(Y(0), Y(1)\big)\perp \!\!\! \perp Z|X\), as we are in the RCT
+setting, we can predict CATE from \begin{align*}
+\tau(x) &= E\{Y(0)\:\vert\:X=x\}-E\{Y(1)\:\vert\:X=x\}\\
+&=E\{Y\:\vert\:X=x, Z=0\}-E\{Y\:\vert\:X=x, Z=1\}
+\end{align*}
+
+Based on an estimate of baseline risk \[
+E\{Y\:\vert\:X=x, Z=0\}=g\big(lp(x)\big)
+\] with \(u=lp(x)=\hat{\beta}^tx\) the linear predictor and \(g\) the
+link function, we predict CATE from \[
+\hat{\tau}(x) = g\big(f(u, 0)\big) - g\big(f(u, 1)\big)
+\] where \(f(u,z)\) describes interactions of the baseline risk linear
+predictor with treatment. In the rest of the paper we use the linear
+predictor \(lp(x)\) as the prognostic index (PI) for the outcome \(Y\).
+
 \hypertarget{simulation-scenarios}{%
@@
 is the true linear predictor in the treatment arm, so that
-\(P(y=1|X, t_x=1) = \text{expit}(lp_1)\). Finally, we simulated
-scenarios where a constant absolute harm is applied across all treated
-patients. In this case we have
-\(P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}\).
+\(P(y=1|X, Z=1) = \text{expit}(lp_1)\). Finally, we simulated scenarios
+where a constant absolute harm is applied across all treated patients.
+In this case we have
+\(P(y=1|X,Z=1) = \text{expit}(lp_1) + \text{harm}\).
 
@@
 predicted from
-\(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).
+\(\hat{\tau}(\bm{x}) = \text{expit}(PI) - \text{expit}(PI +\log(\text{OR}))\).
 

### 2022-04-08 submission/supplement.rmd
update manuscript and supplement

@@
   c(
-    here("figures/scenario_251.png"),
-    here("figures/scenario_324.png")
+    here("figures/scenario_251.tiff"),
+    here("figures/scenario_324.tiff")
   )
@@
   c(
-    here("figures/scenario_406.png"),
-    here("figures/scenario_422.png")
+    here("figures/scenario_406.tiff"),
+    here("figures/scenario_422.tiff")
   )

### 2022-11-13 submission/supplement.rmd
add sensitivity analyses

--- 2022-03-22 submission/supplement.rmd
@@
 
+# Notation
+We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
+treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
+covariates measured. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
+outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
+predicting the conditional average treatment effect (CATE), 
+$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
+Assuming that $(Z, X, Y)$ is a random sample from the target population and that
+$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
+CATE from
+\begin{align*}
+\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
+&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
+\end{align*}
+
+Based on an estimate of baseline risk
+$$
+E\{Y\given X=x, Z=0\}=g\big(\hat{lp}(x)\big)
+$$
+with $\hat{u}=\hat{lp}(x)=x^t\hat{\beta}$ the linear predictor and $g$ the link function,
+we predict CATE from
+$$
+\hat{\tau}(x) = g\big(f(\hat{u}, 0)\big) - g\big(f(\hat{u}, 1)\big)
+$$
+where $f(u,z)$ describes interactions of the baseline risk linear predictor with
+treatment.
+
 # Simulation settings
-For all patients we observe covariates $x_1,\dots,x_8$, of which $4$ are
+For all patients we observe covariates $X_1,\dots,X_8$, of which $4$ are
 continuous and $4$ are binary. More specifically,
@@
 \begin{equation*}
-x_1,\dots,x_4 \sim N(0, 1)
+X_1,\dots,X_4 \sim N(0, 1)
 \end{equation*}
 \begin{equation*}
-x_5,\dots,x_8 \sim B(1, 0.2)
+X_5,\dots,X_8 \sim B(1, 0.2)
 \end{equation*}
 
-We first, generate the binary outcomes $y$ for the untreated patients ($t_x=0$),
+We first, generate the binary outcomes $Y$ for the untreated patients ($Z=0$),
 based on
@@
 \begin{equation} 
-P(y\given \bm{x}, t_x=0) = g(\beta_0 + \beta_1x_1+\dots+\beta_8x_8) = g(lp_0),
+P(Y(0)=1\given X=x) = g(\beta_0 + \beta_1x_1+\dots+\beta_8x_8) = g(lp_0),
 (\#eq:p0)
@@
 \begin{equation}
-P(y\given\bm{x}, t_x=1) = g(lp_1)
+P(Y = 1\given X=x, Z=1) = g(lp_1)
 (\#eq:p1)
@@
 development of the prediction model we use the model defined in \@ref(eq:p0)
-including a constant treatment effect. When doing predictions, $t_x$ is set to
-$0$. The value of the true $\bm{\beta}$ is such that the above prediction model
+including a constant treatment effect. When doing predictions, $Z$ is set to
+$0$. The value of the true $\beta$ is such that the above prediction model
 has an AUC of $0.75$.
 
-The previously defined targets are achieved when $\bm{\beta}=(-2.08,
+The previously defined targets are achieved when $\beta=(-2.08,
 0.49,\dots,0.49)^t$. For the derivations in the treatment arm we use
-$\bm{\gamma}=(\log(0.8), 1, 0)^t$.
+$\gamma=(\log(0.8), 1, 0)^t$.
 
@@
 ) %>%
-  select(-"type")
+  select(-c("type", "averageTrueBenefit", "averageObservedBenefit"))
 kableExtra::kbl(
@@
   longtable   = TRUE,
-  align       = "r",
+  align       = "rlrrlr",
   booktabs    = TRUE,
@@
     paste0("g", 0:2),
-    "c",
-    "Before harms",
-    "After harms"
+    "c"
   ),
@@
       "Baseline risk"         = 9,
-      "True treatment effect" = 4,
-      "Benefit"               = 2
+      "True treatment effect" = 4
       )
@@
 
-```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials.
+```{r, cache=TRUE, echo=FALSE, warning=FALSE, fig.cap="Simulation scenarios that closely follow trials.
 In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
@@
 
-```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials.
+```{r, cache=TRUE, echo=FALSE, warning = FALSE, fig.cap="Simulation scenarios that closely follow trials.
 In this case, we see increasing absolute benefits with increasing baseline risk up to the third risk quarter.
@@
 \begin{equation}
-P(y=1\given\bm{x}, t_x) = g(\bm{x}^t\bm{\beta} + \gamma t_x)
+P(Y=1\given X=x,Z=z) = g(x^t\beta + \delta_0 z)
 \end{equation}
@@
 
-The risk linear predictor is then 
-$lp(\bm{x}; \bm{\beta}) = \bm{x}^t\bm{\beta}$. 
-We derive baseline risk predictions for patients by setting $t_x=0$ in
+We derive baseline risk predictions for patients by setting $Z=0$ in
 \@ref(eq:risk). All methods for individualizing benefit predictions are 2-stage
@@
 \begin{equation*}
-\hat{lp} = lp(\bm{x};\hat{\bm{\beta}}) = \bm{x}^t\hat{\bm{\beta}}
+\hat{lp} = lp(x;\hat{\beta}) = x^t\hat{\beta}
 \end{equation*}
@@
 Assuming a constant relative treatment effect, fit the adjusted model in
-\@ref(eq:risk). Then, an estimate of absolute benefit can be derived from
-$$\tau(lp;  \hat{\bm{\beta}}, \hat{\gamma}) = g(\hat{lp}) - g(\hat{lp}+\hat{\gamma}) $$
+\@ref(eq:risk). Then, predict absolute benefit using
+
+\begin{equation}
+\hat{\tau}(x;\hat{\beta},\hat{\gamma})=g(f(\hat{lp}, 0)) - g(f(\hat{lp},1)), 
+(\#eq:main)
+\end{equation}
+
+where $f(\hat{lp}, z) = \hat{lp}+\hat{\delta}_0z$, with $\hat{\delta}_0$ the
+estimated relative treatment effect (log odds ratio).
 
 ## Linear interaction
-The assumption of constant relative treatment effect is relaxed modeling a
-linear interaction of treatment with the risk linear predictor:
+We relax the assumption of a constant relative treatment effect in
+\@ref(eq:main) by setting
 
-$$P(y=1\given lp, t_x;\hat{\bm{\beta}}) = g(\delta_0 + \delta_1\hat{lp} +
-\gamma_0t_x + \gamma_1\hat{lp}t_x)$$
-
-Then the predicted absolute benefit is 
-$$\tau(lp;\hat{\bm{\beta}},\bm{\gamma}, \bm{\delta}) = g(\delta_0 +
-\delta_1\hat{lp}) - g\big((\delta_0+\gamma_0) + (\delta_1+\gamma_1)\hat{lp}\big)$$
+$$ f(\hat{lp}, z) = \delta_0+\delta_1z+\delta_2\hat{lp}+\delta_3z\hat{lp} $$
 
@@
 smoothing with restricted cubic splines with $k=3, 4$ and $5$ knots. More
-specifically, we fit the model:
+specifically, we set:
 
-$$P(y=1|lp, t_x;\hat{\bm{\beta}}, \bm{\delta}, f_{RCS}) = g\big(\delta_0 + \gamma
-t_x +f_{RCS}(\hat{lp})\times t_x\big)$$
+$$ f(\hat{lp}, z) = \delta_0 + \delta_1z+zs(\hat{lp}) $$
 where
-$$f_{RCS}(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$
+$$s(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$
 with $h_1(x)=x$ and for $j=2,\dots,k-2$
@@
 where 
-$t_1,\dots,t_k$ are the selected knots. We predict absolute benefit from 
-$$ \tau(lp; \hat{\bm{\beta}}, \delta_0, f_{RCS}) = P(y=1\given\hat{lp},
-t_x=0) - P(y=1\given \hat{lp}, t_x=1) $$
+$t_1,\dots,t_k$ are the selected knots [@harrell2017regression].
 
@@
 # Discrimination and calibration for benefit
+The c-for-benefit represents the probability that from two randomly chosen
+matched patient pairs with unequal observed benefit, the pair with greater
+observed benefit also has a higher predicted benefit. To be able to calculate
+observed benefit, patients in each treatment arm are ranked based on their
+predicted benefit and then matched 1:1 across treatment arms. Observed treatment
+benefit is defined as the difference of observed outcomes between the untreated
+and the treated patient of each matched patient pair. Predicted benefit is
+defined as the average of predicted benefit within each matched patient pair.
+
+We evaluated calibration in a similar manner, using the integrated calibration
+index (ICI) for benefit [@Austin2019]. The observed benefits are regressed on the
+predicted benefits using a locally weighted scatterplot smoother (loess). The
+ICI-for-benefit is the average absolute difference between predicted and smooth
+observed benefit. Values closer to represent better calibration.
 
@@
 considered: 1) 4 true weak positive interactions
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.83$); 2) 4 strong positive
-interactions ($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$); 3) 2 weak and 2
+($\text{OR}_{Z=1} / \text{OR}_{Z=0}=0.83$); 2) 4 strong positive
+interactions ($\text{OR}_{Z=1} / \text{OR}_{Z=0}=0.61$); 3) 2 weak and 2
 strong positive interactions; 4) 4 weak negative interactions 
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=1.17$); 5) 4 strong negative interactions 
-($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=1.39$); 6) 2 weak and 2 strong negative
+($\text{OR}_{Z=1} / \text{OR}_{Z=0}=1.17$); 5) 4 strong negative interactions 
+($\text{OR}_{Z=1} / \text{OR}_{Z=0}=1.39$); 6) 2 weak and 2 strong negative
 interactions; 7) combined positive and negative strong interactions. We also
@@
   here("data/processed/analysisIdsInteractions.csv")
-)
+) %>%
+  select(-c("averageTrueBenefit", "averageObservedBenefit"))
 kableExtra::kbl(
@@
   longtable = TRUE,
-  align = "r",
+  align = "rllrrlr",
   booktabs = TRUE,
@@
     paste0("b", 0:8),
-    "g0", "g1", "g2", "g5", "g6",
-    "Before harms",
-    "After harms"
+    "g0", "g1", "g2", "g5", "g6"
   ),
@@
     c(
-      "Analysis ID"           = 5,
+      "Analysis ID"           = 6,
       "Baseline risk"         = 9,
-      "True treatment effect" = 5,
-      "Benefit"               = 2
+      "True treatment effect" = 5
       )
@@
 grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_combined.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+\newpage
+
+# Correlated covariates
+
+We analyzed the sensitivity of our results to correlation between baseline
+characteristics by incuding additional simulation scenarios. We sampled
+covariates $W_1,\dots,W_8\sim N(0, \Sigma)$. We generated four continuous
+baseline covariates $X_1=W_1,\dots,X_4=W_4$ and four binary covariates with 20\%
+prevalence $X_5=I(W_5>z_{0.8}), X_8=I(W_8>z_{0.8})$, where $I$ is the indicator
+function. We selected the covariance matrix $\Sigma$ such that
+$\text{cor}(X_i,X_j)=0.5$, for any $i\neq j$. More precisely, we set $\Sigma$ as
+can be seen below:
+
+\begin{equation*}
+\Sigma = 
+\begin{pmatrix}
+1       & 0.5   & 0.5   & 0.5   & 0.708 & 0.708 & 0.708 & 0.708 \\
+0.5     & 1     & 0.5   & 0.5   & 0.708 & 0.708 & 0.708 & 0.708 \\
+0.5     & 0.5   & 1     & 0.5   & 0.708 & 0.708 & 0.708 & 0.708 \\
+0.5     & 0.5   & 0.5   & 1     & 0.708 & 0.708 & 0.708 & 0.708 \\
+0.708   & 0.708 & 0.708 & 0.708 & 1     & 0.745 & 0.745 & 0.745 \\
+0.708   & 0.708 & 0.708 & 0.708 & 0.745 & 1     & 0.745 & 0.745 \\
+0.708   & 0.708 & 0.708 & 0.708 & 0.745 & 0.745 & 1     & 0.745 \\
+0.708   & 0.708 & 0.708 & 0.708 & 0.745 & 0.745 & 0.745 & 1
+\end{pmatrix}
+\end{equation*}
+
+In order to ensure that the simulated datasets were comparable to the original
+main simulation scenarios, i.e. control arm event rate of 20\% and true risk
+model c-statistic (AUC) equal to the target, we needed to adjust the
+coefficients of the true risk model. The exact settings of the simlation
+scenarios for the sensitivity analyses can be found in Table \@ref(tab:tab3).
+
+```{r tab3, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
+asd <- read_csv(
+  here("data/processed/analysisIdsSensitivity.csv")
+) %>%
+  select(-c("type", "averageTrueBenefit", "averageObservedBenefit"))
+kableExtra::kbl(
+  asd,
+  format = "latex",
+  escape = FALSE,
+  longtable   = TRUE,
+  align       = "rlrrlr",
+  booktabs    = TRUE,
+  caption     = "Scenario settings of the entire simulation study.",
+  col.names   = c(
+    "Scenario",
+    "Base",
+    "N",
+    "AUC",
+    "Treatment-related harm",
+    paste0("b", 0:8),
+    paste0("g", 0:2),
+    "c"
+  ),
+  linesep = c('', '', '', '\\addlinespace'),
+  digits      = 3,
+  format.args = list(
+    big.mar      = ",",
+    decimal.mark = "."
+  )
+) %>%
+  kableExtra::landscape() %>%
+  kableExtra::add_header_above(
+    c(
+      "Analysis ID"           = 5,
+      "Baseline risk"         = 9,
+      "True treatment effect" = 4
+      )
+  ) %>%
+  kableExtra::kable_styling(font_size = 7)
+```
+
+The results of the sensitivity analyses can be explored in Figures ... We found
+no noticeable differences between methods for individualizing treatment benefit
+predictions compared to the results of the simulation scenarios where baseline
+covariates were assumed to be statistically independent.
+
+```{r rmseBaseSensitivity, cache=TRUE, echo=FALSE, fig.cap="RMSE of the sensitivity analyses assuming correlated baseline covariates.
+All considered methods were derived using simulated samples of size 4,250 and true prediction c-statistic of 0.75", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base_sensitivity.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r rmseSampleSizeSensitivity, cache=TRUE, echo=FALSE, fig.cap="RMSE of the sensitivity analyses assuming correlated baseline covariates.
+All considered methods were derived using simulated samples of size 17,000 and true prediction c-statistic of 0.75", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size_sensitivity.tiff")))
+# knitr::include_graphics(here::here("figures/calibration_base.png"))
+```
+
+```{r rmseAucSensitivity, cache=TRUE, echo=FALSE, fig.cap="RMSE of the sensitivity analyses assuming correlated baseline covariates.
+All considered methods were derived using simulated samples of size 4,250 and true prediction c-statistic of 0.85", fig.show="hold", out.width = '100%'}
+grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc_sensitivity.tiff")))
 # knitr::include_graphics(here::here("figures/calibration_base.png"))

### 2022-11-13 code/MergeResults.R

@@ -3,11 +3,11 @@
 library(tidyverse)
 source("code/helpers/TmpFiles.R")
 
-scenarios <- length(list.files("data/raw", pattern = "scenario"))
+scenarios <- list.files("data/raw", pattern = "scenario", full.names = T)
 if (!dir.exists(".scratch/tmp")) {
   dir.create(".scratch/tmp", recursive = TRUE)
 }
-lapply(1:scenarios, createTmpFiles, scenarioDir = "data/raw/", tmpDir = ".scratch/tmp/")
+lapply(scenarios, createTmpFiles, tmpDir = ".scratch/tmp/")
 lapply(
   c("rmse", "discrimination", "calibration", "concordance", "adaptiveModel"), 
   mergeTmpFiles, 

### 2022-11-13 code/PlotRmse.R

@@ -1,380 +1,0 @@
-#!/usr/bin/env Rscript
-
-# ===================================================
-# Description:
-#   Generates the RMSE plots for the base case
-# Input:
-#   - sample size
-#   - auc
-#   - value
-# Output:
-#   - figures/rmse_base.tiff
-# Depends:
-#   - data/processed/analysisIds.csv
-#   - data/processed/rmse.csv
-#   - code/helpers/CreateManuscriptPlots.R
-#   - code/helpers/PlotResult.R
-#   - code/helpers/Absolute.R
-# ===================================================
-
-args <- commandArgs(trailingOnly = TRUE)
-args_base <- as.character(args[1])
-args_sampleSize <- as.numeric(args[2])
-args_auc <- as.numeric(args[3])
-args_value <- as.character(args[4])
-
-library(tidyverse)
-library(glue)
-library(ggtext)
-library(gridExtra)
-library(grid)
-library(ggside)
-
-# --------------------------------------
-# Sourcing helper files:
-#   1. Generates single boxplot
-#   2. Generates the boxplot list
-#   3. Generates the absolute plots
-# --------------------------------------
-source("code/helpers/CreateManuscriptPlots.R") 
-source("code/helpers/PlotResult.R")
-source("code/helpers/Absolute.R")
-
-scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
-  filter(
-    base == args_base,
-    !(type %in% c("quadratic-moderate", "linear-moderate")),
-    sampleSize == args_sampleSize,
-    auc == args_auc
-  ) 
-
-metric    <- "rmse"
-
-titles <- scenarioIds %>%
-  mutate(
-    title = ifelse(
-      type == "constant",
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
-      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
-    )
-  ) %>%
-  select(title) %>%
-  unlist() %>%
-  unique()
-
-names(titles) <- NULL
-
-titlePrefix <- paste0(
-  "**",
-  LETTERS[1:5],
-  ".**"
-)
-
-titles <- paste(
-  titlePrefix,
-  titles
-)
-
-
-metricFile <- paste(metric, "csv", sep = ".")
-
-f <- function(x) x * 100
-
-limitsHigh <- ifelse(
-  args_base == "high",
-  yes = .2,
-  no = .15
-)
-
-processed <- readr::read_csv(
[…]

### 2022-11-13 code/SimulationScript.R

@@ -32,11 +32,29 @@ analysisIdsInteractions <- readr::read_csv(
   "data/processed/analysisIdsInteractions.csv",
 )
 
+analysisIdsSensitivity <- readr::read_csv(
+  "data/processed/analysisIdsSensitivity.csv"
+)
+
 selectedScenario <- as.numeric(
   stringr::str_extract(args[1], "[0-9]+")
 )
 
 if (selectedScenario <= 648) {
+  databaseSettings <- SimulateHte::createDatabaseSettings(
+    numberOfObservations = as.numeric(as.character(idSettings$sampleSize)),
+    numberOfCovariates = 8,
+    covariateDistributionSettings = list(
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createBinomialDistributionSettings(prob = .2),
+      SimulateHte::createBinomialDistributionSettings(prob = .2),
+      SimulateHte::createBinomialDistributionSettings(prob = .2),
+      SimulateHte::createBinomialDistributionSettings(prob = .2)
+    )
+  )
   idSettings <- analysisIds %>%
     filter(scenario == selectedScenario)
   
@@ -74,8 +92,22 @@ if (selectedScenario <= 648) {
       )
     )
   )
-} else {
-  idSettings <- analysisIdsInteractions %>%
+} else if (selectedScenario <= 676) {
+  databaseSettings <- SimulateHte::createDatabaseSettings(
+    numberOfObservations = as.numeric(as.character(idSettings$sampleSize)),
+    numberOfCovariates = 8,
+    covariateDistributionSettings = list(
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createNormalDistributionSettings(),
+      SimulateHte::createBinomialDistributionSettings(prob = .2),
+      SimulateHte::createBinomialDistributionSettings(prob = .2),
+      SimulateHte::createBinomialDistributionSettings(prob = .2),
+      SimulateHte::createBinomialDistributionSettings(prob = .2)
+    )
+  )
+  idSettings <- analysisIdsSensitivity %>%
[…]

### 2022-11-13 code/helpers/TmpFiles.R

@@ -1,18 +1,15 @@
 library(tidyverse)
 
-createTmpFiles <- function(scenarioId, scenarioDir, tmpDir) {
+createTmpFiles <- function(scenarioDir, tmpDir) {
   
   evaluation <- readRDS(
     file.path(
       scenarioDir,
-      paste(
-        "scenario",
-        scenarioId,
-        sep = "_"
-      ),
       "evaluation.rds"
     )
   )
+  
+  scenarioId <- as.numeric(str_extract(scenarioDir, "(?<=\\_).*"))
   
   evaluation <- evaluation %>% map(~mutate(., scenarioId = scenarioId))
   
